### 2019-12-03 rmds/linear_regression.Rmd
add rmds folder

---
title: "Linear Regression (TMAs)"
author: "Raven McKnight"
date: "12/3/2019"
output: 
  html_document: 
    toc: true
    toc_float: true
    theme: paper
    code_folding: hide
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(tigris)
library(ggplot2)
library(leaps)

options(tigris_class = "sf")
options(tigris_use_cache = TRUE)

counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
bgs <- block_groups("MN", counties, 2016)

# should find better way to read data from rmd folder
lm_dat <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/modeling-dat/lm_dat_logged.RDS')
```

# Introduction

The simplest model we can fit is a linear regression. This model was inspired by Metro Transit's Transit Market Areas. Here, we'll let $Y_i$ equal the log of average number of weekday boardings per stop in census block group $i$ in 2017. We'll use four basic predictors, inspired by the original Transit Market Areas: population density, employment density, percent of households with no vehicle, and a walkability metric derived from walk isochrones. We also include number of stops made in block group $i$ as a control for level of service.

# The Model

```{r}
lm1 <- lm(log_daily_boards ~ pop_density + emp_density + walkability + perc_no_veh + log_daily_stops, lm_dat, na.action = na.exclude)

summary(lm1)
```

I'm guessing population density is insignificant because it's so correlated with employment.

# Results
## Transit Market Areas

```{r}
lm_dat[, index := -3.79 + 0.06*pop_density + 0.22*emp_density + 0.21*walkability + 0.19*perc_no_veh + 1.18*log_daily_stops]

lm_dat[, predicted_boardings := exp(index)]

lm_dat_sf <- dplyr::left_join(bgs, lm_dat)
```

We can use the model above to define updated Transit Market Areas based on the number of boardings a block group can support, rather than an arbitary index value. The distribution of predicted boardings per stop is below:

```{r}
ggplot(lm_dat, aes(x=predicted_boardings)) + 
  geom_histogram() +
  theme_minimal() +
  labs(title = "Predicted Boardings")
```

```{r}
ggplot(lm_dat, aes(x=daily_boards, y = predicted_boardings)) +
  geom_point() +
  geom_smooth()

ggplot(lm_dat[daily_boards < 200], aes(x=daily_boards, y = predicted_boardings)) +
  geom_point() +
  geom_smooth()
```

There are several ways we can split this distribution to achieve 5 distinct market areas. However, a map of the continuous estimate is also informative:

```{r}
ggplot(lm_dat_sf %>% dplyr::filter(!is.na(predicted_boardings))) +
  geom_sf(aes(fill = predicted_boardings), lwd = 0) +
  scale_fill_viridis_c() +
  theme_minimal()
```

Here we can see the "bullseye" pattern we might expect

## Residuals

### 2019-12-03 rmds/poisson_regression.Rmd
add rmds folder

---
title: "Poisson Regression"
author: "Raven McKnight"
date: "12/3/2019"
output: 
  html_document: 
    toc: true
    toc_float: true
    theme: paper
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Before moving on to spatial models, we can work with a simpler intermediate step: Bayesian Poisson regression. We expect Poisson regression to be "better" than a simple linear regression for several reasons. First, ridership is a count variable so using a Poisson is more ture to the distribution of our response.

Second, it allows us to use an offset term. Offset terms are covariates with their parameters set to 1. They are used to scale the mean of a Poisson variable, like a rate. In disease mapping studies, the offset is often set to the total population of areal unit $i$, which is a natural maximum to disease counts. In our case, there is no obvious bounding maximum, but total population and total times a bus makes a stop in block group $i$ are natural places to start.

# The Model

Let $Y_i$ equal the average number of boardings in block group $i$ in 2017, and let $E_i$ equal the average number of times a bus stops in that block group. Then the Poisson regression can be written

\begin{align}
Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
\eta_i = log(\lambda_i) &= \alpha + \sum_{n=1}^{4}x_n\beta
\end{align}

where $\lambda_i$ is the relative "risk" of ridership given exposure to covariates $x\beta$.

We'll use the same four covariates as we used in the linear regression: population density, employment density, percent of households with no vehicle, and a walkability metric.

# Results

## Residuals

### 2019-12-03 rmds/spatial_mod.Rmd
add rmds folder

---
title: "Spatial Model"
author: "Raven McKnight"
date: "12/3/2019"
output: 
  html_document: 
    toc: true
    toc_float: true
    theme: paper
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

### 2019-12-03 scripts/03_get-mod-dat.R

@@ -18,7 +18,6 @@ options(tigris_class = 'sf')
 apc <- readRDS('data/metro-transit/apc-interpolated.RDS')
 setDT(apc)
 
-apc <- apc[, year := year(ymd(date_key))]
 apc[, site_id := as.character(site_id)]
 
 # assign stops to block groups ------------------

### 2019-12-03 scripts/07_join-mod-dat.R

@@ -3,14 +3,17 @@
 library(data.table)
 library(tigris)
 library(ggplot2)
+library(lubridate)
 
 options(tigris_class = "sf")
+options(tigris_use_cache = TRUE)
 
 counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
 bgs <- block_groups("MN", counties, 2016)
 
 mod_dat <- readRDS('data/modeling-dat/basic_mod_dat.RDS')
 setDT(mod_dat)
+mod_dat[, year := year(ymd(date_key))]
 
 ## linear regression -------------
 
@@ -20,6 +23,14 @@ lm_dat <- lm_dat[, .(daily_boards = mean(daily_boards, na.rm = T),
                      daily_stops = mean(daily_stops, na.rm = T)), keyby = .(GEOID)]
 lm_dat <- unique(lm_dat)
 
-lm_dat[, daily_boards_per_stop := daily_boards/daily_stops]
+cov <- readRDS('data/modeling-dat/ag_2017_logged_mod.RDS') # using unscaled, unlogged for interpretability for TMA
+setDT(cov)
 
-cov <- readRDS('data/modeling-dat/ag_2017_scaled_mod.RDS')
+lm_dat <- merge(lm_dat, cov, on = 'GEOID', all.x = TRUE)
+lm_dat[daily_boards == 0, daily_boards := 0.01]
+lm_dat[daily_stops == 0, daily_stops := 0.01]
+
+lm_dat[, log_daily_boards := log(daily_boards)]
+lm_dat[, log_daily_stops := log(daily_stops)]
+
+saveRDS(lm_dat, 'data/modeling-dat/lm_dat_logged.RDS')

### 2019-12-10 rmds/linear_regression.Rmd
work from weekend at home

@@
 ```{r setup}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
 
@@
 # should find better way to read data from rmd folder
-lm_dat <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/modeling-dat/lm_dat_logged.RDS')
+lm_dat <- readRDS('/Users/raven/Documents/honors/honors-project/data/modeling-dat/lm_dat_scaled.RDS')
 ```
@@
 We also include number of stops made in block group $i$ as a control for level of service.
+Each variable is scaled for comparison.
 
@@
 ```{r}
-lm1 <- lm(log_daily_boards ~ pop_density + emp_density + walkability + perc_no_veh + log_daily_stops, lm_dat, na.action = na.exclude)
+lm_dat <- lm_dat[!is.na(daily_boards) & !is.na(pop_density) & !is.na(emp_density) & !is.na(perc_no_veh) & !is.na(walkability) & !is.na(log_daily_stops)]
+lm1 <- lm(scale(log_daily_boards) ~ pop_density + emp_density + perc_no_veh + walkability + scale(log_daily_stops), lm_dat, na.action = na.exclude)
 
@@
 
-I'm guessing population density is insignificant because it's so correlated with employment.
+# Results
 
-# Results
-## Transit Market Areas
+## Predictions
 
 ```{r}
-lm_dat[, index := -3.79 + 0.06*pop_density + 0.22*emp_density + 0.21*walkability + 0.19*perc_no_veh + 1.18*log_daily_stops]
-
-lm_dat[, predicted_boardings := exp(index)]
-
-lm_dat_sf <- dplyr::left_join(bgs, lm_dat)
+lm_dat$predicted_log_scale_boards <- lm1$fitted.values
+ggplot(lm_dat, aes(x=scale(log_daily_boards), y = predicted_log_scale_boards)) +
+  geom_point() +
+  labs(title = "Actual versus fitted") +
+  theme_minimal() + 
+  xlim(-3, 3) + ylim(-3, 3)
 ```
 
-We can use the model above to define updated Transit Market Areas based on the number of boardings a block group can support, rather than an arbitary index value.
-The distribution of predicted boardings per stop is below:
-
-```{r}
-ggplot(lm_dat, aes(x=predicted_boardings)) + 
-  geom_histogram() +
-  theme_minimal() +
-  labs(title = "Predicted Boardings")
-```
-
-```{r}
-ggplot(lm_dat, aes(x=daily_boards, y = predicted_boardings)) +
-  geom_point() +
-  geom_smooth()
-
-ggplot(lm_dat[daily_boards < 200], aes(x=daily_boards, y = predicted_boardings)) +
-  geom_point() +
-  geom_smooth()
-```
-
+## Transit Market Areas
 There are several ways we can split this distribution to achieve 5 distinct market areas. However, a map of the continuous estimate is also informative:
@@
 ```{r}
-ggplot(lm_dat_sf %>% dplyr::filter(!is.na(predicted_boardings))) +
-  geom_sf(aes(fill = predicted_boardings), lwd = 0) +
+lm_dat$resid <- residuals(lm1)
+lm_dat_sf <- dplyr::left_join(bgs, lm_dat)
+
+ggplot(lm_dat_sf %>% dplyr::filter(!is.na(predicted_log_scale_boards))) +
+  geom_sf(aes(fill = predicted_log_scale_boards), lwd = 0) +
   scale_fill_viridis_c() +
@@
 
-Here we can see the "bullseye" pattern we might expect
+Here we can see the "bullseye" pattern we might expect.
 
 ## Residuals
+
+```{r}
+ggplot(lm_dat_sf %>% dplyr::filter(!is.na(resid))) + 
+  geom_sf(aes(fill = resid), lwd = 0) +
+  scale_fill_viridis_c() + 
+  theme_minimal() +
+  labs(title = "residuals")
+```
+
+```{r}
+library(spdep)
+resid_dat <- lm_dat_sf %>%
+  dplyr::filter(!is.na(resid))
+
+nb <- poly2nb(resid_dat)
+W <- nb2mat(nb, style = "B", zero.policy = TRUE)
+listW <- nb2listw(nb, zero.policy = TRUE)
+
+moran.mc(resid_dat$log_daily_boards, listW, nsim = 9999, zero.policy = TRUE)
+```
+
+```{r}
+moran.mc(resid_dat$resid, listW, nsim = 9999, zero.policy = TRUE) # completely spatially random
+```
+
+Moran's I suggests some spatial clustering.
+The "strength" of clustering is less than we might expect because transit ridership occurs on linear paths rather than in "bullseye" clusters.
+That said, the observed rank in the Moran simulation above suggests that the remaining clustering is worth addressing with an inherently spatial model!

### 2019-12-10 rmds/poisson_regression.Rmd
work from weekend at home

@@
 knitr::opts_chunk$set(echo = TRUE)
+
+#install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
+library(INLA)
+library(rstan)
+library(shinystan)
+library(data.table)
+library(ggplot2)
+library(rgdal)
+library(sf)
+library(tigris)
+library(bayesplot)
+
+options(mc.cores = parallel::detectCores())
+options(tigris_class = "sf")
+options(tigris_use_cache = TRUE)
+
+mod_dat <- readRDS('/Users/raven/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
+setDT(mod_dat)
+
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+bgs <- block_groups("MN", counties, 2016)
 ```
@@
 
-# The Model
+# The Simplest Model
 
-Let $Y_i$ equal the average number of boardings in block group $i$ in 2017, and let $E_i$ equal the average number of times a bus stops in that block group. Then the Poisson regression can be written
+Let $Y_i$ equal the average number of weekday boardings in block group $i$ in 2017, and let $E_i$ equal the average number of times a bus stops in that block group.
+Then the Poisson regression can be written
 
@@
 Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
-\eta_i = log(\lambda_i) &= \alpha + \sum_{n=1}^{4}x_n\beta
+\eta_i = log(\lambda_i) &= \mu + \sum_{n=1}^{4}x_n\beta\\
+\theta & \sim N(0, 1)
 \end{align}
 
-where $\lambda_i$ is the relative "risk" of ridership given exposure to covariates $x\beta$.
+where $\lambda_i$ is the relative "risk" of ridership given exposure to covariates $x\beta$ and \theta models extra-Poisson variance.
 
@@
 
+```{r}
+mod_dat <- mod_dat[!is.na(daily_boards) & !is.na(pop_density) & !is.na(emp_density) & !is.na(perc_no_veh) & !is.na(walkability) & !is.na(daily_stops)]
+
+y <- mod_dat$daily_boards
+E <- mod_dat$daily_stops
+
+N <- nrow(mod_dat)
+
+K <- 4
+x <- matrix(ncol = 4, c(mod_dat$pop_density, mod_dat$emp_density, mod_dat$perc_no_veh, mod_dat$walkability))
+
+standat1 <- list(y = y, E = E, x = x, K = K, N = N)
+
+poisson <- "/Users/raven/Documents/honors/honors-project/stan/poisson.stan"
+
+#poisson_fitsimp <- stan(poisson, data = standat1, warmup = 1000, iter = 2000)
+poisson_fitsimp <- readRDS("/Users/raven/Documents/honors/honors-project/fits/poisson_fitsimp.RDS")
+
+print(poisson_fitsimp, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]"), probs=c(0.025, 0.5, 0.975))
+#launch_shinystan(poisson_fitsimp)
+```
+
+## Predictions
+```{r}
+posterior <- extract(poisson_fitsimp)
+yrep <- posterior$y_rep
+yrep <- yrep[sample(nrow(yrep), 50), ]
+
+pp_check(y, yrep, ppc_dens_overlay) + 
+  theme_minimal() +
+  xlim(0, 200) +
+  labs(title = "Actual versus fitted")
+```
+
+```{r}
+mod_dat[, predicted_boards := rpois(lambda = daily_stops * exp(-1.13 + 0.06*pop_density + 0.04*emp_density + 0.16*perc_no_veh), n = nrow(mod_dat))]
+
+mod_dat_sf <- dplyr::left_join(bgs, mod_dat)
+
+ggplot(mod_dat_sf, aes(x=daily_boards, y = predicted_boards)) +
+  geom_point() +
+  theme_minimal() +
+  labs(title = "Actual versus predicted", subtitle = "Clearly overstimating outliers")
+
+ggplot(mod_dat_sf %>% dplyr::filter(daily_boards < 500), aes(x=daily_boards, y = predicted_boards)) +
+  geom_point() +
+  theme_minimal() +
+  labs(title = "Actual versus predicted", subtitle = "Clearly overstimating outliers") +
+  xlim(0, 500) + ylim(0, 500)
+```
+
+```{r}
+a <- ggplot(mod_dat_sf %>% dplyr::filter(!is.na(daily_boards) & daily_boards != 0), aes(fill=log(daily_boards))) + 
+  geom_sf(lwd = 0) +
+  scale_fill_viridis_c() + 
+  theme_minimal() +
+  labs(title = "Observed log daily boardings") +
+  theme(legend.position = 'none')
+
+b <- ggplot(mod_dat_sf %>% dplyr::filter(!is.na(daily_boards)), aes(fill=log(predicted_boards))) + 
+  geom_sf(lwd = 0) +
+  scale_fill_viridis_c() + 
+  theme_minimal() +
+  labs(title = "Predicted log daily boardings") +
+  theme(legend.position = 'none')
+
+gridExtra::grid.arrange(a, b, ncol = 2)
+```
+
+It looks like the simple Poisson regression is beginning to capture the correct spatial pattern, though it's underestimating some areas on the suburbs.
+
 ## Residuals
+
+```{r}
+ggplot(mod_dat_sf %>% dplyr::filter(!is.na(daily_boards)), aes(fill=log(daily_boards) - log(predicted_boards))) + 
+  geom_sf(lwd = 0) +
+  scale_fill_viridis_c() + 
+  theme_minimal() +
+  labs(title = "Residuals") +
+  theme(legend.position = 'none')
+```
+
+In general, it seems the model is failing in a few ways: overestimating ridership in high-ridership block groups, and underestimating the number of block groups with low ridership.
+
+Let's see if these residuals are spatially autocorrelated.
+
+```{r}
+library(spdep)
+
+resid_dat <- mod_dat_sf %>% 
+  dplyr::filter(!is.na(daily_boards))
+
+nb <- poly2nb(resid_dat)
+W <- nb2mat(nb, style = "B", zero.policy = TRUE)
+listW <- nb2listw(nb, zero.policy = TRUE)
+
+resid_dat <- resid_dat %>%
+  dplyr::mutate(resid = daily_boards - predicted_boards)
+
+moran.mc(resid_dat$resid, listW, nsim = 9999, zero.policy = TRUE)
+```
+
+# An Improvement
+
+One way we could improve this model is by explicitly modeling the extra-Poisson variance (or overdispersion) which may be contributing the the poor fit above.
+We can add a $\theta$ parameter to address this variation as follows
+
+\begin{align}
+Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
+\eta_i = log(\lambda_i) &= \mu + \sum_{n=1}^{4}x_n\beta + \theta \\
+\theta & \sim N(0, 1)
+\end{align}
+
+## Results
+
+```{r}
+poisson_theta <- "/Users/raven/Documents/honors/honors-project/stan/poisson_theta.stan"
+
+poisson_theta_fit <- stan(poisson_theta, data = standat1, warmup = 7000, iter = 8000)
+saveRDS(poisson_theta_fit, "/Users/raven/Documents/honors/honors-project/stan/poisson_theta_test.stan")
+
+beepr::beep("fanfare")
+
+print(poisson_theta_fit, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]", "theta[5]", "eta[5]", "lambda[5]"), probs=c(0.025, 0.5, 0.975))
+launch_shinystan(poisson_theta_fit)
+```
+
+```{r}
+fit <- rstan::extract(poisson_theta_fit)
+test <- rpois(fit$lambda, n = 1513)
+
+ggplot(mod_dat, aes(x=test)) +
+  geom_density(color = "red") +
+  geom_density(aes(x=daily_boards*365), color = "blue") + xlim(0, 500)
+```
+
+So this isn't great but I think it might just need more iterations? This is on the "run overnight" list.
+
+# Testing total annual rides
+
+```{r}
+standat1 <- list(y = y*365, E = E, x = x, K = K, N = N)
+
+poisson <- "/Users/raven/Documents/honors/honors-project/stan/poisson.stan"
+
+poisson_fit_year <- stan(poisson, data = standat1, warmup = 1000, iter = 2000)
+#poisson_fitsimp <- readRDS("/Users/raven/Documents/honors/honors-project/fits/poisson_fitsimp.RDS")
+beepr::beep("fanfare")
+print(poisson_fit_year, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]"), probs=c(0.025, 0.5, 0.975))
+```
+
+```{r}
+fityear <- rstan::extract(poisson_fit_year)
+testyear <- rpois(fityear$lambda, n = 1513)
+
+ggplot(mod_dat, aes(x=testyear)) +
+  geom_density(color = "red") +
+  geom_density(aes(x=daily_boards*365), color = "blue") + xlim(0, 3000)
+```
+
+# Testing negative
+
+```{r}
+negbinom <- "/Users/raven/Documents/honors/honors-project/stan/negbinom.stan"
+
+nb_fit <- stan(negbinom, data = standat1, warmup = 1000, iter = 2000, control = list(adapt_delta = 0.99))
+
+print(nb_fit, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]", "eta[5]", "lambda[5]", "scale"), probs=c(0.025, 0.5, 0.975))
+
+launch_shinystan(nb_fit)
+```

### 2019-12-10 rmds/spatial_mod.Rmd
work from weekend at home

@@
 knitr::opts_chunk$set(echo = TRUE)
+
+#install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
+library(INLA)
+library(rstan)
+library(shinystan)
+library(data.table)
+library(ggplot2)
+library(rgdal)
+library(sf)
+
+options(mc.cores = parallel::detectCores())
+options(tigris_class = "sf")
+options(tigris_use_cache = TRUE)
+
+mod_dat <- readRDS('/Users/raven/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
+setDT(mod_dat)
+
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+bgs <- block_groups("MN", counties)
+
+spat_dat <- dplyr::left_join(bgs, mod_dat)
+spat_dat <- spat_dat %>%
+  dplyr::filter(!is.na(daily_boards) & !is.na(walkability) & !is.na(perc_no_veh) & !is.na(pop_density) & !is.na(emp_density))
 ```
@@
 # Introduction
+
+We can improve the Poisson regression by modeling the residuals.
+As we saw, there is some spatial autocorrelation in the residuals.
+Additionally, the distribution of $Y_i$ is overdispersed, meaning the variance is greater than we would expect from a Poisson distribution.
+Therefore, we'll add *two* parameters to account for both pure, uncorrelated overdispersion ($\theta$) and spatially autocorrelated variance ($\phi$).
+
+## The Model
+For areal unit $i$, the spatial model can be defined
+
+Let $Y_i$ represent the average number of daily boardings in census block group $i$ (rounded to an integer). Then the spatial model can be defined
+
+\begin{align}
+Y_i|\theta_i & \sim \text{Poisson}(E_i\lambda_i) \\
+\eta_i = \text{log}(\lambda_i) &= \mu + x\beta + ((\sqrt{\rho / s})\phi + (\sqrt{1-\rho})\theta)\sigma  \\
+\phi &\sim N(0, Q^{-1}) \\
+\theta &\sim N(0, n) \\
+\rho & \sim \text{beta}(1/2, 1/2) \\
+\sigma &\sim \text{exponential }
+
+\end{align}
+
+where
+
+* E is an offset term, sometimes also called the baseline or "expected value" of Y (same as Poisson regression)
+* $\lambda_i$ is the relative "risk", and $\eta_i$ is the log relative risk
+* $\mu$ is the overall "risk" level, an intercept term
+* $x\beta$ are covariates which vary spatially
+* $\phi$ is the intrinsic conditional autoregressive (ICAR) model which accounts for spatially structured errors. In this case, we are using the BYM2 model
+* $\theta$ accounts for pure overdispersion. This parameter is the complement to $\phi$ and accounts for non-spatial errors. 
+* $\rho$ is a mixing parameter.
+It determines how much variance comes from spatially correlated versus independent error terms.
+The introduction of $\rho$ is a key difference between the BYM and BYM2 models.
+* $s$ is a scaling factor such that $Var(\phi_i) \approx 1$. We set this variance so that $\sigma$ is the actual standard deviation of the errors.
+* $n$ is the number of fully connected subgraphs in the neighborhood structure. $n=1$ for study areas with no regions with 0 neighbors. 
+* $\sigma \geq 0$ is the overall standard deviation for combined error terms
+
+# Results
+
+## One covariate, Ramsey County
+```{r}
+spat_dat <- spat_dat %>% 
+  dplyr::filter(COUNTYFP == 123)
+
+y = spat_dat$daily_boards
+E = spat_dat$daily_stops
+
+K = 1
+x = matrix(ncol = 1, c(spat_dat$emp_density))
+
+nb <- poly2nb(spat_dat)
+
+source("nb_helper_funs.R")
+nbs <- nb2graph(nb)
+N <- nbs$N
+node1 <- nbs$node1
+node2 <- nbs$node2
+N_edges <- nbs$N_edges
+scaling_factor <- scale_nb_components(nb)[1]
+
+standat1 <- list(N = N, N_edges = N_edges, node1 = node1, node2 = node2, y = y, E = E, scaling_factor = scaling_factor, K = K, x = x)
+
+bym2 <- "/Users/raven/Documents/honors/honors-project/stan/bym2.stan"
+
+#bym2_fit1 <- stan(bym2, data = standat1, warmup = 5000, iter = 6000, verbose = T)
+#saveRDS(bym2_fit1, "/Users/raven/Documents/honors/honors-project/fits/bym2_fit1.RDS")
+
+bym2_fit1 <- readRDS("/Users/raven/Documents/honors/honors-project/fits/bym2_fit1.RDS")
+```
+
+```{r}
+print(bym2_fit1, pars=c("beta0", "betas[1]", "rho", "sigma", "logit_rho", "mu[5]", "phi[5]", "theta[5]"), probs=c(0.025, 0.5, 0.975))
+#launch_shinystan(bym2_fit1)
+```
+
+I'm not sure if the problem with this is that I didn't run it enough, it's parameterized poorly, etc. Actually only took ~10 minutes to run.
+
+# All four covariates, Ramsey
+
+Using not enough iterations right now so I can see if it works
+```{r}
+x2 <- matrix(ncol = 4, c(spat_dat$pop_density, spat_dat$emp_density, spat_dat$perc_no_veh, spat_dat$walkability))
+K2 <- 4
+
+standat2 <- list(N = N, N_edges = N_edges, node1 = node1, node2 = node2, y = y, E = E, scaling_factor = scaling_factor, K = K2, x = x2)
+
+#bym2_fit2 <- stan(bym2, data = standat2, warmup = 2000, iter = 3000, thin = 1, verbose = T)
+
+#saveRDS(bym2_fit2, "/Users/raven/Documents/honors/honors-project/fits/bym2_fit2.RDS")
+
+bym2_fit2 <- readRDS("/Users/raven/Documents/honors/honors-project/fits/bym2_fit2.RDS")
+```
+
+```{r}
+#launch_shinystan(bym2_fit2)
+print(bym2_fit2, pars=c("beta0", "betas[1]", "betas[2]", "betas[3]", "betas[4]", "rho", "sigma", "logit_rho", "mu[5]", "phi[5]", "theta[5]"), probs=c(0.025, 0.5, 0.975))
+```
+
+Everything except $\rho$ seems to be alright..

### 2019-12-10 scripts/06_clean-covariates.R

@@ -102,9 +102,16 @@ setDT(logged_dat)
 saveRDS(logged_dat, 'data/modeling-dat/ag_2017_logged_mod.RDS')
 
 # scale
-scaled_dat <- lapply(logged_dat[, -c('GEOID')], scale)
+scaled_dat <- lapply(small_dat[, -c('GEOID')], scale)
 scaled_dat$GEOID <- logged_dat$GEOID
 scaled_dat <- as.data.table(scaled_dat)
 
 saveRDS(scaled_dat, 'data/modeling-dat/ag_2017_scaled_mod.RDS')
 
+
+# scale
+lscaled_dat <- lapply(logged_dat[, -c('GEOID')], scale)
+lscaled_dat$GEOID <- logged_dat$GEOID
+lscaled_dat <- as.data.table(lscaled_dat)
+
+saveRDS(scaled_dat, 'data/modeling-dat/ag_2017_logscaled_mod.RDS')

### 2019-12-10 scripts/07_join-mod-dat.R

@@ -23,7 +23,7 @@ lm_dat <- lm_dat[, .(daily_boards = mean(daily_boards, na.rm = T),
                      daily_stops = mean(daily_stops, na.rm = T)), keyby = .(GEOID)]
 lm_dat <- unique(lm_dat)
 
-cov <- readRDS('data/modeling-dat/ag_2017_logged_mod.RDS') # using unscaled, unlogged for interpretability for TMA
+cov <- readRDS('data/modeling-dat/ag_2017_scaled_mod.RDS') # using scaled but not logged
 setDT(cov)
 
 lm_dat <- merge(lm_dat, cov, on = 'GEOID', all.x = TRUE)
@@ -33,4 +33,14 @@ lm_dat[daily_stops == 0, daily_stops := 0.01]
 lm_dat[, log_daily_boards := log(daily_boards)]
 lm_dat[, log_daily_stops := log(daily_stops)]
 
-saveRDS(lm_dat, 'data/modeling-dat/lm_dat_logged.RDS')
+saveRDS(lm_dat, 'data/modeling-dat/lm_dat_scaled.RDS')
+
+## poisson regression ------------
+
+p_dat <- lm_dat[, -c('log_daily_boards', 'log_daily_stops')]
+p_dat[, daily_boards := as.integer(daily_boards)]
+p_dat[, daily_stops := as.integer(daily_stops)]
+
+saveRDS(p_dat, 'data/modeling-dat/p_dat_scaled.RDS')
+
+# unscaled for easier prediction?

### 2019-12-10 scripts/03_get-mod-dat.R
update data pull script

@@ -12,6 +12,7 @@ invisible(lapply(packages, library, character.only = TRUE))
 
 rm(miss_pkgs, packages)
 
+options(tigris_use_cache = TRUE)
 options(tigris_class = 'sf')
 
 # Read "raw" data, from "gap-fill.R" script
@@ -70,8 +71,10 @@ apc <- apc[line_id != 902 & line_id != 901]
 # exclude things outside of 7 county
 apc <- apc[!is.na(GEOID)]
 
+apc[, boardings_per_stop := mean(board, na.rm = T), keyby = .(date_key, site_id)]
+
 apc_ag <- apc[, .(daily_boards = sum(board, na.rm = T), daily_alights = sum(alight, na.rm = T), num_interpolated = sum(interpolated), 
-                  num_routes = length(unique(line_id)), daily_stops = .N), keyby = .(date_key, GEOID)]
+                  num_routes = length(unique(line_id)), daily_stops = .N, boardings_per_stop = mean(boardings_per_stop)), keyby = .(date_key, GEOID)]
 
 # more modeling specific things
 apc_ag[, wday := wday(ymd(date_key))]

### 2019-12-10 rmds/linear_regression.Rmd
some work with a slightly different response

@@
 # should find better way to read data from rmd folder
-lm_dat <- readRDS('/Users/raven/Documents/honors/honors-project/data/modeling-dat/lm_dat_scaled.RDS')
+lm_dat <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/modeling-dat/lm_dat_scaled.RDS')
 ```
@@
 lm_dat <- lm_dat[!is.na(daily_boards) & !is.na(pop_density) & !is.na(emp_density) & !is.na(perc_no_veh) & !is.na(walkability) & !is.na(log_daily_stops)]
-lm1 <- lm(scale(log_daily_boards) ~ pop_density + emp_density + perc_no_veh + walkability + scale(log_daily_stops), lm_dat, na.action = na.exclude)
+lm_dat[daily_boards_per_stop == 0, daily_boards_per_stop := 0.01]
+
+lm1 <- lm((scale(log(daily_boards_per_stop))) ~ pop_density + emp_density + perc_no_veh + walkability + scale(log_daily_stops), lm_dat, na.action = na.exclude)
 
@@
 lm_dat$predicted_log_scale_boards <- lm1$fitted.values
-ggplot(lm_dat, aes(x=scale(log_daily_boards), y = predicted_log_scale_boards)) +
-  geom_point() +
+ggplot(lm_dat, aes(x=scale(log(daily_boards_per_stop)), y = predicted_log_scale_boards)) +
+  geom_point(alpha = 0.5) +
+  geom_abline(yintercept = 0, slope = 1) +
   labs(title = "Actual versus fitted") +
-  theme_minimal() + 
-  xlim(-3, 3) + ylim(-3, 3)
+  theme_minimal() 
 ```

### 2019-12-10 rmds/poisson_regression.Rmd
some work with a slightly different response

@@
 
-mod_dat <- readRDS('/Users/raven/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
+mod_dat <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
 setDT(mod_dat)
@@
 ```{r}
-mod_dat <- mod_dat[!is.na(daily_boards) & !is.na(pop_density) & !is.na(emp_density) & !is.na(perc_no_veh) & !is.na(walkability) & !is.na(daily_stops)]
+mod_dat <- mod_dat[!is.na(daily_boards_per_stop) & !is.na(pop_density) & !is.na(emp_density) & !is.na(perc_no_veh) & !is.na(walkability) & !is.na(daily_stops)]
 
-y <- mod_dat$daily_boards
+y <- mod_dat$daily_boards_per_stop
 E <- mod_dat$daily_stops
@@
 
-poisson <- "/Users/raven/Documents/honors/honors-project/stan/poisson.stan"
+poisson <- "/Users/mcknigri/Documents/honors/honors-project/stan/poisson.stan"
 
-#poisson_fitsimp <- stan(poisson, data = standat1, warmup = 1000, iter = 2000)
-poisson_fitsimp <- readRDS("/Users/raven/Documents/honors/honors-project/fits/poisson_fitsimp.RDS")
+poisson_fitsimp <- stan(poisson, data = standat1, warmup = 1000, iter = 2000)
+#poisson_fitsimp <- readRDS("/Users/raven/Documents/honors/honors-project/fits/poisson_fitsimp.RDS")
 
 print(poisson_fitsimp, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]"), probs=c(0.025, 0.5, 0.975))
-#launch_shinystan(poisson_fitsimp)
+launch_shinystan(poisson_fitsimp)
 ```
@@
 ```{r}
-posterior <- extract(poisson_fitsimp)
+posterior <- rstan::extract(poisson_fitsimp)
 yrep <- posterior$y_rep
-yrep <- yrep[sample(nrow(yrep), 50), ]
+yrep <- yrep[sample(nrow(yrep), 100), ]
 
@@
   xlim(0, 200) +
-  labs(title = "Actual versus fitted")
+  labs(title = "Actual versus fitted") + xlim(-1, 5)
 ```
@@
 ```{r}
-poisson_theta <- "/Users/raven/Documents/honors/honors-project/stan/poisson_theta.stan"
+poisson_theta <- "/Users/mcknigri/Documents/honors/honors-project/stan/poisson_theta.stan"
 
-poisson_theta_fit <- stan(poisson_theta, data = standat1, warmup = 7000, iter = 8000)
+poisson_theta_fit <- stan(poisson_theta, data = standat1, warmup = 1000, iter = 2000)
 saveRDS(poisson_theta_fit, "/Users/raven/Documents/honors/honors-project/stan/poisson_theta_test.stan")
@@
 fit <- rstan::extract(poisson_theta_fit)
-test <- rpois(fit$lambda, n = 1513)
+theta_yrep <- fit$y_rep
+theta_yrep <- theta_yrep[sample(nrow(theta_yrep), 100), ]
 
-ggplot(mod_dat, aes(x=test)) +
-  geom_density(color = "red") +
-  geom_density(aes(x=daily_boards*365), color = "blue") + xlim(0, 500)
+pp_check(y, theta_yrep, ppc_dens_overlay) + 
+  theme_minimal() +
+  xlim(0, 200) +
+  labs(title = "Actual versus fitted") + xlim(-1, 5)
 ```

### 2019-12-10 scripts/07_join-mod-dat.R

@@ -20,7 +20,8 @@ mod_dat[, year := year(ymd(date_key))]
 lm_dat <- mod_dat[year == 2017]
 lm_dat <- lm_dat[wday != 6 & wday != 7]
 lm_dat <- lm_dat[, .(daily_boards = mean(daily_boards, na.rm = T), 
-                     daily_stops = mean(daily_stops, na.rm = T)), keyby = .(GEOID)]
+                     daily_stops = mean(daily_stops, na.rm = T), 
+                     daily_boards_per_stop = mean(boardings_per_stop)), keyby = .(GEOID)]
 lm_dat <- unique(lm_dat)
 
 cov <- readRDS('data/modeling-dat/ag_2017_scaled_mod.RDS') # using scaled but not logged
@@ -40,6 +41,7 @@ saveRDS(lm_dat, 'data/modeling-dat/lm_dat_scaled.RDS')
 p_dat <- lm_dat[, -c('log_daily_boards', 'log_daily_stops')]
 p_dat[, daily_boards := as.integer(daily_boards)]
 p_dat[, daily_stops := as.integer(daily_stops)]
+p_dat[, daily_boards_per_stop := as.integer(daily_boards_per_stop)]
 
 saveRDS(p_dat, 'data/modeling-dat/p_dat_scaled.RDS')
 

### 2019-12-13 rmds/poisson_regression.Rmd
work on poisson & spatial, negbinom.stan not in use

--- 2019-12-10 rmds/poisson_regression.Rmd
@@
 
-where $\lambda_i$ is the relative "risk" of ridership given exposure to covariates $x\beta$ and \theta models extra-Poisson variance.
+where $\lambda_i$ is the relative "risk" of ridership given exposure to covariates $x\beta$.
 
@@
 ```{r}
-mod_dat <- mod_dat[!is.na(daily_boards) & !is.na(pop_density) & !is.na(emp_density) & !is.na(perc_no_veh) & !is.na(walkability) & !is.na(daily_stops)]
+mod_dat <- mod_dat[!is.na(daily_boards_per_stop) & !is.na(pop_density) & !is.na(emp_density) & !is.na(perc_no_veh) & !is.na(walkability) & !is.na(daily_stops)]
 
@@
 #poisson_fitsimp <- stan(poisson, data = standat1, warmup = 1000, iter = 2000)
+#saveRDS(poisson_fitsimp, "/Users/raven/Documents/honors/honors-project/fits/poisson_fitsimp.RDS")
+
 poisson_fitsimp <- readRDS("/Users/raven/Documents/honors/honors-project/fits/poisson_fitsimp.RDS")
@@
 ```{r}
-posterior <- extract(poisson_fitsimp)
-yrep <- posterior$y_rep
-yrep <- yrep[sample(nrow(yrep), 50), ]
+posterior <- rstan::extract(poisson_fitsimp)
+yrep <- matrix(ncol = 1513, nrow = 100)
+
+for(i in 1:100){
+  yrep[i, ] <- rpois(1513, posterior$lambda*E)
+}
 
@@
   xlim(0, 200) +
-  labs(title = "Actual versus fitted")
+  labs(title = "Actual versus fitted") + xlim(0, 100)
 ```
@@
 
-Let's see if these residuals are spatially autocorrelated.
+Let's see if these residuals are spatially autocorrelated:
 
@@
 
+According to the Moran simulation, they're not spatially clustered, but we still haven't found a good way to measure "linear" clustering.
+
 # An Improvement
@@
 One way we could improve this model is by explicitly modeling the extra-Poisson variance (or overdispersion) which may be contributing the the poor fit above.
-We can add a $\theta$ parameter to address this variation as follows
+We can add a $\theta$ parameter (scaled by $\sigma$, following the bym2 model) to address this variation as follows
 
@@
 Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
-\eta_i = log(\lambda_i) &= \mu + \sum_{n=1}^{4}x_n\beta + \theta \\
+\eta_i = log(\lambda_i) &= \mu + \sum_{n=1}^{4}x_n\beta + \theta*\sigma \\
 \theta & \sim N(0, 1)
@@
 
-poisson_theta_fit <- stan(poisson_theta, data = standat1, warmup = 7000, iter = 8000)
-saveRDS(poisson_theta_fit, "/Users/raven/Documents/honors/honors-project/stan/poisson_theta_test.stan")
+#poisson_theta_fit <- stan(poisson_theta, data = standat1, warmup = 1000, iter = 2000)
+#saveRDS(poisson_theta_fit, "/Users/raven/Documents/honors/honors-project/fits/poisson_theta_test.RDS")
 
-beepr::beep("fanfare")
+poisson_theta_fit <- readRDS("/Users/raven/Documents/honors/honors-project/fits/poisson_theta_test.RDS")
 
 print(poisson_theta_fit, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]", "theta[5]", "eta[5]", "lambda[5]"), probs=c(0.025, 0.5, 0.975))
+
 launch_shinystan(poisson_theta_fit)
@@
 
+This also has a lot of autocorrelation in the chains, needs to be run for more iterations.
+
 ```{r}
-fit <- rstan::extract(poisson_theta_fit)
-test <- rpois(fit$lambda, n = 1513)
+posterior2 <- rstan::extract(poisson_theta_fit)
+yrep2 <- matrix(ncol = 1513, nrow = 100)
 
-ggplot(mod_dat, aes(x=test)) +
-  geom_density(color = "red") +
-  geom_density(aes(x=daily_boards*365), color = "blue") + xlim(0, 500)
+for(i in 1:100){
+  yrep2[i, ] <- rpois(1513, posterior2$lambda[i,]*E)
+}
+
+pp_check(y, yrep2, ppc_dens_overlay) + 
+  theme_minimal() +
+  xlim(0, 200) +
+  labs(title = "Actual versus fitted") + xlim(0, 100)
 ```
 
-So this isn't great but I think it might just need more iterations? This is on the "run overnight" list.
-
-# Testing total annual rides
-
-```{r}
-standat1 <- list(y = y*365, E = E, x = x, K = K, N = N)
-
-poisson <- "/Users/raven/Documents/honors/honors-project/stan/poisson.stan"
-
-poisson_fit_year <- stan(poisson, data = standat1, warmup = 1000, iter = 2000)
-#poisson_fitsimp <- readRDS("/Users/raven/Documents/honors/honors-project/fits/poisson_fitsimp.RDS")
-beepr::beep("fanfare")
-print(poisson_fit_year, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]"), probs=c(0.025, 0.5, 0.975))
-```
-
-```{r}
-fityear <- rstan::extract(poisson_fit_year)
-testyear <- rpois(fityear$lambda, n = 1513)
-
-ggplot(mod_dat, aes(x=testyear)) +
-  geom_density(color = "red") +
-  geom_density(aes(x=daily_boards*365), color = "blue") + xlim(0, 3000)
-```
-
-# Testing negative
-
-```{r}
-negbinom <- "/Users/raven/Documents/honors/honors-project/stan/negbinom.stan"
-
-nb_fit <- stan(negbinom, data = standat1, warmup = 1000, iter = 2000, control = list(adapt_delta = 0.99))
-
-print(nb_fit, pars = c("mu", "betas[1]", "betas[2]", "betas[3]", "betas[4]", "eta[5]", "lambda[5]", "scale"), probs=c(0.025, 0.5, 0.975))
-
-launch_shinystan(nb_fit)
-```
+The next improvement would be to determine roughly how much of the variation currently modeled by $\theta * \sigma$ is due to unmeasured covariates vs pure overdispersion vs spatial variance.

### 2019-12-13 rmds/spatial_mod.Rmd
work on poisson & spatial, negbinom.stan not in use

@@
 
-```{r setup, include=FALSE}
+```{r setup, include=FALSE, warning = FALSE, message = FALSE}
 knitr::opts_chunk$set(echo = TRUE)
@@
 library(sf)
+library(tigris)
+library(spdep)
+library(bayesplot)
 
@@
 
-bym2 <- "/Users/raven/Documents/honors/honors-project/stan/bym2.stan"
-
-#bym2_fit1 <- stan(bym2, data = standat1, warmup = 5000, iter = 6000, verbose = T)
-#saveRDS(bym2_fit1, "/Users/raven/Documents/honors/honors-project/fits/bym2_fit1.RDS")
-
-bym2_fit1 <- readRDS("/Users/raven/Documents/honors/honors-project/fits/bym2_fit1.RDS")
+# bym2 <- "/Users/raven/Documents/honors/honors-project/stan/bym2.stan"
+# 
+# #bym2_fit1 <- stan(bym2, data = standat1, warmup = 5000, iter = 6000, verbose = T)
+# #saveRDS(bym2_fit1, "/Users/raven/Documents/honors/honors-project/fits/bym2_fit1.RDS")
+# 
+# bym2_fit1 <- readRDS("/Users/raven/Documents/honors/honors-project/fits/bym2_fit1.RDS")
 ```
 
-```{r}
-print(bym2_fit1, pars=c("beta0", "betas[1]", "rho", "sigma", "logit_rho", "mu[5]", "phi[5]", "theta[5]"), probs=c(0.025, 0.5, 0.975))
-#launch_shinystan(bym2_fit1)
-```
+<!-- ```{r} -->
+<!-- print(bym2_fit1, pars=c("beta0", "betas[1]", "rho", "sigma", "logit_rho", "mu[5]", "phi[5]", "theta[5]"), probs=c(0.025, 0.5, 0.975)) -->
+<!-- #launch_shinystan(bym2_fit1) -->
+<!-- ``` -->
 
-I'm not sure if the problem with this is that I didn't run it enough, it's parameterized poorly, etc. Actually only took ~10 minutes to run.
+<!-- I'm not sure if the problem with this is that I didn't run it enough, it's parameterized poorly, etc. Actually only took ~10 minutes to run.  -->
 
-# All four covariates, Ramsey
+# All four covariates, Ramsey County
 
-Using not enough iterations right now so I can see if it works
 ```{r}
@@
 ```{r}
-#launch_shinystan(bym2_fit2)
+launch_shinystan(bym2_fit2)
 print(bym2_fit2, pars=c("beta0", "betas[1]", "betas[2]", "betas[3]", "betas[4]", "rho", "sigma", "logit_rho", "mu[5]", "phi[5]", "theta[5]"), probs=c(0.025, 0.5, 0.975))
@@
 
-Everything except $\rho$ seems to be alright..
+Some of these terms haven't fully converged but I've only run the chains for 2000 iterations.
+Using `shinystan`, I can see that there's a lot of autocorrelation in the chains.
+I think if there are no other red flags I can run this for many more iterations for a good result.
+
+```{r}
+posteriorbym2 <- rstan::extract(bym2_fit2)
+yrepbym2 <- matrix(ncol = 357, nrow = 100)
+
+for(i in 1:100){
+  yrepbym2[i, ] <- rpois(357, posteriorbym2$mu[i, ]) # reparameterize
+}
+
+pp_check(y, yrepbym2, ppc_dens_overlay) + 
+  theme_minimal() +
+  xlim(0, 200) +
+  labs(title = "Actual versus fitted") + xlim(0, 100)
+```
+
+This fit should also be improved once I finish variable selection.

### 2019-12-13 rmds/poisson_regression.Rmd
first attempt at bayesian lasso - not working yet

@@
 
-y <- mod_dat$daily_boards
-E <- mod_dat$daily_stops
+xdat <- mod_dat[, -c("GEOID", "daily_boards_per_stop", "sqkm", "perc_transit_comm")] # still need to fix that var
+xdat <- na.omit(xdat)
 
-N <- nrow(mod_dat)
+y <- xdat$daily_boards
+E <- xdat$daily_stops
 
-K <- 4
-x <- matrix(ncol = 4, c(mod_dat$pop_density, mod_dat$emp_density, mod_dat$perc_no_veh, mod_dat$walkability))
+N <- nrow(xdat)
+
+K <- 27
+x <- as.matrix(xdat)
 
@@
 ```{r}
-poisson_theta <- "/Users/raven/Documents/honors/honors-project/stan/poisson_theta.stan"
+poisson_theta <- "/Users/mcknigri/Documents/honors/honors-project/stan/poisson_theta.stan"
 
-#poisson_theta_fit <- stan(poisson_theta, data = standat1, warmup = 1000, iter = 2000)
+poisson_theta_fit_LASSO <- stan(poisson_theta, data = standat1, warmup = 7000, iter = 8000)
 #saveRDS(poisson_theta_fit, "/Users/raven/Documents/honors/honors-project/fits/poisson_theta_test.RDS")
@@
 
-launch_shinystan(poisson_theta_fit)
+launch_shinystan(poisson_theta_fit_LASSO)
 ```

### 2020-01-09 scripts/aws/poissonLASSO.R
update pathname

@@ -5,7 +5,7 @@ library(dplyr)
 options(mc.cores = parallel::detectCores())
 
 ## data
-mod_dat <- readRDS('~/honors-project/data/modeling-dat/p_dat_scaled.RDS')
+mod_dat <- readRDS('~/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
 
 # prep for stan
 xdat <- mod_dat %>%
@@ -25,6 +25,6 @@ standat1 <- list(y = y, E = E, x = x, K = K, N = N)
 ## fitting
 
 # the model code
-poisson_theta_lasso <- "~/honors-project/stan/poisson_theta_lasso.stan"
+poisson_theta_lasso <- "~/Documents/honors/honors-project/stan/poisson_theta_lasso.stan"
 
-poisson_theta_fit_LASSO <- stan(poisson_theta_lasso, data = standat1, warmup = 10, iter = 20, verbose = T)
+poisson_theta_fit_LASSO <- stan(poisson_theta_lasso, data = standat1, warmup = 1000, iter = 2000, verbose = T)

### 2020-01-09 rmds/linear_regression.Rmd
small updates

@@
 # should find better way to read data from rmd folder
-lm_dat <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/modeling-dat/lm_dat_scaled.RDS')
+lm_dat <- readRDS('~/Documents/honors/honors-project/data/modeling-dat/lm_dat_scaled.RDS')
 ```
@@
 lm_dat[daily_boards_per_stop == 0, daily_boards_per_stop := 0.01]
+
+lm_dat_small <- lm_dat %>% 
+  dplyr::select(-c(GEOID, daily_boards_per_stop, sqkm, perc_transit_comm, log_daily_boards, log_daily_stops))
 

### 2020-01-09 scripts/aws/poissonLASSO.R

@@ -27,4 +27,4 @@ standat1 <- list(y = y, E = E, x = x, K = K, N = N)
 # the model code
 poisson_theta_lasso <- "~/Documents/honors/honors-project/stan/poisson_theta_lasso.stan"
 
-poisson_theta_fit_LASSO <- stan(poisson_theta_lasso, data = standat1, warmup = 1000, iter = 2000, verbose = T)
+poisson_theta_fit_LASSO <- stan(poisson_theta_lasso, data = standat1, warmup = 100, iter = 200, verbose = T)

### 2020-01-09 scripts/aws/poissonLASSO.R
add poisson_lasso script

@@ -24,6 +24,10 @@ standat1 <- list(y = y, E = E, x = x, K = K, N = N)
 
 ## fitting
 
+# simpler
+poisson_lasso <- "~/Documents/honors/honors-project/stan/poisson_lasso.stan"
+poisson_fit_LASSO <- stan(poisson_theta, data = standat1, warmup = 100, iter = 200, verbose = T)
+
 # the model code
 poisson_theta_lasso <- "~/Documents/honors/honors-project/stan/poisson_theta_lasso.stan"
 

### 2020-01-10 scripts/aws/poissonLASSO.R
non-working updates

@@ -3,6 +3,7 @@
 library(rstan)
 library(dplyr)
 options(mc.cores = parallel::detectCores())
+rstan_options(autowrite = T)
 
 ## data
 mod_dat <- readRDS('~/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
@@ -22,11 +23,19 @@ x <- as.matrix(xdat)
 
 standat1 <- list(y = y, E = E, x = x, K = K, N = N)
 
+standat2 <- list(y = y, E = E, x = x, K = K, N = N,
+                 scale_alpha = 10, df_tau = 1, scale_tau = 1, rate_sigma = 1)
+
 ## fitting
 
 # simpler
 poisson_lasso <- "~/Documents/honors/honors-project/stan/poisson_lasso.stan"
-poisson_fit_LASSO <- stan(poisson_theta, data = standat1, warmup = 100, iter = 200, verbose = T)
+poisson_fit_LASSO <- stan(poisson_lasso, data = standat2, warmup = 1000, iter = 2000, verbose = T)
+
+# ridge?
+
+poisson_ridge <- "~/Documents/honors/honors-project/stan/poisson_ridge.stan"
+poisson_fit_ridge <- stan(poisson_ridge, data = standat1, warmup = 100, iter = 200, verbose = T)
 
 # the model code
 poisson_theta_lasso <- "~/Documents/honors/honors-project/stan/poisson_theta_lasso.stan"

### 2020-01-10 scripts/aws/poisson.R
switching to poisson_log_glm

@@ -24,7 +24,11 @@ standat1 <- list(y = y, E = E, x = x, K = K, N = N)
 
 ## fitting
 
-# simpler
+# simplest
+poisson <- "~/Documents/honors/honors-project/stan/poisson.stan"
+poisson_fit <- stan(poisson, data = standat1, warmup = 1000, iter = 2000, verbose = T)
+
+# simple lasso
 poisson_lasso <- "~/Documents/honors/honors-project/stan/poisson_lasso.stan"
 poisson_fit_LASSO <- stan(poisson_theta, data = standat1, warmup = 100, iter = 200, verbose = T)
 

### 2020-01-16 scripts/04_get-census.R
add lots of work with stan over past few days

@@ -169,7 +169,7 @@ saveRDS(ten_veh, 'data/covariates/housing-and-vehicles.RDS')
 acs_emp <- map_dfr(
   years,
   ~ get_acs(
-    geography = "block group",
+    geography = "tract",
     variables = c(transit_commute = "B08006_008",
                   total_commute = "B08006_001"),
     state = "MN",

### 2020-01-16 scripts/06_clean-covariates.R

@@ -24,7 +24,6 @@ house_veh <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariat
 language <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/language.RDS')
 nativity <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/nativity.RDS')
 wac <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/wac/all-wac.RDS')
-rac <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/rac/all-rac.RDS')
 acs_emp <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/acs-emp.RDS')
 walk <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/walkability.RDS')
 setDT(educ)
@@ -32,7 +31,6 @@ setDT(house_veh)
 setDT(language)
 setDT(nativity)
 setDT(wac)
-setDT(rac)
 setDT(acs_emp)
 
 acs[, year := NULL]
@@ -51,19 +49,15 @@ wac <- wac[year == 2017]
 wac <- wac[, c("w_total_jobs_here", "GEOID", "w_perc_jobs_white", "w_perc_jobs_men", "w_perc_jobs_no_college", 
                 "w_perc_jobs_less40", "w_perc_jobs_age_less30")]
  
-rac <- rac[year == 2017]
-rac <- rac[, c("GEOID", "tot_jobs", "perc_jobs_white", "perc_jobs_men", "perc_jobs_no_college", 
-                "perc_jobs_less40", "perc_jobs_age_less30")]
- 
 mod_dat <- merge(mod_dat, wac, by = 'GEOID', all.x = TRUE)
-mod_dat <- merge(mod_dat, rac, by = 'GEOID', all.x = TRUE)
+
 acs_emp <- acs_emp[year == 3, c("GEOID", "perc_transit_comm")]
 setnames(acs_emp, "GEOID", "tract_GEOID")
 mod_dat <- merge(mod_dat, acs_emp, by = 'tract_GEOID', all.x = TRUE)
 
-# for some reason, currently only have 1314 walkability observations?
-mod_dat <- merge(mod_dat, walk[, c('GEOID', 'walkability')], by = 'GEOID', all.x = TRUE)
-mod_dat[, walkability := as.numeric(walkability)]
[…]

### 2020-01-16 scripts/07_join-mod-dat.R

@@ -20,8 +20,7 @@ mod_dat[, year := year(ymd(date_key))]
 lm_dat <- mod_dat[year == 2017]
 lm_dat <- lm_dat[wday != 6 & wday != 7]
 lm_dat <- lm_dat[, .(daily_boards = mean(daily_boards, na.rm = T), 
-                     daily_stops = mean(daily_stops, na.rm = T), 
-                     daily_boards_per_stop = mean(boardings_per_stop)), keyby = .(GEOID)]
+                     daily_stops = mean(daily_stops, na.rm = T)), keyby = .(GEOID)]
 lm_dat <- unique(lm_dat)
 
 cov <- readRDS('data/modeling-dat/ag_2017_scaled_mod.RDS') # using scaled but not logged
@@ -41,8 +40,5 @@ saveRDS(lm_dat, 'data/modeling-dat/lm_dat_scaled.RDS')
 p_dat <- lm_dat[, -c('log_daily_boards', 'log_daily_stops')]
 p_dat[, daily_boards := as.integer(daily_boards)]
 p_dat[, daily_stops := as.integer(daily_stops)]
-p_dat[, daily_boards_per_stop := as.integer(daily_boards_per_stop)]
 
 saveRDS(p_dat, 'data/modeling-dat/p_dat_scaled.RDS')
-
-# unscaled for easier prediction?

### 2020-01-16 scripts/aws/poisson.R

@@ -2,46 +2,86 @@
 
 library(rstan)
 library(dplyr)
+library(data.table)
 options(mc.cores = parallel::detectCores())
 rstan_options(autowrite = T)
 
 ## data
 mod_dat <- readRDS('~/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
+mod_dat <- na.omit(mod_dat)
+setDT(mod_dat)
 
-# prep for stan
-xdat <- mod_dat %>%
-  dplyr::select(-c(GEOID, daily_boards_per_stop, sqkm, perc_transit_comm))
-xdat <- na.omit(xdat)
+xdat <- mod_dat[, -c('GEOID', 'daily_boards', 'daily_stops', 'sqkm', 'estimate_tot_pop')]
 
-y <- xdat$daily_boards
-E <- xdat$daily_stops
+## regular poisson regression -------------------
 
-N <- nrow(xdat)
+y <- mod_dat$daily_boards
+E <- mod_dat$daily_stops
 
-K <- 27
-x <- as.matrix(xdat)
+N <- nrow(mod_dat)
+
+K <- ncol(xdat)
+x <- as.matrix(ncol = K, xdat)
 
 standat1 <- list(y = y, E = E, x = x, K = K, N = N)
 
-standat2 <- list(y = y, E = E, x = x, K = K, N = N,
-                 scale_alpha = 10, df_tau = 1, scale_tau = 1, rate_sigma = 1)
+## basic poisson --------------------------------
 
-## fitting
-
-# simplest
 poisson <- "~/Documents/honors/honors-project/stan/poisson.stan"
 poisson_fit <- stan(poisson, data = standat1, warmup = 1000, iter = 2000, verbose = T)
 
-# simple lasso
-poisson_lasso <- "~/Documents/honors/honors-project/stan/poisson_lasso.stan"
-poisson_fit_LASSO <- stan(poisson_lasso, data = standat2, warmup = 1000, iter = 2000, verbose = T)
+poisson_fit <- readRDS("~/Documents/honors/honors-project/final-fits/poisson.RDS")
 
-# ridge?
+#shinystan::launch_shinystan(poisson_fit)
 
-poisson_ridge <- "~/Documents/honors/honors-project/stan/poisson_ridge.stan"
-poisson_fit_ridge <- stan(poisson_ridge, data = standat1, warmup = 100, iter = 200, verbose = T)
+## add overdispersion parameter -----------------
 
-# the model code
[…]

### 2020-01-16 chapters/01_introduction.Rmd
add chapters folder

---
title: "01_introduction"
author: "Raven McKnight"
date: "1/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2020-01-16 scripts/03_get-mod-dat.R
thursday at work

@@ -71,17 +71,11 @@ apc <- apc[line_id != 902 & line_id != 901]
 # exclude things outside of 7 county
 apc <- apc[!is.na(GEOID)]
 
-apc[, boardings_per_stop := mean(board, na.rm = T), keyby = .(date_key, site_id)]
+apc_ag <- apc[year(ymd(date_key)) == 2017, .(year_boards = sum(board, na.rm = T), year_alights = sum(alight, na.rm = T), num_interpolated = sum(interpolated), 
+                  num_routes = length(unique(line_id)), year_stops = .N), keyby = .(GEOID)]
 
-apc_ag <- apc[, .(daily_boards = sum(board, na.rm = T), daily_alights = sum(alight, na.rm = T), num_interpolated = sum(interpolated), 
-                  num_routes = length(unique(line_id)), daily_stops = .N, boardings_per_stop = mean(boardings_per_stop)), keyby = .(date_key, GEOID)]
+apc_ag[, year_activity := year_boards + year_alights, keyby = .(GEOID)]
 
-# more modeling specific things
-apc_ag[, wday := wday(ymd(date_key))]
-apc_ag[, daily_activity := daily_boards + daily_alights, keyby = .(date_key, GEOID)]
-apc_ag[, total_daily_activity := sum(daily_activity, na.rm =T), keyby = .(date_key)]
-
-
-saveRDS(apc_ag, 'data/modeling-dat/basic_mod_dat.RDS')
+saveRDS(apc_ag, 'data/modeling-dat/basic_mod_dat_ann17.RDS')
 
 

### 2020-01-16 scripts/04_get-census.R

@@ -204,72 +204,7 @@ saveRDS(tot_jobs, 'data/covariates/tot_jobs.RDS')
 
 # maybe better is rac data
 
-## get RAC data -------------------------------------------
 
-# residence area characteristics
-# type = JT00 is all jobs
-# year = most recent year
-
-rac_urls <- c('https://lehd.ces.census.gov/data/lodes/LODES7/mn/rac/mn_rac_S000_JT00_2017.csv.gz',
-              'https://lehd.ces.census.gov/data/lodes/LODES7/mn/rac/mn_rac_S000_JT00_2016.csv.gz',
-              'https://lehd.ces.census.gov/data/lodes/LODES7/mn/rac/mn_rac_S000_JT00_2015.csv.gz')
-
-for(i in 1:length(years)){
-  download.file(rac_urls[i], paste0('data/covariates/rac/mn_rac_', years[i], '.csv.gz'))
-}
-
-rac_files <- list.files(path = 'data/covariates/rac/')
-l <- lapply(paste0('data/covariates/rac/', rac_files), fread)
-
-l[[1]]$year <- 2017
-l[[2]]$year <- 2016
-l[[3]]$year <- 2015
-
-for(i in 1:3){
-  colnames(l[[i]]) <- c('h_geocode', 'tot_jobs', 'age_29', 'age_30_54', 'age_55',
-                      'wage_1250', 'wage_1250_3333', 'wage_3333', 'ag', 'mining',
-                      'utils', 'construction', 'manuf', 'whole_trade', 'ret_trade',
-                      'transp', 'info', 'finance', 'real_estate', 'prof_serv', 
-                      'manage', 'admin', 'educ', 'health', 'arts', 'ac_food', 
-                      'other', 'pub_admin', 'white_alone', 'black', 'amind', 
-                      'asian', 'native', 'two_or_more', 'not_hisp', 'hisp', 
-                      'less_than_hs', 'hs', 'some_college', 'degree', 'male', 'female', 
-                      'createdate', 'year')
-}
-
-# i think i'll mostly be interested in the service industry
-
-# aggregate blocks to bgs
-
-for(i in 1:3){
-  l[[i]][, GEOID := substr(h_geocode, 1, 12)]
-}
-
-for(i in 1:3){
-  l[[i]]$createdate <- NULL
[…]

### 2020-01-16 scripts/06_clean-covariates.R

@@ -15,17 +15,17 @@ rm(miss_pkgs, packages)
 options(tigris_class = 'sf')
 
 # add basic acs
-acs <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/basic_acs.RDS')
+acs <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/covariates/basic_acs.RDS')
 setDT(acs)
 acs <- acs[year == 3]
 
-educ <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/education.RDS')
-house_veh <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/housing-and-vehicles.RDS')
-language <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/language.RDS')
-nativity <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/nativity.RDS')
-wac <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/wac/all-wac.RDS')
-acs_emp <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/acs-emp.RDS')
-walk <- readRDS('/Users/raven/Documents/honors/honors-project/data/covariates/walkability.RDS')
+educ <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/covariates/education.RDS')
+house_veh <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/covariates/housing-and-vehicles.RDS')
+language <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/covariates/language.RDS')
+nativity <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/covariates/nativity.RDS')
+wac <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/covariates/wac/all-wac.RDS')
+acs_emp <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/covariates/acs-emp.RDS')
+walk <- readRDS('/Users/mcknigri/Documents/honors/honors-project/data/covariates/walkability.RDS')
 setDT(educ)
 setDT(house_veh)
 setDT(language)

### 2020-01-16 scripts/07_join-mod-dat.R

@@ -11,34 +11,25 @@ options(tigris_use_cache = TRUE)
 counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
 bgs <- block_groups("MN", counties, 2016)
 
-mod_dat <- readRDS('data/modeling-dat/basic_mod_dat.RDS')
+mod_dat <- readRDS('data/modeling-dat/basic_mod_dat_ann17.RDS')
 setDT(mod_dat)
-mod_dat[, year := year(ymd(date_key))]
 
-## linear regression -------------
-
-lm_dat <- mod_dat[year == 2017]
-lm_dat <- lm_dat[wday != 6 & wday != 7]
-lm_dat <- lm_dat[, .(daily_boards = mean(daily_boards, na.rm = T), 
-                     daily_stops = mean(daily_stops, na.rm = T)), keyby = .(GEOID)]
-lm_dat <- unique(lm_dat)
 
 cov <- readRDS('data/modeling-dat/ag_2017_scaled_mod.RDS') # using scaled but not logged
 setDT(cov)
 
-lm_dat <- merge(lm_dat, cov, on = 'GEOID', all.x = TRUE)
-lm_dat[daily_boards == 0, daily_boards := 0.01]
-lm_dat[daily_stops == 0, daily_stops := 0.01]
+lm_dat <- merge(mod_dat, cov, on = 'GEOID', all.x = TRUE)
+lm_dat[year_boards == 0, year_boards := 0.01]
+lm_dat[year_stops == 0, year_stops := 0.01]
 
-lm_dat[, log_daily_boards := log(daily_boards)]
-lm_dat[, log_daily_stops := log(daily_stops)]
+lm_dat[, log_year_boards := log(year_boards)]
+lm_dat[, log_year_stops := log(year_stops)]
 
 saveRDS(lm_dat, 'data/modeling-dat/lm_dat_scaled.RDS')
 
 ## poisson regression ------------
 
-p_dat <- lm_dat[, -c('log_daily_boards', 'log_daily_stops')]
-p_dat[, daily_boards := as.integer(daily_boards)]
-p_dat[, daily_stops := as.integer(daily_stops)]
+p_dat <- lm_dat[, -c('log_year_boards', 'log_year_stops')]
+
 
 saveRDS(p_dat, 'data/modeling-dat/p_dat_scaled.RDS')

### 2020-01-16 scripts/aws/poisson.R

@@ -11,35 +11,33 @@ mod_dat <- readRDS('~/Documents/honors/honors-project/data/modeling-dat/p_dat_sc
 mod_dat <- na.omit(mod_dat)
 setDT(mod_dat)
 
-xdat <- mod_dat[, -c('GEOID', 'daily_boards', 'daily_stops', 'sqkm', 'estimate_tot_pop')]
+xdat <- mod_dat[, -c('GEOID', 'year_boards', 'year_stops', 'sqkm', 'estimate_tot_pop', 'year_alights', 'num_interpolated', 'num_routes', 'year_activity')]
 
 ## regular poisson regression -------------------
 
-y <- mod_dat$daily_boards
-E <- mod_dat$daily_stops
+y <- mod_dat$year_boards
+E <- mod_dat$year_stops
 
 N <- nrow(mod_dat)
 
 K <- ncol(xdat)
 x <- as.matrix(ncol = K, xdat)
 
-standat1 <- list(y = y, E = E, x = x, K = K, N = N)
+standat1 <- list(y = as.integer(y/365), E = E, x = x, K = K, N = N)
 
 ## basic poisson --------------------------------
 
 poisson <- "~/Documents/honors/honors-project/stan/poisson.stan"
 poisson_fit <- stan(poisson, data = standat1, warmup = 1000, iter = 2000, verbose = T)
 
-poisson_fit <- readRDS("~/Documents/honors/honors-project/final-fits/poisson.RDS")
-
-#shinystan::launch_shinystan(poisson_fit)
+saveRDS(poisson_fit, '~/Documents/honors/honors-project/final-fits/poisson_fit.RDS')
 
 ## add overdispersion parameter -----------------
 
 poisson_theta <- "~/Documents/honors/honors-project/stan/poisson_theta.stan"
-poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 10000, verbose = T)
+poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 2000, verbose = T)
 
-#shinystan::launch_shinystan(poisson_theta_fit)
+ #shinystan::launch_shinystan(poisson_theta_fit)
 
 ## add horseshoe priors ------------------------0
 

### 2020-01-21 chapters/03_methods.Rmd
add chapters folder
add initial writing

--- 2020-01-16 chapters/01_introduction.Rmd
@@
 ---
-title: "01_introduction"
+title: "Methods"
 author: "Raven McKnight"
 date: "1/16/2020"
-output: html_document
+output: 
+  html_document:
+    toc: true
+    toc_float: true
+    theme: paper
 ---
@@
 
-## R Markdown
+# Poisson Regression
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+Each model presented in this honors is built upon a simple Poisson regression.
+Poisson regressions are used to model count data and assume that the natural log of the mean of response variable $Y$ can be modeled by a linear combination of predictors.
+Poisson regressions are a form of generalized linear models with the natural log as its link function.
+For response variable $Y$ and covariates $x_1, x_2, ..., x_k$, the basic Poisson regression can be written
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+\begin{align}
+\tag{1}
+Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
+\eta_i  = log(\lambda_i) & = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k \\
+\end{align}
 
-```{r cars}
-summary(cars)
-```
+where $\beta_0$ is an optional intercept term and $E_i$ is an offset term.
+The offset $E_i$ is often termed "exposure" in applications such as disease risk modeling where the offset may be the number of observed cases in a previous year, etc.
 
-## Including Plots
+The most obvious assumption made in a Poisson regression is that response variable $Y$ follows a Poisson distribution.
+The primary assumption this makes about our data is that $\text{Var}(Y) = \text{E}(Y)$.
+When this assumption is not met, we can either use a generalization of the Poisson distribution, such as a Negative-Binomial distribution, or add a parameter to account for extra-Poisson variance.
 
-You can also embed plots, for example:
+For the purposes of this study, we can add a set of heterogeneous random effects $\theta$ to account for overdispersion (where Var($Y$) > E($Y$)).
+Following **cite mitzi morris**, the second line of **(1)** can be written
 
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
+\begin{align}
+\tag{2}
+\eta_i = log(\lambda_i) & = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + \theta \sigma
+\end{align}
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+where $\theta$ and $\sigma$ are given normal priors centered around 0, such as $\theta \sim$ Normal(0, 1) and $\sigma \sim$ Normal(0, 5).
+The addition of these random effects improve fits on data with higher-than-expected variance.
+
+# Horseshoe Priors and Variable Selection
+
+Often, the coefficients $\beta_1, ..., \beta_k$ are given simple normal priors such as $\beta_k \sim$ Normal(0, 1) in the case of standardized parameters.
+When parameter vector $x$ contains many possible covariates, we may reasonably assume that some entries in $\beta$ are 0 -- in other words, that some of the parameters in $x$ have no effect on $Y$.
+In such a case, it is beneficial to use *sparsity inducing priors* on $\beta$ to ensure that irrelevant parameter coefficients are shrunk to 0.
+
+In frequentist applications, methods such as the LASSO (Least Absolute Shrinkage and Selection Operator) are employed for this type of variable selection.
+In the Bayesian setting, shrinkage and regularization can be performed using a variety of different priors on $\beta$.
+**Cite van Erp etc** provide an introduction to many such continuous priors.
+Ridge regression and the Bayesian LASSO are two of the most approchable sparse regression priors.
+The horseshoe prior, proposed by **Carvalho et al., 2010**, is another popular choice in Bayesian literature **()** (somehow this accidentally came out sounding exactly like van Erp).
+Following the notation of **Vehtari**, the horsehoe prior can be written
+
+\begin{align}
+\tag{3}
+\beta_k \text{ | } \lambda_k, \tau & \sim \text{Normal}(0, \tau^2\lambda_k^2) \\
+\lambda_k & \sim \text{Half-Cauchy}(0, 1)
+\end{align}
+
+for $k = 1, 2, ..., K$. The horseshoe prior is so-named because for fixed values $\tau = \lambda = 1$, the prior resemebles a Beta(1/2, 1/2) distribution, or a horseshoe.
+
+The horseshoe is often favored because it is a global-local shrinkage prior.
+This means that $\tau$ shrinks *all* parameters towards 0 while the local parameter $\lambda_k$ and the heavy Cauchy tails allow larger coefficients to remain unshrunk.
+While this is precisely the goal of a sparsity inducing prior, the horseshoe prior can fail to regularize lrage coefficients *at all*, which can be a problem when parameters are weakly identified by data.
+Additionally, there is no consensus in the literature for assigning priors to $tau$.
+**Vehatari..** introduced the *regularized horseshoe* to address both of these shortcomings.
+The regularized horseshoe builds upon **(3)** such that
+
+\begin{align}
+\beta_k \text{ | } \lambda_j, \tau, c & \sim \text{Normal}(0, \tau^2 \tilde{\lambda_k^2}) \\
+\tilde{\lambda_k^2} & = \frac{c^2\lambda_k^2}{c^2 + \tau^2\lambda_k^2} \\
+\lambda_k & \sim \text{Half-Cauchy}(0, 1) \\
+c & \sim ~ \text{Inverse-Gamma}(a, b)
+\end{align}
+
+where $c$ > 0 helps to regularize $\beta_k$ far from zero. When $c$ approaches infinity, the regularized horseshoe becomes the standard horseshoe. **This is a place where my understanding is weak**
+
+# Spatial Structure
+
+In general, we expect areas with

### 2020-01-21 chapters/03_methods.Rmd
push new stan to run

@@
 When parameter vector $x$ contains many possible covariates, we may reasonably assume that some entries in $\beta$ are 0 -- in other words, that some of the parameters in $x$ have no effect on $Y$.
-In such a case, it is beneficial to use *sparsity inducing priors* on $\beta$ to ensure that irrelevant parameter coefficients are shrunk to 0.
+In frequentist applications, we might use a method such as the LASSO (Least Absolute Shrinkage and Selection Operator) to identify relevant variables.
 
-In frequentist applications, methods such as the LASSO (Least Absolute Shrinkage and Selection Operator) are employed for this type of variable selection.
-In the Bayesian setting, shrinkage and regularization can be performed using a variety of different priors on $\beta$.
-**Cite van Erp etc** provide an introduction to many such continuous priors.
-Ridge regression and the Bayesian LASSO are two of the most approchable sparse regression priors.
-The horseshoe prior, proposed by **Carvalho et al., 2010**, is another popular choice in Bayesian literature **()** (somehow this accidentally came out sounding exactly like van Erp).
-Following the notation of **Vehtari**, the horsehoe prior can be written
+## Spike and Slab
+
+In the Bayesian setting, there are two primary methods for variable selection.
+The first is called the "spike-and-slab" prior **(cite)** and has often been considered the "gold standard" for sparse Bayesian regression, or variable selection **(Vehtari)**.
+This prior is often expressed as
 
@@
 \tag{3}
+\beta_j | \lambda_j, c, \epsilon & \sim \lambda_j \text{Normal}(0, c^2) + (1-\lambda_j)\text{Normal}(0, \epsilon^2) \\
+\lambda_j & \sim \text{Bernoulli}(\pi)
+\end{align}
+
+for $j = 1, 2, ..., J$ where $\lambda_j \in {0, 1}$ indicates whether $\beta_j$ is from the "spike" (ie, near zero) or the "slab" (ie, nonzero).
+Priors must be set for the width of the slab $c$ and the prior inclusion probability $\pi$.
+Here, $\pi$ reflects our prior understanding of the sparsity of $\beta$.
+
+## Horseshoe
+
+The second method for Bayesian sparse regression is to give $\beta$ some continuous *sparsity inducing prior*.
+**Van Erp et al** provide an introduction to many such continous priors.
+Ridge regression and the Bayesian LASSO are two of the most approachable priors, but all sparsity inducing priors utilize the same logic: give $\beta$ a prior with *most* of its mass near 0, to shrink irrelevant coefficients, and the rest of its mass "far" from 0.
+Identifying "far" depends more or less on the specific data and model in question.
+
+The horseshoe prior, proposed by **Carvalho et all, 2010** is a popular choice in Bayesian literature, in part because it is similar to the "gold standard" spike-and-slab.
+Following the notation of **Vehtari & Pironeen**, the horseshoe prior can be expressed
+
+\begin{align}
+\tag{4}
 \beta_k \text{ | } \lambda_k, \tau & \sim \text{Normal}(0, \tau^2\lambda_k^2) \\
@@
 
-for $k = 1, 2, ..., K$. The horseshoe prior is so-named because for fixed values $\tau = \lambda = 1$, the prior resemebles a Beta(1/2, 1/2) distribution, or a horseshoe.
+for $k = 1, 2, ..., K$. The horseshoe prior is so-named because for fixed values $\tau = \lambda_k = 1$, the prior resemebles a Beta(1/2, 1/2) distribution, or a horseshoe.
 
@@
 This means that $\tau$ shrinks *all* parameters towards 0 while the local parameter $\lambda_k$ and the heavy Cauchy tails allow larger coefficients to remain unshrunk.
-While this is precisely the goal of a sparsity inducing prior, the horseshoe prior can fail to regularize lrage coefficients *at all*, which can be a problem when parameters are weakly identified by data.
+While this is precisely the goal of a sparsity inducing prior, the horseshoe prior can fail to regularize large coefficients *at all*, which can be a problem when parameters are weakly identified by data.
 Additionally, there is no consensus in the literature for assigning priors to $tau$.
@@
 \begin{align}
+\tag{5}
 \beta_k \text{ | } \lambda_j, \tau, c & \sim \text{Normal}(0, \tau^2 \tilde{\lambda_k^2}) \\
@@
 
-where $c$ > 0 helps to regularize $\beta_k$ far from zero. When $c$ approaches infinity, the regularized horseshoe becomes the standard horseshoe. **This is a place where my understanding is weak**
+where $c$ > 0 helps to regularize $\beta_k$ far from zero.
+When $c$ approaches infinity, the regularized horseshoe becomes the standard horseshoe.
+**This is a place where my understanding is weak** Using the spike-and-slab as intuition, c corresponds to the width of the slab.
+**Double check this, but I think** This is, in essence, how we describe "far" from 0.
 

### 2020-01-21 chapters/04_application.Rmd
add chapters folder
push new stan to run

--- 2020-01-16 chapters/01_introduction.Rmd
@@
 ---
-title: "01_introduction"
+title: "04_application"
 author: "Raven McKnight"
 date: "1/16/2020"
-output: html_document
+output: 
+  html_document:
+    toc: true
+    toc_float: true
+    theme: paper
 ---
@@
 
-## R Markdown
+# Case Study
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+## Background
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+Metro Transit is the primary transit provider in the Minneapolis-Saint Paul metropolitan area.
+The agency is interested in predicting transit demand across the region in order to guide route planning and service provision.
+Currently, Metro Transit uses a set of 5 Transit Market Areas to estimate spatial demand.
+The Transit Market Areas are writen into the official Transportation Planning Policy (TPP) and are calculated using a simple linear regression.
+Following the notation of the TPP, the formula for determing Transit Market Areas is expressed
 
-```{r cars}
-summary(cars)
-```
+\begin{align}
+\text{Transit Market Index }  = & 0.64\text{ * population density} + 0.23\text{ * intersection density} +  \\
+& 0.20\text{ * employment_density} + 0.11\text{ * automobile availability}
+\end{align}
 
-## Including Plots
+where each predictor is logged and scaled by developed land acreage per census block group.
+Here, automobile availability refers to the number of adults over age 16 less the total number of automobiles available in a block group (scaled by developed land acreage).
+There is an additional indicator variable, omitted in this notation, for the census block group containing the MSP International Airport.
+For more documentation of the official Transit Market Areas, refere to the TPPs [Appendix G](https://metrocouncil.org/Transportation/Planning-2/Key-Transportation-Planning-Documents/Transportation-Policy-Plan/The-Adopted-2040-TPP-(1)/Final-2040-Transportation-Policy-Plan/2040-TPP-Appendix-G-Transit-Design-and-Perf-Standa.aspx).
 
-You can also embed plots, for example:
+Census block groups are split into 5 market areas based on the Transit Market Index described above, where the highest market area (Market Area 1) is expected to support high-frequency, all-day service and the lowest market area (Market Area 5) is expected to support peak commuter express service, if that.
 
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
+The Transit Market Index values are geographically smoothed to create more-or-less concentric Transit Market Areas.
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+## Data Description
+
+## Data Preperation
+
+## Modeling
+
+### Poisson Regression
+
+### Horseshoe Prior
+
+### Overdispersion
+
+### Spatial Structure

### 2020-01-21 scripts/aws/poisson.R

@@ -23,14 +23,25 @@ N <- nrow(mod_dat)
 K <- ncol(xdat)
 x <- as.matrix(ncol = K, xdat)
 
-standat1 <- list(y = as.integer(y/365), E = E, x = x, K = K, N = N)
+standat1 <- list(y = y, E = E, x = x, K = K, N = N)
+
+horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N, 
+                      scale_icept = 10, scale_global = 0.2, 
+                      nu_global = 1, nu_local = 1,
+                      slab_scale = 2, slab_df = 4)
 
 ## basic poisson --------------------------------
 
 poisson <- "~/Documents/honors/honors-project/stan/poisson.stan"
-poisson_fit <- stan(poisson, data = standat1, warmup = 1000, iter = 2000, verbose = T)
+poisson_fit <- stan(poisson, data = standat1, warmup = 100, iter = 200, verbose = T)
 
 saveRDS(poisson_fit, '~/Documents/honors/honors-project/final-fits/poisson_fit.RDS')
+
+## add horseshoe priors -------------------------
+
+poisson_hrs <- "~/Documents/honors/honors-project/stan/poisson_horseshoe.stan"
+poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, warmup = 100, iter = 200, verbose = T)
+
 
 ## add overdispersion parameter -----------------
 
@@ -43,10 +54,7 @@ poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 2000, verbose =
 
 poisson_horseshoe <- "~/Documents/honors/honors-project/stan/poisson_theta_horseshoe.stan"
 
-horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N, 
-                  scale_icept = 10, scale_global = 0.2, 
-                  nu_global = 1, nu_local = 1,
-                  slab_scale = 2, slab_df = 4)
+
 
 poisson_horseshoe_fit <- stan(poisson_horseshoe, data = horseshoe_dat, iter = 200, verbose = T, 
                               control = list(adapt_delta = 0.99))

### 2020-01-21 chapters/04_application.Rmd
revert to avg daily boardings

@@
 
+### Metro Transit Data
+
+### Covariates
+
 ## Data Preperation
@@
 
+The baseline Poisson regression can be fit easily in Stan. This simplest model in this case study can be written
+
+\begin{align}
+Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
+\eta_i  = log(\lambda_i) & = \beta_0 + \sum_{k=1}^{19}x_k^T\beta_k \\
+\beta_0, \beta & \sim \text{Normal}(0, 1) \\
+\end{align}
+
+where $Y_i$ is the number of boardings in census block group $i$ in 2017, $E_i$ is the number of bus stops made in block group $i$ in 2017, and parameter vector $x$ corresponds to the covariates discussed above.
+**Figure ...** shows this model written in Stan.
+
+```{r}
+# FIND A BETTER WAY TO PRINT THIS
+writeLines(readLines("~/Documents/honors/honors-project/stan/poisson.stan"))
+```
+
 ### Horseshoe Prior
+
+The model above does a decent job predicting ridership but there are many possible improvements.
+One shortcoming of the simplest Poisson model is that the normal priors on $\beta$ make it challenging to determine which coefficients are truly significantly different from 0.
+This is where the regularized horseshoe prior comes into play.
+
+This model can be expressed in Stan as pictured in **Figure...**. The Stan program for this model was adapted from **Vehtari..**.
+
+```{r}
+writeLines(readLines("~/Documents/honors/honors-project/stan/poisson_horseshoe.stan"))
+```
 

### 2020-01-21 scripts/07_join-mod-dat.R

@@ -1,35 +1,29 @@
 # Goal: export the data needed for each model I'll fit
 
 library(data.table)
-library(tigris)
 library(ggplot2)
 library(lubridate)
 
-options(tigris_class = "sf")
-options(tigris_use_cache = TRUE)
-
-counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
-bgs <- block_groups("MN", counties, 2016)
-
-mod_dat <- readRDS('data/modeling-dat/basic_mod_dat_ann17.RDS')
+mod_dat <- readRDS('data/modeling-dat/basic_mod_dat.RDS')
 setDT(mod_dat)
-
+mod_dat[, year := year(ymd(date_key))]
+mod_dat <- mod_dat[year == 2017]
 
 cov <- readRDS('data/modeling-dat/ag_2017_scaled_mod.RDS') # using scaled but not logged
 setDT(cov)
 
-lm_dat <- merge(mod_dat, cov, on = 'GEOID', all.x = TRUE)
-lm_dat[year_boards == 0, year_boards := 0.01]
-lm_dat[year_stops == 0, year_stops := 0.01]
+lm_dat <- mod_dat[wday != 6 & wday != 7]
+lm_dat <- lm_dat[, .(daily_boards = mean(daily_boards, na.rm = T), 
+                     daily_stops = mean(daily_stops, na.rm = T)), keyby = .(GEOID)]
+lm_dat <- unique(lm_dat)
 
-lm_dat[, log_year_boards := log(year_boards)]
-lm_dat[, log_year_stops := log(year_stops)]
+lm_dat <- merge(lm_dat, cov, by = 'GEOID', all.x = TRUE)
 
 saveRDS(lm_dat, 'data/modeling-dat/lm_dat_scaled.RDS')
 
 ## poisson regression ------------
 
-p_dat <- lm_dat[, -c('log_year_boards', 'log_year_stops')]
 
 
-saveRDS(p_dat, 'data/modeling-dat/p_dat_scaled.RDS')
+
+saveRDS(lm_dat, 'data/modeling-dat/p_dat_scaled.RDS')

### 2020-01-21 scripts/aws/poisson.R

@@ -11,12 +11,12 @@ mod_dat <- readRDS('~/Documents/honors/honors-project/data/modeling-dat/p_dat_sc
 mod_dat <- na.omit(mod_dat)
 setDT(mod_dat)
 
-xdat <- mod_dat[, -c('GEOID', 'year_boards', 'year_stops', 'sqkm', 'estimate_tot_pop', 'year_alights', 'num_interpolated', 'num_routes', 'year_activity')]
+xdat <- mod_dat[, -c('GEOID', 'daily_boards', 'daily_stops', 'sqkm', 'estimate_tot_pop', 'daily_alights', 'num_interpolated', 'num_routes', 'daily_activity')]
 
 ## regular poisson regression -------------------
 
-y <- mod_dat$year_boards
-E <- mod_dat$year_stops
+y <- as.integer(mod_dat$daily_boards)
+E <- mod_dat$daily_stops
 
 N <- nrow(mod_dat)
 
@@ -28,12 +28,12 @@ standat1 <- list(y = y, E = E, x = x, K = K, N = N)
 horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N, 
                       scale_icept = 10, scale_global = 0.2, 
                       nu_global = 1, nu_local = 1,
-                      slab_scale = 2, slab_df = 4)
+                      slab_scale = 1, slab_df = 4)
 
 ## basic poisson --------------------------------
 
 poisson <- "~/Documents/honors/honors-project/stan/poisson.stan"
-poisson_fit <- stan(poisson, data = standat1, warmup = 100, iter = 200, verbose = T)
+poisson_fit <- stan(poisson, data = standat1, iter = 2000, verbose = T)
 
 saveRDS(poisson_fit, '~/Documents/honors/honors-project/final-fits/poisson_fit.RDS')
 
@@ -46,6 +46,7 @@ poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, warmup = 100, iter = 
 ## add overdispersion parameter -----------------
 
 poisson_theta <- "~/Documents/honors/honors-project/stan/poisson_theta.stan"
+
 poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 2000, verbose = T)
 
  #shinystan::launch_shinystan(poisson_theta_fit)

### 2020-01-21 scripts/aws/poisson.R
prep to fit horseshoe

@@ -40,7 +40,8 @@ saveRDS(poisson_fit, '~/Documents/honors/honors-project/final-fits/poisson_fit.R
 ## add horseshoe priors -------------------------
 
 poisson_hrs <- "~/Documents/honors/honors-project/stan/poisson_horseshoe.stan"
-poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, warmup = 100, iter = 200, verbose = T)
+poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, iter = 2000, verbose = T,
+                        control = list(adapt_delta = 0.9999, max_treedepth = 15))
 
 
 ## add overdispersion parameter -----------------

### 2020-01-21 scripts/aws/poisson.R
last work from MT on Tuesday

@@ -15,7 +15,7 @@ xdat <- mod_dat[, -c('GEOID', 'daily_boards', 'daily_stops', 'sqkm', 'estimate_t
 
 ## regular poisson regression -------------------
 
-y <- as.integer(mod_dat$daily_boards)
+y <- round(mod_dat$daily_boards)
 E <- mod_dat$daily_stops
 
 N <- nrow(mod_dat)
@@ -26,22 +26,32 @@ x <- as.matrix(ncol = K, xdat)
 standat1 <- list(y = y, E = E, x = x, K = K, N = N)
 
 horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N, 
-                      scale_icept = 10, scale_global = 0.2, 
+                      scale_icept = 3, scale_global = 0.2, 
                       nu_global = 1, nu_local = 1,
-                      slab_scale = 1, slab_df = 4)
+                      slab_scale = 2, slab_df = 4)
 
 ## basic poisson --------------------------------
 
 poisson <- "~/Documents/honors/honors-project/stan/poisson.stan"
-poisson_fit <- stan(poisson, data = standat1, iter = 2000, verbose = T)
+#poisson_fit <- stan(poisson, data = standat1, iter = 2000, verbose = T)
 
-saveRDS(poisson_fit, '~/Documents/honors/honors-project/final-fits/poisson_fit.RDS')
+#saveRDS(poisson_fit, '~/Documents/honors/honors-project/final-fits/poisson_fit.RDS')
 
 ## add horseshoe priors -------------------------
 
 poisson_hrs <- "~/Documents/honors/honors-project/stan/poisson_horseshoe.stan"
-poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, iter = 2000, verbose = T,
-                        control = list(adapt_delta = 0.9999, max_treedepth = 15))
+poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, iter = 10000, verbose = T,
+                        control = list(adapt_delta = 0.99, max_treedepth = 15))
+
+## test the code
+
+horseshoe_dat_test <- list(y = log(y), E = E, x = x, K = K, N = N, 
+                      scale_icept = 10, scale_global = 0.2, 
+                      nu_global = 1, nu_local = 1,
+                      slab_scale = 1, slab_df = 4)
+
+hrs_test <- "~/Documents/honors/honors-project/stan/vehtari_alt.stan"
[…]

### 2020-01-22 chapters/04_application.Rmd
update script for final fits, some notes thru chapters

@@
 
+#### Divergences
+
+After running four chains for 10,000 iterations each, this model does produce some *divergences*.
+**This means I should explain what a divergence is** To be exact, this model produces 30 divergent transtions out of 20,000 post-warmup transitions, for a divergent rate of 0.15%.
+Divergencies are a diagnostic particular to the HMC and NUTS samplers used by Stan.
+
+#### Variables Selected
+
+Perhaps surprisingly, the use of  horseshoe priors only suggests the removal of four variables: __, __, __, __.
+This is likely because the initial set of covariates were selected based on domain knowledge and therefore, it was unlikely for $\beta$ to be exceedingly sparse.
+The application of the horseshoe prior would be more informative in a data set with more parameters, and with less insight into the selection of the initial covariate vector $x$.
+
 ### Overdispersion
 
+For computational efficiency, we can drop the horseshoe priors for this model. Armed with the information provided by our fist two models, we can simply give $\beta$ tigher normal priors.
+
 ### Spatial Structure

### 2020-01-22 scripts/aws/fitting.R

@@ -1,0 +1,114 @@
+## packages
+
+library(rstan)
+library(dplyr)
+library(data.table)
+library(shinystan)
+library(ggplot2)
+library(bayesplot)
+options(mc.cores = parallel::detectCores())
+rstan_options(autowrite = T)
+
+## data
+mod_dat <- readRDS('~/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
+mod_dat <- na.omit(mod_dat)
+setDT(mod_dat)
+
+xdat <- mod_dat[, -c('GEOID', 'daily_boards', 'daily_stops', 'sqkm', 'estimate_tot_pop', 'daily_alights', 'num_interpolated', 'num_routes', 'daily_activity')]
+
+## regular poisson regression -------------------
+
+y <- round(mod_dat$daily_boards)
+E <- mod_dat$daily_stops
+
+N <- nrow(mod_dat)
+
+K <- ncol(xdat)
+x <- as.matrix(ncol = K, xdat)
+
+standat1 <- list(y = y, E = E, x = x, K = K, N = N)
+
+horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N, 
+                      scale_icept = 3, scale_global = 0.2, 
+                      nu_global = 1, nu_local = 1,
+                      slab_scale = 2, slab_df = 4)
+
+## basic poisson --------------------------------
+
+poisson <- "~/Documents/honors/honors-project/stan/final-stan/poisson.stan"
+poisson_fit <- stan(poisson, data = standat1, iter = 10000, verbose = T)
+
+saveRDS(poisson_fit, '~/Documents/honors/honors-project/final-fits/poisson_fit.RDS')
+#poisson_fit <- readRDS("~/Documents/honors/honors-project/final-fits/poisson_fit.RDS")
+
+post <- rstan::extract(poisson_fit)
+lat <- post$f
+samp <- matrix(0, nrow=4000, ncol=1495)
+
+for(i in 1:nrow(samp)){
+  samp[i, ] <- rpois(1495, exp(lat[i, ]))
+}
+
+setDT(as.data.frame(samp))
+
+
+pois <- pp_check(y, samp[1:100, ], ppc_dens_overlay) + 
+  theme_minimal() +
+  labs(title = "Poisson Actual versus fitted") + xlim(0, 200)
+
[…]

### 2020-01-22 scripts/aws/poisson.R

@@ -1,105 +1,0 @@
-## packages
-
-library(rstan)
-library(dplyr)
-library(data.table)
-options(mc.cores = parallel::detectCores())
-rstan_options(autowrite = T)
-
-## data
-mod_dat <- readRDS('~/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
-mod_dat <- na.omit(mod_dat)
-setDT(mod_dat)
-
-xdat <- mod_dat[, -c('GEOID', 'daily_boards', 'daily_stops', 'sqkm', 'estimate_tot_pop', 'daily_alights', 'num_interpolated', 'num_routes', 'daily_activity')]
-
-## regular poisson regression -------------------
-
-y <- round(mod_dat$daily_boards)
-E <- mod_dat$daily_stops
-
-N <- nrow(mod_dat)
-
-K <- ncol(xdat)
-x <- as.matrix(ncol = K, xdat)
-
-standat1 <- list(y = y, E = E, x = x, K = K, N = N)
-
-horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N, 
-                      scale_icept = 3, scale_global = 0.2, 
-                      nu_global = 1, nu_local = 1,
-                      slab_scale = 2, slab_df = 4)
-
-## basic poisson --------------------------------
-
-poisson <- "~/Documents/honors/honors-project/stan/poisson.stan"
-#poisson_fit <- stan(poisson, data = standat1, iter = 2000, verbose = T)
-
-#saveRDS(poisson_fit, '~/Documents/honors/honors-project/final-fits/poisson_fit.RDS')
-
-## add horseshoe priors -------------------------
-
-poisson_hrs <- "~/Documents/honors/honors-project/stan/poisson_horseshoe.stan"
-poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, iter = 10000, verbose = T,
-                        control = list(adapt_delta = 0.99, max_treedepth = 15))
-
-## test the code
-
-horseshoe_dat_test <- list(y = log(y), E = E, x = x, K = K, N = N, 
-                      scale_icept = 10, scale_global = 0.2, 
-                      nu_global = 1, nu_local = 1,
-                      slab_scale = 1, slab_df = 4)
-
-hrs_test <- "~/Documents/honors/honors-project/stan/vehtari_alt.stan"
[…]

### 2020-01-26 scripts/aws/fitting.R
prep for AWS

@@ -28,10 +28,30 @@ x <- as.matrix(ncol = K, xdat)
 
 standat1 <- list(y = y, E = E, x = x, K = K, N = N)
 
-horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N, 
-                      scale_icept = 3, scale_global = 0.2, 
+
+# original fit
+# horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N, 
+#                       scale_icept = 2, scale_global = 0.2, 
+#                       nu_global = 1, nu_local = 1,
+#                       slab_scale = 2, slab_df = 4) 
+
+# new params
+# horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N, 
+#                       scale_icept = 2, scale_global = 0.2, 
+#                       nu_global = 1, nu_local = 1,
+#                       slab_scale = 1/2, slab_df = 1) 
+
+# new params 2
+# horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N, 
+#                       scale_icept = 2, scale_global = 1, 
+#                       nu_global = 1, nu_local = 1,
+#                       slab_scale = 1/2, slab_df = 1) 
+
+# new params 3
+horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N,
+                      scale_icept = 2, scale_global = 0.002,
                       nu_global = 1, nu_local = 1,
-                      slab_scale = 2, slab_df = 4)
+                      slab_scale = 1, slab_df = 24)
 
 ## basic poisson --------------------------------
 
@@ -61,7 +81,7 @@ pois <- pp_check(y, samp[1:100, ], ppc_dens_overlay) +
 poisson_hrs <- "~/Documents/honors/honors-project/stan/final-stan/poisson_horseshoe.stan"
 poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, iter = 10000, verbose = T,
                         control = list(adapt_delta = 0.99, max_treedepth = 15))
-saveRDS(poisson_hrs_fit, '~/Documents/honors/honors-project/final-fits/poisson_hrs_fit.RDS')
+saveRDS(poisson_hrs_fit, '~/Documents/honors/honors-project/final-fits/poisson_hrs_fit_new_params_3.RDS')
 
 #poisson_hrs_fit <- readRDS('~/Documents/honors/honors-project/final-fits/poisson_hrs_fit.RDS')
 

### 2020-01-27 chapters/03_methods.Rmd
revert to avg daily boardings
update script for final fits, some notes thru chapters
update written chapters

@@
 ---
-title: "Methods"
+title: "Methods/Lit Review"
 author: "Raven McKnight"
-date: "1/16/2020"
 output:
+  word_document:
+    toc: yes
+  pdf_document: null
   html_document:
-    toc: true
-    toc_float: true
     theme: paper
+    toc: yes
+    toc_float: yes
 ---
@@
 Each model presented in this honors is built upon a simple Poisson regression.
-Poisson regressions are used to model count data and assume that the natural log of the mean of response variable $Y$ can be modeled by a linear combination of predictors.
+Poisson regressions are used to model count data.
+The model assumes that the natural log of the mean of response variable $Y$ can be modeled by a linear combination of predictors.
 Poisson regressions are a form of generalized linear models with the natural log as its link function.
@@
 \begin{align}
-\tag{1}
 Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
-\eta_i  = log(\lambda_i) & = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k \\
+\eta_i  = log(\lambda_i) & = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k 
 \end{align}
@@
 where $\beta_0$ is an optional intercept term and $E_i$ is an offset term.
+
 The offset $E_i$ is often termed "exposure" in applications such as disease risk modeling where the offset may be the number of observed cases in a previous year, etc.
+More technically, the offset term scales the Poisson output to be a rate rather than a count.
+This is appropriate when observations $i$ have different potentials for response $Y$.
+For example, a county with higher population will naturally have a higher count of patients with asthma than a county with a smaller population.
+Mathematically, the offset is a covariate with parameter set equal to 1.
+Recall, the Poisson regression assumes we can model the mean of response $Y$, $\bar{Y}$ with a combination of linear predictors:
+
+$$
+log(\bar{Y}) = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k \\
+$$
+Therefore, if we wish to model the rate $Y/E$, the equation is rewritten
+
+\begin{align}
+log(\bar{Y}/E) & = \beta^{'}_0 + \sum_{k=1}^{K}x_k^T\beta^{'}_k \\
+log(\bar{Y}) & = log(E) + \beta^{'}_0 + \sum_{k=1}^{K}x_k^T\beta^{'}_k
+\end{align}
 
@@
 For the purposes of this study, we can add a set of heterogeneous random effects $\theta$ to account for overdispersion (where Var($Y$) > E($Y$)).
-Following **cite mitzi morris**, the second line of **(1)** can be written
+Following Morris et al (2019), the second line of **1** can be written
 
 \begin{align}
-\tag{2}
-\eta_i = log(\lambda_i) & = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + \theta \sigma
+\eta_i = log(\lambda_i) & = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + \theta
 \end{align}
 
-where $\theta$ and $\sigma$ are given normal priors centered around 0, such as $\theta \sim$ Normal(0, 1) and $\sigma \sim$ Normal(0, 5).
+where $\theta$ is given a vague normal prior centered around 0, such as $\theta \sim$ Normal(0, 1).
 The addition of these random effects improve fits on data with higher-than-expected variance.
+With sufficiently vague priors, $\theta$ can account for most or all overdispersion not accounted for by $\beta_0 + \sum_{k=1}^{K}x_k^T\beta_k$.
 
@@
 
-Often, the coefficients $\beta_1, ..., \beta_k$ are given simple normal priors such as $\beta_k \sim$ Normal(0, 1) in the case of standardized parameters.
+Often, the coefficients $\beta_1, ..., \beta_k$ are given simple normal priors such as $\beta \sim$ Normal(0, 1) in the case of standardized parameters.
 When parameter vector $x$ contains many possible covariates, we may reasonably assume that some entries in $\beta$ are 0 -- in other words, that some of the parameters in $x$ have no effect on $Y$.
 In frequentist applications, we might use a method such as the LASSO (Least Absolute Shrinkage and Selection Operator) to identify relevant variables.
+There are several Bayesian alternatives for variable selection which generally consist of applying particularl priors to $\beta$.
 
@@
 In the Bayesian setting, there are two primary methods for variable selection.
-The first is called the "spike-and-slab" prior **(cite)** and has often been considered the "gold standard" for sparse Bayesian regression, or variable selection **(Vehtari)**.
+The first is called the "spike-and-slab" prior (Mitchell and Beauchamp, 1988; George and McCulloch, 1999) and has often been considered the "gold standard" for sparse Bayesian regression, or variable selection (Piironen and Vehtari, 2017).
 This prior is often expressed as
@@
 \begin{align}
-\tag{3}
 \beta_j | \lambda_j, c, \epsilon & \sim \lambda_j \text{Normal}(0, c^2) + (1-\lambda_j)\text{Normal}(0, \epsilon^2) \\
@@
 for $j = 1, 2, ..., J$ where $\lambda_j \in {0, 1}$ indicates whether $\beta_j$ is from the "spike" (ie, near zero) or the "slab" (ie, nonzero).
+The "spike" is the area where most of the prior density for $\beta$ is centered (generally around 0) while the "slab" refers to the low-density extent of the prior.
+The spike-and-slab prior encourages all $\beta_j$ towards zero; only $\beta_j$ sufficiently far from 0 will be estimated to be from the slab.
+
+The primary challenge with the spike-and-slab (and other Bayesian shrinkage methods) is the selction of priors.
 Priors must be set for the width of the slab $c$ and the prior inclusion probability $\pi$.
 Here, $\pi$ reflects our prior understanding of the sparsity of $\beta$.
+In practice, we rarely have strong prior knowledge of the number of predictors we expect to be distinguishable from zero.
+This can make the implementation of sparse regression more challenging than, say, the frequentist LASSO.
 
@@
 The second method for Bayesian sparse regression is to give $\beta$ some continuous *sparsity inducing prior*.
-**Van Erp et al** provide an introduction to many such continous priors.
+**Van Erp et al (2019)** provide an introduction to many such continous priors.
 Ridge regression and the Bayesian LASSO are two of the most approachable priors, but all sparsity inducing priors utilize the same logic: give $\beta$ a prior with *most* of its mass near 0, to shrink irrelevant coefficients, and the rest of its mass "far" from 0.
@@
 \begin{align}
-\tag{4}
 \beta_k \text{ | } \lambda_k, \tau & \sim \text{Normal}(0, \tau^2\lambda_k^2) \\
@@
 While this is precisely the goal of a sparsity inducing prior, the horseshoe prior can fail to regularize large coefficients *at all*, which can be a problem when parameters are weakly identified by data.
+With no regularization applied to the largest coefficients, there is a risk of overfitting.
 Additionally, there is no consensus in the literature for assigning priors to $tau$.
-**Vehatari..** introduced the *regularized horseshoe* to address both of these shortcomings.
-The regularized horseshoe builds upon **(3)** such that
+Piironen and Vehtari (2017) introduced the *regularized horseshoe* to address both of these shortcomings of the horseshoe.
+The regularized horseshoe builds upon **need to figure out equation numbering** such that
 
 \begin{align}
-\tag{5}
 \beta_k \text{ | } \lambda_j, \tau, c & \sim \text{Normal}(0, \tau^2 \tilde{\lambda_k^2}) \\
@@
 When $c$ approaches infinity, the regularized horseshoe becomes the standard horseshoe.
-**This is a place where my understanding is weak** Using the spike-and-slab as intuition, c corresponds to the width of the slab.
-**Double check this, but I think** This is, in essence, how we describe "far" from 0.
+Additionally, $c$ in the regularized horseshoe corresponds to the slab width in the spike-and-slab prior.
+
+As with the spike-and-slab, the selection of priors is a challenge with the regularized horseshoe...
 
@@
 
-In general, we expect areas with
+When modeling data with a spatial component, we generally expect adjacent areas to be more similar than areas which are far apart.
+While this is a straightforward assumption, it can improve model fits in several ways.
+First, it can encode prior information about response $Y$ not otherwise represented by covariates $x$ (ie, that nearby observations are more similar than far flung ones).
+Second, it can provide geographic smoothing when observations are sparse or noisy.
+This is often the case in small areas or when observing events which can only occur at specific locations, such as air quality measured by sensors.
+
+## Conditional Autoregressive Priors
+
+### This whole section follows Morris et al (probably too) closely right now
+
+We can encode spatial information into Bayesian models in several ways.
+Conditional Autoregressive (CAR) priors are one of the most widespread Bayesian methods for modeling spatial autocorrelation.
+Conditional Autoregressive models were introduced in Besag 1974 and remain perhaps the primary method for Bayesian areal data modeling.
+CAR models are used as $priors$, not as the actual observation model.
+
+Areal data corresponds to finitely many discrete areal units, such as counties or census tracts.
+Counts in areal units tend to be noisy, particularly when the counted event is rare, the population of areal unit $i$ is small, or the phsyical boundaries of areal units form unusual patterns in the data.
+Conditonal Autoregressive models combine information from areal unit $i$ as well as its neighbors to smooth noisy counts.
+
+Generally, CAR models use binary neighborhood relationships encoded in a neighborhood matrix.
+Areal units $i$ and $j$ are considered neighbors if they share a boundary.
+For strictly rectangualar lattice data, we often choose either a "Queens" or "Rooks" neighborhood structures **(include graphic)**.
+For irregularly shaped areal units, such as census tracts, we generally default to a queen neighborhood structure wherein areal units that share *any* points of contact are deemed neighbors.
+For $M$ regions represented in a $M$x$M$ neighborhood matrix $W$, $w_{ij} = 1$ if regions $i$ and $j$ are neighbors and 0 otherwise.
+Matrix $W$ is symmetric.
+Note that regions $i$ are not considered neighbors with themselves.
+
+The spatial interactions between $i$ and $j$ are represented by random variable $\phi$.
+Here, $\phi$ is a vector $\phi = \phi_1, \phi_2, ..., \phi_M$.
+The distribution of each $\phi_i$ is determined by the sum of its neighbors values such that
+
+# having a notation crisis
+
+$$
+\phi_i \text{ | } \phi_j, j \neq i \sim \text{Normal}(\sum_{i=1}^{M}w_{ij}\phi_j, \sigma^2) 
+$$
+
+Besag (1974) proved that the joint distribution of $\phi$ is a multivariate normal centered at 0 where its variance is given by a symmetric positive definite precision matrix $Q$.
+Matrix $Q$ is simply the inverse of the covariance matrix $\sum$.
+This finding simplifies the above to
+
+$$
+\phi \sim \text{Normal}(0, Q^{-1})
+$$
+
+Precision matrix $Q$ can be defined for multivariate normal $\phi$ using two other $M$x$M$ matrices: the neighborhood matrix $W$ as well as diagonal matrix $D$ in which $d_{ii}$ indicates the number of neighbors region $i$ has.
+All off-diagonal entries are 0.
+Using these matrices, $Q$ is defined
+
+$$
+Q = D(I - \alpha W)
+$$
+
+where $I$ is the identity matrix and $\alpha \in (0, 1)$ determines the amount of spatial autocorrelation present in $Y$.
+Parameter $\alpha$ is a key component of CAR models.
+At $\alpha = 0$, the model assumes spatial independence and at $\alpha = 1$, perfect spatial autocorrelation.
+
+Following Morris et al, the log probability density of $\phi$ is proportional to
+
+$$
+\frac{M}{2}\text{log(det}(Q)) - \frac{1}{2}\phi^TQ\phi
+$$
+
+Calculating this value is computationally expensive. For models with 1,000 areal units, for example, calculating the determinant requires 1 billion operations (Morris et al, 2019).
+
+## Intrinsic Conditional Autoregressive
+
+The Intrinsic Conditional Autoregressive (ICAR) prior is a slight simplifation of a CAR prior. ICAR priors set $\alpha = 1$ which simplifies the definition of Q
+
+$$
+Q = D-A
+$$
+
+thereby setting $det(Q) = 0$. As a result, the first term in **equation numbering** can be simplified to
+
+$$
+-\frac{1}{2}\phi^TQ\phi
+$$
+
+This reduces the computational expense of computing CAR models significantly. Notably, the ICAR prior is improper but yields proper posteriors (Morris et al, 2019).
+
+The ICAR specifies each $\phi_i$ to be normally distributed with its mean equal to the mean of its neighbors values.
+This is how the CAR/ICAR model "borrows strength" from geographic neighbors to smooth noisy estimates.
+Intuitively, the variance of each $\phi_i$ decreases as its number of neighbors $d_i$ increases.
+The conditional distribution for each $\phi_i$ can be written
+
+$$
+\text{p}(\phi_i \text{ | } \phi_{i \sim j}) = \text{Normal}(\frac{\sum_{i \sim j}\phi_i}{d_i}, \frac{\sigma_i^2}{d_i})
+$$
+
+Here, the variance $\sigma$ is unknown.
+The conditional specification of each $\phi_i$ is helpful in building intuition about the assumptions the ICAR component makes and how it performs its "smoothing".
+The joint distribution can be rewritten
+
+$$
+\text{p}(\phi) \propto \text{exp}(-\frac{1}{2}\sum_{i \sim j}(\phi_i - \phi_j)^2)
+$$
+
+This pairwise difference specification is also helpful in building intuition about how ICAR models perform smoothing.
+
+## BYM and BYM2
+
+The specific ICAR model used in this analysis is the Besag-York-Mollie (BYM) Poisson model.
+The BYM model contains random effects and a spatial ICAR term to account for both spatial and non-spatial heterogeneity.
+Note that in **section??*, we introduced a random effect term $\theta$.
+Therefore, the BYM model can be though of as decomposing $\theta$ into spatial and non-spatial components.
+
+Specifically, the BYM model is written
+
+$$
+\eta_i = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + \theta + \phi
+$$
+
+where $\phi$ is the addition: an ICAR component.
+
+The BYM model is appealing in its simplicity.
+However, the lack of hyperpriors specified for $\theta$ and $\phi$ can make the model incredibly challenging to fit.
+In theory, extra-Poisson variance could be explained 100% by $\theta$, 100% by $\phi$, or anywhere in between.
+In *practice*, we likely have limited information about where that ratio falls.
+Riebler et al (2016) further explain the sampling issues faced by the BYM model faced with no hyperpriors.
+In short, the sampler is forced to explore all possible combinations of $\theta$ and $\phi$, no matter how unlikely a particular combination my be.
+In the context of fitting models with Stan, this is liekly to cause uneccesarily long computation times and perhaps to yield biased posterior estimates.
+
+There are some suggestions in the literature for setting hyperpriors for $\theta$ and $\phi$ (Besag and Mollie, 1991; Clayton and Montomoli, 1995).
+The existing methods, however, tend to rely on the specific data at hand which makes the selection of hyperparameters time unnecessarily time consuming.
+
+The BYM2 model proposed by Simpson et al and further described in Riebler et al (2016) aims to solve these sampling problems.
+The primary difference between the BYM and BYM2 models is the addition of a mixing parameter, $\rho$.
+The BYM2 model rewrites the BYM as
+
+$$
+\eta_i = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + ((\sqrt{\frac{\rho}{s}})\theta^{*} + ((\sqrt{1-\rho})\phi)\sigma
+$$
+
+where $\rho \in (0, 1)$ determines how much variance/overdispersion is caused by spatial versus non-spatial error terms.
+Like the popular Leroux CAR prior, proposed by Leroux et al (2009), the BYM2 scales both $\theta$ and $\phi$ by $\sigma$, the standard deviation of the combined error terms (Morris et al, 2019).
+In this parameterization, $s$ is a scaling factor such that Var($\theta_i$) $\approx$ Var($\phi_i$) $\approx$ 1.
+The equal unit variance is necessary for $\sigma$ to truly be the standard deviation of the error terms.
+
+Setting priors is somewhat more straightforward for the BYM2.
+The ICAR component, $\phi^{*}$ remains unchanged.
+Riebler and Morris recommend the prior $\theta \sim$ Normal($0, n)$ where $n$ is the number of connected graphs in the neighborhood graph.
+In many cases, such as modeling data for all counties in a state, $n=1$.
+When $n>1$, the variance is different in each subgraph which affects $\sigma$ and $s$.
+It is possible to fit the model in this case, although computation time is much longer.
+
+A relatively vague Normal(0, 1) prior is appropriate for $\sigma$. Mixing parameter $\rho$ is given a Beta(1/2, 1/2) prior. Similar to the horseshoe,
+
+# Stan
+
+The models above can all be fit using the Stan Programming Langauge (cite). Stan is probabilistic programming language used to compute joint log probability densities.
+
+## HMC/NUTS Sampling
+
+Stan utilizes the Hamiltonian Monte Carlo (HMC) and specialized No U-Turn Sampler (NUTS).
+
+Include parameters we'll set : treedepth, adapt_delta, what they correspond to mathematically
+
+## Divergent Transitions and Other Diagnostics
+
+Stan output provides several diagnostics which are helpful in confirming that the HMC/NUTS algorithm has converged and identifying problems with the models.
+There are a few we will refer to in the case study portion of this honors.
+
+The quickest diagnostic Stan provides is $\hat{R}$, or `Rhat`.
+$\hat{R}$ indicates how well the Markov chains have mixed by comparing between- and within-chain estimates for each parameter.
+At model convergence, $\hat{R} = 1$.
+According to Stan best practices, $\hat{R} < 1.1$ generally indicates sufficient mixing.
+Mixing (and therefore, $\hat{R}$) can often be improved by running chains for more iterations.
+
+Stan output `n_eff` is the estimated Effective Sample Size (ESS).
+ESS is the number of independent samples our sample is equivalent to.
+In other words, if `n_eff` = 1,000, we expect to gain the same amount of information from our sample as we would from 1,000 independent samples.
+Stan also provides Bulk ESS and Tail ESS estimates to determine how well Stan is exploring the most dense sections of the posterior (Bulk) versus the 5% and 95% quantiles (Tail).
+When `n_eff` is small, posterior estimates are unreliable.
+When `n_eff` is significantly less than the actual number of samples $n$, there is likely unusually high autocorrelation between samples.
+Through their GUI ShinyStan, Stan gives a warning when `n_eff` is less than 10% of $n$.
+However, if `n_eff` increases linearly with $n$, there is likely no "problem" with the model and a suitable effective sample size can be reached by increasing the number of iterations chains run for.
+
+Diagnostics $\hat{R}$ and `n_eff` are both helpful "gut-checks" for model convergence and performance.
+However, neither are sufficient for diagnosing convergence and they should be treated only as initial checks.
+
+Divergent transitions are perhaps the most important "diagnostic", or warning, Stan provides.
+Divergent transitions occur when the simulated Hamiltonian departs from the actual Hamiltonian trajectory **(need citation)**.
+When this occurs...
+In general, if Stan produces divergent transitions, model output must be discarded.
+Divergent transitions introduce bias which renders posterior estimates unusable.
+While $\hat{R}$ and `n_eff` can often be improved by running the chains for more iterations, divergent transitions will persist unless sampling parameters such as the target acceptance rate are changed.
+Often, divergent transitions call for reparameterizing the model.
+
+# Graphic Diagnostics, PPcheck
+
+Various graphical checks of priors and posteriors are helpful tools in building, improving, and comparing Bayesian models.
+The application portion of this honors follows Michael Betancourt's "Towards a Principled Bayesian Workflow" as well as ______ in terms of using visual aids for diagnosing issues with models and comparing quality of fit between models.
+Only the most "useful" plots are included below, for more on using graphics throughout a Bayesian workflow see Betancourt or Gabry.

### 2020-01-27 chapters/04_application.Rmd
update written chapters

@@
 ---
-title: "04_application"
+title: "Case Study/Application"
 author: "Raven McKnight"
@@
 output:
+  pdf_document:
+    toc: yes
   html_document:
-    toc: true
-    toc_float: true
     theme: paper
+    toc: yes
+    toc_float: yes
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
+library(knitr)
+library(kableExtra)
+library(rstan)
+library(bayesplot)
+library(ggplot2)
+library(data.table)
 ```
@@
 
-\begin{align}
-\text{Transit Market Index }  = & 0.64\text{ * population density} + 0.23\text{ * intersection density} +  \\
-& 0.20\text{ * employment_density} + 0.11\text{ * automobile availability}
-\end{align}
+$$
+\begin{aligned}
+\text{Transit Market Index } = & \text{ 0.64 * Population Density + 0.20 * Employment Density +} \\
+& \text{0.23 * Intersection Density + 0.11 * Automobile Availability}
+\end{aligned}
+$$
 
@@
 
-## Data Preperation
+## Data Preparation
 
@@
 
-### Poisson Regression
+### Model 1: Poisson Regression
 
@@
 
-\begin{align}
+$$
+\begin{aligned}
 Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
@@
 \beta_0, \beta & \sim \text{Normal}(0, 1) \\
-\end{align}
+\end{aligned}
+$$
 
@@
 **Figure ...** shows this model written in Stan.
+The model was fit with four chains of 10,000 iterations each, although convergence was confirmed after only 2,000 iterations.
+After burn-in, or the warmup period, Model 1 contains 20,000 samples.
 
@@
 # FIND A BETTER WAY TO PRINT THIS
-writeLines(readLines("~/Documents/honors/honors-project/stan/poisson.stan"))
+writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/poisson.stan"))
+```
+
+Table 1 reports 95% confidence intervals for each of the 19 parameters included in $x$ as well as diagnostics $\hat{R}$ and $n_eff$.
+For Model 1, $n_eff > N$, which indicates "better than independent" (https://discourse.mc-stan.org/t/n-eff-bda3-vs-stan/2608/50, citing a forum?
+by stan developers?) draws.
+Simply put, the diagnostics suggest that we can trust this model.
+
+Note that many $\beta_k$ are close to zero, but few contain zero within their 95% confidence interval.
+The small magnitude of $\beta_k$ estimates are because parameters $x$ are normalized to have standard deviation equal to 1.
+
+\begin{table}[!h]
+\centering
+\begin{tabular}{lrrrrrrr}
+  \hline
+Parameter & Rhat & n\_eff & mean & sd & 2.5\% & 50\% & 97.5\% \\ 
+  \hline
+beta\_0 & 1.00 & 25223 & -1.38 & 0.00 & -1.39 & -1.38 & -1.37 \\ 
+  beta[1] & 1.00 & 20629 & 0.05 & 0.01 & 0.04 & 0.05 & 0.07 \\ 
+  beta[2] & 1.00 & 22559 & 0.07 & 0.00 & 0.06 & 0.07 & 0.08 \\ 
+  beta[3] & 1.00 & 22078 & -0.01 & 0.00 & -0.02 & -0.01 & 0.00 \\ 
+  beta[4] & 1.00 & 20946 & -0.02 & 0.00 & -0.02 & -0.02 & -0.01 \\ 
+  beta[5] & 1.00 & 20568 & -0.02 & 0.00 & -0.03 & -0.02 & -0.01 \\ 
+  beta[6] & 1.00 & 16294 & 0.21 & 0.01 & 0.20 & 0.21 & 0.22 \\ 
+  beta[7] & 1.00 & 20239 & 0.04 & 0.00 & 0.03 & 0.04 & 0.05 \\ 
+  beta[8] & 1.00 & 18051 & -0.01 & 0.00 & -0.02 & -0.01 & -0.00 \\ 
+  beta[9] & 1.00 & 21361 & 0.10 & 0.00 & 0.10 & 0.10 & 0.11 \\ 
+  beta[10] & 1.00 & 16390 & -0.01 & 0.00 & -0.01 & -0.01 & -0.00 \\ 
+  beta[11] & 1.00 & 17378 & 0.10 & 0.00 & 0.09 & 0.10 & 0.10 \\ 
+  beta[12] & 1.00 & 21918 & -0.11 & 0.00 & -0.12 & -0.11 & -0.11 \\ 
+  beta[13] & 1.00 & 26748 & -0.08 & 0.00 & -0.09 & -0.08 & -0.07 \\ 
+  beta[14] & 1.00 & 20707 & -0.00 & 0.01 & -0.01 & -0.00 & 0.01 \\ 
+  beta[15] & 1.00 & 15443 & -0.02 & 0.00 & -0.03 & -0.02 & -0.01 \\ 
+  beta[16] & 1.00 & 15498 & 0.13 & 0.00 & 0.12 & 0.13 & 0.14 \\ 
+  beta[17] & 1.00 & 23302 & -0.09 & 0.00 & -0.09 & -0.09 & -0.08 \\ 
+  beta[18] & 1.00 & 26155 & 0.03 & 0.00 & 0.02 & 0.03 & 0.04 \\ 
+  beta[19] & 1.00 & 24639 & 0.01 & 0.00 & 0.01 & 0.01 & 0.02 \\ 
+   \hline
+\end{tabular}
+\caption{Table 1: Parameter estimates and HMC/NUTS diagnostics for Model 1, the simple Poisson GLM} 
+\end{table}
+
+Based on the diagnostics above, we have reasonable confirmation that the MCMC algorithm fit the model appropriately.
+Beyond convergence, however, we need to examine model fit.
+**Figure  __** shows the actual distribution of $Y_i$ in dark blue with 100 samples from values of $\hat{Y}_i$ simulated by Model 1 in light blue.
+Immediately, it is clear that Model 1 is underperforming.
+Model 1 underestimates the number of block groups with low ridership and overestimates the number of block groups with high ridership, and the ridership within those block groups.
+Table 2 reports minima and maxima of $Y_i$ and $\hat{Y}_i$ for comparison.
+
+```{r}
+mod_dat <- readRDS('~/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
+mod_dat <- na.omit(mod_dat)
+setDT(mod_dat)
+
+xdat <- mod_dat[, -c('GEOID', 'daily_boards', 'daily_stops', 'sqkm', 'estimate_tot_pop', 'daily_alights', 'num_interpolated', 'num_routes', 'daily_activity')]
+
+y <- round(mod_dat$daily_boards)
+E <- mod_dat$daily_stops
+
+N <- nrow(mod_dat)
+
+K <- ncol(xdat)
+x <- as.matrix(ncol = K, xdat)
+
+poisson_fit <- readRDS("~/Documents/honors/honors-project/final-fits/poisson_fit.RDS")
+
+post <- rstan::extract(poisson_fit)
+lat <- post$f
+samp <- matrix(0, nrow=4000, ncol=1495)
+
+for(i in 1:nrow(samp)){
+  samp[i, ] <- rpois(1495, exp(lat[i, ]))
+}
+
+setDT(as.data.frame(samp))
+
+pois <- pp_check(y, samp[1:100, ], ppc_dens_overlay) + 
+  theme_minimal() +
+  labs(title = "Poisson Actual versus fitted") + xlim(0, 300) + ylim(0, 0.025)
+
+pois
 ```
@@
 ```{r}
-writeLines(readLines("~/Documents/honors/honors-project/stan/poisson_horseshoe.stan"))
+#writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/poisson_horseshoe.stan"))
 ```

### 2020-01-27 chapters/03_methods.Rmd
update methods, big picture mostly complete

@@
 output:
+  pdf_document:
+    number_sections: true
   word_document:
-    toc: yes
-  pdf_document: null
   html_document:
@@
 Each model presented in this honors is built upon a simple Poisson regression.
-Poisson regressions are used to model count data.
+Poisson regressions are a form of generalized linear models with the natural log as its link function.
+They are generally used to model count data.
 The model assumes that the natural log of the mean of response variable $Y$ can be modeled by a linear combination of predictors.
-Poisson regressions are a form of generalized linear models with the natural log as its link function.
 For response variable $Y$ and covariates $x_1, x_2, ..., x_k$, the basic Poisson regression can be written
@@
 
+In the basic Poisson regression, we often give $\beta_0$ and $\beta$ vague priors such as Normal($0, 1$). The exposure term $E_i$ comes from the data and does not recieve a prior.
+
+## Overdispersion
+
 The most obvious assumption made in a Poisson regression is that response variable $Y$ follows a Poisson distribution.
-The primary assumption this makes about our data is that $\text{Var}(Y) = \text{E}(Y)$.
-When this assumption is not met, we can either use a generalization of the Poisson distribution, such as a Negative-Binomial distribution, or add a parameter to account for extra-Poisson variance.
+In addition to requiring integer counts, this assumption requires that $\text{Var}(Y) = \text{E}(Y)$.
+This is often not true in practice.
+In the case of the Metro Transit ridership data, $\text{Var}(Y) \approx 7*\text{E}(Y)$.
+In this case, we call the unexpectedly large variance "overdispersion" or "extra-Poisson variance."
 
-For the purposes of this study, we can add a set of heterogeneous random effects $\theta$ to account for overdispersion (where Var($Y$) > E($Y$)).
-Following Morris et al (2019), the second line of **1** can be written
+Data with this feature can be difficult to model directly with a Poisson regression.
+One option is to use a generalization of the Poisson distribution, such as a Negative-Binomial distribution, as the observation model.
+For the Metro Transit case study, we instead modify the latent function to include a seperate term to account for overdispersion.
+As such, the Poisson regression above can be rewritten to include
 
@@
 
-where $\theta$ is given a vague normal prior centered around 0, such as $\theta \sim$ Normal(0, 1).
-The addition of these random effects improve fits on data with higher-than-expected variance.
-With sufficiently vague priors, $\theta$ can account for most or all overdispersion not accounted for by $\beta_0 + \sum_{k=1}^{K}x_k^T\beta_k$.
+where $\theta$ is a set of heterogenous random effects.
+The addition of this random effects term improves fits on data with overdispersion.
+With sufficiently vague priors, $\theta$ can account for most or all of the overdispersion not accounted for by $\beta_0 + \sum_{k=1}^{K}x_k^T\beta_k$.
 
@@
 
-## Spike and Slab
+## Spike and Slab -- needs graphics
 
@@
 
-## Horseshoe
+## Horseshoe and Regularized Horseshoe -- needs graphics
 
@@
 
-The horseshoe prior, proposed by **Carvalho et all, 2010** is a popular choice in Bayesian literature, in part because it is similar to the "gold standard" spike-and-slab.
-Following the notation of **Vehtari & Pironeen**, the horseshoe prior can be expressed
+The horseshoe prior, proposed by **Carvalho et al, 2010** is a popular choice in Bayesian literature, in part because it is similar to the "gold standard" spike-and-slab.
+Following the notation of **Vehtari & Piironen**, the horseshoe prior can be expressed
 
@@
 
-# Spatial Structure
+# Spatial Structure -- needs many graphics
 
@@
 
-### This whole section follows Morris et al (probably too) closely right now
-
+### This whole section follows Morris et al (probably too) closely right now -- encorporate more from Riebler, Simpson, Besag
 We can encode spatial information into Bayesian models in several ways.
@@
 For irregularly shaped areal units, such as census tracts, we generally default to a queen neighborhood structure wherein areal units that share *any* points of contact are deemed neighbors.
-For $M$ regions represented in a $M$x$M$ neighborhood matrix $W$, $w_{ij} = 1$ if regions $i$ and $j$ are neighbors and 0 otherwise.
+For $N$ regions represented in a $N$x$N$ neighborhood matrix $W$, $w_{ij} = 1$ if regions $i$ and $j$ are neighbors and 0 otherwise.
 Matrix $W$ is symmetric.
@@
 The spatial interactions between $i$ and $j$ are represented by random variable $\phi$.
-Here, $\phi$ is a vector $\phi = \phi_1, \phi_2, ..., \phi_M$.
+Here, $\phi$ is a vector $\phi = \phi_1, \phi_2, ..., \phi_N$.
 The distribution of each $\phi_i$ is determined by the sum of its neighbors values such that
 
-# having a notation crisis
-
 $$
-\phi_i \text{ | } \phi_j, j \neq i \sim \text{Normal}(\sum_{i=1}^{M}w_{ij}\phi_j, \sigma^2) 
+\phi_i \text{ | } \phi_j, j \neq i \sim \text{Normal}(\sum_{i=1}^{N}w_{ij}\phi_j, \sigma^2) 
 $$
@@
 
-The ICAR specifies each $\phi_i$ to be normally distributed with its mean equal to the mean of its neighbors values.
+The ICAR prior specifies each $\phi_i$ to be normally distributed with its mean equal to the mean of its neighbor's values.
 This is how the CAR/ICAR model "borrows strength" from geographic neighbors to smooth noisy estimates.
@@
 The conditional specification of each $\phi_i$ is helpful in building intuition about the assumptions the ICAR component makes and how it performs its "smoothing".
+The ICAR prior matches our intuition about spatial autocorrelation: areal unit $i$ is similar to the areal units surrounding it, and the more areal units surround it, the more confident we feel in that similarity.
 The joint distribution can be rewritten
@@
 
-This pairwise difference specification is also helpful in building intuition about how ICAR models perform smoothing.
+This pairwise difference specification is also helpful in building intuition about how ICAR models perform smoothing. Each $(\phi_i - \phi_j)^2$ works as a penalty term: the ???
 
@@
 The specific ICAR model used in this analysis is the Besag-York-Mollie (BYM) Poisson model.
-The BYM model contains random effects and a spatial ICAR term to account for both spatial and non-spatial heterogeneity.
-Note that in **section??*, we introduced a random effect term $\theta$.
+The BYM is a "classical" spatial Bayesian method (Riebler et al, 2016).
+It is generally used in disease mapping studies to model rare events (ie, disease occurrence) in small areal units.
+The BYM capitalizes on the CAR prior's ability to smooth noisy estimates and share information across geographic units.
+
+The latent function in the BYM model contains heterogeneous random effects as well as a spatial ICAR term in order to account for both spatial and non-spatial heterogeneity.
+Note that in Section 1.1, we introduced a random effect term $\theta$.
 Therefore, the BYM model can be though of as decomposing $\theta$ into spatial and non-spatial components.
 
-Specifically, the BYM model is written
+Specifically, the BYM model replaces $\eta_i$ from equation *1* with
 
@@
 where $\phi$ is the addition: an ICAR component.
+Decomposing overdispersion as such is helpful in furthering our understanding of response $Y$.
+When we use a simple random effect $\theta$ alone, we can accomplish a good model fit.
+However, $\theta$ provides no additional information about $Y$ or why it varies so much.
+The decomposition of $\theta$ in the BYM allows us to better quantify how much variance is white noise (ordinary random effects) versus some unmeasured confounding correlated across space.
 
-The BYM model is appealing in its simplicity.
-However, the lack of hyperpriors specified for $\theta$ and $\phi$ can make the model incredibly challenging to fit.
+The BYM model as written above is appealing in its simplicity.
+However, the lack of informative hyperpriors specified for $\theta$ and $\phi$ can make the model incredibly challenging to fit.
 In theory, extra-Poisson variance could be explained 100% by $\theta$, 100% by $\phi$, or anywhere in between.
@@
 There are some suggestions in the literature for setting hyperpriors for $\theta$ and $\phi$ (Besag and Mollie, 1991; Clayton and Montomoli, 1995).
-The existing methods, however, tend to rely on the specific data at hand which makes the selection of hyperparameters time unnecessarily time consuming.
+The existing methods, however, tend to rely on the specific data at hand which makes the selection of hyperparameters unnecessarily time consuming.
 
 The BYM2 model proposed by Simpson et al and further described in Riebler et al (2016) aims to solve these sampling problems.
+The BYM2 takes a more "fully Bayesian" approach to hyperprior/hyperparameter selection.
 The primary difference between the BYM and BYM2 models is the addition of a mixing parameter, $\rho$.
@@
 $$
-\eta_i = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + ((\sqrt{\frac{\rho}{s}})\theta^{*} + ((\sqrt{1-\rho})\phi)\sigma
+\eta_i = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + ((\sqrt{\frac{\rho}{s}})\theta^{*} + ((\sqrt{1-\rho})\phi^{*})\sigma
 $$
@@
 
-A relatively vague Normal(0, 1) prior is appropriate for $\sigma$. Mixing parameter $\rho$ is given a Beta(1/2, 1/2) prior. Similar to the horseshoe,
+A relatively vague Normal(0, 1) prior is appropriate for $\sigma$.
+Mixing parameter $\rho$ is given a Beta(1/2, 1/2) prior.
+Riebler et al (2016) also propose a more complex prior for $\rho$ such that ____ Similar to the horseshoe,
 
@@
 
-The models above can all be fit using the Stan Programming Langauge (cite). Stan is probabilistic programming language used to compute joint log probability densities.
+The models above can all be fit using the Stan Programming Langauge (cite).
+Stan is probabilistic programming language used to compute joint log probability densities.
+The analysis in this honors uses Stan Version ____.
 
@@
 
-Stan utilizes the Hamiltonian Monte Carlo (HMC) and specialized No U-Turn Sampler (NUTS).
+Stan utilizes Hamiltonian Monte Carlo (HMC) and specialized No U-Turn Sampler (NUTS).
+The technicalities of Stan's sampling process is beyond the scope of this paper.
+However, there are a few sampling parameters which are important to understand in order to optimize model fits.
+Specifically, the parameters `max_treedepth` and `adapt_delta` appear in the application portion of this honors.
+
+The `max_treedepth` argument specifies how many steps the sampler is allowed to make on a given transition.
+Stan's default max treedepth is 10.
+When the default is used, the sampler makes up to 10 steps.
+If it reaches 10 leapfrog steps without ___, the sampler stops for that iteration and Stan informs the user that maximum treedepth has been exceeded.
+When a model experiences many iterations reaching maximum treedepth, there are several considerations to make.
+
+The `adapt_delta` argument works hand in hand with `max_treedepth`.
 

### 2020-01-27 functions/ppcheck.R
add start of ppcheck function

@@ -1,0 +1,57 @@
+library(bayesplot)
+library(rstan)
+library(loo)
+library(data.table)
+library(ggplot2)
+
+## for testing
+
+post <- readRDS("~/Documents/honors/honors-project/final-fits/poisson_fit.RDS")
+n_obs = 1495
+n_iter = 20000
+
+# function will take stanfit, observed y, number observations, number iterations, pars for trace, 
+
+ppcheck <- function(post, y, n_obs, n_iter){
+
+## get posterior draws
+
+  samp <- rstan::extract(post)
+  samp2 <- as.array(post)
+  f <- samp$f
+  
+  yrep <- matrix(0, nrow=n_iter, ncol=n_obs)
+  
+  for(i in 1:nrow(yrep)){
+    yrep[i, ] <- rpois(n_obs, exp(f[i, ]))
+  }
+  
+  setDT(as.data.frame(yrep))
+  
+  ppoverlay <- pp_check(y, yrep[sample(100), ], ppc_dens_overlay) + 
+    theme_minimal() +
+    labs(title = "Poisson Actual versus fitted") 
+  
+  ## get traceplot
+  
+  trace <- mcmc_trace(samp2, pars = c("beta_0"))
+  
+  ## get parcoord
+  
+  np <-  nuts_params(post)
+  
+  pcoord <- mcmc_parcoord(
+    samp2,
+    regex_pars = "beta",
+    transform = function(x) {(x - mean(x)) / sd(x)},
+    size = 0.25,
+    alpha = 0.1,
+    np = np
+  )
+  
+  ## loo
+  
+  
+  ## map
+}
+

### 2020-01-27 chapters/03_methods.Rmd
work with testing ppcheck function, etc

@@
 
+## Spatial Autocorrelation
+
 ## Conditional Autoregressive Priors
 
-### This whole section follows Morris et al (probably too) closely right now -- encorporate more from Riebler, Simpson, Besag
+### This whole section follows Morris et al (probably too) closely right now -- incorporate more from Riebler, Simpson, Besag
+
 We can encode spatial information into Bayesian models in several ways.
@@
 
-Precision matrix $Q$ can be defined for multivariate normal $\phi$ using two other $M$x$M$ matrices: the neighborhood matrix $W$ as well as diagonal matrix $D$ in which $d_{ii}$ indicates the number of neighbors region $i$ has.
+Precision matrix $Q$ can be defined for multivariate normal $\phi$ using two other $N$x$N$ matrices: the neighborhood matrix $W$ as well as diagonal matrix $D$ in which $d_{ii}$ indicates the number of neighbors region $i$ has.
 All off-diagonal entries are 0.
@@
 Various graphical checks of priors and posteriors are helpful tools in building, improving, and comparing Bayesian models.
-The application portion of this honors follows Michael Betancourt's "Towards a Principled Bayesian Workflow" as well as ______ in terms of using visual aids for diagnosing issues with models and comparing quality of fit between models.
+The application portion of this honors follows Michael Betancourt's "Towards a Principled Bayesian Workflow" as well as Gabry et al's "Visualization in a Bayesian Workflow" in terms of using visual aids for diagnosing issues with models and comparing quality of fit between models.
 Only the most "useful" plots are included below, for more on using graphics throughout a Bayesian workflow see Betancourt or Gabry.

### 2020-01-27 chapters/04_application.Rmd
work with testing ppcheck function, etc

@@
 author: "Raven McKnight"
-date: "1/16/2020"
 output:
   pdf_document:
-    toc: yes
+    number_sections: true
   html_document:
@@
 
-# Case Study
-
-## Background
+# Background
 
@@
 
-## Data Description
+# Data Description
 
-### Metro Transit Data
+## Metro Transit Data
 
-### Covariates
+## Covariates
 
-## Data Preparation
+# Data Preparation
 
-## Modeling
+# Modeling
 
-### Model 1: Poisson Regression
+## Model 1: Poisson Regression
 
@@
 
-### Horseshoe Prior
+## Horseshoe Prior
 
@@
 
-#### Divergences
+## Divergences
 
@@
 
-#### Variables Selected
+## Variables Selected
 
@@
 
-### Overdispersion
+## Overdispersion
 
@@
 
-### Spatial Structure
+## Spatial Structure

### 2020-01-27 functions/ppcheck.R

@@ -6,9 +6,9 @@ library(ggplot2)
 
 ## for testing
 
-post <- readRDS("~/Documents/honors/honors-project/final-fits/poisson_fit.RDS")
+post <- readRDS("~/Documents/honors/honors-project/test_post.RDS")
 n_obs = 1495
-n_iter = 20000
+n_iter = 2000
 
 # function will take stanfit, observed y, number observations, number iterations, pars for trace, 
 
@@ -51,7 +51,21 @@ ppcheck <- function(post, y, n_obs, n_iter){
   
   ## loo
   
+  loglik <- as.matrix(post, pars = "log_lik")
+  loo <- loo(loglik)
+  psis <- psislw(-loglik)
+  pit <- rstantools::loo_pit(object = yrep, y=y, lw = as.matrix(psis$lw_smooth))
+  
+  unifs <- matrix(runif(length(pit) * 100), nrow = 100)
+  coords <- coord_cartesian(ylim = c(0, 2), xlim = c(0.1, 0.9))
+  
+  ppc_dens_overlay(pit, unifs) + legend_none() + coords
+  
+  
+  ## quadrant
   
   ## map
+  
+  
 }
 

### 2020-02-05 scripts/aws/fitting.R
update fitting script thru bym2

@@ -6,15 +6,21 @@ library(data.table)
 library(shinystan)
 library(ggplot2)
 library(bayesplot)
+library(tigris)
+library(spdep)
+library(sf)
 options(mc.cores = parallel::detectCores())
+options(tigris_use_cache = TRUE)
 rstan_options(autowrite = T)
+options(tigris_class = "sf")
 
 ## data
 mod_dat <- readRDS('~/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
 mod_dat <- na.omit(mod_dat)
 setDT(mod_dat)
 
-xdat <- mod_dat[, -c('GEOID', 'daily_boards', 'daily_stops', 'sqkm', 'estimate_tot_pop', 'daily_alights', 'num_interpolated', 'num_routes', 'daily_activity')]
+xdat <- mod_dat[, -c('GEOID', 'daily_boards', 'daily_stops', 'sqkm', 'estimate_tot_pop', 'daily_alights', 
+                     'num_interpolated', 'num_routes', 'daily_activity', 'w_total_jobs_here')]
 
 ## regular poisson regression -------------------
 
@@ -47,33 +53,55 @@ poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, iter = 10000, verbose
                         control = list(adapt_delta = 0.99, max_treedepth = 15), seed = 1997)
 saveRDS(poisson_hrs_fit, '~/Documents/honors/honors-project/final-fits/poisson_hrs_fit.RDS')
 
+#so... it's run with no divergences before, but with a predictor i don't want to include.... i think i'll use this with a disclaimer
 
 ## add overdispersion parameter -----------------
 
-poisson_theta <- "~/Documents/honors/honors-project/stan/final-stan/poisson_theta.stan"
+# for computational efficiently, we'll drop the "0" covariates now
+xdat[, 8]
+xdat[, 1]
 
-xdat2 <- xdat[, c(3, 8, 10, 14) := NULL]
-x2 <- as.matrix(ncol = K, xdat2)
+xdat2 <- xdat[, -c(1, 8)]
 K2 <- ncol(xdat2)
+x2 <- as.matrix(xdat2, ncol = K2)
 
 standat2 <- list(y = y, E = E, x = x2, K = K2, N = N)
 
[…]

### 2020-02-05 scripts/aws/nb_data_funs.R

@@ -1,0 +1,278 @@
+library(INLA);
+library(spdep);
+
+
+# validateNb
+# check that nbObject is symmetric, has connected components
+#
+validateNb = function(x) {
+  if (is.symmetric.nb(x) && n.comp.nb(x)[[1]] < length(x)) return(TRUE);
+  return(FALSE);
+}
+
+
+# isDisconnected
+# check that nbObject has more than 1 component
+#
+isDisconnected = function(x) {
+  return(n.comp.nb(x)[[1]] > 1);
+}
+
+
+# indexByComponent
+#
+# input: vector of component ids
+# returns: vector of per-component consecutive node ids
+#
+indexByComponent = function(x) {
+  y = x;
+  comps = as.matrix(table(x));
+  num_comps = nrow(comps);
+  for (i in 1:nrow(comps)) {
+    idx = 1;
+    rel_idx = 1;
+    while (idx <= length(x)) {
+      if (x[idx] == i) {
+        y[idx] = rel_idx;
+        rel_idx = rel_idx + 1;
+      }
+      idx = idx + 1;
+    }
+  }
+  return(y);
+}
+
+
+# nb2graph
+#
+# input: nb_object
+# returns: dataframe containing num nodes, num edges,
+#          and a list of graph edges from node1 to node2.
+#
+nb2graph = function(x) {
+  N = length(x);
+  n_links = 0;
+  for (i in 1:N) {
+    if (x[[i]][1] != 0) {
+      n_links = n_links + length(x[[i]]);
+    }
+  }
+  N_edges = n_links / 2;
+  node1 = vector(mode="numeric", length=N_edges);
+  node2 = vector(mode="numeric", length=N_edges);
+  idx = 0;
+  for (i in 1:N) {
+    if (x[[i]][1] > 0) {
+      for (j in 1:length(x[[i]])) {
+        n2 = unlist(x[[i]][j]);
+        if (i < n2) {
+          idx = idx + 1;
+          node1[idx] = i;
+          node2[idx] = n2;
+        }
+      }
+    }
+  }
+  return (list("N"=N,"N_edges"=N_edges,"node1"=node1,"node2"=node2));
+}
+
+
+# nb2subgraph
+# for a given subcomponent, return graph as lists of node1, node2 pairs
+#
+# inputs:
+# x: nb object
+# c_id: subcomponent id
+# comp_ids: vector of subcomponent ids
+# offsets: vector of subcomponent node numberings
+# returns: list of node1, node2 ids
+#
[…]

### 2020-02-05 scripts/aws/fitting.R
last updates from wed @ work

@@ -105,3 +105,5 @@ bym_dat <- list(y = y2, E = E2, x = x2, K = K2,
 bym_fit <- stan(bym2, data = bym_dat, iter = 10000, verbose = T, seed = 1997, 
                 control = list(max_treedepth = 12))
 saveRDS(bym_fit, "~/Documents/honors/honors-project/final-fits/bym.RDS")
+
+## bym with horseshoe?? -------------------------

### 2020-02-08 chapters/04_application.Rmd
testing bymhrs

@@
 
-Census block groups are split into 5 market areas based on the Transit Market Index described above, where the highest market area (Market Area 1) is expected to support high-frequency, all-day service and the lowest market area (Market Area 5) is expected to support peak commuter express service, if that.
+Census block groups are split into 5 market areas based on the Transit Market Index described above, where the highest market area (Market Area 1) is expected to support high-frequency, all-day service and the lowest market area (Market Area 5) is expected to support peak commuter express service and park-and-rides, if that.
 
 The Transit Market Index values are geographically smoothed to create more-or-less concentric Transit Market Areas.
+The linear regression is an intuitive way to think about transit ridership across space.
+The four predictors used are common-sense indicators of transit ridership: high population and employment density are characteristic of trip origins and destinations, automobile availability is reasonably assumed to be related to transit ridership, and intersection density is a proxy for the "walkability" of an area.
+However, the exsiting Transit Market Area model may be over-simplifying the complex question of transit ridership.
+
+Transit ridership has been in decline in the Twin Cities, and across the country, for several years (Cite).
 
@@
 
+This analysis relies on several Metro Transit provided data sets. The two primary sources are automatically reported by in-service vehicles, yielding billions of rows of observations.
+
+**Automatic Passenger Count (APC)** data is reported every time a bus opens its doors.
+Two beams of light detect movement through the doors of the bus and counts the number of passengers getting on and off at each bus stop.
+Naturally, this data source is flawed: sometimes, vehicles fail to report data, and the sensors can be easily tricked.
+Someone getting off the bus pulling a suitcase behind them, for example, would likely be counted as two passengers exiting.
+However, APC is the most granular ridership data available, giving us the most control over our spatial aggregations of ridership.
+This is the primary data source used in this analysis.
+
+**Automatic Vehicle Location (AVL)** data is similarly reported by in-service vehicles (approximately) every 8 seconds.
+The vehicle reports in GPS location continuously while in service.
+We use AVL data to adjust for missing APC data.
+
+**Metro Transit's Schedule** as *planned*, rather than as *run*, is also used to correct missing APC data.
+
+This analysis began by pulling all APC, AVL, and schedule data from 2015-2018. The analysis presented here focuses on 2017 data.
+
+### Data Interpolation
+
+In theory, we have APC data for each run bus trip in 2017.
+However, we know (how to cite Metro Transit knowledge?) that, for a variety of reasons, this is untrue.
+The automatic data reporting technology on Metro Transit's buses ///
+
 ## Covariates
-
-# Data Preparation
 

### 2020-02-08 07_figures.R

@@ -1,0 +1,20 @@
+library(ggplot2)
+library(sf)
+library(tigris)
+library(dplyr)
+library(viridis)
+
+options(tigris_class = "sf")
+options(tigris_use_cache = TRUE)
+
+mod_dat <- readRDS('~/Documents/honors/honors-project/data/modeling-dat/p_dat_scaled.RDS')
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+bgs <- block_groups("MN", counties, 2017)
+
+sfdat <- left_join(bgs, mod_dat, on = 'GEOID')
+
+ggplot(sfdat) + 
+  geom_sf(aes(fill = log(daily_boards)), color = "grey", lwd = .1) +
+  theme_minimal() +
+  scale_fill_viridis(na.value = "transparent", option = "plasma") +
+  labs(fill = "Daily Boardings")

### 2020-02-08 scripts/aws/fitting.R
duh push script to actually fit new mod

@@ -42,16 +42,16 @@ horseshoe_dat <- list(y = y, E = E, x = x, K = K, N = N,
 ## basic poisson --------------------------------
 
 poisson <- "~/Documents/honors/honors-project/stan/final-stan/poisson.stan"
-poisson_fit <- stan(poisson, data = standat1, iter = 10000, verbose = T, seed = 1997)
+#poisson_fit <- stan(poisson, data = standat1, iter = 10000, verbose = T, seed = 1997)
 
-saveRDS(poisson_fit, '~/Documents/honors/honors-project/final-fits/poisson_fit.RDS')
+#saveRDS(poisson_fit, '~/Documents/honors/honors-project/final-fits/poisson_fit.RDS')
 
 ## add horseshoe priors -------------------------
 
 poisson_hrs <- "~/Documents/honors/honors-project/stan/final-stan/poisson_horseshoe.stan"
-poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, iter = 10000, verbose = T,
+#poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, iter = 10000, verbose = T,
                         control = list(adapt_delta = 0.99, max_treedepth = 15), seed = 1997)
-saveRDS(poisson_hrs_fit, '~/Documents/honors/honors-project/final-fits/poisson_hrs_fit.RDS')
+#saveRDS(poisson_hrs_fit, '~/Documents/honors/honors-project/final-fits/poisson_hrs_fit.RDS')
 
 #so... it's run with no divergences before, but with a predictor i don't want to include.... i think i'll use this with a disclaimer
 
@@ -70,8 +70,8 @@ standat2 <- list(y = y, E = E, x = x2, K = K2, N = N)
 poisson_theta <- "~/Documents/honors/honors-project/stan/final-stan/poisson_theta.stan"
 
 
-poisson_theta_fit <- stan(poisson_theta, data = standat2, iter = 20000, verbose = T, seed = 1997)
-saveRDS(poisson_theta_fit, "~/Documents/honors/honors-project/final-fits/poisson_theta.RDS")
+#poisson_theta_fit <- stan(poisson_theta, data = standat2, iter = 20000, verbose = T, seed = 1997)
+#saveRDS(poisson_theta_fit, "~/Documents/honors/honors-project/final-fits/poisson_theta.RDS")
 
 ## and the bym ! --------------------------------
 
[…]

### 2020-02-10 chapters/03_methods.Rmd
add messy draft

@@
 While this is a straightforward assumption, it can improve model fits in several ways.
-First, it can encode prior information about response $Y$ not otherwise represented by covariates $x$ (ie, that nearby observations are more similar than far flung ones).
+First, it can encode prior information about response $Y$ not otherwise represented by covariates $x$ (ie, that nearby observations are more similar than far-flung ones).
 Second, it can provide geographic smoothing when observations are sparse or noisy.

### 2020-02-10 chapters/article-draft.Rmd
add messy draft

---
title: "Honors Draft"
author: "Raven McKnight"
header-includes:
- \usepackage{booktabs}
- \usepackage{makecell}
- \usepackage{float}
date: "2/9/2020"
output:
  pdf_document:
    number_sections: true
  html_document:
    theme: paper
    toc: yes
    toc_float: yes
abstract: "abstract"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Public transit ridership has been in decline around the country for years, but particularly since the mid 2010s (). Following a spike in ridership related to the 2008 housing market crash, ridership has been steadily trending downwards in most metropolitan areas (). Bus ridership is particularly affected compared to more "desirable" light rail and commuter rail. Transit agencies are naturally interested in understanding the specific causes of these declines, as well as more generally being able to predict transit demand throughout space and time.

Transit ridership, or demand, is often forecasted using models from the policy and planning world, such as the classic four-step model. Four-step transportation demand models consist of simulating trips and their spatial distribution before assigning them modes (ie bus versus car versus bicycle) and routes. Other models, called activity-based models predict transportation demand based on when and where riders are likely to conduct various tasks (commuting, shopping, recreation, etc). Others still are based largely on land-use characteristics under the general premise that different types of development (or lack thereof) are more or less likely to draw riders.

It is somewhat rare to see more traditional statistical models applied to transit ridership. In the context of predicting future ridership, the above methods are often sufficient. Additionally, they are the industry standard and allow for the integration of expert planning knowledge. However, in the context of understanding *past* ridership, we can apply more explanatory methods. Specifically, we can use hierarchical spatial Bayesian models to incorporate many covariates in addition to the spatial structure that necessarily underpins transit ridership (ie, boardings can only occur at existing bus stops).

Hierarchical spatial models allow us to understand the patterns of ridership within a city with great granularity. Some existing literature explores the decline of transit ridership *between* cities (). Few studies have been conducted on the various patterns of ridership *within* a single metropolitan area. Additionally, few studies incorporate many demographic predictors. Naturally, we cannot collect demographic information for every rider of a transit system. Instead, we can use demographic predictors associated with spatial units such as census block groups to explore the question of *who* rides -- or, at the very least, *where* do they ride.

# Background

## Metro Transit

Metro Transit is the primary transit provider in the Minneapolis-Saint Paul metropolitan area. The agency operates one commuter rail line (Northstar), two light rail lines (Blue Line and Green Line), and over 100 bus routes. Since the 2014 opening of the Green Line, rail ridership has increased each year (). Bus ridership, however, has been in decline. According to the Metro Transit Riders' Almanac blog, much of the decline is bus ridership can be attributed to the busiest urban local routes as well as situational factors, such as lower-than-average gas prices and a system-wide fare increase in 2017.

The agency is naturally interested in understanding with more detail the reasons for and nature of these declines.

Conducted in tandem with Metro Transit's Network Next bus system redesign, this study aims to further Metro Transit's understanding of their bus network. Specifically, we aim to describe the existing pattern of transit ridership via demographic predictors and spatial patterns. Future work will incorporate temporal trends to better understand declining ridership.

## Transit Market Areas

Metro is interested in predicting transit demand across the region in order to guide route planning and service provision. Currently, Metro Transit uses a set of 5 Transit Market Areas to estimate spatial demand. The Transit Market Areas are writen into the official Transportation Planning Policy (TPP) and are calculated using a simple linear regression. Following the notation of the TPP, the formula for determing Transit Market Areas is expressed

$$
\begin{aligned}
\text{Transit Market Index } = & \text{ 0.64 * Population Density + 0.20 * Employment Density +} \\
& \text{0.23 * Intersection Density + 0.11 * Automobile Availability}
\end{aligned}
$$

where each predictor is logged and scaled by developed land acreage per census block group. Here, automobile availability refers to the number of adults over age 16 less the total number of automobiles available in a block group (scaled by developed land acreage). There is an additional indicator variable, omitted in this notation, for the census block group containing the MSP International Airport. For more documentation of the official Transit Market Areas, refer to the TPPs [Appendix G](https://metrocouncil.org/Transportation/Planning-2/Key-Transportation-Planning-Documents/Transportation-Policy-Plan/The-Adopted-2040-TPP-(1)/Final-2040-Transportation-Policy-Plan/2040-TPP-Appendix-G-Transit-Design-and-Perf-Standa.aspx).

\begin{figure}[htb]
  \centering
  \includegraphics[width = 5in]{TMA.png}
  \caption{Metro Transit's Transit Market Areas from Metropolitan Council's Transportation Planning Policy, Appendix G. }
\end{figure}

Census block groups are split into 5 market areas based on the Transit Market Index described above, where the highest market area (Market Area 1) is expected to support high-frequency, all-day service and the lowest market area (Market Area 5) is expected to support peak commuter express service and park-and-rides, if that.

The Transit Market Index values are geographically smoothed to create more-or-less concentric Transit Market Areas. The linear regression is an intuitive way to think about transit ridership across space. The four predictors used are common-sense indicators of transit ridership: high population and employment density are characteristic of trip origins and destinations, automobile availability is reasonably assumed to be related to transit ridership, and intersection density is a proxy for the "walkability" of an area. However, the exsiting Transit Market Area model may be over-simplifying the complex question of transit ridership.

# Data

## Metro Transit Data

This analysis relies on several Metro Transit provided data sets. The two primary sources are automatically reported by in-service vehicles, yielding billions of rows of observations.

**Automatic Passenger Count (APC)** data is reported every time a bus opens its doors. Two beams of light detect movement through the doors of the bus and counts the number of passengers getting on and off at each bus stop. Naturally, this data source is flawed: sometimes, vehicles fail to report data, and the sensors can be easily tricked. Someone getting off the bus pulling a suitcase behind them, for example, would likely be counted as two passengers exiting. However, APC is the most granular ridership data available, giving us the most control over our spatial aggregations of ridership. This is the primary data source used in this analysis.

**Automatic Vehicle Location (AVL)** data is similarly reported by in-service vehicles (approximately) every 8 seconds. The vehicle reports in GPS location continuously while in service. We use AVL data to adjust for missing APC data.

**Metro Transit's Schedule** as *planned*, rather than as *run*, is also used to correct missing APC data.

This analysis began by pulling all APC, AVL, and schedule data from 2015-2018. The initial data pull consisted of approximately 496 *billion* rows of data. The analysis presented here focuses on 2017 data. Figure *___* shows the average weekday boardings by census block group in 2017.

\begin{figure}[H]
  \centering
  \includegraphics[width = 5in]{weekdayboardings.png}
  \caption{Average weekday boardings by census block group, the response variable for this analysis.}
\end{figure}

### Data Interpolation

In theory, we have APC data for each run bus trip in 2017. However, we know that, for a variety of reasons, this is untrue. The automatic data reporting technology on Metro Transit buses is flawed and can fail to report data unpredictably. Additionally, the trips missing data may not be random: all buses of one "series", for example, may fail to report at once. This could lead to the APC dataset understimating boardings on a particular set of routes or locations.T herefore, we use the three data sets described above to create a more complete augmented APC data set.

The algorithm used to interpolate data works essentially as follows:

1. Compares scheduled trips to trips we have APC data for. If a trip was scheduled but has no APC records, we consider that trip missing. 
2. For missing trips, check AVL data. If there are no AVL records, we consider the trip cut (ie, we assume the trip was not run). 
3. For missing trips with AVL data, we estimate boardings for that trip by taking the average number of boardings for trips of that nature (trips on the same route, day of week, season, location, etc).

### Data Aggregation

For the models presented in this study, we aggregate boardings at individual bus stops to the census block group. This is in large part to match the granularity of the American Community Survey covariates we are interested in using. Future work could explore using areal covariates to model ridership at specific points (ie, bus stops).

Additionally, we aggregate boardings counts to average weekday boardings. This a) greatly reduces the number of rows of data in play, and therefore shortens computation time significantly, and b) allows us to build a spatial model without simultaneously incorporating temporal trends. The data interpolation described above will be more impactful in future studies incorporating temporal trends.

## Other data sources

All other data sources used in this study are from outside sources. Census Block Groups are used as the areal unit of analysis. Covariates of interest are primarily from the 2017 American Community Survey (ACS) 5-Year Estimates. Additional variables related to employment are sourced from the Longitudinal Origin-Destination Employment Statistics (LODES) dataset. A full list of covariates used in this study is below.

<!-- \begin{table}[h] -->
<!-- \centering -->
<!-- \begin{tabular}{lr} -->
<!--   \toprule -->
<!-- Parameter & meaning \\ -->
<!--   \midrule -->
<!-- beta[1] &  median household income \\ -->
<!--   beta[2] &  percent residents identified as white alone \\ -->
<!--   beta[3] &  area of 10 minute walk isochrone from population-weighted block group centroid \\ -->
<!--   beta[4] &  percent residents with a high school diploma \\ -->
<!--   beta[5] &  percent residents with a bachelor's degree \\ -->
<!--   beta[6] &  percent of residents who rent their residence \\ -->
<!--   beta[7] &  percent of residents without a vehicle \\ -->
<!--   beta[8] &  percent of residents who speak only English \\ -->
<!--   beta[9] &  percent of residents identified as foreign \\ -->
<!--   beta[10] & employment density \\ -->
<!--   beta[11] & percent of jobs in block group for white employees \\ -->
<!--   beta[12] & percent of jobs in block group for male employees \\ -->
<!--   beta[13] & percent of jobs in block group for employees with no college degree \\ -->
<!--   beta[14] &  percent of jobs in block group for employees who make less than $40,000 \\ -->
<!--   beta[15] &  percent of jobs in block group for employees under age 30 \\ -->
<!--   beta[16] &  median age \\ -->
<!--   beta[17] &  percent residents who use transit to commute to work \\ -->
<!--   beta[18] &  population density \\ -->
<!--    \bottomrule -->
<!-- \end{tabular} -->
<!-- \caption{List of covariates} -->
<!-- \end{table} -->

Note that `beta[3]` is a measure of walkability. A 10-minute walk isochrone, generated using OpenTripPlanner, shows the area a person on foot can cover in 10 minutes on the existing street network. In theory, we expect more walkable areas, such as the downtown Saint Paul and Minneapolis, to have larger walk-isochrones. Unfortunately, the metric can be fooled by rural areas where you can walk uninterruped down a single roadway, for example. An improved "walkability" metric could improve models of transit ridership.

All covariates in the table above are standardized to have mean 0 and standard deviation 1. Intuitively, we expect many of the covariates to be correlated. Therefore, we will utilize variable selection to determine which covariates to keep in our analysis.

# Models

## Poisson Regression

For each block group $i = 1, 2, 3, ..., 1495$ with any ridership, the "baseline" model is a simple Poisson regression. Poisson regressions are a type of generalized linear model with the natural log as its link function. For ridership $Y_i$ and covariates $x_1, ..., x_k$ discussed above, the Poisson regression is written

$$
\begin{aligned}
Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
\eta_i  = log(\lambda_i) & = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k \\
\beta_0, \beta &\sim \text{Normal}(0, 1)
\end{aligned}
$$

where $\beta_0$ is the intercept and $E_i$ is an offset term.

The offset $E_i$ is often termed "exposure" in applications such as disease risk modeling where the offset may be the number of observed cases in a previous year, for example. More technically, the offset term scales the Poisson output to be a rate rather than a count. This is appropriate when observations $i$ have different potentials for response $Y$. For example, a county with higher population will naturally have a higher count of patients with asthma than a county with a smaller population. Mathematically, the offset is a covariate with parameter set equal to 1. Recall, the Poisson regression assumes we can model the mean of response $Y$, $\bar{Y}$ with a combination of linear predictors:

$$
log(\bar{Y}) = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k \\
$$
Therefore, if we wish to model the rate $Y/E$, the equation is rewritten

$$
\begin{aligned}
log(\bar{Y}/E) & = \beta^{'}_0 + \sum_{k=1}^{K}x_k^T\beta^{'}_k \\
log(\bar{Y}) & = log(E) + \beta^{'}_0 + \sum_{k=1}^{K}x_k^T\beta^{'}_k
\end{aligned}
$$

In this study, we define $E_i$ to equal the number of times a bus stops in block group $i$. This allows us to control for block groups with more or less supply.

In the basic Poisson regression, we often give $\beta_0$ and $\beta$ vague priors such as Normal($0, 1$). The exposure term $E_i$ comes from the data and does not recieve a prior.

## Overdispersed Poisson Regression

The most obvious assumption made in a Poisson regression is that response variable $Y$ follows a Poisson distribution. In addition to requiring integer counts, this assumption requires that $\text{Var}(Y) = \text{E}(Y)$. This is often not true in practice. In the case of the Metro Transit ridership data, $\text{Var}(Y) >>\text{E}(Y)$. In this case, we call the unexpectedly large variance "overdispersion" or "extra-Poisson variance."

Data with this feature can be difficult to model directly with a Poisson regression. One option is to use a generalization of the Poisson distribution, such as a Negative-Binomial distribution, as the observation model. For the Metro Transit application, we instead modify the latent function to include a seperate term to account for overdispersion. As such, the Poisson regression above can be rewritten to include

$$
\begin{aligned}
\eta_i = log(\lambda_i) & = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + \theta\sigma \\
\theta &\sim \text{Normal}(0, 1) \\
\sigma &\sim \text{Normal}(0, 5)
\end{aligned}
$$

where $\theta$ is a set of heterogenous random effects. In practice, $\theta$ is scaled by its variance parameter $\sigma$ to allow eassier fitting in Stan. The addition of this set of random effects improves fits on data with overdispersion. With sufficiently vague priors, $\theta$ and $\sigma$ can account for most or all of the overdispersion not accounted for by $\beta_0 + \sum_{k=1}^{K}x_k^T\beta_k$.

## Horseshoe Priors and Variable Selection

In the baseline model, we give coefficients $\beta_1, ..., \beta_k$ simple normal priors. With a large set of possible covariates, however, we may reasonably expect some of the $\beta_k$ coefficients to be equal to zero -- in other words, that some of the parameters $x_k$ have no effect on response $Y$. In frequentist applications, we might use a method such as the LASSO (Least Absolute Shrinkage and Selection Operator) to identify relevant variables. There are several Bayesian alternatives for variable selection with generally consist of applying particular priors to $\beta$.

### Spike and Slab - needs graphics

In the Bayesian setting, there are two primary methods for variable selection. The first is called the "spike-and-slab" prior (Mitchell and Beauchamp, 1988; George and McCulloch, 1999) and has often been considered the "gold standard" for sparse Bayesian regression, or variable selection (Piironen and Vehtari, 2017). This prior is often expressed as

$$
\begin{aligned}
\beta_j | \lambda_j, c, \epsilon & \sim \lambda_j \text{Normal}(0, c^2) + (1-\lambda_j)\text{Normal}(0, \epsilon^2) \\
\lambda_j & \sim \text{Bernoulli}(\pi)
\end{aligned}
$$

for $j = 1, 2, ..., J$ where $\lambda_j \in {0, 1}$ indicates whether $\beta_j$ is from the "spike" (ie, near zero) or the "slab" (ie, nonzero). The "spike" is the area where most of the prior density for $\beta$ is centered (generally around 0) while the "slab" refers to the low-density extent of the prior. The spike-and-slab prior encourages all $\beta_j$ towards zero; only $\beta_j$ sufficiently far from 0 will be estimated to be from the slab.

The primary challenge with the spike-and-slab (and other Bayesian shrinkage methods) is the selction of priors. Priors must be set for the width of the slab $c$ and the prior inclusion probability $\pi$. Here, $\pi$ reflects our prior understanding of the sparsity of $\beta$. In practice, we rarely have strong prior knowledge of the number of predictors we expect to be distinguishable from zero. This can make the implementation of sparse regression more challenging than, say, the frequentist LASSO.

### Horseshoe - needs graphics

The second method for Bayesian sparse regression is to give $\beta$ some continuous *sparsity inducing prior*. **Van Erp et al (2019)** provide an introduction to many such continous priors. Ridge regression and the Bayesian LASSO are two of the most approachable priors, but all sparsity inducing priors utilize the same logic as the spike-and-slab: give $\beta$ a prior with *most* of its mass near 0, to shrink irrelevant coefficients, and the rest of its mass "far" from 0. Identifying "far" depends more or less on the specific data and model in question.

The horseshoe prior, proposed by **Carvalho et al, 2010** is a popular choice in Bayesian literature, in part because it is similar to the "gold standard" spike-and-slab. Following the notation of **Vehtari & Piironen**, the horseshoe prior can be expressed

$$
\begin{aligned}
\beta_k \text{ | } \lambda_k, \tau & \sim \text{Normal}(0, \tau^2\lambda_k^2) \\
\lambda_k & \sim \text{Half-Cauchy}(0, 1)
\end{aligned}
$$

for $k = 1, 2, ..., K$. The horseshoe prior is so-named because for fixed values $\tau = \lambda_k = 1$, the prior resemebles a Beta(1/2, 1/2) distribution, or a horseshoe.

The horseshoe is often favored because it is a global-local shrinkage prior. This means that $\tau$ shrinks *all* parameters towards 0 while the local parameter $\lambda_k$ and the heavy Cauchy tails allow larger coefficients to remain unshrunk. While this is precisely the goal of a sparsity inducing prior, the horseshoe prior can fail to regularize large coefficients *at all*, which can be a problem when parameters are weakly identified by data. With no regularization applied to the largest coefficients, there is a risk of overfitting. Additionally, there is no consensus in the literature for assigning priors to $tau$. Piironen and Vehtari (2017) introduced the *regularized horseshoe* to address both of these shortcomings of the horseshoe.

### Regularized Horseshoe - needs graphics

 The regularized horseshoe builds upon the horseshoe prior such that

$$
\begin{aligned}
\beta_k \text{ | } \lambda_j, \tau, c & \sim \text{Normal}(0, \tau^2 \tilde{\lambda_k^2}) \\
\tilde{\lambda_k^2} & = \frac{c^2\lambda_k^2}{c^2 + \tau^2\lambda_k^2} \\
\lambda_k & \sim \text{Half-Cauchy}(0, 1) \\
c & \sim ~ \text{Inverse-Gamma}(a, b) \\
\tau & \sim 
\end{aligned}
$$

where $c$ > 0 helps to regularize $\beta_k$ far from zero. When $c$ approaches infinity, the regularized horseshoe becomes the standard horseshoe. Additionally, $c$ in the regularized horseshoe corresponds to the slab width in the spike-and-slab prior.

As with the spike-and-slab, the selection of priors is a challenge with the regularized horseshoe. Piironen and Vehtari (2017) provide a set of recommendations for setting priors which we follow in this study. In practice, we tune the priors for $\lambda_k$, $c$, and $\tau$ using simulated data with features similar to our observed data (ie, parameters with coefficients of similar magnitude).

## Adding Spatial Structure

When modeling data with a spatial component, we generally expect adjacent areas to be more similar than areas which are far apart. While this is a straightforward assumption, it can improve model fits in several ways. First, it can encode prior information about response $Y$ not otherwise represented by covariates $x$ (ie, that nearby observations are more similar than far-flung ones). Second, it can provide geographic smoothing when observations are sparse or noisy. This is often the case in small areas or when observing events which can only occur at specific locations, such as air quality measured by sensors. In the case of the Metro Transit ridership data, this smoothing can help avoid overfitting areas such as the outliers in downtown Minneapolis and better represent block groups with no existing bus stops.

In the case of Metro Transit's ridership data, we expect encoding spatial structure to improve the model fit in several ways. First, we know that assigning transit ridership to census geography is inherently flawed. This is because census geographies, including block groups, are generally bounded by major roads (in addition to natural features such as lakes and political boundaries such as city limits). Naturally, bus stops tend to exist along those same major roadways.

This bounding convention means that boardings are often assigned to different block groups depending on which direction the passenger is travelling. For example, consider the A-Line running along Snelling Avenue. Boarding at Snelling and Grand *northbound* puts you in block group ____ while boarding at the same intersection but *southbound* puts you in block group ____. Integrating spatial structure will help to smooth ridership across block group bounds. **will add a map here**

Additionally, we know intuitively that transit riders don't necessarily board in the block groups they live in. Particularly in dense areas, it's very likely for riders to board in a block group other than the block group they reside in. For example, I live in block group ____ but walk about 10 minutes to board the A-Line southbound at Snelling and Randolph. Commuting patterns such as these can assign ridership to different block groups than we might expect. Geographic smoothing will allow neighboring block groups to share information, better representing these ridership patterns.

### Conditional Autoregressive Priors

Conditional Autoregressive (CAR) priors are one of the most widespread Bfayesian methods for modeling spatial autocorrelation. CAR models were introduced in Besag 1974 and remain perhaps the primary method for Bayesian areal data modeling. Note that CAR models are recommended for use as *priors* rather than as an observation model.

Areal data corresponds to finitely many discrete areal units, such as counties or census tracts. CAR priors are often used in the context of small-area count data. This sort of response variable tends to be noisy, particularly when the counted event is rare, the population of areal unit $i$ is small, or the physical boundaries of areal units present challenges. CAR priors were designed in part to smooth such noisy counts. As such, they are often applied in epidemiological disease risk modeling in which disease occurences may be rare.

Generally, CAR models rely on a binary neighborhood structure. Areal units $i$ and $j$ are considered neighbors if they share a boundary. For strictly rectangular lattice data, we often choose either a "Queens" or "Rooks" neighborood structure. **insert graphics** For irregularly shaped areal data, such as our 1,495 census block groups, we generally default to a queen neighborhood structure wherein areal units sharing *any* points of contact are deemed neighbors.

For $n$ regions, the neighorhood relationships are encoded in an $n \times n$ neighborhood matrix $W$. Matrix $W$ is defined such that $w_{ij} = 1$ if if regions $i$ and $j$ are neighbors (denoted $i \sim j$) and 0 otherwise. This matrix is symmetric. Note that regions are not considered neighbors of themselves, ie the diagonal entries $w_{ii} = 0$.

The spatial interactions are modeled by a random variable $\phi$. Here, $\phi$ is a vector $\phi = \phi_1, \phi_2, ..., \phi_n$. The distribution of each $\phi$ is determined by the sum of its neighbors' values such that

$$
\phi_i \text{ | } \phi_j, j \neq i \sim \text{Normal}(\sum_{i=1}^{N}w_{ij}\phi_j, \sigma^2) 
$$

The conditional distribution of each $\phi_i$ is helpful in building intuition about the functionality of the CAR prior. Basing the value of each $\phi_i$ off of its neighbors is how the CAR prior performs spatial smoothing.

Besag (1974) proved that the joint distribution of $\phi$ is a multivariate normal centered at 0 where its variance is given by a symmetric positive definite precision matrix $Q$. Matrix $Q$ is simply the inverse of the covariance matrix $\sum$. This finding simplifies the above to

$$
\phi \sim \text{Normal}(0, Q^{-1})
$$

Precision matrix $Q$ can be defined for multivariate normal $\phi$ using two other $n \times n$ matrices: the neighborhood matrix $W$ and the diagonal matrix $D$ in which $d_{ii}$ indicates the number of neighbors region $i$ has. All off-diagonal entries are 0. Using these matrices, $Q$ is defined

$$
Q = D(I - \alpha W)
$$

where $I$ is the identity matrix and $\alpha \in (0, 1)$ determines the amount of spatial autocorrelation present in $Y$. Parameter $\alpha$ is a key component of CAR models. At $\alpha = 0$, the model assumes spatial independence and at $\alpha = 1$, perfect spatial autocorrelation.

Following Morris et al, the log probability density of $\phi$ is proportional to

$$
\frac{M}{2}\text{log(det}(Q)) - \frac{1}{2}\phi^TQ\phi
$$

Calculating this value is computationally expensive. For models with 1,000 areal units, for example, calculating the determinant requires 1 billion operations (Morris et al, 2019). Given that the Metro Transit ridership data corresponds to 1,495 areal units, a strict CAR prior may be too computationally inefficient for our models.

### Intrinsic Conditional Autoregressive

The Intrinsic Conditional Autoregressive (ICAR) prior is a slight simplifation of a CAR prior. ICAR priors set $\alpha = 1$ which simplifies the definition of Q

$$
Q = D-A
$$

thereby setting $det(Q) = 0$. As a result, the first term in **equation ___** can be simplified to

$$
-\frac{1}{2}\phi^TQ\phi
$$

This reduces the computational expense of computing CAR models significantly. Notably, the ICAR prior is improper but yields proper posteriors (Morris et al, 2019).

The ICAR prior specifies each $\phi_i$ to be normally distributed with its mean equal to the mean of its neighbor's values. This is how the CAR/ICAR model "borrows strength" from geographic neighbors to smooth noisy estimates. Intuitively, the variance of each $\phi_i$ decreases as its number of neighbors $d_i$ increases. The conditional distribution for each $\phi_i$ can be written

$$
\text{p}(\phi_i \text{ | } \phi_{i \sim j}) = \text{Normal}(\frac{\sum_{i \sim j}\phi_i}{d_i}, \frac{\sigma_i^2}{d_i})
$$

Here, the variance $\sigma$ is unknown. The conditional specification of each $\phi_i$ is helpful in building intuition about the assumptions the ICAR component makes and how it performs its "smoothing". The ICAR prior matches our intuition about spatial autocorrelation: areal unit $i$ is similar to the areal units surrounding it, and the more areal units surround it, the more confident we feel in that similarity. The joint distribution can be rewritten

$$
\text{p}(\phi) \propto \text{exp}(-\frac{1}{2}\sum_{i \sim j}(\phi_i - \phi_j)^2)
$$

This pairwise difference specification is more computationally efficient to fit in Stan (Morris et al, 2019).

### Besag-York-Mollie (BYM)

The specific ICAR model used in this analysis is the Besag-York-Mollie (BYM) Poisson model. The BYM is a "classical" spatial Bayesian method (Riebler et al, 2016). It is generally used in disease mapping studies to model rare events (ie, disease occurrence) in small areal units. The BYM capitalizes on the ICAR prior's ability to smooth noisy estimates and share information across geographic units.

The latent function in the BYM model contains heterogeneous random effects as well as a spatial ICAR term in order to account for both spatial and non-spatial heterogeneity. Note that in Section ___, we introduced a random effect term $\theta$. Therefore, the BYM model can be though of as decomposing $\theta$ into spatial and non-spatial components.

Specifically, the BYM model replaces $\eta_i$ from equation ___ with

$$
\eta_i = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + \theta + \phi
$$

where $\phi$ is the addition: an ICAR component. Decomposing overdispersion as such is helpful in furthering our understanding of response $Y$. When we use a simple random effect $\theta$ alone, we can accomplish a good model fit. However, $\theta$ provides no additional information about $Y$. The decomposition of $\theta$ in the BYM allows us to better quantify how much variance is white noise (ordinary random effects) versus some unmeasured confounding correlated across space.

The BYM model as written above is appealing in its simplicity. However, the lack of informative hyperpriors specified for $\theta$ and $\phi$ can make the model incredibly challenging to fit. In theory, extra-Poisson variance could be explained 100% by $\theta$, 100% by $\phi$, or anywhere in between. In *practice*, we likely have limited information about where that ratio falls. Riebler et al (2016) further explain the sampling issues faced by the BYM model faced with no hyperpriors. In short, the sampler is forced to explore all possible combinations of $\theta$ and $\phi$, no matter how unlikely a particular combination my be. In the context of fitting models with Stan, this is liekly to cause uneccesarily long computation times and perhaps to yield biased posterior estimates.

There are some suggestions in the literature for setting hyperpriors for $\theta$ and $\phi$ (Besag and Mollie, 1991; Clayton and Montomoli, 1995). The existing methods, however, tend to rely on the specific data at hand which makes the selection of hyperparameters unnecessarily time consuming.

### BYM2

The BYM2 model proposed by Simpson et al and further described in Riebler et al (2016) aims to solve these sampling problems. The BYM2 takes a more "fully Bayesian" approach to hyperprior/hyperparameter selection. The primary difference between the BYM and BYM2 models is the addition of a mixing parameter, $\rho$. The BYM2 model rewrites the BYM as

$$
\eta_i = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + ((\sqrt{\frac{\rho}{s}})\theta^{*} + ((\sqrt{1-\rho})\phi^{*})\sigma
$$

where $\rho \in (0, 1)$ determines how much variance/overdispersion is caused by spatial versus non-spatial error terms. Like the popular Leroux CAR prior, proposed by Leroux et al (2009), the BYM2 scales both $\theta$ and $\phi$ by $\sigma$, the standard deviation of the combined error terms (Morris et al, 2019). In this parameterization, $s$ is a scaling factor such that Var($\theta_i$) $\approx$ Var($\phi_i$) $\approx$ 1. The equal unit variance is necessary for $\sigma$ to truly be the standard deviation of the error terms.

Setting priors is somewhat more straightforward for the BYM2. The ICAR component, $\phi^{*}$ remains unchanged. Riebler and Morris recommend the prior $\theta \sim$ Normal($0, n)$ where $n$ is the number of connected graphs in the neighborhood graph. In many cases, such as modeling data for all counties in a state, $n=1$. When $n>1$, the variance is different in each subgraph which affects $\sigma$ and $s$. It is possible to fit the model in this case, although computation time is much longer.

A relatively vague Normal(0, 1) prior is appropriate for $\sigma$. Riebler et al (2016) also propose either a Beta(1/2, 1/2) or more specialized, complex prior for mixing parameter $\rho$. For this study, we use the simpler Beta(1/2, 1/2) prior.

## Stan

All of the models discussed above can be fit using the programming language Stan. Stan is a probabilistic language which uses Hamiltonian Monte Carlo (HMC) and a specialized No U-Turn Sampler (NUTS) to compute joint log probability densities. The specifics of Stan are beyond the scope of this paper. However, a few of its particular diagnostics are helpful in understanding the differences between the model fits below.

* $\hat{R}$ or `Rhat` indicates how well the Markov chains have mixed by comparing between- and within-chain estimates. At convergence, $\hat{R} = 1$. When $\hat{R} > 1$, the Markov chains have not mixed well and the model likely needs to be run for more iterations.  
* `n_eff` is the estimated Effective Sample Size (ESS). ESS is the number of independent samples our sample is equivalent to. In other words, if `n_eff` = 1,000, we expect to gain the same amount of information from our sample as we would from 1,000 independent samples. In a "perfect" scenario, `n_eff` should equal the actual number of samples. When `n_eff` is significantly below the actual number of samples, there may be a problem with the model and posterior estimates may be biased. 
* **Divergent transitions** occur when the simulated Hamiltonian departs from the actual Hamiltonian trajectory. In general, if stan gives any divergent transition warnings, we should consider the posterior estimates unreliable.  
* `max_treedepth` is a parameter we set when fitting a model in Stan. Treedepth refers to the number of steps the sampler has to take at each iteration. The default is 10. For extremely "curved" or complex posteriors, the treedepth may be much higher. When `max_treedepth` is exceeded, the sampler quits prematurely. Therefore, if we have *many* iterations exceeding `max_treedepth`, posterior estimates may be unreliable. Additionally, setting high treedepth causes the sampler to perform much more slowly. 
* `adapt_delta` is also a sampling parameter; it corresponds to the Mean Metropolis Acceptance Rate (see ____). The default is 0.8. Setting a higher `adapt_delta` slows computation time but *can* eliminate "false alarm" divergent transitions.

# Results

In this study, we fit several increasingly complex models building upon the baseline Poisson regression, Model 1. Model 2 incorporates a simple set of random effects to account for overdispersion. Model 3 incorporates regularized horseshoe priors for variable selection. Model 4 includes the spatial ICAR priors but no horseshoe priors, and Model 5 includes both regularized horseshoe priors and ICAR priors.

All five models were run for at least 10,000 iterations and "pass" the basic diagnostic checks: $\hat{R} = 1$, sufficiently high `n_eff`, and no concerning `max_treedepth` or divergence warnings.

Model 1 uses all 18 covariates with simple Normal(0, 1) priors and no overdispersion parameters. The Stan program used to fit Model 1, as well as the other four models, can be found in the appendix. Model 1 fits quickly and efficiently. The Stan output for $\beta_0$ and $\beta$ is below.

\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrr}
  \toprule
Parameter & Rhat & n\_eff & mean & sd & 2.5\% & 97.5\% \\ 
  \midrule
beta\_0 & 1.000 & 19365 & -1.376 & 0.005 & -1.385 & -1.366 \\ 
  beta[1] & 1.000 & 18085 & 0.006 & 0.007 & -0.007 & 0.019 \\ 
  beta[2] & 1.000 & 19330 & 0.063 & 0.005 & 0.054 & 0.072 \\ 
  beta[3] & 1.000 & 18752 & -0.027 & 0.004 & -0.035 & -0.019 \\ 
  beta[4] & 1.000 & 18026 & -0.029 & 0.004 & -0.038 & -0.021 \\ 
  beta[5] & 1.000 & 18380 & -0.013 & 0.005 & -0.022 & -0.004 \\ 
  beta[6] & 1.000 & 15809 & 0.241 & 0.005 & 0.231 & 0.251 \\ 
  beta[7] & 1.000 & 18930 & 0.014 & 0.003 & 0.007 & 0.020 \\ 
  beta[8] & 1.000 & 17331 & -0.003 & 0.005 & -0.012 & 0.006 \\ 
  beta[9] & 1.000 & 19360 & 0.074 & 0.004 & 0.067 & 0.082 \\ 
  beta[10] & 1.000 & 21329 & 0.066 & 0.001 & 0.065 & 0.068 \\ 
  beta[11] & 1.000 & 20224 & -0.121 & 0.004 & -0.128 & -0.114 \\ 
  beta[12] & 1.000 & 22539 & -0.069 & 0.004 & -0.076 & -0.061 \\ 
  beta[13] & 1.000 & 18372 & -0.011 & 0.005 & -0.021 & -0.001 \\ 
  beta[14] & 1.000 & 15602 & -0.026 & 0.005 & -0.035 & -0.017 \\ 
  beta[15] & 1.000 & 15480 & 0.113 & 0.005 & 0.104 & 0.122 \\ 
  beta[16] & 1.000 & 19904 & -0.085 & 0.004 & -0.093 & -0.078 \\ 
  beta[17] & 1.000 & 21169 & 0.047 & 0.003 & 0.042 & 0.053 \\ 
  beta[18] & 1.000 & 22032 & -0.025 & 0.003 & -0.031 & -0.020 \\ 
   \bottomrule
\end{tabular}
\caption{Stan output for Model 1} 
\end{table}

Model 1 produces coefficient estimates with tight confidence intervals; few of the $\beta_k$ 95% intervals cross zero. Based on initial diagnostics and precise parameter estimates, Model 1 appears to be a "good" fit. In this way, Model 1 illustrates the importance of diagnostics beyond the default Stan output.

\begin{figure}[H]
  \centering
  \includegraphics[width = 5in]{ppcheck_mod1.png}
  \caption{Posterior predictive check for Model 1.}
\end{figure}

Model 1 clearly underestimates the number of block groups with low ridership. This is due in large part to the extra-Poisson variance in Metro Transit's ridership data.

Model 2 builds upon Model 1 by incorporating and overdispersion parameter. Model 2 is *significantly* more computationally expensive to fit than Model 1. This is due to relatively large autocorrelation between each iteration, visible in the traceplots below (Figure ___) as well as in the significantly lower `n_eff` for Model 2. The parameter estimates for $\beta_0$ and $\beta$ for Model 2 are below.

Because we gave $\theta$ and $\sigma$ vague priors, we can see a *much* better fit in the posterior predictive checks for Model 2. One *could* stop analysis here. However, the addition of $\theta \sigma$ only improves the *fit* of the model, not our understanding of the data. Therefore, for this study, we will decompose $\theta \sigma$ into spatial and non-spatial random effects.

First, however, we implement the horseshoe priors to determine if any of our 18 covariates $x_k$ should be removed from the model.

<!-- Notably, Model 3 experienced one divergent transition after warmup. However, the model was fitted several times with no divergent transitions. A "false alarm" divergent transition can occur when the cutoff for determining a divergent transition is too low; this can be true in the case of particularly complex posteriors. Because only 1 out of 20,000 iterations produced a divergent transition, we can proceed with caution using Model 3.  -->

# Discussion

- bymhrs is probably not "worth it", needs to be reparameterized/coded by someone better than me
- incorporating overdispersion is the most important improvement in terms of model fit

# Appendix

## Model 1
```{r}
writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/poisson.stan"))
```

## Model 2

```{r}
writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/poisson_theta.stan"))
```

## Model 3
```{r}
writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/poisson_horseshoe.stan"))
```

## Model 4

```{r}
writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/bym2.stan"))
```

## Model 5
```{r}
writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/bym_horseshoe.stan"))

```

### 2020-02-10 07_figures.R

@@ -16,5 +16,5 @@ sfdat <- left_join(bgs, mod_dat, on = 'GEOID')
 ggplot(sfdat) + 
   geom_sf(aes(fill = log(daily_boards)), color = "grey", lwd = .1) +
   theme_minimal() +
-  scale_fill_viridis(na.value = "transparent", option = "plasma") +
-  labs(fill = "Daily Boardings")
+  scale_fill_viridis(na.value = "transparent") +
+  labs(fill = "Boardings", title = "Average Weekday Boardings, 2017", subtitle = "Log scale, enumerated by block group")

### 2020-02-10 functions/ppcheck.R

@@ -4,15 +4,12 @@ library(loo)
 library(data.table)
 library(ggplot2)
 
-## for testing
-
-post <- readRDS("~/Documents/honors/honors-project/test_post.RDS")
+ 
+post <- poisson_fit
 n_obs = 1495
-n_iter = 2000
+n_iter = 20000
+y <- na.omit(p_dat_scaled)$daily_boards
 
-# function will take stanfit, observed y, number observations, number iterations, pars for trace, 
-
-ppcheck <- function(post, y, n_obs, n_iter){
 
 ## get posterior draws
 
@@ -29,8 +26,10 @@ ppcheck <- function(post, y, n_obs, n_iter){
   setDT(as.data.frame(yrep))
   
   ppoverlay <- pp_check(y, yrep[sample(100), ], ppc_dens_overlay) + 
-    theme_minimal() +
-    labs(title = "Poisson Actual versus fitted") 
+    theme_minimal() 
+  
+  ppoverlay + xlim(0, 200) + ylim(0, 0.025) + 
+    labs(title = "Observed versus simulated ridership for Model 1")
   
   ## get traceplot
   
@@ -67,5 +66,5 @@ ppcheck <- function(post, y, n_obs, n_iter){
   ## map
   
   
-}
+
 

### 2020-02-10 scripts/aws/fitting.R

@@ -49,7 +49,7 @@ poisson <- "~/Documents/honors/honors-project/stan/final-stan/poisson.stan"
 ## add horseshoe priors -------------------------
 
 poisson_hrs <- "~/Documents/honors/honors-project/stan/final-stan/poisson_horseshoe.stan"
-#poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, iter = 10000, verbose = T,
+poisson_hrs_fit <- stan(poisson_hrs, data = horseshoe_dat, iter = 10000, verbose = T,
                         control = list(adapt_delta = 0.99, max_treedepth = 15), seed = 1997)
 #saveRDS(poisson_hrs_fit, '~/Documents/honors/honors-project/final-fits/poisson_hrs_fit.RDS')
 
@@ -70,7 +70,7 @@ standat2 <- list(y = y, E = E, x = x2, K = K2, N = N)
 poisson_theta <- "~/Documents/honors/honors-project/stan/final-stan/poisson_theta.stan"
 
 
-#poisson_theta_fit <- stan(poisson_theta, data = standat2, iter = 20000, verbose = T, seed = 1997)
+poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 20000, verbose = T, seed = 1997)
 #saveRDS(poisson_theta_fit, "~/Documents/honors/honors-project/final-fits/poisson_theta.RDS")
 
 ## and the bym ! --------------------------------
@@ -113,5 +113,7 @@ bymhrs_dat <- list(y = y, E = E, x = x, K = K,
                    N = N2, node1 = node1, node2 = node2, N_edges = N_edges, scaling_factor = scaling_factor, 
                    scale_icept = 2, scale_global = .002, nu_global = 1, nu_local = 1, slab_scale = 1, slab_df = 24)
 
-bymhrs <- stan(bymhrs, data = bymhrs_dat, iter = 10000, verbose = T, seed = 1997, 
+bymrs <- stan(bymhrs, data = bymhrs_dat, iter = 50, verbose = T, seed = 1997, 
                control = list(max_treedepth = 15, adapt_delta = 0.99))
+
+launch_shinystan(bymrs)

### 2020-02-10 chapters/article-draft.Rmd
updates to article draft

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
@@
 
+## Stan
+
+All of the models discussed above can be fit using the programming language Stan.
+Stan is a probabilistic language which uses Hamiltonian Monte Carlo (HMC) and a specialized No U-Turn Sampler (NUTS) to compute joint log probability densities.
+The specifics of Stan are beyond the scope of this paper.
+However, a few of its particular diagnostics are helpful in understanding the differences between the model fits below.
+
+* $\hat{R}$ or `Rhat` indicates how well the Markov chains have mixed by comparing between- and within-chain estimates.
+At convergence, $\hat{R} = 1$.
+When $\hat{R} > 1$, the Markov chains have not mixed well and the model likely needs to be run for more iterations.
+* `n_eff` is the estimated Effective Sample Size (ESS).
+ESS is the number of independent samples our sample is equivalent to.
+In other words, if `n_eff` = 1,000, we expect to gain the same amount of information from our sample as we would from 1,000 independent samples.
+In a "perfect" scenario, `n_eff` should equal the actual number of samples.
+When `n_eff` is significantly below the actual number of samples, there may be a problem with the model and posterior estimates may be biased.
+* **Divergent transitions** occur when the simulated Hamiltonian departs from the actual Hamiltonian trajectory.
+In general, if stan gives any divergent transition warnings, we should consider the posterior estimates unreliable.
+* `max_treedepth` is a parameter we set when fitting a model in Stan.
+Treedepth refers to the number of steps the sampler has to take at each iteration.
+The default is 10.
+For extremely "curved" or complex posteriors, the treedepth may be much higher.
+When `max_treedepth` is exceeded, the sampler quits prematurely.
+Therefore, if we have *many* iterations exceeding `max_treedepth`, posterior estimates may be unreliable.
+Additionally, setting high treedepth causes the sampler to perform much more slowly.
+* `adapt_delta` is also a sampling parameter; it corresponds to the Mean Metropolis Acceptance Rate (see ____).
+The default is 0.8.
+Setting a higher `adapt_delta` slows computation time but *can* eliminate "false alarm" divergent transitions.
+
 ## Poisson Regression
@@
 
+### Model 1
+
+We will call this baseline Poisson regression Model 1.
+Model 1 uses all 18 covariates with simple Normal(0, 1) priors and no overdispersion parameters.
+The Stan program used to fit Model 1, as well as the other four models, can be found in the appendix.
+Model 1 fits quickly and efficiently.
+The Stan output for $\beta_0$ and $\beta$ is below.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrrrr}
+  \toprule
+Parameter & Rhat & n\_eff & mean & sd & 2.5\% & 97.5\% \\ 
+  \midrule
+beta\_0 & 1.000 & 19365 & -1.376 & 0.005 & -1.385 & -1.366 \\ 
+  beta[1] & 1.000 & 18085 & 0.006 & 0.007 & -0.007 & 0.019 \\ 
+  beta[2] & 1.000 & 19330 & 0.063 & 0.005 & 0.054 & 0.072 \\ 
+  beta[3] & 1.000 & 18752 & -0.027 & 0.004 & -0.035 & -0.019 \\ 
+  beta[4] & 1.000 & 18026 & -0.029 & 0.004 & -0.038 & -0.021 \\ 
+  beta[5] & 1.000 & 18380 & -0.013 & 0.005 & -0.022 & -0.004 \\ 
+  beta[6] & 1.000 & 15809 & 0.241 & 0.005 & 0.231 & 0.251 \\ 
+  beta[7] & 1.000 & 18930 & 0.014 & 0.003 & 0.007 & 0.020 \\ 
+  beta[8] & 1.000 & 17331 & -0.003 & 0.005 & -0.012 & 0.006 \\ 
+  beta[9] & 1.000 & 19360 & 0.074 & 0.004 & 0.067 & 0.082 \\ 
+  beta[10] & 1.000 & 21329 & 0.066 & 0.001 & 0.065 & 0.068 \\ 
+  beta[11] & 1.000 & 20224 & -0.121 & 0.004 & -0.128 & -0.114 \\ 
+  beta[12] & 1.000 & 22539 & -0.069 & 0.004 & -0.076 & -0.061 \\ 
+  beta[13] & 1.000 & 18372 & -0.011 & 0.005 & -0.021 & -0.001 \\ 
+  beta[14] & 1.000 & 15602 & -0.026 & 0.005 & -0.035 & -0.017 \\ 
+  beta[15] & 1.000 & 15480 & 0.113 & 0.005 & 0.104 & 0.122 \\ 
+  beta[16] & 1.000 & 19904 & -0.085 & 0.004 & -0.093 & -0.078 \\ 
+  beta[17] & 1.000 & 21169 & 0.047 & 0.003 & 0.042 & 0.053 \\ 
+  beta[18] & 1.000 & 22032 & -0.025 & 0.003 & -0.031 & -0.020 \\ 
+   \bottomrule
+\end{tabular}
+\caption{Stan output for Model 1} 
+\end{table}
+
+Model 1 produces coefficient estimates with tight confidence intervals; few of the $\beta_k$ 95% intervals cross zero.
+Based on initial diagnostics and precise parameter estimates, Model 1 appears to be a "good" fit.
+In this way, Model 1 illustrates the importance of diagnostics beyond the default Stan output.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width = 6in]{ppcheck_mod1.png}
+  \caption{Posterior predictive check for Model 1.}
+\end{figure}
+
+Model 1 clearly underestimates the number of block groups with low ridership. This is due in large part to extra-Poisson variance present in Metro Transit's ridership data.
+
 ## Overdispersed Poisson Regression
@@
 With sufficiently vague priors, $\theta$ and $\sigma$ can account for most or all of the overdispersion not accounted for by $\beta_0 + \sum_{k=1}^{K}x_k^T\beta_k$.
+
+### Model 2
+
+Model 2 builds upon Model 1 by incorporating the set of random effects described above.
+Model 2 is *significantly* more computationally expensive to fit than Model 1.
+This is due to relatively large autocorrelation between each iteration, visible in the traceplots below (Figure ___).
+This autocorrelation made it necessary to run Model 2 on four chains for 20,000 iterations each.
+Despite this, `n_eff` is significantly lower for Model 2 than Model 1.
+Reparameterization could likely reduce autocorrelation between iterations.
+
+The parameter estimates for $\beta_0$ and $\beta$ for Model 2 are below.
+
+Because we gave $\theta$ and $\sigma$ vague priors, we can see a *much* better fit in the posterior predictive checks for Model 2.
+One *could* stop analysis here.
+However, the addition of $\theta \sigma$ only improves the *fit* of the model, not our understanding of the data.
+Therefore, for this study, we will decompose $\theta \sigma$ into spatial and non-spatial random effects.
+First, however, we will implement a Bayesian variable selection process to determine which, if any, of our covariates $x_k$ should be removed from the analysis.
 
@@
 In practice, we tune the priors for $\lambda_k$, $c$, and $\tau$ using simulated data with features similar to our observed data (ie, parameters with coefficients of similar magnitude).
+
+### Model 3
+
+For this study, we implement the regularize horseshoe using code modified from Piironen & Vehtari.
+
+Notably, Model 3 experienced one divergent transition after warmup.
+However, the model was fitted several times with no divergent transitions.
+A "false alarm" divergent transition can occur when the cutoff for determining a divergent transition is too low; this can be true in the case of particularly complex posteriors.
+Because only 1 out of 20,000 iterations produced a divergent transition, we can proceed with caution using Model 3.
 
@@
 
-## Stan
+### Model 4
 
-All of the models discussed above can be fit using the programming language Stan.
-Stan is a probabilistic language which uses Hamiltonian Monte Carlo (HMC) and a specialized No U-Turn Sampler (NUTS) to compute joint log probability densities.
-The specifics of Stan are beyond the scope of this paper.
-However, a few of its particular diagnostics are helpful in understanding the differences between the model fits below.
+Model 4 uses all 18 covariates of interest, no horseshoe priors, and decomposes $\theta\sigma$ into a set of random effects plus a spatial ICAR component.
+This model converges much more efficiently than Model 2 or Model 3 due to careful reparameterization of the BYM by Riebler et al (2016).
+Future would could potentially improve the speed of fitting Models 2 and 3.
 
-* $\hat{R}$ or `Rhat` indicates how well the Markov chains have mixed by comparing between- and within-chain estimates.
-At convergence, $\hat{R} = 1$.
-When $\hat{R} > 1$, the Markov chains have not mixed well and the model likely needs to be run for more iterations.
-* `n_eff` is the estimated Effective Sample Size (ESS).
-ESS is the number of independent samples our sample is equivalent to.
-In other words, if `n_eff` = 1,000, we expect to gain the same amount of information from our sample as we would from 1,000 independent samples.
-In a "perfect" scenario, `n_eff` should equal the actual number of samples.
-When `n_eff` is significantly below the actual number of samples, there may be a problem with the model and posterior estimates may be biased.
-* **Divergent transitions** occur when the simulated Hamiltonian departs from the actual Hamiltonian trajectory.
-In general, if stan gives any divergent transition warnings, we should consider the posterior estimates unreliable.
-* `max_treedepth` is a parameter we set when fitting a model in Stan.
-Treedepth refers to the number of steps the sampler has to take at each iteration.
-The default is 10.
-For extremely "curved" or complex posteriors, the treedepth may be much higher.
-When `max_treedepth` is exceeded, the sampler quits prematurely.
-Therefore, if we have *many* iterations exceeding `max_treedepth`, posterior estimates may be unreliable.
-Additionally, setting high treedepth causes the sampler to perform much more slowly.
-* `adapt_delta` is also a sampling parameter; it corresponds to the Mean Metropolis Acceptance Rate (see ____).
-The default is 0.8.
-Setting a higher `adapt_delta` slows computation time but *can* eliminate "false alarm" divergent transitions.
+Model 4 was run on four chains for 10,000 iterations each. The parameter estimates for Model 4 are below. Note that `n_eff` is lower than for Model 1 but significantly higher than Model 2 or 3.
 
-# Results
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrrrr}
+  \toprule
+Parameter & Rhat & n\_eff & mean & sd & 2.5\% & 97.5\% \\ 
+  \midrule
+beta0 & 1.001 & 4770 & -1.880 & 0.028 & -1.937 & -1.825 \\ 
+  sigma & 1.003 &  520 & 2.698 & 0.372 & 1.993 & 3.451 \\ 
+  rho & 1.002 &  514 & 0.888 & 0.038 & 0.793 & 0.941 \\ 
+  theta[5] & 1.000 & 17797 & -0.179 & 0.777 & -1.688 & 1.336 \\ 
+  beta[1] & 1.000 & 4322 & 0.034 & 0.047 & -0.056 & 0.127 \\ 
+  beta[2] & 1.001 & 3801 & 0.061 & 0.035 & -0.009 & 0.131 \\ 
+  beta[3] & 1.000 & 4203 & -0.003 & 0.039 & -0.080 & 0.073 \\ 
+  beta[4] & 1.001 & 4330 & -0.049 & 0.044 & -0.134 & 0.037 \\ 
+  beta[5] & 1.000 & 4706 & 0.209 & 0.042 & 0.127 & 0.292 \\ 
+  beta[6] & 1.000 & 4919 & 0.065 & 0.038 & -0.009 & 0.138 \\ 
+  beta[7] & 1.002 & 3929 & 0.029 & 0.038 & -0.046 & 0.104 \\ 
+  beta[8] & 1.000 & 6278 & 0.145 & 0.032 & 0.083 & 0.207 \\ 
+  beta[9] & 1.000 & 4607 & -0.088 & 0.031 & -0.150 & -0.026 \\ 
+  beta[10] & 1.000 & 6145 & 0.014 & 0.029 & -0.043 & 0.070 \\ 
+  beta[11] & 1.000 & 6978 & -0.022 & 0.031 & -0.083 & 0.040 \\ 
+  beta[12] & 1.000 & 6125 & -0.068 & 0.034 & -0.136 & -0.002 \\ 
+  beta[13] & 1.000 & 5875 & 0.094 & 0.033 & 0.029 & 0.158 \\ 
+  beta[14] & 1.000 & 4743 & -0.089 & 0.035 & -0.157 & -0.019 \\ 
+  beta[15] & 1.001 & 2720 & 0.076 & 0.039 & 0.001 & 0.151 \\ 
+  beta[16] & 1.000 & 6053 & -0.029 & 0.031 & -0.091 & 0.031 \\ 
+   \bottomrule
+\end{tabular}
+\caption{Stan output for Model 4} 
+\end{table}
+
+Like Models 2 and 3, the BYM2 model performs well in the posterior predictive check (Figure __).
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width = 6in]{ppcheckmod4.png}
+  \caption{Posterior predictive check for Model 1.}
+\end{figure}
+
+### Model 5
+
+Model 5 builds upon Model 4 by incorporating regularized horsehoe priors in addition to spatial ICAR priors.
+This model has high autocorrelation between iterations and took *significantly* longer than any of the other models to fit in Stan (upwards of 12 hours).
+That said, the `n_eff` and $\hat{R}$ diagnostics indicate convergence and the parameter estimates (Table __) match our expectations in terms of sign and magnitude based on the previous 4 models.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrrrr}
+  \toprule
+Parameter & Rhat & n\_eff & mean & sd & 2.5\% & 97.5\% \\ 
+  \midrule
+beta0 & 1.0 & 5837 & -1.9 & 0.0 & -1.9 & -1.8 \\ 
+  sigma & 1.0 & 721 & 2.8 & 0.4 & 2.1 & 3.6 \\ 
+  rho & 1.0 & 713 & 0.9 & 0.0 & 0.8 & 0.9 \\ 
+  beta[1] & 1.0 & 3637 & -0.1 & 0.1 & -0.2 & -0.0 \\ 
+  beta[2] & 1.0 & 8170 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
+  beta[3] & 1.0 & 3987 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
+  beta[4] & 1.0 & 8531 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
+  beta[5] & 1.0 & 7619 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
+  beta[6] & 1.0 & 2999 & 0.2 & 0.1 & 0.1 & 0.3 \\ 
+  beta[7] & 1.0 & 3726 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
+  beta[8] & 1.0 & 7956 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
+  beta[9] & 1.0 & 5622 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
+  beta[10] & 1.0 & 7530 & 0.1 & 0.0 & 0.1 & 0.2 \\ 
+  beta[11] & 1.0 & 4558 & -0.1 & 0.0 & -0.1 & 0.0 \\ 
+  beta[12] & 1.0 & 11150 & 0.0 & 0.0 & -0.0 & 0.0 \\ 
+  beta[13] & 1.0 & 8289 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
+  beta[14] & 1.0 & 5469 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
+  beta[15] & 1.0 & 4822 & 0.1 & 0.0 & -0.0 & 0.1 \\ 
+  beta[16] & 1.0 & 3863 & -0.1 & 0.0 & -0.1 & 0.0 \\ 
+  beta[17] & 1.0 & 2094 & 0.1 & 0.0 & -0.0 & 0.1 \\ 
+  beta[18] & 1.0 & 6949 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
+   \bottomrule
+\end{tabular}
+\caption{Stan output for Model 5} 
+\end{table}
+
+# Discussion
 
@@
 
-Model 1 uses all 18 covariates with simple Normal(0, 1) priors and no overdispersion parameters.
-The Stan program used to fit Model 1, as well as the other four models, can be found in the appendix.
-Model 1 fits quickly and efficiently.
-The Stan output for $\beta_0$ and $\beta$ is below.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrrrrrr}
-  \toprule
-Parameter & Rhat & n\_eff & mean & sd & 2.5\% & 97.5\% \\ 
-  \midrule
-beta\_0 & 1.000 & 19365 & -1.376 & 0.005 & -1.385 & -1.366 \\ 
-  beta[1] & 1.000 & 18085 & 0.006 & 0.007 & -0.007 & 0.019 \\ 
-  beta[2] & 1.000 & 19330 & 0.063 & 0.005 & 0.054 & 0.072 \\ 
-  beta[3] & 1.000 & 18752 & -0.027 & 0.004 & -0.035 & -0.019 \\ 
-  beta[4] & 1.000 & 18026 & -0.029 & 0.004 & -0.038 & -0.021 \\ 
-  beta[5] & 1.000 & 18380 & -0.013 & 0.005 & -0.022 & -0.004 \\ 
-  beta[6] & 1.000 & 15809 & 0.241 & 0.005 & 0.231 & 0.251 \\ 
-  beta[7] & 1.000 & 18930 & 0.014 & 0.003 & 0.007 & 0.020 \\ 
-  beta[8] & 1.000 & 17331 & -0.003 & 0.005 & -0.012 & 0.006 \\ 
-  beta[9] & 1.000 & 19360 & 0.074 & 0.004 & 0.067 & 0.082 \\ 
-  beta[10] & 1.000 & 21329 & 0.066 & 0.001 & 0.065 & 0.068 \\ 
-  beta[11] & 1.000 & 20224 & -0.121 & 0.004 & -0.128 & -0.114 \\ 
-  beta[12] & 1.000 & 22539 & -0.069 & 0.004 & -0.076 & -0.061 \\ 
-  beta[13] & 1.000 & 18372 & -0.011 & 0.005 & -0.021 & -0.001 \\ 
-  beta[14] & 1.000 & 15602 & -0.026 & 0.005 & -0.035 & -0.017 \\ 
-  beta[15] & 1.000 & 15480 & 0.113 & 0.005 & 0.104 & 0.122 \\ 
-  beta[16] & 1.000 & 19904 & -0.085 & 0.004 & -0.093 & -0.078 \\ 
-  beta[17] & 1.000 & 21169 & 0.047 & 0.003 & 0.042 & 0.053 \\ 
-  beta[18] & 1.000 & 22032 & -0.025 & 0.003 & -0.031 & -0.020 \\ 
-   \bottomrule
-\end{tabular}
-\caption{Stan output for Model 1} 
-\end{table}
-
-Model 1 produces coefficient estimates with tight confidence intervals; few of the $\beta_k$ 95% intervals cross zero.
-Based on initial diagnostics and precise parameter estimates, Model 1 appears to be a "good" fit.
-In this way, Model 1 illustrates the importance of diagnostics beyond the default Stan output.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width = 5in]{ppcheck_mod1.png}
-  \caption{Posterior predictive check for Model 1.}
-\end{figure}
-
-Model 1 clearly underestimates the number of block groups with low ridership. This is due in large part to the extra-Poisson variance in Metro Transit's ridership data.
-
-Model 2 builds upon Model 1 by incorporating and overdispersion parameter.
-Model 2 is *significantly* more computationally expensive to fit than Model 1.
-This is due to relatively large autocorrelation between each iteration, visible in the traceplots below (Figure ___) as well as in the significantly lower `n_eff` for Model 2.
-The parameter estimates for $\beta_0$ and $\beta$ for Model 2 are below.
-
-Because we gave $\theta$ and $\sigma$ vague priors, we can see a *much* better fit in the posterior predictive checks for Model 2.
-One *could* stop analysis here.
-However, the addition of $\theta \sigma$ only improves the *fit* of the model, not our understanding of the data.
-Therefore, for this study, we will decompose $\theta \sigma$ into spatial and non-spatial random effects.
-
-First, however, we implement the horseshoe priors to determine if any of our 18 covariates $x_k$ should be removed from the model.
-
-<!-- Notably, Model 3 experienced one divergent transition after warmup.
-However, the model was fitted several times with no divergent transitions.
-A "false alarm" divergent transition can occur when the cutoff for determining a divergent transition is too low; this can be true in the case of particularly complex posteriors.
-Because only 1 out of 20,000 iterations produced a divergent transition, we can proceed with caution using Model 3.
--->
-
-# Discussion
+Naturally, the baseline Model 1 converges most quickly and has the highest `n_eff` diagnostics.
+The simplicity of the Model, however, yields a poor model fit.
+Perhaps unexpectedly, Model 4, the BYM2, is the second-most efficient in terms of convergence and Effective Sample Size.
 
@@
 
-# Appendix
+# Appendix: Stan Code
 
 ## Model 1
+
+The baseline 
 ```{r}

### 2020-02-10 functions/ppcheck.R

@@ -5,7 +5,7 @@ library(data.table)
 library(ggplot2)
 
  
-post <- poisson_fit
+post <- bym
 n_obs = 1495
 n_iter = 20000
 y <- na.omit(p_dat_scaled)$daily_boards
@@ -28,10 +28,11 @@ y <- na.omit(p_dat_scaled)$daily_boards
   ppoverlay <- pp_check(y, yrep[sample(100), ], ppc_dens_overlay) + 
     theme_minimal() 
   
-  ppoverlay + xlim(0, 200) + ylim(0, 0.025) + 
-    labs(title = "Observed versus simulated ridership for Model 1")
+  ppoverlay + xlim(0, 200) + ylim(0, 0.026) + 
+    labs(title = "Observed versus simulated ridership for Model 4")
+  ggsave("~/Documents/honors/honors-projects/chapters/ppcheckmod4.png", width = 6, units = "in")
   
-  ## get traceplot
+   ## get traceplot
   
   trace <- mcmc_trace(samp2, pars = c("beta_0"))
   

### 2020-02-10 chapters/article-draft.Rmd
fix table

--- 2020-02-10 chapters/article-draft.Rmd
@@
 
-<!-- \begin{table}[h] -->
-<!-- \centering -->
-<!-- \begin{tabular}{lr} -->
-<!--   \toprule -->
-<!-- Parameter & meaning \\ -->
-<!--   \midrule -->
-<!-- beta[1] &  median household income \\ -->
-<!--   beta[2] &  percent residents identified as white alone \\ -->
-<!--   beta[3] &  area of 10 minute walk isochrone from population-weighted block group centroid \\ -->
-<!--   beta[4] &  percent residents with a high school diploma \\ -->
-<!--   beta[5] &  percent residents with a bachelor's degree \\ -->
-<!--   beta[6] &  percent of residents who rent their residence \\ -->
-<!--   beta[7] &  percent of residents without a vehicle \\ -->
-<!--   beta[8] &  percent of residents who speak only English \\ -->
-<!--   beta[9] &  percent of residents identified as foreign \\ -->
-<!--   beta[10] & employment density \\ -->
-<!--   beta[11] & percent of jobs in block group for white employees \\ -->
-<!--   beta[12] & percent of jobs in block group for male employees \\ -->
-<!--   beta[13] & percent of jobs in block group for employees with no college degree \\ -->
-<!--   beta[14] &  percent of jobs in block group for employees who make less than $40,000 \\ -->
-<!--   beta[15] &  percent of jobs in block group for employees under age 30 \\ -->
-<!--   beta[16] &  median age \\ -->
-<!--   beta[17] &  percent residents who use transit to commute to work \\ -->
-<!--   beta[18] &  population density \\ -->
-<!--    \bottomrule -->
-<!-- \end{tabular} -->
-<!-- \caption{List of covariates} -->
-<!-- \end{table} -->
+\begin{table}[H]
+\centering
+\begin{tabular}{ll} \toprule
+    Parameter & Meaning \\ \midrule
+    beta[1] & median household income \\ 
+    beta[2] & percent residents identified as white alone \\ 
+    beta[3] &  area of 10 minute walk isochrone from population-weighted block group centroid \\
+    beta[4] &  percent residents with a high school diploma \\
+    beta[5] &  percent residents with a bachelor's degree \\
+    beta[6] &  percent of residents who rent their residence \\
+    beta[7] &  percent of residents without a vehicle \\
+    beta[8] &  percent of residents who speak only English \\
+    beta[9] &  percent of residents identified as foreign \\
+    beta[10] & employment density \\
+    beta[11] & percent of jobs in block group for white employees \\
+    beta[12] & percent of jobs in block group for male employees \\
+    beta[13] & percent of jobs in block group for employees with no college degree \\
+    beta[14] & percent of jobs in block group for employees who make less than \$40,000 \\
+    beta[15] & percent of jobs in block group for employees under age 30 \\
+    beta[16] & median age \\
+    beta[17] & percent residents who use transit to commute to work \\
+    beta[18] & population density \\
+    \bottomrule
+ \hline
+\end{tabular}
+\caption{List of covariates}
+\end{table}
 

### 2020-02-10 chapters/article-draft.Rmd
checking to see if pdf will compile

@@
 author: "Raven McKnight"
-header-includes:
-- \usepackage{booktabs}
-- \usepackage{makecell}
-- \usepackage{float}
 date: "2/9/2020"
@@
 Despite this, `n_eff` is significantly lower for Model 2 than Model 1.
+Model 2 also experienced 2 transitions exceeding maixmum treedepth: not a problem in terms of model fit, but a signal that the posterior is more challenging to fit.
 Reparameterization could likely reduce autocorrelation between iterations.
@@
 The parameter estimates for $\beta_0$ and $\beta$ for Model 2 are below.
+Notably, the direction and magnitude of the coefficient estimates are similar to those produced by Model 1.
+However, the coverage intervals are much wider.
+Based on "test runs" of Model 2, we expect increasing the number of iterations to decrease the width of these coverage intervals (we observe that the converage interval is proportional to the number of iterations in this setting).
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lrrrrrr}
+  \toprule
+Parameter & Rhat & n\_eff & mean & sd & 2.5\% & 97.5\% \\ 
+  \midrule
+beta\_0 & 1.0 & 1354 & -1.9 & 0.0 & -2.0 & -1.8 \\ 
+  sigma & 1.0 & 1800 & 1.0 & 0.0 & 0.9 & 1.0 \\ 
+  beta[1] & 1.0 & 1546 & -0.1 & 0.0 & -0.2 & -0.0 \\ 
+  beta[2] & 1.0 & 1095 & 0.0 & 0.1 & -0.1 & 0.1 \\ 
+  beta[3] & 1.0 & 1322 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
+  beta[4] & 1.0 & 1208 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
+  beta[5] & 1.0 & 1282 & -0.0 & 0.0 & -0.1 & 0.1 \\ 
+  beta[6] & 1.0 & 988 & 0.1 & 0.0 & 0.1 & 0.2 \\ 
+  beta[7] & 1.0 & 1019 & 0.1 & 0.0 & -0.0 & 0.1 \\ 
+  beta[8] & 1.0 & 1178 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
+  beta[9] & 1.0 & 1125 & 0.0 & 0.0 & -0.1 & 0.1 \\ 
+  beta[10] & 1.0 & 1796 & 0.1 & 0.0 & 0.1 & 0.2 \\ 
+  beta[11] & 1.0 & 1152 & -0.1 & 0.0 & -0.1 & -0.0 \\ 
+  beta[12] & 1.0 & 1629 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
+  beta[13] & 1.0 & 2101 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
+  beta[14] & 1.0 & 1415 & -0.1 & 0.0 & -0.1 & -0.0 \\ 
+  beta[15] & 1.0 & 1581 & 0.1 & 0.0 & 0.0 & 0.2 \\ 
+  beta[16] & 1.0 & 1160 & -0.1 & 0.0 & -0.2 & -0.0 \\ 
+  beta[17] & 1.0 & 997 & 0.1 & 0.0 & 0.0 & 0.2 \\ 
+  beta[18] & 1.0 & 1589 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
+   \bottomrule
+\end{tabular}
+\caption{Stan output for Model 2} 
+\end{table}
+
+<!-- \begin{figure}[H] -->
+<!--   \centering -->
+<!--   \includegraphics[width = 6in]{ppcheckmod2.png} -->
+<!--   \caption{Posterior predictive check for Model 1.} -->
+<!-- \end{figure} -->
 
@@
 
-The baseline 
+The Stan program for the baseline model is below.
+This model could also have been fit using Stan's more efficient built-in Poisson GLM function.
+For this study, we write the GLM more "longhand" for ease of comparison between models.
 ```{r}
@@
 
+The Stan program for Model 2 differs from Model 1 only in the latent variable, `f`. 
 ```{r}
@@
 ## Model 3
+
+The regularized horseshoe prior used to fit Model 3 is adopted from Piironen and Vehtari. 
 ```{r}
@@
 
+The BYM2 model is fit using code from Morris et al.
+
 ```{r}
@@
 ## Model 5
+
+Finally, the code to fit Model 5 quite literally combines the programs for Models 3 and 4. 
 ```{r}

### 2020-02-10 chapters/article-draft.tex
checking to see if pdf will compile

--- 2020-02-10 chapters/article-draft.Rmd
@@
----
-title: "Honors Draft"
-author: "Raven McKnight"
-date: "2/9/2020"
-output:
-  pdf_document:
-    number_sections: true
-  html_document:
-    theme: paper
-    toc: yes
-    toc_float: yes
-abstract: "abstract"
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
-```
-
-# Introduction
-
-Public transit ridership has been in decline around the country for years, but particularly since the mid 2010s ().
-Following a spike in ridership related to the 2008 housing market crash, ridership has been steadily trending downwards in most metropolitan areas ().
-Bus ridership is particularly affected compared to more "desirable" light rail and commuter rail.
-Transit agencies are naturally interested in understanding the specific causes of these declines, as well as more generally being able to predict transit demand throughout space and time.
-
-Transit ridership, or demand, is often forecasted using models from the policy and planning world, such as the classic four-step model.
-Four-step transportation demand models consist of simulating trips and their spatial distribution before assigning them modes (ie bus versus car versus bicycle) and routes.
-Other models, called activity-based models predict transportation demand based on when and where riders are likely to conduct various tasks (commuting, shopping, recreation, etc).
-Others still are based largely on land-use characteristics under the general premise that different types of development (or lack thereof) are more or less likely to draw riders.
-
-It is somewhat rare to see more traditional statistical models applied to transit ridership.
-In the context of predicting future ridership, the above methods are often sufficient.
-Additionally, they are the industry standard and allow for the integration of expert planning knowledge.
-However, in the context of understanding *past* ridership, we can apply more explanatory methods.
-Specifically, we can use hierarchical spatial Bayesian models to incorporate many covariates in addition to the spatial structure that necessarily underpins transit ridership (ie, boardings can only occur at existing bus stops).
-
-Hierarchical spatial models allow us to understand the patterns of ridership within a city with great granularity.
-Some existing literature explores the decline of transit ridership *between* cities ().
-Few studies have been conducted on the various patterns of ridership *within* a single metropolitan area.
-Additionally, few studies incorporate many demographic predictors.
-Naturally, we cannot collect demographic information for every rider of a transit system.
-Instead, we can use demographic predictors associated with spatial units such as census block groups to explore the question of *who* rides -- or, at the very least, *where* do they ride.
-
-# Background
-
-## Metro Transit
-
-Metro Transit is the primary transit provider in the Minneapolis-Saint Paul metropolitan area.
-The agency operates one commuter rail line (Northstar), two light rail lines (Blue Line and Green Line), and over 100 bus routes.
-Since the 2014 opening of the Green Line, rail ridership has increased each year ().
-Bus ridership, however, has been in decline.
-According to the Metro Transit Riders' Almanac blog, much of the decline is bus ridership can be attributed to the busiest urban local routes as well as situational factors, such as lower-than-average gas prices and a system-wide fare increase in 2017.
-
-The agency is naturally interested in understanding with more detail the reasons for and nature of these declines.
-
-Conducted in tandem with Metro Transit's Network Next bus system redesign, this study aims to further Metro Transit's understanding of their bus network.
-Specifically, we aim to describe the existing pattern of transit ridership via demographic predictors and spatial patterns.
-Future work will incorporate temporal trends to better understand declining ridership.
-
-## Transit Market Areas
-
-Metro is interested in predicting transit demand across the region in order to guide route planning and service provision.
-Currently, Metro Transit uses a set of 5 Transit Market Areas to estimate spatial demand.
-The Transit Market Areas are writen into the official Transportation Planning Policy (TPP) and are calculated using a simple linear regression.
-Following the notation of the TPP, the formula for determing Transit Market Areas is expressed
-
-$$
+\documentclass[]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage[]{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
+\hypersetup{
+            pdftitle={Honors Draft},
+            pdfauthor={Raven McKnight},
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+\usepackage[margin=1in]{geometry}
+\usepackage{graphicx,grffile}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{5}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
+\title{Honors Draft}
+\author{Raven McKnight}
+\date{2/9/2020}
+
+\begin{document}
+\maketitle
+\begin{abstract}
+abstract
+\end{abstract}
+
+\section{Introduction}\label{introduction}
+
+Public transit ridership has been in decline around the country for
+years, but particularly since the mid 2010s (). Following a spike in
+ridership related to the 2008 housing market crash, ridership has been
+steadily trending downwards in most metropolitan areas (). Bus ridership
+is particularly affected compared to more ``desirable'' light rail and
+commuter rail. Transit agencies are naturally interested in
+understanding the specific causes of these declines, as well as more
+generally being able to predict transit demand throughout space and
+time.
+
+Transit ridership, or demand, is often forecasted using models from the
+policy and planning world, such as the classic four-step model.
+Four-step transportation demand models consist of simulating trips and
+their spatial distribution before assigning them modes (ie bus versus
+car versus bicycle) and routes. Other models, called activity-based
+models predict transportation demand based on when and where riders are
+likely to conduct various tasks (commuting, shopping, recreation, etc).
+Others still are based largely on land-use characteristics under the
+general premise that different types of development (or lack thereof)
+are more or less likely to draw riders.
+
+It is somewhat rare to see more traditional statistical models applied
+to transit ridership. In the context of predicting future ridership, the
+above methods are often sufficient. Additionally, they are the industry
+standard and allow for the integration of expert planning knowledge.
+However, in the context of understanding \emph{past} ridership, we can
+apply more explanatory methods. Specifically, we can use hierarchical
+spatial Bayesian models to incorporate many covariates in addition to
+the spatial structure that necessarily underpins transit ridership (ie,
+boardings can only occur at existing bus stops).
+
+Hierarchical spatial models allow us to understand the patterns of
+ridership within a city with great granularity. Some existing literature
+explores the decline of transit ridership \emph{between} cities (). Few
+studies have been conducted on the various patterns of ridership
+\emph{within} a single metropolitan area. Additionally, few studies
+incorporate many demographic predictors. Naturally, we cannot collect
+demographic information for every rider of a transit system. Instead, we
+can use demographic predictors associated with spatial units such as
+census block groups to explore the question of \emph{who} rides -- or,
+at the very least, \emph{where} do they ride.
+
+\section{Background}\label{background}
+
+\subsection{Metro Transit}\label{metro-transit}
+
+Metro Transit is the primary transit provider in the Minneapolis-Saint
+Paul metropolitan area. The agency operates one commuter rail line
+(Northstar), two light rail lines (Blue Line and Green Line), and over
+100 bus routes. Since the 2014 opening of the Green Line, rail ridership
+has increased each year (). Bus ridership, however, has been in decline.
+According to the Metro Transit Riders' Almanac blog, much of the decline
+is bus ridership can be attributed to the busiest urban local routes as
+well as situational factors, such as lower-than-average gas prices and a
+system-wide fare increase in 2017.
+
+The agency is naturally interested in understanding with more detail the
+reasons for and nature of these declines.
+
+Conducted in tandem with Metro Transit's Network Next bus system
+redesign, this study aims to further Metro Transit's understanding of
+their bus network. Specifically, we aim to describe the existing pattern
+of transit ridership via demographic predictors and spatial patterns.
+Future work will incorporate temporal trends to better understand
+declining ridership.
+
+\subsection{Transit Market Areas}\label{transit-market-areas}
+
+Metro is interested in predicting transit demand across the region in
+order to guide route planning and service provision. Currently, Metro
+Transit uses a set of 5 Transit Market Areas to estimate spatial demand.
+The Transit Market Areas are writen into the official Transportation
+Planning Policy (TPP) and are calculated using a simple linear
+regression. Following the notation of the TPP, the formula for determing
+Transit Market Areas is expressed
+
+\[
 \begin{aligned}
@@
 \end{aligned}
-$$
-
-where each predictor is logged and scaled by developed land acreage per census block group.
-Here, automobile availability refers to the number of adults over age 16 less the total number of automobiles available in a block group (scaled by developed land acreage).
-There is an additional indicator variable, omitted in this notation, for the census block group containing the MSP International Airport.
-For more documentation of the official Transit Market Areas, refer to the TPPs [Appendix G](https://metrocouncil.org/Transportation/Planning-2/Key-Transportation-Planning-Documents/Transportation-Policy-Plan/The-Adopted-2040-TPP-(1)/Final-2040-Transportation-Policy-Plan/2040-TPP-Appendix-G-Transit-Design-and-Perf-Standa.aspx).
+\]
+
+where each predictor is logged and scaled by developed land acreage per
+census block group. Here, automobile availability refers to the number
+of adults over age 16 less the total number of automobiles available in
+a block group (scaled by developed land acreage). There is an additional
+indicator variable, omitted in this notation, for the census block group
+containing the MSP International Airport. For more documentation of the
+official Transit Market Areas, refer to the TPPs
+\href{https://metrocouncil.org/Transportation/Planning-2/Key-Transportation-Planning-Documents/Transportation-Policy-Plan/The-Adopted-2040-TPP-(1)/Final-2040-Transportation-Policy-Plan/2040-TPP-Appendix-G-Transit-Design-and-Perf-Standa.aspx}{Appendix
+G}.
 
@@
 
-Census block groups are split into 5 market areas based on the Transit Market Index described above, where the highest market area (Market Area 1) is expected to support high-frequency, all-day service and the lowest market area (Market Area 5) is expected to support peak commuter express service and park-and-rides, if that.
-
-The Transit Market Index values are geographically smoothed to create more-or-less concentric Transit Market Areas.
-The linear regression is an intuitive way to think about transit ridership across space.
-The four predictors used are common-sense indicators of transit ridership: high population and employment density are characteristic of trip origins and destinations, automobile availability is reasonably assumed to be related to transit ridership, and intersection density is a proxy for the "walkability" of an area.
-However, the exsiting Transit Market Area model may be over-simplifying the complex question of transit ridership.
-
-# Data
-
-## Metro Transit Data
-
-This analysis relies on several Metro Transit provided data sets. The two primary sources are automatically reported by in-service vehicles, yielding billions of rows of observations.
-
-**Automatic Passenger Count (APC)** data is reported every time a bus opens its doors.
-Two beams of light detect movement through the doors of the bus and counts the number of passengers getting on and off at each bus stop.
-Naturally, this data source is flawed: sometimes, vehicles fail to report data, and the sensors can be easily tricked.
-Someone getting off the bus pulling a suitcase behind them, for example, would likely be counted as two passengers exiting.
-However, APC is the most granular ridership data available, giving us the most control over our spatial aggregations of ridership.
-This is the primary data source used in this analysis.
-
-**Automatic Vehicle Location (AVL)** data is similarly reported by in-service vehicles (approximately) every 8 seconds.
-The vehicle reports in GPS location continuously while in service.
-We use AVL data to adjust for missing APC data.
-
-**Metro Transit's Schedule** as *planned*, rather than as *run*, is also used to correct missing APC data.
-
-This analysis began by pulling all APC, AVL, and schedule data from 2015-2018.
-The initial data pull consisted of approximately 496 *billion* rows of data.
-The analysis presented here focuses on 2017 data.
-Figure *___* shows the average weekday boardings by census block group in 2017.
+Census block groups are split into 5 market areas based on the Transit
+Market Index described above, where the highest market area (Market Area
+1) is expected to support high-frequency, all-day service and the lowest
+market area (Market Area 5) is expected to support peak commuter express
+service and park-and-rides, if that.
+
+The Transit Market Index values are geographically smoothed to create
+more-or-less concentric Transit Market Areas. The linear regression is
+an intuitive way to think about transit ridership across space. The four
+predictors used are common-sense indicators of transit ridership: high
+population and employment density are characteristic of trip origins and
+destinations, automobile availability is reasonably assumed to be
+related to transit ridership, and intersection density is a proxy for
+the ``walkability'' of an area. However, the exsiting Transit Market
+Area model may be over-simplifying the complex question of transit
+ridership.
+
+\section{Data}\label{data}
+
+\subsection{Metro Transit Data}\label{metro-transit-data}
+
+This analysis relies on several Metro Transit provided data sets. The
+two primary sources are automatically reported by in-service vehicles,
+yielding billions of rows of observations.
+
+\textbf{Automatic Passenger Count (APC)} data is reported every time a
+bus opens its doors. Two beams of light detect movement through the
+doors of the bus and counts the number of passengers getting on and off
+at each bus stop. Naturally, this data source is flawed: sometimes,
+vehicles fail to report data, and the sensors can be easily tricked.
+Someone getting off the bus pulling a suitcase behind them, for example,
+would likely be counted as two passengers exiting. However, APC is the
+most granular ridership data available, giving us the most control over
+our spatial aggregations of ridership. This is the primary data source
+used in this analysis.
+
+\textbf{Automatic Vehicle Location (AVL)} data is similarly reported by
+in-service vehicles (approximately) every 8 seconds. The vehicle reports
+in GPS location continuously while in service. We use AVL data to adjust
+for missing APC data.
+
+\textbf{Metro Transit's Schedule} as \emph{planned}, rather than as
+\emph{run}, is also used to correct missing APC data.
+
+This analysis began by pulling all APC, AVL, and schedule data from
+2015-2018. The initial data pull consisted of approximately 496
+\emph{billion} rows of data. The analysis presented here focuses on 2017
+data. Figure *\_\_\_* shows the average weekday boardings by census
+block group in 2017.
 
@@
 
-### Data Interpolation
-
-In theory, we have APC data for each run bus trip in 2017.
-However, we know that, for a variety of reasons, this is untrue.
-The automatic data reporting technology on Metro Transit buses is flawed and can fail to report data unpredictably.
-Additionally, the trips missing data may not be random: all buses of one "series", for example, may fail to report at once.
-This could lead to the APC dataset understimating boardings on a particular set of routes or locations.T herefore, we use the three data sets described above to create a more complete augmented APC data set.
+\subsubsection{Data Interpolation}\label{data-interpolation}
+
+In theory, we have APC data for each run bus trip in 2017. However, we
+know that, for a variety of reasons, this is untrue. The automatic data
+reporting technology on Metro Transit buses is flawed and can fail to
+report data unpredictably. Additionally, the trips missing data may not
+be random: all buses of one ``series'', for example, may fail to report
+at once. This could lead to the APC dataset understimating boardings on
+a particular set of routes or locations.T herefore, we use the three
+data sets described above to create a more complete augmented APC data
+set.
 
@@
 
-1. Compares scheduled trips to trips we have APC data for. If a trip was scheduled but has no APC records, we consider that trip missing. 
-2. For missing trips, check AVL data. If there are no AVL records, we consider the trip cut (ie, we assume the trip was not run). 
-3.
-For missing trips with AVL data, we estimate boardings for that trip by taking the average number of boardings for trips of that nature (trips on the same route, day of week, season, location, etc).
-
-### Data Aggregation
-
-For the models presented in this study, we aggregate boardings at individual bus stops to the census block group.
-This is in large part to match the granularity of the American Community Survey covariates we are interested in using.
-Future work could explore using areal covariates to model ridership at specific points (ie, bus stops).
-
-Additionally, we aggregate boardings counts to average weekday boardings.
-This a) greatly reduces the number of rows of data in play, and therefore shortens computation time significantly, and b) allows us to build a spatial model without simultaneously incorporating temporal trends.
-The data interpolation described above will be more impactful in future studies incorporating temporal trends.
-
-## Other data sources
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Compares scheduled trips to trips we have APC data for. If a trip was
+  scheduled but has no APC records, we consider that trip missing.
+\item
+  For missing trips, check AVL data. If there are no AVL records, we
+  consider the trip cut (ie, we assume the trip was not run).
+\item
+  For missing trips with AVL data, we estimate boardings for that trip
+  by taking the average number of boardings for trips of that nature
+  (trips on the same route, day of week, season, location, etc).
+\end{enumerate}
+
+\subsubsection{Data Aggregation}\label{data-aggregation}
+
+For the models presented in this study, we aggregate boardings at
+individual bus stops to the census block group. This is in large part to
+match the granularity of the American Community Survey covariates we are
+interested in using. Future work could explore using areal covariates to
+model ridership at specific points (ie, bus stops).
+
+Additionally, we aggregate boardings counts to average weekday
+boardings. This a) greatly reduces the number of rows of data in play,
+and therefore shortens computation time significantly, and b) allows us
+to build a spatial model without simultaneously incorporating temporal
+trends. The data interpolation described above will be more impactful in
+future studies incorporating temporal trends.
+
+\subsection{Other data sources}\label{other-data-sources}
 
 All other data sources used in this study are from outside sources.
-Census Block Groups are used as the areal unit of analysis.
-Covariates of interest are primarily from the 2017 American Community Survey (ACS) 5-Year Estimates.
-Additional variables related to employment are sourced from the Longitudinal Origin-Destination Employment Statistics (LODES) dataset.
-A full list of covariates used in this study is below.
+Census Block Groups are used as the areal unit of analysis. Covariates
+of interest are primarily from the 2017 American Community Survey (ACS)
+5-Year Estimates. Additional variables related to employment are sourced
+from the Longitudinal Origin-Destination Employment Statistics (LODES)
+dataset. A full list of covariates used in this study is below.
 
@@
 
-Note that `beta[3]` is a measure of walkability.
-A 10-minute walk isochrone, generated using OpenTripPlanner, shows the area a person on foot can cover in 10 minutes on the existing street network.
-In theory, we expect more walkable areas, such as the downtown Saint Paul and Minneapolis, to have larger walk-isochrones.
-Unfortunately, the metric can be fooled by rural areas where you can walk uninterruped down a single roadway, for example.
-An improved "walkability" metric could improve models of transit ridership.
-
-All covariates in the table above are standardized to have mean 0 and standard deviation 1.
-Intuitively, we expect many of the covariates to be correlated.
-Therefore, we will utilize variable selection to determine which covariates to keep in our analysis.
-
-# Models
-
-## Stan
-
-All of the models discussed above can be fit using the programming language Stan.
-Stan is a probabilistic language which uses Hamiltonian Monte Carlo (HMC) and a specialized No U-Turn Sampler (NUTS) to compute joint log probability densities.
-The specifics of Stan are beyond the scope of this paper.
-However, a few of its particular diagnostics are helpful in understanding the differences between the model fits below.
-
-* $\hat{R}$ or `Rhat` indicates how well the Markov chains have mixed by comparing between- and within-chain estimates.
-At convergence, $\hat{R} = 1$.
-When $\hat{R} > 1$, the Markov chains have not mixed well and the model likely needs to be run for more iterations.
-* `n_eff` is the estimated Effective Sample Size (ESS).
-ESS is the number of independent samples our sample is equivalent to.
-In other words, if `n_eff` = 1,000, we expect to gain the same amount of information from our sample as we would from 1,000 independent samples.
-In a "perfect" scenario, `n_eff` should equal the actual number of samples.
-When `n_eff` is significantly below the actual number of samples, there may be a problem with the model and posterior estimates may be biased.
-* **Divergent transitions** occur when the simulated Hamiltonian departs from the actual Hamiltonian trajectory.
-In general, if stan gives any divergent transition warnings, we should consider the posterior estimates unreliable.
-* `max_treedepth` is a parameter we set when fitting a model in Stan.
-Treedepth refers to the number of steps the sampler has to take at each iteration.
-The default is 10.
-For extremely "curved" or complex posteriors, the treedepth may be much higher.
-When `max_treedepth` is exceeded, the sampler quits prematurely.
-Therefore, if we have *many* iterations exceeding `max_treedepth`, posterior estimates may be unreliable.
-Additionally, setting high treedepth causes the sampler to perform much more slowly.
-* `adapt_delta` is also a sampling parameter; it corresponds to the Mean Metropolis Acceptance Rate (see ____).
-The default is 0.8.
-Setting a higher `adapt_delta` slows computation time but *can* eliminate "false alarm" divergent transitions.
-
-## Poisson Regression
-
-For each block group $i = 1, 2, 3, ..., 1495$ with any ridership, the "baseline" model is a simple Poisson regression.
-Poisson regressions are a type of generalized linear model with the natural log as its link function.
-For ridership $Y_i$ and covariates $x_1, ..., x_k$ discussed above, the Poisson regression is written
-
-$$
+Note that \texttt{beta{[}3{]}} is a measure of walkability. A 10-minute
+walk isochrone, generated using OpenTripPlanner, shows the area a person
+on foot can cover in 10 minutes on the existing street network. In
+theory, we expect more walkable areas, such as the downtown Saint Paul
+and Minneapolis, to have larger walk-isochrones. Unfortunately, the
+metric can be fooled by rural areas where you can walk uninterruped down
+a single roadway, for example. An improved ``walkability'' metric could
+improve models of transit ridership.
+
+All covariates in the table above are standardized to have mean 0 and
+standard deviation 1. Intuitively, we expect many of the covariates to
+be correlated. Therefore, we will utilize variable selection to
+determine which covariates to keep in our analysis.
+
+\section{Models}\label{models}
+
+\subsection{Stan}\label{stan}
+
+All of the models discussed above can be fit using the programming
+language Stan. Stan is a probabilistic language which uses Hamiltonian
+Monte Carlo (HMC) and a specialized No U-Turn Sampler (NUTS) to compute
+joint log probability densities. The specifics of Stan are beyond the
+scope of this paper. However, a few of its particular diagnostics are
+helpful in understanding the differences between the model fits below.
+
+\begin{itemize}
+\tightlist
+\item
+  \(\hat{R}\) or \texttt{Rhat} indicates how well the Markov chains have
+  mixed by comparing between- and within-chain estimates. At
+  convergence, \(\hat{R} = 1\). When \(\hat{R} > 1\), the Markov chains
+  have not mixed well and the model likely needs to be run for more
+  iterations.\\
+\item
+  \texttt{n\_eff} is the estimated Effective Sample Size (ESS). ESS is
+  the number of independent samples our sample is equivalent to. In
+  other words, if \texttt{n\_eff} = 1,000, we expect to gain the same
+  amount of information from our sample as we would from 1,000
+  independent samples. In a ``perfect'' scenario, \texttt{n\_eff} should
+  equal the actual number of samples. When \texttt{n\_eff} is
+  significantly below the actual number of samples, there may be a
+  problem with the model and posterior estimates may be biased.
+\item
+  \textbf{Divergent transitions} occur when the simulated Hamiltonian
+  departs from the actual Hamiltonian trajectory. In general, if stan
+  gives any divergent transition warnings, we should consider the
+  posterior estimates unreliable.\\
+\item
+  \texttt{max\_treedepth} is a parameter we set when fitting a model in
+  Stan. Treedepth refers to the number of steps the sampler has to take
+  at each iteration. The default is 10. For extremely ``curved'' or
+  complex posteriors, the treedepth may be much higher. When
+  \texttt{max\_treedepth} is exceeded, the sampler quits prematurely.
+  Therefore, if we have \emph{many} iterations exceeding
+  \texttt{max\_treedepth}, posterior estimates may be unreliable.
+  Additionally, setting high treedepth causes the sampler to perform
+  much more slowly.
+\item
+  \texttt{adapt\_delta} is also a sampling parameter; it corresponds to
+  the Mean Metropolis Acceptance Rate (see \_\_\_\_). The default is
+  0.8. Setting a higher \texttt{adapt\_delta} slows computation time but
+  \emph{can} eliminate ``false alarm'' divergent transitions.
+\end{itemize}
+
+\subsection{Poisson Regression}\label{poisson-regression}
+
+For each block group \(i = 1, 2, 3, ..., 1495\) with any ridership, the
+``baseline'' model is a simple Poisson regression. Poisson regressions
+are a type of generalized linear model with the natural log as its link
+function. For ridership \(Y_i\) and covariates \(x_1, ..., x_k\)
+discussed above, the Poisson regression is written
+
+\[
 \begin{aligned}
@@
 \end{aligned}
-$$
-
-where $\beta_0$ is the intercept and $E_i$ is an offset term.
-
-The offset $E_i$ is often termed "exposure" in applications such as disease risk modeling where the offset may be the number of observed cases in a previous year, for example.
-More technically, the offset term scales the Poisson output to be a rate rather than a count.
-This is appropriate when observations $i$ have different potentials for response $Y$.
-For example, a county with higher population will naturally have a higher count of patients with asthma than a county with a smaller population.
-Mathematically, the offset is a covariate with parameter set equal to 1.
-Recall, the Poisson regression assumes we can model the mean of response $Y$, $\bar{Y}$ with a combination of linear predictors:
-
-$$
+\]
+
+where \(\beta_0\) is the intercept and \(E_i\) is an offset term.
+
+The offset \(E_i\) is often termed ``exposure'' in applications such as
+disease risk modeling where the offset may be the number of observed
+cases in a previous year, for example. More technically, the offset term
+scales the Poisson output to be a rate rather than a count. This is
+appropriate when observations \(i\) have different potentials for
+response \(Y\). For example, a county with higher population will
+naturally have a higher count of patients with asthma than a county with
+a smaller population. Mathematically, the offset is a covariate with
+parameter set equal to 1. Recall, the Poisson regression assumes we can
+model the mean of response \(Y\), \(\bar{Y}\) with a combination of
+linear predictors:
+
+\[
 log(\bar{Y}) = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k \\
-$$
-Therefore, if we wish to model the rate $Y/E$, the equation is rewritten
-
-$$
+\] Therefore, if we wish to model the rate \(Y/E\), the equation is
+rewritten
+
+\[
 \begin{aligned}
@@
 \end{aligned}
-$$
-
-In this study, we define $E_i$ to equal the number of times a bus stops in block group $i$. This allows us to control for block groups with more or less supply.
-
-In the basic Poisson regression, we often give $\beta_0$ and $\beta$ vague priors such as Normal($0, 1$). The exposure term $E_i$ comes from the data and does not recieve a prior.
-
-### Model 1
-
-We will call this baseline Poisson regression Model 1.
-Model 1 uses all 18 covariates with simple Normal(0, 1) priors and no overdispersion parameters.
-The Stan program used to fit Model 1, as well as the other four models, can be found in the appendix.
-Model 1 fits quickly and efficiently.
-The Stan output for $\beta_0$ and $\beta$ is below.
+\]
+
+In this study, we define \(E_i\) to equal the number of times a bus
+stops in block group \(i\). This allows us to control for block groups
+with more or less supply.
+
+In the basic Poisson regression, we often give \(\beta_0\) and \(\beta\)
+vague priors such as Normal(\(0, 1\)). The exposure term \(E_i\) comes
+from the data and does not recieve a prior.
+
+\subsubsection{Model 1}\label{model-1}
+
+We will call this baseline Poisson regression Model 1. Model 1 uses all
+18 covariates with simple Normal(0, 1) priors and no overdispersion
+parameters. The Stan program used to fit Model 1, as well as the other
+four models, can be found in the appendix. Model 1 fits quickly and
+efficiently. The Stan output for \(\beta_0\) and \(\beta\) is below.
 
@@
 
-Model 1 produces coefficient estimates with tight confidence intervals; few of the $\beta_k$ 95% intervals cross zero.
-Based on initial diagnostics and precise parameter estimates, Model 1 appears to be a "good" fit.
-In this way, Model 1 illustrates the importance of diagnostics beyond the default Stan output.
+Model 1 produces coefficient estimates with tight confidence intervals;
+few of the \(\beta_k\) 95\% intervals cross zero. Based on initial
+diagnostics and precise parameter estimates, Model 1 appears to be a
+``good'' fit. In this way, Model 1 illustrates the importance of
+diagnostics beyond the default Stan output.
 
@@
 
-Model 1 clearly underestimates the number of block groups with low ridership. This is due in large part to extra-Poisson variance present in Metro Transit's ridership data.
-
-## Overdispersed Poisson Regression
-
-The most obvious assumption made in a Poisson regression is that response variable $Y$ follows a Poisson distribution.
-In addition to requiring integer counts, this assumption requires that $\text{Var}(Y) = \text{E}(Y)$.
-This is often not true in practice.
-In the case of the Metro Transit ridership data, $\text{Var}(Y) >>\text{E}(Y)$.
-In this case, we call the unexpectedly large variance "overdispersion" or "extra-Poisson variance."
-
-Data with this feature can be difficult to model directly with a Poisson regression.
-One option is to use a generalization of the Poisson distribution, such as a Negative-Binomial distribution, as the observation model.
-For the Metro Transit application, we instead modify the latent function to include a seperate term to account for overdispersion.
-As such, the Poisson regression above can be rewritten to include
-
-$$
+Model 1 clearly underestimates the number of block groups with low
+ridership. This is due in large part to extra-Poisson variance present
+in Metro Transit's ridership data.
+
+\subsection{Overdispersed Poisson
+Regression}\label{overdispersed-poisson-regression}
+
+The most obvious assumption made in a Poisson regression is that
+response variable \(Y\) follows a Poisson distribution. In addition to
+requiring integer counts, this assumption requires that
+\(\text{Var}(Y) = \text{E}(Y)\). This is often not true in practice. In
+the case of the Metro Transit ridership data,
+\(\text{Var}(Y) >>\text{E}(Y)\). In this case, we call the unexpectedly
+large variance ``overdispersion'' or ``extra-Poisson variance.''
+
+Data with this feature can be difficult to model directly with a Poisson
+regression. One option is to use a generalization of the Poisson
+distribution, such as a Negative-Binomial distribution, as the
+observation model. For the Metro Transit application, we instead modify
+the latent function to include a seperate term to account for
+overdispersion. As such, the Poisson regression above can be rewritten
+to include
+
+\[
 \begin{aligned}
@@
 \end{aligned}
-$$
-
-where $\theta$ is a set of heterogenous random effects.
-In practice, $\theta$ is scaled by its variance parameter $\sigma$ to allow eassier fitting in Stan.
-The addition of this set of random effects improves fits on data with overdispersion.
-With sufficiently vague priors, $\theta$ and $\sigma$ can account for most or all of the overdispersion not accounted for by $\beta_0 + \sum_{k=1}^{K}x_k^T\beta_k$.
-
-### Model 2
-
-Model 2 builds upon Model 1 by incorporating the set of random effects described above.
-Model 2 is *significantly* more computationally expensive to fit than Model 1.
-This is due to relatively large autocorrelation between each iteration, visible in the traceplots below (Figure ___).
-This autocorrelation made it necessary to run Model 2 on four chains for 20,000 iterations each.
-Despite this, `n_eff` is significantly lower for Model 2 than Model 1.
-Model 2 also experienced 2 transitions exceeding maixmum treedepth: not a problem in terms of model fit, but a signal that the posterior is more challenging to fit.
-Reparameterization could likely reduce autocorrelation between iterations.
-
-The parameter estimates for $\beta_0$ and $\beta$ for Model 2 are below.
-Notably, the direction and magnitude of the coefficient estimates are similar to those produced by Model 1.
-However, the coverage intervals are much wider.
-Based on "test runs" of Model 2, we expect increasing the number of iterations to decrease the width of these coverage intervals (we observe that the converage interval is proportional to the number of iterations in this setting).
+\]
+
+where \(\theta\) is a set of heterogenous random effects. In practice,
+\(\theta\) is scaled by its variance parameter \(\sigma\) to allow
+eassier fitting in Stan. The addition of this set of random effects
+improves fits on data with overdispersion. With sufficiently vague
+priors, \(\theta\) and \(\sigma\) can account for most or all of the
+overdispersion not accounted for by
+\(\beta_0 + \sum_{k=1}^{K}x_k^T\beta_k\).
+
+\subsubsection{Model 2}\label{model-2}
+
+Model 2 builds upon Model 1 by incorporating the set of random effects
+described above. Model 2 is \emph{significantly} more computationally
+expensive to fit than Model 1. This is due to relatively large
+autocorrelation between each iteration, visible in the traceplots below
+(Figure \_\_\_). This autocorrelation made it necessary to run Model 2
+on four chains for 20,000 iterations each. Despite this, \texttt{n\_eff}
+is significantly lower for Model 2 than Model 1. Model 2 also
+experienced 2 transitions exceeding maixmum treedepth: not a problem in
+terms of model fit, but a signal that the posterior is more challenging
+to fit. Reparameterization could likely reduce autocorrelation between
+iterations.
+
+The parameter estimates for \(\beta_0\) and \(\beta\) for Model 2 are
+below. Notably, the direction and magnitude of the coefficient estimates
+are similar to those produced by Model 1. However, the coverage
+intervals are much wider. Based on ``test runs'' of Model 2, we expect
+increasing the number of iterations to decrease the width of these
+coverage intervals (we observe that the converage interval is
+proportional to the number of iterations in this setting).
 
@@
 
-<!-- \begin{figure}[H] -->
-<!--   \centering -->
-<!--   \includegraphics[width = 6in]{ppcheckmod2.png} -->
-<!--   \caption{Posterior predictive check for Model 1.} -->
-<!-- \end{figure} -->
-
-Because we gave $\theta$ and $\sigma$ vague priors, we can see a *much* better fit in the posterior predictive checks for Model 2.
-One *could* stop analysis here.
-However, the addition of $\theta \sigma$ only improves the *fit* of the model, not our understanding of the data.
-Therefore, for this study, we will decompose $\theta \sigma$ into spatial and non-spatial random effects.
-First, however, we will implement a Bayesian variable selection process to determine which, if any, of our covariates $x_k$ should be removed from the analysis.
-
-## Horseshoe Priors and Variable Selection
-
-In the baseline model, we give coefficients $\beta_1, ..., \beta_k$ simple normal priors.
-With a large set of possible covariates, however, we may reasonably expect some of the $\beta_k$ coefficients to be equal to zero -- in other words, that some of the parameters $x_k$ have no effect on response $Y$.
-In frequentist applications, we might use a method such as the LASSO (Least Absolute Shrinkage and Selection Operator) to identify relevant variables.
-There are several Bayesian alternatives for variable selection with generally consist of applying particular priors to $\beta$.
-
-### Spike and Slab - needs graphics
-
-In the Bayesian setting, there are two primary methods for variable selection.
-The first is called the "spike-and-slab" prior (Mitchell and Beauchamp, 1988; George and McCulloch, 1999) and has often been considered the "gold standard" for sparse Bayesian regression, or variable selection (Piironen and Vehtari, 2017).
-This prior is often expressed as
-
-$$
+Because we gave \(\theta\) and \(\sigma\) vague priors, we can see a
+\emph{much} better fit in the posterior predictive checks for Model 2.
+One \emph{could} stop analysis here. However, the addition of
+\(\theta \sigma\) only improves the \emph{fit} of the model, not our
+understanding of the data. Therefore, for this study, we will decompose
+\(\theta \sigma\) into spatial and non-spatial random effects. First,
+however, we will implement a Bayesian variable selection process to
+determine which, if any, of our covariates \(x_k\) should be removed
+from the analysis.
+
+\subsection{Horseshoe Priors and Variable
+Selection}\label{horseshoe-priors-and-variable-selection}
+
+In the baseline model, we give coefficients \(\beta_1, ..., \beta_k\)
+simple normal priors. With a large set of possible covariates, however,
+we may reasonably expect some of the \(\beta_k\) coefficients to be
+equal to zero -- in other words, that some of the parameters \(x_k\)
+have no effect on response \(Y\). In frequentist applications, we might
+use a method such as the LASSO (Least Absolute Shrinkage and Selection
+Operator) to identify relevant variables. There are several Bayesian
+alternatives for variable selection with generally consist of applying
+particular priors to \(\beta\).
+
+\subsubsection{Spike and Slab - needs
+graphics}\label{spike-and-slab---needs-graphics}
+
+In the Bayesian setting, there are two primary methods for variable
+selection. The first is called the ``spike-and-slab'' prior (Mitchell
+and Beauchamp, 1988; George and McCulloch, 1999) and has often been
+considered the ``gold standard'' for sparse Bayesian regression, or
+variable selection (Piironen and Vehtari, 2017). This prior is often
+expressed as
+
+\[
 \begin{aligned}
@@
 \end{aligned}
-$$
-
-for $j = 1, 2, ..., J$ where $\lambda_j \in {0, 1}$ indicates whether $\beta_j$ is from the "spike" (ie, near zero) or the "slab" (ie, nonzero).
-The "spike" is the area where most of the prior density for $\beta$ is centered (generally around 0) while the "slab" refers to the low-density extent of the prior.
-The spike-and-slab prior encourages all $\beta_j$ towards zero; only $\beta_j$ sufficiently far from 0 will be estimated to be from the slab.
-
-The primary challenge with the spike-and-slab (and other Bayesian shrinkage methods) is the selction of priors.
-Priors must be set for the width of the slab $c$ and the prior inclusion probability $\pi$.
-Here, $\pi$ reflects our prior understanding of the sparsity of $\beta$.
-In practice, we rarely have strong prior knowledge of the number of predictors we expect to be distinguishable from zero.
-This can make the implementation of sparse regression more challenging than, say, the frequentist LASSO.
-
-### Horseshoe - needs graphics
-
-The second method for Bayesian sparse regression is to give $\beta$ some continuous *sparsity inducing prior*.
-**Van Erp et al (2019)** provide an introduction to many such continous priors.
-Ridge regression and the Bayesian LASSO are two of the most approachable priors, but all sparsity inducing priors utilize the same logic as the spike-and-slab: give $\beta$ a prior with *most* of its mass near 0, to shrink irrelevant coefficients, and the rest of its mass "far" from 0.
-Identifying "far" depends more or less on the specific data and model in question.
-
-The horseshoe prior, proposed by **Carvalho et al, 2010** is a popular choice in Bayesian literature, in part because it is similar to the "gold standard" spike-and-slab.
-Following the notation of **Vehtari & Piironen**, the horseshoe prior can be expressed
-
-$$
+\]
+
+for \(j = 1, 2, ..., J\) where \(\lambda_j \in {0, 1}\) indicates
+whether \(\beta_j\) is from the ``spike'' (ie, near zero) or the
+``slab'' (ie, nonzero). The ``spike'' is the area where most of the
+prior density for \(\beta\) is centered (generally around 0) while the
+``slab'' refers to the low-density extent of the prior. The
+spike-and-slab prior encourages all \(\beta_j\) towards zero; only
+\(\beta_j\) sufficiently far from 0 will be estimated to be from the
+slab.
+
+The primary challenge with the spike-and-slab (and other Bayesian
+shrinkage methods) is the selction of priors. Priors must be set for the
+width of the slab \(c\) and the prior inclusion probability \(\pi\).
+Here, \(\pi\) reflects our prior understanding of the sparsity of
+\(\beta\). In practice, we rarely have strong prior knowledge of the
+number of predictors we expect to be distinguishable from zero. This can
+make the implementation of sparse regression more challenging than, say,
+the frequentist LASSO.
+
+\subsubsection{Horseshoe - needs
+graphics}\label{horseshoe---needs-graphics}
+
+The second method for Bayesian sparse regression is to give \(\beta\)
+some continuous \emph{sparsity inducing prior}. \textbf{Van Erp et al
+(2019)} provide an introduction to many such continous priors. Ridge
+regression and the Bayesian LASSO are two of the most approachable
+priors, but all sparsity inducing priors utilize the same logic as the
+spike-and-slab: give \(\beta\) a prior with \emph{most} of its mass near
+0, to shrink irrelevant coefficients, and the rest of its mass ``far''
+from 0. Identifying ``far'' depends more or less on the specific data
+and model in question.
+
+The horseshoe prior, proposed by \textbf{Carvalho et al, 2010} is a
+popular choice in Bayesian literature, in part because it is similar to
+the ``gold standard'' spike-and-slab. Following the notation of
+\textbf{Vehtari \& Piironen}, the horseshoe prior can be expressed
+
+\[
 \begin{aligned}
@@
 \end{aligned}
-$$
-
-for $k = 1, 2, ..., K$. The horseshoe prior is so-named because for fixed values $\tau = \lambda_k = 1$, the prior resemebles a Beta(1/2, 1/2) distribution, or a horseshoe.
-
-The horseshoe is often favored because it is a global-local shrinkage prior.
-This means that $\tau$ shrinks *all* parameters towards 0 while the local parameter $\lambda_k$ and the heavy Cauchy tails allow larger coefficients to remain unshrunk.
-While this is precisely the goal of a sparsity inducing prior, the horseshoe prior can fail to regularize large coefficients *at all*, which can be a problem when parameters are weakly identified by data.
-With no regularization applied to the largest coefficients, there is a risk of overfitting.
-Additionally, there is no consensus in the literature for assigning priors to $tau$.
-Piironen and Vehtari (2017) introduced the *regularized horseshoe* to address both of these shortcomings of the horseshoe.
-
-### Regularized Horseshoe - needs graphics
+\]
+
+for \(k = 1, 2, ..., K\). The horseshoe prior is so-named because for
+fixed values \(\tau = \lambda_k = 1\), the prior resemebles a Beta(1/2,
+1/2) distribution, or a horseshoe.
+
+The horseshoe is often favored because it is a global-local shrinkage
+prior. This means that \(\tau\) shrinks \emph{all} parameters towards 0
+while the local parameter \(\lambda_k\) and the heavy Cauchy tails allow
+larger coefficients to remain unshrunk. While this is precisely the goal
+of a sparsity inducing prior, the horseshoe prior can fail to regularize
+large coefficients \emph{at all}, which can be a problem when parameters
+are weakly identified by data. With no regularization applied to the
+largest coefficients, there is a risk of overfitting. Additionally,
+there is no consensus in the literature for assigning priors to \(tau\).
+Piironen and Vehtari (2017) introduced the \emph{regularized horseshoe}
+to address both of these shortcomings of the horseshoe.
+
+\subsubsection{Regularized Horseshoe - needs
+graphics}\label{regularized-horseshoe---needs-graphics}
 
@@
 
-$$
+\[
 \begin{aligned}
@@
 \end{aligned}
-$$
-
-where $c$ > 0 helps to regularize $\beta_k$ far from zero.
-When $c$ approaches infinity, the regularized horseshoe becomes the standard horseshoe.
-Additionally, $c$ in the regularized horseshoe corresponds to the slab width in the spike-and-slab prior.
-
-As with the spike-and-slab, the selection of priors is a challenge with the regularized horseshoe.
-Piironen and Vehtari (2017) provide a set of recommendations for setting priors which we follow in this study.
-In practice, we tune the priors for $\lambda_k$, $c$, and $\tau$ using simulated data with features similar to our observed data (ie, parameters with coefficients of similar magnitude).
-
-### Model 3
-
-For this study, we implement the regularize horseshoe using code modified from Piironen & Vehtari.
+\]
+
+where \(c\) \textgreater{} 0 helps to regularize \(\beta_k\) far from
+zero. When \(c\) approaches infinity, the regularized horseshoe becomes
+the standard horseshoe. Additionally, \(c\) in the regularized horseshoe
+corresponds to the slab width in the spike-and-slab prior.
+
+As with the spike-and-slab, the selection of priors is a challenge with
+the regularized horseshoe. Piironen and Vehtari (2017) provide a set of
+recommendations for setting priors which we follow in this study. In
+practice, we tune the priors for \(\lambda_k\), \(c\), and \(\tau\)
+using simulated data with features similar to our observed data (ie,
+parameters with coefficients of similar magnitude).
+
+\subsubsection{Model 3}\label{model-3}
+
+For this study, we implement the regularize horseshoe using code
+modified from Piironen \& Vehtari.
 
 Notably, Model 3 experienced one divergent transition after warmup.
-However, the model was fitted several times with no divergent transitions.
-A "false alarm" divergent transition can occur when the cutoff for determining a divergent transition is too low; this can be true in the case of particularly complex posteriors.
-Because only 1 out of 20,000 iterations produced a divergent transition, we can proceed with caution using Model 3.
-
-## Adding Spatial Structure
-
-When modeling data with a spatial component, we generally expect adjacent areas to be more similar than areas which are far apart.
-While this is a straightforward assumption, it can improve model fits in several ways.
-First, it can encode prior information about response $Y$ not otherwise represented by covariates $x$ (ie, that nearby observations are more similar than far-flung ones).
-Second, it can provide geographic smoothing when observations are sparse or noisy.
-This is often the case in small areas or when observing events which can only occur at specific locations, such as air quality measured by sensors.
-In the case of the Metro Transit ridership data, this smoothing can help avoid overfitting areas such as the outliers in downtown Minneapolis and better represent block groups with no existing bus stops.
-
-In the case of Metro Transit's ridership data, we expect encoding spatial structure to improve the model fit in several ways.
-First, we know that assigning transit ridership to census geography is inherently flawed.
-This is because census geographies, including block groups, are generally bounded by major roads (in addition to natural features such as lakes and political boundaries such as city limits).
-Naturally, bus stops tend to exist along those same major roadways.
-
-This bounding convention means that boardings are often assigned to different block groups depending on which direction the passenger is travelling.
-For example, consider the A-Line running along Snelling Avenue.
-Boarding at Snelling and Grand *northbound* puts you in block group ____ while boarding at the same intersection but *southbound* puts you in block group ____.
-Integrating spatial structure will help to smooth ridership across block group bounds.
-**will add a map here**
-
-Additionally, we know intuitively that transit riders don't necessarily board in the block groups they live in.
-Particularly in dense areas, it's very likely for riders to board in a block group other than the block group they reside in.
-For example, I live in block group ____ but walk about 10 minutes to board the A-Line southbound at Snelling and Randolph.
-Commuting patterns such as these can assign ridership to different block groups than we might expect.
-Geographic smoothing will allow neighboring block groups to share information, better representing these ridership patterns.
-
-### Conditional Autoregressive Priors
-
-Conditional Autoregressive (CAR) priors are one of the most widespread Bfayesian methods for modeling spatial autocorrelation.
-CAR models were introduced in Besag 1974 and remain perhaps the primary method for Bayesian areal data modeling.
-Note that CAR models are recommended for use as *priors* rather than as an observation model.
-
-Areal data corresponds to finitely many discrete areal units, such as counties or census tracts.
-CAR priors are often used in the context of small-area count data.
-This sort of response variable tends to be noisy, particularly when the counted event is rare, the population of areal unit $i$ is small, or the physical boundaries of areal units present challenges.
-CAR priors were designed in part to smooth such noisy counts.
-As such, they are often applied in epidemiological disease risk modeling in which disease occurences may be rare.
-
-Generally, CAR models rely on a binary neighborhood structure.
-Areal units $i$ and $j$ are considered neighbors if they share a boundary.
-For strictly rectangular lattice data, we often choose either a "Queens" or "Rooks" neighborood structure.
-**insert graphics** For irregularly shaped areal data, such as our 1,495 census block groups, we generally default to a queen neighborhood structure wherein areal units sharing *any* points of contact are deemed neighbors.
-
-For $n$ regions, the neighorhood relationships are encoded in an $n \times n$ neighborhood matrix $W$.
-Matrix $W$ is defined such that $w_{ij} = 1$ if if regions $i$ and $j$ are neighbors (denoted $i \sim j$) and 0 otherwise.
-This matrix is symmetric.
-Note that regions are not considered neighbors of themselves, ie the diagonal entries $w_{ii} = 0$.
-
-The spatial interactions are modeled by a random variable $\phi$.
-Here, $\phi$ is a vector $\phi = \phi_1, \phi_2, ..., \phi_n$.
-The distribution of each $\phi$ is determined by the sum of its neighbors' values such that
-
-$$
+However, the model was fitted several times with no divergent
+transitions. A ``false alarm'' divergent transition can occur when the
+cutoff for determining a divergent transition is too low; this can be
+true in the case of particularly complex posteriors. Because only 1 out
+of 20,000 iterations produced a divergent transition, we can proceed
+with caution using Model 3.
+
+\subsection{Adding Spatial Structure}\label{adding-spatial-structure}
+
+When modeling data with a spatial component, we generally expect
+adjacent areas to be more similar than areas which are far apart. While
+this is a straightforward assumption, it can improve model fits in
+several ways. First, it can encode prior information about response
+\(Y\) not otherwise represented by covariates \(x\) (ie, that nearby
+observations are more similar than far-flung ones). Second, it can
+provide geographic smoothing when observations are sparse or noisy. This
+is often the case in small areas or when observing events which can only
+occur at specific locations, such as air quality measured by sensors. In
+the case of the Metro Transit ridership data, this smoothing can help
+avoid overfitting areas such as the outliers in downtown Minneapolis and
+better represent block groups with no existing bus stops.
+
+In the case of Metro Transit's ridership data, we expect encoding
+spatial structure to improve the model fit in several ways. First, we
+know that assigning transit ridership to census geography is inherently
+flawed. This is because census geographies, including block groups, are
+generally bounded by major roads (in addition to natural features such
+as lakes and political boundaries such as city limits). Naturally, bus
+stops tend to exist along those same major roadways.
+
+This bounding convention means that boardings are often assigned to
+different block groups depending on which direction the passenger is
+travelling. For example, consider the A-Line running along Snelling
+Avenue. Boarding at Snelling and Grand \emph{northbound} puts you in
+block group \_\_\_\_ while boarding at the same intersection but
+\emph{southbound} puts you in block group \_\_\_\_. Integrating spatial
+structure will help to smooth ridership across block group bounds.
+\textbf{will add a map here}
+
+Additionally, we know intuitively that transit riders don't necessarily
+board in the block groups they live in. Particularly in dense areas,
+it's very likely for riders to board in a block group other than the
+block group they reside in. For example, I live in block group \_\_\_\_
+but walk about 10 minutes to board the A-Line southbound at Snelling and
+Randolph. Commuting patterns such as these can assign ridership to
+different block groups than we might expect. Geographic smoothing will
+allow neighboring block groups to share information, better representing
+these ridership patterns.
+
+\subsubsection{Conditional Autoregressive
+Priors}\label{conditional-autoregressive-priors}
+
+Conditional Autoregressive (CAR) priors are one of the most widespread
+Bfayesian methods for modeling spatial autocorrelation. CAR models were
+introduced in Besag 1974 and remain perhaps the primary method for
+Bayesian areal data modeling. Note that CAR models are recommended for
+use as \emph{priors} rather than as an observation model.
+
+Areal data corresponds to finitely many discrete areal units, such as
+counties or census tracts. CAR priors are often used in the context of
+small-area count data. This sort of response variable tends to be noisy,
+particularly when the counted event is rare, the population of areal
+unit \(i\) is small, or the physical boundaries of areal units present
+challenges. CAR priors were designed in part to smooth such noisy
+counts. As such, they are often applied in epidemiological disease risk
+modeling in which disease occurences may be rare.
+
+Generally, CAR models rely on a binary neighborhood structure. Areal
+units \(i\) and \(j\) are considered neighbors if they share a boundary.
+For strictly rectangular lattice data, we often choose either a
+``Queens'' or ``Rooks'' neighborood structure. \textbf{insert graphics}
+For irregularly shaped areal data, such as our 1,495 census block
+groups, we generally default to a queen neighborhood structure wherein
+areal units sharing \emph{any} points of contact are deemed neighbors.
+
+For \(n\) regions, the neighorhood relationships are encoded in an
+\(n \times n\) neighborhood matrix \(W\). Matrix \(W\) is defined such
+that \(w_{ij} = 1\) if if regions \(i\) and \(j\) are neighbors (denoted
+\(i \sim j\)) and 0 otherwise. This matrix is symmetric. Note that
+regions are not considered neighbors of themselves, ie the diagonal
+entries \(w_{ii} = 0\).
+
+The spatial interactions are modeled by a random variable \(\phi\).
+Here, \(\phi\) is a vector \(\phi = \phi_1, \phi_2, ..., \phi_n\). The
+distribution of each \(\phi\) is determined by the sum of its neighbors'
+values such that
+
+\[
 \phi_i \text{ | } \phi_j, j \neq i \sim \text{Normal}(\sum_{i=1}^{N}w_{ij}\phi_j, \sigma^2) 
-$$
-
-The conditional distribution of each $\phi_i$ is helpful in building intuition about the functionality of the CAR prior.
-Basing the value of each $\phi_i$ off of its neighbors is how the CAR prior performs spatial smoothing.
-
-Besag (1974) proved that the joint distribution of $\phi$ is a multivariate normal centered at 0 where its variance is given by a symmetric positive definite precision matrix $Q$.
-Matrix $Q$ is simply the inverse of the covariance matrix $\sum$.
-This finding simplifies the above to
-
-$$
+\]
+
+The conditional distribution of each \(\phi_i\) is helpful in building
+intuition about the functionality of the CAR prior. Basing the value of
+each \(\phi_i\) off of its neighbors is how the CAR prior performs
+spatial smoothing.
+
+Besag (1974) proved that the joint distribution of \(\phi\) is a
+multivariate normal centered at 0 where its variance is given by a
+symmetric positive definite precision matrix \(Q\). Matrix \(Q\) is
+simply the inverse of the covariance matrix \(\sum\). This finding
+simplifies the above to
+
+\[
 \phi \sim \text{Normal}(0, Q^{-1})
-$$
-
-Precision matrix $Q$ can be defined for multivariate normal $\phi$ using two other $n \times n$ matrices: the neighborhood matrix $W$ and the diagonal matrix $D$ in which $d_{ii}$ indicates the number of neighbors region $i$ has.
-All off-diagonal entries are 0.
-Using these matrices, $Q$ is defined
-
-$$
+\]
+
+Precision matrix \(Q\) can be defined for multivariate normal \(\phi\)
+using two other \(n \times n\) matrices: the neighborhood matrix \(W\)
+and the diagonal matrix \(D\) in which \(d_{ii}\) indicates the number
+of neighbors region \(i\) has. All off-diagonal entries are 0. Using
+these matrices, \(Q\) is defined
+
+\[
 Q = D(I - \alpha W)
-$$
-
-where $I$ is the identity matrix and $\alpha \in (0, 1)$ determines the amount of spatial autocorrelation present in $Y$.
-Parameter $\alpha$ is a key component of CAR models.
-At $\alpha = 0$, the model assumes spatial independence and at $\alpha = 1$, perfect spatial autocorrelation.
-
-Following Morris et al, the log probability density of $\phi$ is proportional to
-
-$$
+\]
+
+where \(I\) is the identity matrix and \(\alpha \in (0, 1)\) determines
+the amount of spatial autocorrelation present in \(Y\). Parameter
+\(\alpha\) is a key component of CAR models. At \(\alpha = 0\), the
+model assumes spatial independence and at \(\alpha = 1\), perfect
+spatial autocorrelation.
+
+Following Morris et al, the log probability density of \(\phi\) is
+proportional to
+
+\[
 \frac{M}{2}\text{log(det}(Q)) - \frac{1}{2}\phi^TQ\phi
-$$
-
-Calculating this value is computationally expensive.
-For models with 1,000 areal units, for example, calculating the determinant requires 1 billion operations (Morris et al, 2019).
-Given that the Metro Transit ridership data corresponds to 1,495 areal units, a strict CAR prior may be too computationally inefficient for our models.
-
-### Intrinsic Conditional Autoregressive
-
-The Intrinsic Conditional Autoregressive (ICAR) prior is a slight simplifation of a CAR prior. ICAR priors set $\alpha = 1$ which simplifies the definition of Q
-
-$$
+\]
+
+Calculating this value is computationally expensive. For models with
+1,000 areal units, for example, calculating the determinant requires 1
+billion operations (Morris et al, 2019). Given that the Metro Transit
+ridership data corresponds to 1,495 areal units, a strict CAR prior may
+be too computationally inefficient for our models.
+
+\subsubsection{Intrinsic Conditional
+Autoregressive}\label{intrinsic-conditional-autoregressive}
+
+The Intrinsic Conditional Autoregressive (ICAR) prior is a slight
+simplifation of a CAR prior. ICAR priors set \(\alpha = 1\) which
+simplifies the definition of Q
+
+\[
 Q = D-A
-$$
-
-thereby setting $det(Q) = 0$. As a result, the first term in **equation ___** can be simplified to
-
-$$
+\]
+
+thereby setting \(det(Q) = 0\). As a result, the first term in
+**equation \_\_\_** can be simplified to
+
+\[
 -\frac{1}{2}\phi^TQ\phi
-$$
-
-This reduces the computational expense of computing CAR models significantly. Notably, the ICAR prior is improper but yields proper posteriors (Morris et al, 2019).
-
-The ICAR prior specifies each $\phi_i$ to be normally distributed with its mean equal to the mean of its neighbor's values.
-This is how the CAR/ICAR model "borrows strength" from geographic neighbors to smooth noisy estimates.
-Intuitively, the variance of each $\phi_i$ decreases as its number of neighbors $d_i$ increases.
-The conditional distribution for each $\phi_i$ can be written
-
-$$
+\]
+
+This reduces the computational expense of computing CAR models
+significantly. Notably, the ICAR prior is improper but yields proper
+posteriors (Morris et al, 2019).
+
+The ICAR prior specifies each \(\phi_i\) to be normally distributed with
+its mean equal to the mean of its neighbor's values. This is how the
+CAR/ICAR model ``borrows strength'' from geographic neighbors to smooth
+noisy estimates. Intuitively, the variance of each \(\phi_i\) decreases
+as its number of neighbors \(d_i\) increases. The conditional
+distribution for each \(\phi_i\) can be written
+
+\[
 \text{p}(\phi_i \text{ | } \phi_{i \sim j}) = \text{Normal}(\frac{\sum_{i \sim j}\phi_i}{d_i}, \frac{\sigma_i^2}{d_i})
-$$
-
-Here, the variance $\sigma$ is unknown.
-The conditional specification of each $\phi_i$ is helpful in building intuition about the assumptions the ICAR component makes and how it performs its "smoothing".
-The ICAR prior matches our intuition about spatial autocorrelation: areal unit $i$ is similar to the areal units surrounding it, and the more areal units surround it, the more confident we feel in that similarity.
-The joint distribution can be rewritten
-
-$$
+\]
+
+Here, the variance \(\sigma\) is unknown. The conditional specification
+of each \(\phi_i\) is helpful in building intuition about the
+assumptions the ICAR component makes and how it performs its
+``smoothing''. The ICAR prior matches our intuition about spatial
+autocorrelation: areal unit \(i\) is similar to the areal units
+surrounding it, and the more areal units surround it, the more confident
+we feel in that similarity. The joint distribution can be rewritten
+
+\[
 \text{p}(\phi) \propto \text{exp}(-\frac{1}{2}\sum_{i \sim j}(\phi_i - \phi_j)^2)
-$$
-
-This pairwise difference specification is more computationally efficient to fit in Stan (Morris et al, 2019).
-
-### Besag-York-Mollie (BYM)
-
-The specific ICAR model used in this analysis is the Besag-York-Mollie (BYM) Poisson model.
-The BYM is a "classical" spatial Bayesian method (Riebler et al, 2016).
-It is generally used in disease mapping studies to model rare events (ie, disease occurrence) in small areal units.
-The BYM capitalizes on the ICAR prior's ability to smooth noisy estimates and share information across geographic units.
-
-The latent function in the BYM model contains heterogeneous random effects as well as a spatial ICAR term in order to account for both spatial and non-spatial heterogeneity.
-Note that in Section ___, we introduced a random effect term $\theta$.
-Therefore, the BYM model can be though of as decomposing $\theta$ into spatial and non-spatial components.
-
-Specifically, the BYM model replaces $\eta_i$ from equation ___ with
-
-$$
+\]
+
+This pairwise difference specification is more computationally efficient
+to fit in Stan (Morris et al, 2019).
+
+\subsubsection{Besag-York-Mollie (BYM)}\label{besag-york-mollie-bym}
+
+The specific ICAR model used in this analysis is the Besag-York-Mollie
+(BYM) Poisson model. The BYM is a ``classical'' spatial Bayesian method
+(Riebler et al, 2016). It is generally used in disease mapping studies
+to model rare events (ie, disease occurrence) in small areal units. The
+BYM capitalizes on the ICAR prior's ability to smooth noisy estimates
+and share information across geographic units.
+
+The latent function in the BYM model contains heterogeneous random
+effects as well as a spatial ICAR term in order to account for both
+spatial and non-spatial heterogeneity. Note that in Section \_\_\_, we
+introduced a random effect term \(\theta\). Therefore, the BYM model can
+be though of as decomposing \(\theta\) into spatial and non-spatial
+components.
+
+Specifically, the BYM model replaces \(\eta_i\) from equation \_\_\_
+with
+
+\[
 \eta_i = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + \theta + \phi
-$$
-
-where $\phi$ is the addition: an ICAR component.
-Decomposing overdispersion as such is helpful in furthering our understanding of response $Y$.
-When we use a simple random effect $\theta$ alone, we can accomplish a good model fit.
-However, $\theta$ provides no additional information about $Y$.
-The decomposition of $\theta$ in the BYM allows us to better quantify how much variance is white noise (ordinary random effects) versus some unmeasured confounding correlated across space.
-
-The BYM model as written above is appealing in its simplicity.
-However, the lack of informative hyperpriors specified for $\theta$ and $\phi$ can make the model incredibly challenging to fit.
-In theory, extra-Poisson variance could be explained 100% by $\theta$, 100% by $\phi$, or anywhere in between.
-In *practice*, we likely have limited information about where that ratio falls.
-Riebler et al (2016) further explain the sampling issues faced by the BYM model faced with no hyperpriors.
-In short, the sampler is forced to explore all possible combinations of $\theta$ and $\phi$, no matter how unlikely a particular combination my be.
-In the context of fitting models with Stan, this is liekly to cause uneccesarily long computation times and perhaps to yield biased posterior estimates.
-
-There are some suggestions in the literature for setting hyperpriors for $\theta$ and $\phi$ (Besag and Mollie, 1991; Clayton and Montomoli, 1995).
-The existing methods, however, tend to rely on the specific data at hand which makes the selection of hyperparameters unnecessarily time consuming.
-
-### BYM2
-
-The BYM2 model proposed by Simpson et al and further described in Riebler et al (2016) aims to solve these sampling problems.
-The BYM2 takes a more "fully Bayesian" approach to hyperprior/hyperparameter selection.
-The primary difference between the BYM and BYM2 models is the addition of a mixing parameter, $\rho$.
-The BYM2 model rewrites the BYM as
-
-$$
+\]
+
+where \(\phi\) is the addition: an ICAR component. Decomposing
+overdispersion as such is helpful in furthering our understanding of
+response \(Y\). When we use a simple random effect \(\theta\) alone, we
+can accomplish a good model fit. However, \(\theta\) provides no
+additional information about \(Y\). The decomposition of \(\theta\) in
+the BYM allows us to better quantify how much variance is white noise
+(ordinary random effects) versus some unmeasured confounding correlated
+across space.
+
+The BYM model as written above is appealing in its simplicity. However,
+the lack of informative hyperpriors specified for \(\theta\) and
+\(\phi\) can make the model incredibly challenging to fit. In theory,
+extra-Poisson variance could be explained 100\% by \(\theta\), 100\% by
+\(\phi\), or anywhere in between. In \emph{practice}, we likely have
+limited information about where that ratio falls. Riebler et al (2016)
+further explain the sampling issues faced by the BYM model faced with no
+hyperpriors. In short, the sampler is forced to explore all possible
+combinations of \(\theta\) and \(\phi\), no matter how unlikely a
+particular combination my be. In the context of fitting models with
+Stan, this is liekly to cause uneccesarily long computation times and
+perhaps to yield biased posterior estimates.
+
+There are some suggestions in the literature for setting hyperpriors for
+\(\theta\) and \(\phi\) (Besag and Mollie, 1991; Clayton and Montomoli,
+1995). The existing methods, however, tend to rely on the specific data
+at hand which makes the selection of hyperparameters unnecessarily time
+consuming.
+
+\subsubsection{BYM2}\label{bym2}
+
+The BYM2 model proposed by Simpson et al and further described in
+Riebler et al (2016) aims to solve these sampling problems. The BYM2
+takes a more ``fully Bayesian'' approach to hyperprior/hyperparameter
+selection. The primary difference between the BYM and BYM2 models is the
+addition of a mixing parameter, \(\rho\). The BYM2 model rewrites the
+BYM as
+
+\[
 \eta_i = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + ((\sqrt{\frac{\rho}{s}})\theta^{*} + ((\sqrt{1-\rho})\phi^{*})\sigma
-$$
-
-where $\rho \in (0, 1)$ determines how much variance/overdispersion is caused by spatial versus non-spatial error terms.
-Like the popular Leroux CAR prior, proposed by Leroux et al (2009), the BYM2 scales both $\theta$ and $\phi$ by $\sigma$, the standard deviation of the combined error terms (Morris et al, 2019).
-In this parameterization, $s$ is a scaling factor such that Var($\theta_i$) $\approx$ Var($\phi_i$) $\approx$ 1.
-The equal unit variance is necessary for $\sigma$ to truly be the standard deviation of the error terms.
-
-Setting priors is somewhat more straightforward for the BYM2.
-The ICAR component, $\phi^{*}$ remains unchanged.
-Riebler and Morris recommend the prior $\theta \sim$ Normal($0, n)$ where $n$ is the number of connected graphs in the neighborhood graph.
-In many cases, such as modeling data for all counties in a state, $n=1$.
-When $n>1$, the variance is different in each subgraph which affects $\sigma$ and $s$.
-It is possible to fit the model in this case, although computation time is much longer.
-
-A relatively vague Normal(0, 1) prior is appropriate for $\sigma$.
-Riebler et al (2016) also propose either a Beta(1/2, 1/2) or more specialized, complex prior for mixing parameter $\rho$.
-For this study, we use the simpler Beta(1/2, 1/2) prior.
-
-### Model 4
-
-Model 4 uses all 18 covariates of interest, no horseshoe priors, and decomposes $\theta\sigma$ into a set of random effects plus a spatial ICAR component.
-This model converges much more efficiently than Model 2 or Model 3 due to careful reparameterization of the BYM by Riebler et al (2016).
-Future would could potentially improve the speed of fitting Models 2 and 3.
-
-Model 4 was run on four chains for 10,000 iterations each. The parameter estimates for Model 4 are below. Note that `n_eff` is lower than for Model 1 but significantly higher than Model 2 or 3.
+\]
+
+where \(\rho \in (0, 1)\) determines how much variance/overdispersion is
+caused by spatial versus non-spatial error terms. Like the popular
+Leroux CAR prior, proposed by Leroux et al (2009), the BYM2 scales both
+\(\theta\) and \(\phi\) by \(\sigma\), the standard deviation of the
+combined error terms (Morris et al, 2019). In this parameterization,
+\(s\) is a scaling factor such that Var(\(\theta_i\)) \(\approx\)
+Var(\(\phi_i\)) \(\approx\) 1. The equal unit variance is necessary for
+\(\sigma\) to truly be the standard deviation of the error terms.
+
+Setting priors is somewhat more straightforward for the BYM2. The ICAR
+component, \(\phi^{*}\) remains unchanged. Riebler and Morris recommend
+the prior \(\theta \sim\) Normal(\(0, n)\) where \(n\) is the number of
+connected graphs in the neighborhood graph. In many cases, such as
+modeling data for all counties in a state, \(n=1\). When \(n>1\), the
+variance is different in each subgraph which affects \(\sigma\) and
+\(s\). It is possible to fit the model in this case, although
+computation time is much longer.
+
+A relatively vague Normal(0, 1) prior is appropriate for \(\sigma\).
+Riebler et al (2016) also propose either a Beta(1/2, 1/2) or more
+specialized, complex prior for mixing parameter \(\rho\). For this
+study, we use the simpler Beta(1/2, 1/2) prior.
+
+\subsubsection{Model 4}\label{model-4}
+
+Model 4 uses all 18 covariates of interest, no horseshoe priors, and
+decomposes \(\theta\sigma\) into a set of random effects plus a spatial
+ICAR component. This model converges much more efficiently than Model 2
+or Model 3 due to careful reparameterization of the BYM by Riebler et al
+(2016). Future would could potentially improve the speed of fitting
+Models 2 and 3.
+
+Model 4 was run on four chains for 10,000 iterations each. The parameter
+estimates for Model 4 are below. Note that \texttt{n\_eff} is lower than
+for Model 1 but significantly higher than Model 2 or 3.
 
@@
 
-Like Models 2 and 3, the BYM2 model performs well in the posterior predictive check (Figure __).
+Like Models 2 and 3, the BYM2 model performs well in the posterior
+predictive check (Figure \_\_).
 
@@
 
-### Model 5
-
-Model 5 builds upon Model 4 by incorporating regularized horsehoe priors in addition to spatial ICAR priors.
-This model has high autocorrelation between iterations and took *significantly* longer than any of the other models to fit in Stan (upwards of 12 hours).
-That said, the `n_eff` and $\hat{R}$ diagnostics indicate convergence and the parameter estimates (Table __) match our expectations in terms of sign and magnitude based on the previous 4 models.
+\subsubsection{Model 5}\label{model-5}
+
+Model 5 builds upon Model 4 by incorporating regularized horsehoe priors
+in addition to spatial ICAR priors. This model has high autocorrelation
+between iterations and took \emph{significantly} longer than any of the
+other models to fit in Stan (upwards of 12 hours). That said, the
+\texttt{n\_eff} and \(\hat{R}\) diagnostics indicate convergence and the
+parameter estimates (Table \_\_) match our expectations in terms of sign
+and magnitude based on the previous 4 models.
 
@@
 
-# Discussion
-
-In this study, we fit several increasingly complex models building upon the baseline Poisson regression, Model 1.
-Model 2 incorporates a simple set of random effects to account for overdispersion.
-Model 3 incorporates regularized horseshoe priors for variable selection.
-Model 4 includes the spatial ICAR priors but no horseshoe priors, and Model 5 includes both regularized horseshoe priors and ICAR priors.
-
-All five models were run for at least 10,000 iterations and "pass" the basic diagnostic checks: $\hat{R} = 1$, sufficiently high `n_eff`, and no concerning `max_treedepth` or divergence warnings.
-
-Naturally, the baseline Model 1 converges most quickly and has the highest `n_eff` diagnostics.
-The simplicity of the Model, however, yields a poor model fit.
-Perhaps unexpectedly, Model 4, the BYM2, is the second-most efficient in terms of convergence and Effective Sample Size.
-
-- bymhrs is probably not "worth it", needs to be reparameterized/coded by someone better than me
-- incorporating overdispersion is the most important improvement in terms of model fit
-
-# Appendix: Stan Code
-
-## Model 1
-
-The Stan program for the baseline model is below.
-This model could also have been fit using Stan's more efficient built-in Poisson GLM function.
-For this study, we write the GLM more "longhand" for ease of comparison between models.
-```{r}
-writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/poisson.stan"))
-```
-
-## Model 2
-
-The Stan program for Model 2 differs from Model 1 only in the latent variable, `f`. 
-```{r}
-writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/poisson_theta.stan"))
-```
-
-## Model 3
-
-The regularized horseshoe prior used to fit Model 3 is adopted from Piironen and Vehtari. 
-```{r}
-writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/poisson_horseshoe.stan"))
-```
-
-## Model 4
+\section{Discussion}\label{discussion}
+
+In this study, we fit several increasingly complex models building upon
+the baseline Poisson regression, Model 1. Model 2 incorporates a simple
+set of random effects to account for overdispersion. Model 3
+incorporates regularized horseshoe priors for variable selection. Model
+4 includes the spatial ICAR priors but no horseshoe priors, and Model 5
+includes both regularized horseshoe priors and ICAR priors.
+
+All five models were run for at least 10,000 iterations and ``pass'' the
+basic diagnostic checks: \(\hat{R} = 1\), sufficiently high
+\texttt{n\_eff}, and no concerning \texttt{max\_treedepth} or divergence
+warnings.
+
+Naturally, the baseline Model 1 converges most quickly and has the
+highest \texttt{n\_eff} diagnostics. The simplicity of the Model,
+however, yields a poor model fit. Perhaps unexpectedly, Model 4, the
+BYM2, is the second-most efficient in terms of convergence and Effective
+Sample Size.
+
+\begin{itemize}
+\tightlist
+\item
+  bymhrs is probably not ``worth it'', needs to be reparameterized/coded
+  by someone better than me
+\item
+  incorporating overdispersion is the most important improvement in
+  terms of model fit
+\end{itemize}
+
+\section{Appendix: Stan Code}\label{appendix-stan-code}
+
+\subsection{Model 1}\label{model-1-1}
+
+The Stan program for the baseline model is below. This model could also
+have been fit using Stan's more efficient built-in Poisson GLM function.
+For this study, we write the GLM more ``longhand'' for ease of
+comparison between models.
+
+\begin{verbatim}
+## data {
+##   // number obs
+##   int<lower=0> N;
+##   // response
+##   int<lower=0> y[N];       
+##   // "offset" (number of stops)
+##   vector<lower=0>[N] E;     
+##   // number of covariates
+##   int<lower=1> K;
+##   // covariates
+##   matrix[N, K] x;
+## }
+## transformed data {
+##   vector[N] log_E = log(E);
+## }
+## parameters {
+##   // intercept
+##   real beta_0;      
+##   // covariates
+##   vector[K] beta;          
+## }
+## transformed parameters{
+##   // latent function variables
+##   vector[N] f;
+##   f = log_E + beta_0 + x*beta;
+## }
+## model {
+##   /// model 
+##   y ~ poisson_log(f); 
+##   // prior on betas
+##   beta_0 ~ normal(0.0, 3);
+##   beta ~ normal(0.0, 1);
+## }
+## generated quantities {
+##   vector[N] log_lik; 
+##     for(i in 1:N) log_lik[i] = poisson_log_lpmf(y[i] | log_E[i] + beta_0 + x[i,]*beta);
+## }
+\end{verbatim}
+
+\subsection{Model 2}\label{model-2-1}
+
+The Stan program for Model 2 differs from Model 1 only in the latent
+variable, \texttt{f}.
+
+\begin{verbatim}
+## data {
+##   // number obs
+##   int<lower=0> N;
+##   // response
+##   int<lower=0> y[N];       
+##   // "offset" (number of stops)
+##   vector<lower=0>[N] E;     
+##   // number of covariates
+##   int<lower=1> K;
+##   // covariates
+##   matrix[N, K] x;
+## }
+## transformed data {
+##   vector[N] log_E = log(E);
+## }
+## parameters {
+##   // intercept
+##   real beta_0;      
+##   // covariates
+##   vector[K] beta;   
+##   // overdispersion
+##   vector[N] theta;
+##   // random effects variance
+##   real<lower=0> sigma;
+## }
+## transformed parameters{
+##   // latent function variables
+##   vector[N] re; 
+##   vector[N] f;
+##   re = theta * sigma; 
+##   f = log_E + beta_0 + x*beta + re;
+## }
+## model {
+##   /// model 
+##   y ~ poisson_log(f); 
+##   // prior on betas
+##   beta_0 ~ normal(0.0, 1);
+##   beta ~ normal(0.0, 0.2);
+##   // overdispersion
+##   theta ~ normal(0, 1);
+##   sigma ~ normal(0, 5);
+## }
+## generated quantities {
+##   vector[N] log_lik; 
+##     for(i in 1:N) log_lik[i] = poisson_log_lpmf(y[i] | log_E[i] + beta_0 + x[i,]*beta + re[i]);
+## }
+\end{verbatim}
+
+\subsection{Model 3}\label{model-3-1}
+
+The regularized horseshoe prior used to fit Model 3 is adopted from
+Piironen and Vehtari.
+
+\begin{verbatim}
+## data {
+##   // number obs
+##   int<lower=0> N;
+##   // response
+##   int<lower=0> y[N];       
+##   // "offset" (number of stops)
+##   vector<lower=0>[N] E;     
+##   // number of covariates
+##   int<lower=1> K;
+##   // covariates
+##   matrix[N, K] x;
+##   // prior sd for intercept
+##   real<lower=0> scale_icept;
+##   // scale for half t for tau
+##   real<lower=0> scale_global;
+##   // df for tau, lambda
+##   real<lower=1> nu_global;
+##   real<lower=1> nu_local;
+##   // slab scale, df for reg horseshoe
+##   real<lower=0> slab_scale;
+##   real<lower=0> slab_df;
+## }
+## transformed data {
+##   vector[N] log_E = log(E);
+## }
+## parameters {
+##   // intercept
+##   real beta_0;
+##   // parameterization from vehtari
+##   vector[K] z;
+##   real<lower=0> aux1_global;
+##   real<lower=0> aux2_global;
+##   vector<lower=0>[K] aux1_local;
+##   vector<lower=0>[K] aux2_local;
+##   real<lower=0> caux;
+## }
+## transformed parameters {
+##   // global shrinkage
+##   real<lower=0> tau;
+##   // local shrinkage
+##   vector<lower=0>[K] lambda;
+##   vector<lower=0>[K] lambda_tilde; 
+##   // slab scale
+##   real<lower=0> c;
+##   // regression coefficients
+##   vector[K] beta;
+##   // latent function variables
+##   vector[N] f;
+##   
+##   lambda = aux1_local .* sqrt(aux2_local);
+##   tau = aux1_global * sqrt(aux2_global) * scale_global;
+##   c = slab_scale * sqrt(caux); 
+##   lambda_tilde = sqrt(c^2 * square(lambda) ./ (c^2 + tau^2 * square(lambda)));
+##   beta = z .* lambda_tilde*tau;
+##   f = log_E + beta_0 + x*beta;
+## }
+## model {
+##   // half t and inverse gamma priors
+##   z ~ normal(0, 1);
+##   aux1_local ~ normal(0, 1);
+##   aux2_local ~ inv_gamma(0.5*nu_local, 0.5*nu_local);
+##   aux1_global ~ normal(0, 1);
+##   aux2_global ~ inv_gamma(0.5*nu_global, 0.5*nu_global);
+##   caux ~ inv_gamma(0.5*slab_df, 0.5*slab_df);
+##   beta_0 ~ normal(0, scale_icept);
+##   // and the model 
+##   y ~ poisson_log(f);
+## }
+## generated quantities {
+##   vector[N] log_lik; 
+##     for(i in 1:N) log_lik[i] = poisson_log_lpmf(y[i] | log_E[i] + beta_0 + x[i,]*beta);
+## }
+\end{verbatim}
+
+\subsection{Model 4}\label{model-4-1}
 
@@
 
-```{r}
-writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/bym2.stan"))
-```
-
-## Model 5
-
-Finally, the code to fit Model 5 quite literally combines the programs for Models 3 and 4. 
-```{r}
-writeLines(readLines("~/Documents/honors/honors-project/stan/final-stan/bym_horseshoe.stan"))
-
-```
+\begin{verbatim}
+## data {
+##   int<lower=0> N;
+##   int<lower=0> N_edges;
+##   int<lower=1, upper=N> node1[N_edges];  // node1[i] adjacent to node2[i]
+##   int<lower=1, upper=N> node2[N_edges];  // and node1[i] < node2[i]
+## 
+##   int<lower=0> y[N];              // count outcomes
+##   vector<lower=0>[N] E;           // exposure
+##   int<lower=1> K;                 // num covariates
+##   matrix[N, K] x;                 // design matrix
+## 
+##   real<lower=0> scaling_factor; // scales the variance of the spatial effects
+## }
+## transformed data {
+##   vector[N] log_E = log(E);
+## }
+## parameters {
+##   real beta0;            // intercept
+##   vector[K] beta;       // covariates
+## 
+##   real<lower=0> sigma;        // overall standard deviation
+##   real<lower=0, upper=1> rho; // proportion unstructured vs. spatially structured variance
+## 
+##   vector[N] theta;       // heterogeneous effects
+##   vector[N] phi;         // spatial effects
+## }
+## transformed parameters {
+##   vector[N] convolved_re;
+##   vector[N] f; 
+##   // variance of each component should be approximately equal to 1
+##   convolved_re =  sqrt(1 - rho) * theta + sqrt(rho / scaling_factor) * phi;
+##   
+##   f = log_E + beta0 + x * beta + convolved_re * sigma;
+## }
+## model {
+##   y ~ poisson_log(f);  // co-variates
+## 
+##   // This is the prior for phi! (up to proportionality)
+##   target += -0.5 * dot_self(phi[node1] - phi[node2]);
+## 
+##   beta0 ~ normal(0.0, 1.0);
+##   beta ~ normal(0.0, 1.0);
+##   theta ~ normal(0.0, 1.0);
+##   sigma ~ normal(0, 1.0);
+##   rho ~ beta(0.5, 0.5);
+##   // soft sum-to-zero constraint on phi)
+##   sum(phi) ~ normal(0, 0.001 * N);  // equivalent to mean(phi) ~ normal(0,0.001)
+## }
+## generated quantities {
+##   vector[N] log_lik; 
+##     for(i in 1:N) log_lik[i] = poisson_log_lpmf(y[i] | log_E[i] + beta0 + x[i, ] * beta + convolved_re * sigma);
+## }
+\end{verbatim}
+
+\subsection{Model 5}\label{model-5-1}
+
+Finally, the code to fit Model 5 quite literally combines the programs
+for Models 3 and 4.
+
+\begin{verbatim}
+## data {
+##   int<lower=0> N;
+##   int<lower=0> N_edges;
+##   int<lower=1, upper=N> node1[N_edges];  // node1[i] adjacent to node2[i]
+##   int<lower=1, upper=N> node2[N_edges];  // and node1[i] < node2[i]
+## 
+##   int<lower=0> y[N];              // count outcomes
+##   vector<lower=0>[N] E;           // exposure
+##   int<lower=1> K;                 // num covariates
+##   matrix[N, K] x;                 // design matrix
+## 
+##   real<lower=0> scaling_factor; // scales the variance of the spatial effects
+##   
+##   //horseshoe
+##   // prior sd for intercept
+##   real<lower=0> scale_icept;
+##   // scale for half t for tau
+##   real<lower=0> scale_global;
+##   // df for tau, lambda
+##   real<lower=1> nu_global;
+##   real<lower=1> nu_local;
+##   // slab scale, df for reg horseshoe
+##   real<lower=0> slab_scale;
+##   real<lower=0> slab_df;
+## }
+## transformed data {
+##   vector[N] log_E = log(E);
+## }
+## parameters {
+##   real beta0;            // intercept
+## 
+##   real<lower=0> sigma;        // overall standard deviation
+##   real<lower=0, upper=1> rho; // proportion unstructured vs. spatially structured variance
+## 
+##   vector[N] theta;       // heterogeneous effects
+##   vector[N] phi;         // spatial effects
+##   
+##   // horseshoe
+##   vector[K] z;
+##   real<lower=0> aux1_global;
+##   real<lower=0> aux2_global;
+##   vector<lower=0>[K] aux1_local;
+##   vector<lower=0>[K] aux2_local;
+##   real<lower=0> caux;
+## }
+## transformed parameters {
+##   vector[N] convolved_re;
+## 
+##   // horseshoe
+##     // global shrinkage
+##   real<lower=0> tau;
+##   // local shrinkage
+##   vector<lower=0>[K] lambda;
+##   vector<lower=0>[K] lambda_tilde; 
+##   // slab scale
+##   real<lower=0> c;
+##   // regression coefficients
+##   vector[K] beta;
+##   // latent function variables
+##   vector[N] f;
+##   
+##   lambda = aux1_local .* sqrt(aux2_local);
+##   tau = aux1_global * sqrt(aux2_global) * scale_global;
+##   c = slab_scale * sqrt(caux); 
+##   lambda_tilde = sqrt(c^2 * square(lambda) ./ (c^2 + tau^2 * square(lambda)));
+##   beta = z .* lambda_tilde*tau;
+##   
+##   // variance of each component should be approximately equal to 1
+##   convolved_re =  sqrt(1 - rho) * theta + sqrt(rho / scaling_factor) * phi;
+##   
+##   f = log_E + beta0 + x * beta + convolved_re * sigma;
+## }
+## model {
+##   y ~ poisson_log(f);  // co-variates
+## 
+##   // This is the prior for phi! (up to proportionality)
+##   target += -0.5 * dot_self(phi[node1] - phi[node2]);
+## 
+##   beta0 ~ normal(0.0, scale_icept);
+##   theta ~ normal(0.0, 1.0);
+##   sigma ~ normal(0, 1.0);
+##   rho ~ beta(0.5, 0.5);
+##   // soft sum-to-zero constraint on phi)
+##   sum(phi) ~ normal(0, 0.001 * N);  // equivalent to mean(phi) ~ normal(0,0.001)
+##   
+##   //horseshoe
+##     // half t and inverse gamma priors
+##   z ~ normal(0, 1);
+##   aux1_local ~ normal(0, 1);
+##   aux2_local ~ inv_gamma(0.5*nu_local, 0.5*nu_local);
+##   aux1_global ~ normal(0, 1);
+##   aux2_global ~ inv_gamma(0.5*nu_global, 0.5*nu_global);
+##   caux ~ inv_gamma(0.5*slab_df, 0.5*slab_df);
+##   beta0 ~ normal(0, scale_icept);
+## }
+## generated quantities {
+##   vector[N] log_lik; 
+##     for(i in 1:N) log_lik[i] = poisson_log_lpmf(y[i] | log_E[i] + beta0 + x[i, ] * beta + convolved_re * sigma);
+## }
+\end{verbatim}
+
+\end{document}

### 2020-02-10 functions/ppcheck.R

@@ -5,9 +5,9 @@ library(data.table)
 library(ggplot2)
 
  
-post <- bym
+post <- poisson_theta_all
 n_obs = 1495
-n_iter = 20000
+n_iter = 40000
 y <- na.omit(p_dat_scaled)$daily_boards
 
 
@@ -29,8 +29,8 @@ y <- na.omit(p_dat_scaled)$daily_boards
     theme_minimal() 
   
   ppoverlay + xlim(0, 200) + ylim(0, 0.026) + 
-    labs(title = "Observed versus simulated ridership for Model 4")
-  ggsave("~/Documents/honors/honors-projects/chapters/ppcheckmod4.png", width = 6, units = "in")
+    labs(title = "Observed versus simulated ridership for Model 2")
+  ggsave("~/Documents/honors/honors-projects/chapters/ppcheckmod2.png", width = 6, units = "in")
   
    ## get traceplot
   

### 2020-02-10 scripts/aws/fitting.R

@@ -71,7 +71,7 @@ poisson_theta <- "~/Documents/honors/honors-project/stan/final-stan/poisson_thet
 
 
 poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 20000, verbose = T, seed = 1997)
-#saveRDS(poisson_theta_fit, "~/Documents/honors/honors-project/final-fits/poisson_theta.RDS")
+saveRDS(poisson_theta_fit, "~/Documents/honors/honors-project/final-fits/poisson_theta_all.RDS")
 
 ## and the bym ! --------------------------------
 

### 2020-02-17 chapters/article-draft.tex
work from weekend

@@
-\documentclass[]{article}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
 \usepackage{lmodern}
@@
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
 \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
+\IfFileExists{microtype.sty}{% use microtype if available
   \usepackage[]{microtype}
@@
 }{}
-\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \hypersetup{
@@
   pdfauthor={Raven McKnight},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 \setcounter{secnumdepth}{5}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
+\usepackage{float}
+\usepackage{booktabs}
 
@@
 
-\section{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 
-\section{Background}\label{background}
-
-\subsection{Metro Transit}\label{metro-transit}
+\hypertarget{background}{%
+\section{Background}\label{background}}
+
+\hypertarget{metro-transit}{%
+\subsection{Metro Transit}\label{metro-transit}}
 
@@
 
-\subsection{Transit Market Areas}\label{transit-market-areas}
+\hypertarget{transit-market-areas}{%
+\subsection{Transit Market Areas}\label{transit-market-areas}}
 
@@
 
-\section{Data}\label{data}
-
-\subsection{Metro Transit Data}\label{metro-transit-data}
+\hypertarget{data}{%
+\section{Data}\label{data}}
+
+\hypertarget{metro-transit-data}{%
+\subsection{Metro Transit Data}\label{metro-transit-data}}
 
@@
 
-\subsubsection{Data Interpolation}\label{data-interpolation}
+\hypertarget{data-interpolation}{%
+\subsubsection{Data Interpolation}\label{data-interpolation}}
 
@@
 
-\subsubsection{Data Aggregation}\label{data-aggregation}
+\hypertarget{data-aggregation}{%
+\subsubsection{Data Aggregation}\label{data-aggregation}}
 
@@
 
-\subsection{Other data sources}\label{other-data-sources}
+\hypertarget{other-data-sources}{%
+\subsection{Other data sources}\label{other-data-sources}}
 
@@
 
-\section{Models}\label{models}
-
-\subsection{Stan}\label{stan}
+\hypertarget{models}{%
+\section{Models}\label{models}}
+
+\hypertarget{stan}{%
+\subsection{Stan}\label{stan}}
 
@@
 
-\subsection{Poisson Regression}\label{poisson-regression}
+\hypertarget{poisson-regression}{%
+\subsection{Poisson Regression}\label{poisson-regression}}
 
@@
 
-\subsubsection{Model 1}\label{model-1}
+\hypertarget{model-1}{%
+\subsubsection{Model 1}\label{model-1}}
 
@@
 
+\hypertarget{overdispersed-poisson-regression}{%
 \subsection{Overdispersed Poisson
-Regression}\label{overdispersed-poisson-regression}
+Regression}\label{overdispersed-poisson-regression}}
 
@@
 
-\subsubsection{Model 2}\label{model-2}
+\hypertarget{model-2}{%
+\subsubsection{Model 2}\label{model-2}}
 
@@
 
+\begin{figure}[H]
+  \centering
+  \includegraphics[width = 6in]{ppcheckmod2.png}
+  \caption{Posterior predictive check for Model 1.}
+\end{figure}
+
 Because we gave \(\theta\) and \(\sigma\) vague priors, we can see a
@@
 
+\hypertarget{horseshoe-priors-and-variable-selection}{%
 \subsection{Horseshoe Priors and Variable
-Selection}\label{horseshoe-priors-and-variable-selection}
+Selection}\label{horseshoe-priors-and-variable-selection}}
 
@@
 
+\hypertarget{spike-and-slab---needs-graphics}{%
 \subsubsection{Spike and Slab - needs
-graphics}\label{spike-and-slab---needs-graphics}
+graphics}\label{spike-and-slab---needs-graphics}}
 
@@
 
+\hypertarget{horseshoe---needs-graphics}{%
 \subsubsection{Horseshoe - needs
-graphics}\label{horseshoe---needs-graphics}
+graphics}\label{horseshoe---needs-graphics}}
 
@@
 
+\hypertarget{regularized-horseshoe---needs-graphics}{%
 \subsubsection{Regularized Horseshoe - needs
-graphics}\label{regularized-horseshoe---needs-graphics}
+graphics}\label{regularized-horseshoe---needs-graphics}}
 
@@
 
-\subsubsection{Model 3}\label{model-3}
-
-For this study, we implement the regularize horseshoe using code
+\hypertarget{model-3}{%
+\subsubsection{Model 3}\label{model-3}}
+
+For this study, we implement the regularized horseshoe using code
 modified from Piironen \& Vehtari.
@@
 of 20,000 iterations produced a divergent transition, we can proceed
-with caution using Model 3.
-
-\subsection{Adding Spatial Structure}\label{adding-spatial-structure}
+with caution using Model 3. Additionally, the parameter estimates
+provided by Model 3 closely match those from Models 1 and 2, suggesting
+that the single divergent transition did not have an adverse effect on
+the posterior estimates.
+
+The posterior predictive checks for Model 3 match Model 2 almost
+perfectly - this is to be expected. The addition of horseshoe priors on
+\(\beta_k\) does not attempt to improve model fit, but to determine
+which covariates are most ``important''.
+
+On that note, we can see that Model 3 \emph{did} shrink two \(\beta_k\)
+exactly to zero.
+
+\hypertarget{adding-spatial-structure}{%
+\subsection{Adding Spatial Structure}\label{adding-spatial-structure}}
 
@@
 
+\hypertarget{conditional-autoregressive-priors}{%
 \subsubsection{Conditional Autoregressive
-Priors}\label{conditional-autoregressive-priors}
+Priors}\label{conditional-autoregressive-priors}}
 
@@
 
+\hypertarget{intrinsic-conditional-autoregressive}{%
 \subsubsection{Intrinsic Conditional
-Autoregressive}\label{intrinsic-conditional-autoregressive}
+Autoregressive}\label{intrinsic-conditional-autoregressive}}
 
@@
 
-\subsubsection{Besag-York-Mollie (BYM)}\label{besag-york-mollie-bym}
+\hypertarget{besag-york-mollie-bym}{%
+\subsubsection{Besag-York-Mollie (BYM)}\label{besag-york-mollie-bym}}
 
@@
 
-\subsubsection{BYM2}\label{bym2}
+\hypertarget{bym2}{%
+\subsubsection{BYM2}\label{bym2}}
 
@@
 
-\subsubsection{Model 4}\label{model-4}
+\hypertarget{model-4}{%
+\subsubsection{Model 4}\label{model-4}}
 
@@
 
-\subsubsection{Model 5}\label{model-5}
+\hypertarget{model-5}{%
+\subsubsection{Model 5}\label{model-5}}
 
@@
 
-\begin{table}[ht]
+\begin{table}[H]
 \centering
@@
   \midrule
-beta0 & 1.0 & 5837 & -1.9 & 0.0 & -1.9 & -1.8 \\ 
-  sigma & 1.0 & 721 & 2.8 & 0.4 & 2.1 & 3.6 \\ 
-  rho & 1.0 & 713 & 0.9 & 0.0 & 0.8 & 0.9 \\ 
-  beta[1] & 1.0 & 3637 & -0.1 & 0.1 & -0.2 & -0.0 \\ 
-  beta[2] & 1.0 & 8170 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
-  beta[3] & 1.0 & 3987 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
-  beta[4] & 1.0 & 8531 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
-  beta[5] & 1.0 & 7619 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
-  beta[6] & 1.0 & 2999 & 0.2 & 0.1 & 0.1 & 0.3 \\ 
-  beta[7] & 1.0 & 3726 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
-  beta[8] & 1.0 & 7956 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
-  beta[9] & 1.0 & 5622 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
-  beta[10] & 1.0 & 7530 & 0.1 & 0.0 & 0.1 & 0.2 \\ 
-  beta[11] & 1.0 & 4558 & -0.1 & 0.0 & -0.1 & 0.0 \\ 
-  beta[12] & 1.0 & 11150 & 0.0 & 0.0 & -0.0 & 0.0 \\ 
-  beta[13] & 1.0 & 8289 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
-  beta[14] & 1.0 & 5469 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
-  beta[15] & 1.0 & 4822 & 0.1 & 0.0 & -0.0 & 0.1 \\ 
-  beta[16] & 1.0 & 3863 & -0.1 & 0.0 & -0.1 & 0.0 \\ 
-  beta[17] & 1.0 & 2094 & 0.1 & 0.0 & -0.0 & 0.1 \\ 
-  beta[18] & 1.0 & 6949 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
+beta0 & 1.000 & 5837 & -1.878 & 0.028 & -1.934 & -1.822 \\ 
+  sigma & 1.010 &  721 & 2.826 & 0.359 & 2.138 & 3.561 \\ 
+  theta[5] & 1.000 & 19566 & -0.188 & 0.787 & -1.722 & 1.349 \\ 
+  rho & 1.012 &  713 & 0.900 & 0.031 & 0.824 & 0.946 \\ 
+  phi[5] & 1.004 & 1652 & 5.974 & 1.573 & 3.264 & 9.527 \\ 
+  beta[1] & 1.001 & 3637 & -0.119 & 0.059 & -0.233 & -0.002 \\ 
+  beta[2] & 1.001 & 8170 & -0.005 & 0.027 & -0.069 & 0.049 \\ 
+  beta[3] & 1.000 & 3987 & 0.024 & 0.030 & -0.019 & 0.096 \\ 
+  beta[4] & 1.000 & 8531 & -0.004 & 0.022 & -0.057 & 0.040 \\ 
+  beta[5] & 1.000 & 7619 & -0.012 & 0.026 & -0.076 & 0.030 \\ 
+  beta[6] & 1.001 & 2999 & 0.164 & 0.055 & 0.051 & 0.268 \\ 
+  beta[7] & 1.000 & 3726 & 0.032 & 0.035 & -0.016 & 0.114 \\ 
+  beta[8] & 1.000 & 7956 & 0.010 & 0.025 & -0.032 & 0.074 \\ 
+  beta[9] & 1.000 & 5622 & 0.019 & 0.028 & -0.023 & 0.089 \\ 
+  beta[10] & 1.000 & 7530 & 0.144 & 0.033 & 0.079 & 0.208 \\ 
+  beta[11] & 1.000 & 4558 & -0.057 & 0.034 & -0.123 & 0.001 \\ 
+  beta[12] & 1.000 & 11150 & 0.004 & 0.018 & -0.031 & 0.046 \\ 
+  beta[13] & 1.000 & 8289 & -0.017 & 0.024 & -0.073 & 0.020 \\ 
+  beta[14] & 1.000 & 5469 & -0.028 & 0.031 & -0.100 & 0.015 \\ 
+  beta[15] & 1.000 & 4822 & 0.060 & 0.035 & -0.001 & 0.130 \\ 
+  beta[16] & 1.001 & 3863 & -0.060 & 0.038 & -0.134 & 0.003 \\ 
+  beta[17] & 1.000 & 2094 & 0.051 & 0.041 & -0.008 & 0.137 \\ 
+  beta[18] & 1.000 & 6949 & -0.010 & 0.021 & -0.061 & 0.026 \\ 
    \bottomrule
@@
 
-\section{Discussion}\label{discussion}
+The posterior predictive checks and mapped residuals for Model 5 are
+very similar to those for Model 4. Like Model 2, the slow sampling time
+is largely caused by autocorrelation between iterations.
+
+The more interesting comparison here is with Model 3. The horseshoe
+priors in Model 5 fail to shrink any parameters \emph{exactly} to zero.
+As previously discussed, running the model for more iterations would
+likely cuase the coverage intervals on the \(\beta_k\) estimates to
+tighten.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  bymhrs is probably not ``worth it'', needs to be reparameterized/coded
-  by someone better than me
-\item
-  incorporating overdispersion is the most important improvement in
-  terms of model fit
-\end{itemize}
-
-\section{Appendix: Stan Code}\label{appendix-stan-code}
-
-\subsection{Model 1}\label{model-1-1}
+While the most complex model \emph{does} converge and provide reasonable
+parameter estimates, its computational inefficiency may not be worth the
+addition of horseshoe priors to the BYM2 model, depending on the
+application. A more efficient approach may be to conduct variable
+selection using horseshoe priors on a simpler model (in our case, Model
+1 or Model 3, for example) before incorporating the ICAR component.
+
+\hypertarget{appendix-stan-code}{%
+\section{Appendix: Stan Code}\label{appendix-stan-code}}
+
+\hypertarget{model-1-1}{%
+\subsection{Model 1}\label{model-1-1}}
 
@@
 
-\subsection{Model 2}\label{model-2-1}
+\hypertarget{model-2-1}{%
+\subsection{Model 2}\label{model-2-1}}
 
@@
 
-\subsection{Model 3}\label{model-3-1}
+\hypertarget{model-3-1}{%
+\subsection{Model 3}\label{model-3-1}}
 
@@
 
-\subsection{Model 4}\label{model-4-1}
-
-The BYM2 model is fit using code from Morris et al.
+\hypertarget{model-4-1}{%
+\subsection{Model 4}\label{model-4-1}}
+
+The BYM2 model is fit using code from Morris et al.~
 
@@
 
-\subsection{Model 5}\label{model-5-1}
+\hypertarget{model-5-1}{%
+\subsection{Model 5}\label{model-5-1}}
 

### 2020-02-17 chapters/draft.Rmd
work from weekend

--- 2020-02-10 chapters/article-draft.Rmd
@@
 date: "2/9/2020"
+header-includes:
+  - \usepackage{float}
+  - \usepackage{booktabs}
 output:
@@
     number_sections: true
-  html_document:
-    theme: paper
-    toc: yes
-    toc_float: yes
 abstract: "abstract"
@@
 ```{r setup, include=FALSE}
+library(knitr)
 knitr::opts_chunk$set(echo = FALSE)
@@
 Currently, Metro Transit uses a set of 5 Transit Market Areas to estimate spatial demand.
-The Transit Market Areas are writen into the official Transportation Planning Policy (TPP) and are calculated using a simple linear regression.
-Following the notation of the TPP, the formula for determing Transit Market Areas is expressed
+The Transit Market Areas are written into the official Transportation Planning Policy (TPP) and are calculated using a simple linear regression.
+Following the notation of the TPP, the formula for determine Transit Market Areas is expressed
 
@@
 
-\begin{figure}[htb]
+\begin{figure}[H]
   \centering
   \includegraphics[width = 5in]{TMA.png}
-  \caption{Metro Transit's Transit Market Areas from Metropolitan Council's Transportation Planning Policy, Appendix G. }
+  \caption{Metro Transit's Transit Market Areas from Metropolitan Council's Transportation Planning Policy, Appendix G.}
 \end{figure}
@@
 The four predictors used are common-sense indicators of transit ridership: high population and employment density are characteristic of trip origins and destinations, automobile availability is reasonably assumed to be related to transit ridership, and intersection density is a proxy for the "walkability" of an area.
-However, the exsiting Transit Market Area model may be over-simplifying the complex question of transit ridership.
+However, the existing Transit Market Area model may be over-simplifying the complex question of transit ridership.
 
@@
 Additionally, the trips missing data may not be random: all buses of one "series", for example, may fail to report at once.
-This could lead to the APC dataset understimating boardings on a particular set of routes or locations.T herefore, we use the three data sets described above to create a more complete augmented APC data set.
+This could lead to the APC dataset underestimating boardings on a particular set of routes or locations.T therefore, we use the three data sets described above to create a more complete augmented APC data set.
 
@@
 In theory, we expect more walkable areas, such as the downtown Saint Paul and Minneapolis, to have larger walk-isochrones.
-Unfortunately, the metric can be fooled by rural areas where you can walk uninterruped down a single roadway, for example.
+Unfortunately, the metric can be fooled by rural areas where you can walk uninterrupted down a single roadway, for example.
 An improved "walkability" metric could improve models of transit ridership.
+
+Nearly all studies of transit ridership incorporate some measure of employment density.
+Few incorporate anything more granular than number of jobs per acre or square mile.
+This study incorporates more detailed employment characteristics from LODES in an attempt to better isolate groups of riders in addition to areas of high ridership.
 
@@
 
-In the basic Poisson regression, we often give $\beta_0$ and $\beta$ vague priors such as Normal($0, 1$). The exposure term $E_i$ comes from the data and does not recieve a prior.
+In the basic Poisson regression, we often give $\beta_0$ and $\beta$ vague priors such as Normal($0, 1$). The exposure term $E_i$ comes from the data and does not receive a prior.
 
@@
 The Stan program used to fit Model 1, as well as the other four models, can be found in the appendix.
-Model 1 fits quickly and efficiently.
+Low autocorrelation between iterations allows Model 1 to fit quickly and efficiently.
 The Stan output for $\beta_0$ and $\beta$ is below.
@@
 Model 1 produces coefficient estimates with tight confidence intervals; few of the $\beta_k$ 95% intervals cross zero.
+
+Notably, $\beta_6$, or percent of residents who rent their residence, has the largest magnitude of any coefficient.
+This finding confirms Metro Transit's belief that "impermanence" is a good predictor of transit ridership.
+Broadly, "impermanence" or "instability" related with being a recent immigrant, a single parent, or even a student is expected to have a strong positive correlation with ridership.
+It is a novel finding within the agency that percent of residents who rent their home is a "stronger" predictor of ridership than more intuitive predictors such as median household income (in this model, the coverage interval for median household income ($\beta[1]$) crosses zero!).
+
+Unfortunately, our 10-minute walk isochrone approximation of walkability $\beta_2$ is shown to have a negative correlation with average daily boardings.
+This suggests that the metric for walkability needs to be improved because we know inherently that walkability should increase ridership.
+
+In addition to our flawed walkability metric, parameters $\beta_4, \beta_5, \beta_8, \beta_{11}, \beta_{12}, \beta_{13}, \beta_{14}, \beta_{16}, \text{and} \beta_{18}$ are negatively associated with daily boardings.
+The fact that $\beta_18$, population density, is negatively correlated with ridership suggests that this model is flawed: population density is widely accepted as one of the most important predictors for transit ridership.
+
 Based on initial diagnostics and precise parameter estimates, Model 1 appears to be a "good" fit.
-In this way, Model 1 illustrates the importance of diagnostics beyond the default Stan output.
+However, posterior predictive checking reveals shortcomings in Model 1's fit.
+Figure __ shows the observed distribution of $Y$ in dark blue overlaid with 100 simulated observations of $Y$ produced by Model 1.
+In a well-fitting model, the simulated $Y$ should closely follow the observed distribution.
 
@@
 Model 1 clearly underestimates the number of block groups with low ridership. This is due in large part to extra-Poisson variance present in Metro Transit's ridership data.
+
+Figure __ maps observed versus simulated ridership.
+Visually, we can detect areas which are markedly different between the actual versus estimated boardings: the model tends to overestimate ridership in "middle ridership" areas (ie, block groups that are in neither the high-activity urban core nor the low-density rural/suburban outskirts of the metro area).
+<!-- More rigorously, we can calculate moran's i -->
 
@@
 One option is to use a generalization of the Poisson distribution, such as a Negative-Binomial distribution, as the observation model.
-For the Metro Transit application, we instead modify the latent function to include a seperate term to account for overdispersion.
+For the Metro Transit application, we instead modify the latent function to include a separate term to account for overdispersion.
 As such, the Poisson regression above can be rewritten to include
@@
 
-where $\theta$ is a set of heterogenous random effects.
-In practice, $\theta$ is scaled by its variance parameter $\sigma$ to allow eassier fitting in Stan.
+where $\theta$ is a set of heterogeneous random effects.
+In practice, $\theta$ is scaled by its variance parameter $\sigma$ to allow easier fitting in Stan.
 The addition of this set of random effects improves fits on data with overdispersion.
@@
 Model 2 is *significantly* more computationally expensive to fit than Model 1.
-This is due to relatively large autocorrelation between each iteration, visible in the traceplots below (Figure ___).
+This is due to relatively large autocorrelation between each iteration, visible in the trace plots below (Figure ___).
 This autocorrelation made it necessary to run Model 2 on four chains for 20,000 iterations each.
 Despite this, `n_eff` is significantly lower for Model 2 than Model 1.
-Model 2 also experienced 2 transitions exceeding maixmum treedepth: not a problem in terms of model fit, but a signal that the posterior is more challenging to fit.
+Model 2 also experienced 2 transitions exceeding maximum treedepth: not a problem in terms of model fit, but a signal that the posterior is more challenging to fit.
 Reparameterization could likely reduce autocorrelation between iterations.
@@
 Notably, the direction and magnitude of the coefficient estimates are similar to those produced by Model 1.
-However, the coverage intervals are much wider.
-Based on "test runs" of Model 2, we expect increasing the number of iterations to decrease the width of these coverage intervals (we observe that the converage interval is proportional to the number of iterations in this setting).
+Specifically, influential parameters such as $\beta_6$ remain influential in this setting.
+However, the coverage intervals for all $\beta_k$ are much wider.
+Based on "test runs" of Model 2, we expect increasing the number of iterations to decrease the width of these coverage intervals (we observe that the width of the coverage interval is proportional to the number of iterations in this setting).
 
@@
 
-<!-- \begin{figure}[H] -->
-<!--   \centering -->
-<!--   \includegraphics[width = 6in]{ppcheckmod2.png} -->
-<!--   \caption{Posterior predictive check for Model 1.} -->
-<!-- \end{figure} -->
+\begin{figure}[H]
+  \centering
+  \includegraphics[height = 6in]{ppcheckmod2.png}
+  \caption{Posterior predictive check for Model 2.}
+\end{figure}
 
 Because we gave $\theta$ and $\sigma$ vague priors, we can see a *much* better fit in the posterior predictive checks for Model 2.
+As above, Figure __ maps the observed ridership by block group versus the ridership simulated by Model 2.
+
 One *could* stop analysis here.
@@
 
-The primary challenge with the spike-and-slab (and other Bayesian shrinkage methods) is the selction of priors.
+The primary challenge with the spike-and-slab (and other Bayesian shrinkage methods) is the Selection of priors.
 Priors must be set for the width of the slab $c$ and the prior inclusion probability $\pi$.
@@
 The second method for Bayesian sparse regression is to give $\beta$ some continuous *sparsity inducing prior*.
-**Van Erp et al (2019)** provide an introduction to many such continous priors.
+**Van Erp et al (2019)** provide an introduction to many such continuous priors.
 Ridge regression and the Bayesian LASSO are two of the most approachable priors, but all sparsity inducing priors utilize the same logic as the spike-and-slab: give $\beta$ a prior with *most* of its mass near 0, to shrink irrelevant coefficients, and the rest of its mass "far" from 0.
@@
 
-for $k = 1, 2, ..., K$. The horseshoe prior is so-named because for fixed values $\tau = \lambda_k = 1$, the prior resemebles a Beta(1/2, 1/2) distribution, or a horseshoe.
+for $k = 1, 2, ..., K$. The horseshoe prior is so-named because for fixed values $\tau = \lambda_k = 1$, the prior resembles a Beta(1/2, 1/2) distribution, or a horseshoe.
 
@@
 c & \sim ~ \text{Inverse-Gamma}(a, b) \\
-\tau & \sim 
 \end{aligned}
 $$
-
+<!-- add \tau & \sim  -->
 where $c$ > 0 helps to regularize $\beta_k$ far from zero.
@@
 
-For this study, we implement the regularize horseshoe using code modified from Piironen & Vehtari.
+For this study, we implement the regularized horseshoe using code modified from Piironen & Vehtari.
+We set priors following Piironen & Vehtari's recommendations.
+The priors assumed a priori that approximately half of the 18 covariates described in section __ are relevant.
+This choice was made based on the assumption that many of the covariates are correlated and therefore may be redundant.
+
+Like Model 2, this model requires 20,000 iterations for each of four chains to converge and produce suitably large Effective Sample Sizes.
+This model is more complex and slower still than Model 2 to fit.
+Both the overdispersion parameter as currently programmed in Stan and the sparsity inducing priors contribute to overall computation time.
 
@@
 Because only 1 out of 20,000 iterations produced a divergent transition, we can proceed with caution using Model 3.
+Additionally, the parameter estimates provided by Model 3 closely match those from Models 1 and 2, suggesting that the single divergent transition did not have an adverse effect on the posterior estimates.
+
+The posterior predictive checks for Model 3 match Model 2 almost perfectly - this is to be expected.
+The addition of horseshoe priors on $\beta_k$ does not attempt to improve model fit, but to determine which covariates are most "important".
+
+We can see that Model 3 *did* shrink two $\beta_k$ exactly to zero.
+Based on our priors, we expected more than two priors to be shrunk.
+The fact that more coefficients were not shrunk suggests that, despite being correlated, nearly all of the 18 selected covariates have some relevance.
+We expect horseshoe priors to be more helpful in modeling contexts with significantly more parameters and less confidence in their importance.
 
@@
 
+**need map to illustrate this section**
 This bounding convention means that boardings are often assigned to different block groups depending on which direction the passenger is travelling.
@@
 Particularly in dense areas, it's very likely for riders to board in a block group other than the block group they reside in.
-For example, I live in block group ____ but walk about 10 minutes to board the A-Line southbound at Snelling and Randolph.
+For example, I live in block group __ but walk about 10 minutes to board the A-Line southbound at Snelling and Randolph in block group __.
 Commuting patterns such as these can assign ridership to different block groups than we might expect.
@@
 
-Conditional Autoregressive (CAR) priors are one of the most widespread Bfayesian methods for modeling spatial autocorrelation.
+Conditional Autoregressive (CAR) priors are one of the most widespread Bayesian methods for modeling spatial autocorrelation.
 CAR models were introduced in Besag 1974 and remain perhaps the primary method for Bayesian areal data modeling.
@@
 CAR priors were designed in part to smooth such noisy counts.
-As such, they are often applied in epidemiological disease risk modeling in which disease occurences may be rare.
+As such, they are often applied in epidemiological disease risk modeling in which disease occurrences may be rare.
 
@@
 Areal units $i$ and $j$ are considered neighbors if they share a boundary.
-For strictly rectangular lattice data, we often choose either a "Queens" or "Rooks" neighborood structure.
+For strictly rectangular lattice data, we often choose either a "Queens" or "Rooks" neighborhood structure.
 **insert graphics** For irregularly shaped areal data, such as our 1,495 census block groups, we generally default to a queen neighborhood structure wherein areal units sharing *any* points of contact are deemed neighbors.
 
-For $n$ regions, the neighorhood relationships are encoded in an $n \times n$ neighborhood matrix $W$.
+For $n$ regions, the neighborhood relationships are encoded in an $n \times n$ neighborhood matrix $W$.
 Matrix $W$ is defined such that $w_{ij} = 1$ if if regions $i$ and $j$ are neighbors (denoted $i \sim j$) and 0 otherwise.
@@
 
-The Intrinsic Conditional Autoregressive (ICAR) prior is a slight simplifation of a CAR prior. ICAR priors set $\alpha = 1$ which simplifies the definition of Q
+The Intrinsic Conditional Autoregressive (ICAR) prior is a slight simplification of a CAR prior. ICAR priors set $\alpha = 1$ which simplifies the definition of Q
 
@@
 In short, the sampler is forced to explore all possible combinations of $\theta$ and $\phi$, no matter how unlikely a particular combination my be.
-In the context of fitting models with Stan, this is liekly to cause uneccesarily long computation times and perhaps to yield biased posterior estimates.
+In the context of fitting models with Stan, this is likely to cause unnecessarily long computation times and perhaps to yield biased posterior estimates.
 
@@
 A relatively vague Normal(0, 1) prior is appropriate for $\sigma$.
-Riebler et al (2016) also propose either a Beta(1/2, 1/2) or more specialized, complex prior for mixing parameter $\rho$.
+Riebler et al (2016) propose either a Beta(1/2, 1/2) or more specialized, complex prior for mixing parameter $\rho$.
 For this study, we use the simpler Beta(1/2, 1/2) prior.
@@
 Model 4 uses all 18 covariates of interest, no horseshoe priors, and decomposes $\theta\sigma$ into a set of random effects plus a spatial ICAR component.
+
 This model converges much more efficiently than Model 2 or Model 3 due to careful reparameterization of the BYM by Riebler et al (2016).
-Future would could potentially improve the speed of fitting Models 2 and 3.
+Future work could likely improve the speed of fitting Models 2 and 3.
 
@@
   \midrule
-beta0 & 1.001 & 4770 & -1.880 & 0.028 & -1.937 & -1.825 \\ 
-  sigma & 1.003 &  520 & 2.698 & 0.372 & 1.993 & 3.451 \\ 
-  rho & 1.002 &  514 & 0.888 & 0.038 & 0.793 & 0.941 \\ 
-  theta[5] & 1.000 & 17797 & -0.179 & 0.777 & -1.688 & 1.336 \\ 
-  beta[1] & 1.000 & 4322 & 0.034 & 0.047 & -0.056 & 0.127 \\ 
-  beta[2] & 1.001 & 3801 & 0.061 & 0.035 & -0.009 & 0.131 \\ 
-  beta[3] & 1.000 & 4203 & -0.003 & 0.039 & -0.080 & 0.073 \\ 
-  beta[4] & 1.001 & 4330 & -0.049 & 0.044 & -0.134 & 0.037 \\ 
-  beta[5] & 1.000 & 4706 & 0.209 & 0.042 & 0.127 & 0.292 \\ 
-  beta[6] & 1.000 & 4919 & 0.065 & 0.038 & -0.009 & 0.138 \\ 
-  beta[7] & 1.002 & 3929 & 0.029 & 0.038 & -0.046 & 0.104 \\ 
-  beta[8] & 1.000 & 6278 & 0.145 & 0.032 & 0.083 & 0.207 \\ 
-  beta[9] & 1.000 & 4607 & -0.088 & 0.031 & -0.150 & -0.026 \\ 
-  beta[10] & 1.000 & 6145 & 0.014 & 0.029 & -0.043 & 0.070 \\ 
-  beta[11] & 1.000 & 6978 & -0.022 & 0.031 & -0.083 & 0.040 \\ 
-  beta[12] & 1.000 & 6125 & -0.068 & 0.034 & -0.136 & -0.002 \\ 
-  beta[13] & 1.000 & 5875 & 0.094 & 0.033 & 0.029 & 0.158 \\ 
-  beta[14] & 1.000 & 4743 & -0.089 & 0.035 & -0.157 & -0.019 \\ 
-  beta[15] & 1.001 & 2720 & 0.076 & 0.039 & 0.001 & 0.151 \\ 
-  beta[16] & 1.000 & 6053 & -0.029 & 0.031 & -0.091 & 0.031 \\ 
+beta0 & 1.000 & 5803 & -1.893 & 0.029 & -1.950 & -1.836 \\ 
+  sigma & 1.004 &  552 & 2.755 & 0.373 & 2.050 & 3.495 \\ 
+  rho & 1.004 &  556 & 0.893 & 0.037 & 0.806 & 0.943 \\ 
+  theta[5] & 1.000 & 20232 & -0.195 & 0.781 & -1.743 & 1.334 \\ 
+  phi[5] & 1.002 & 1150 & 6.340 & 1.699 & 3.526 & 10.018 \\ 
+  beta[1] & 1.000 & 5929 & -0.134 & 0.051 & -0.234 & -0.034 \\ 
+  beta[2] & 1.001 & 4670 & 0.009 & 0.052 & -0.093 & 0.112 \\ 
+  beta[3] & 1.000 & 4406 & 0.054 & 0.036 & -0.017 & 0.125 \\ 
+  beta[4] & 1.000 & 4538 & -0.032 & 0.040 & -0.112 & 0.047 \\ 
+  beta[5] & 1.000 & 4585 & -0.044 & 0.044 & -0.130 & 0.043 \\ 
+  beta[6] & 1.000 & 5071 & 0.151 & 0.048 & 0.057 & 0.244 \\ 
+  beta[7] & 1.001 & 4621 & 0.050 & 0.038 & -0.025 & 0.125 \\ 
+  beta[8] & 1.001 & 5089 & 0.057 & 0.047 & -0.033 & 0.148 \\ 
+  beta[9] & 1.000 & 4603 & 0.049 & 0.042 & -0.034 & 0.131 \\ 
+  beta[10] & 1.000 & 6915 & 0.146 & 0.031 & 0.085 & 0.208 \\ 
+  beta[11] & 1.001 & 6255 & -0.084 & 0.031 & -0.145 & -0.023 \\ 
+  beta[12] & 1.000 & 6891 & 0.014 & 0.029 & -0.043 & 0.071 \\ 
+  beta[13] & 1.000 & 8267 & -0.027 & 0.031 & -0.088 & 0.035 \\ 
+  beta[14] & 1.001 & 6420 & -0.069 & 0.034 & -0.135 & -0.001 \\ 
+  beta[15] & 1.000 & 6277 & 0.090 & 0.033 & 0.027 & 0.155 \\ 
+  beta[16] & 1.000 & 5411 & -0.093 & 0.035 & -0.162 & -0.025 \\ 
+  beta[17] & 1.001 & 2951 & 0.067 & 0.040 & -0.012 & 0.146 \\ 
+  beta[18] & 1.001 & 6516 & -0.027 & 0.031 & -0.088 & 0.034 \\ 
    \bottomrule
@@
 
-Model 5 builds upon Model 4 by incorporating regularized horsehoe priors in addition to spatial ICAR priors.
+Model 5 builds upon Model 4 by incorporating regularized horseshoe priors in addition to spatial ICAR priors.
 This model has high autocorrelation between iterations and took *significantly* longer than any of the other models to fit in Stan (upwards of 12 hours).
@@
   \midrule
-beta0 & 1.0 & 5837 & -1.9 & 0.0 & -1.9 & -1.8 \\ 
-  sigma & 1.0 & 721 & 2.8 & 0.4 & 2.1 & 3.6 \\ 
-  rho & 1.0 & 713 & 0.9 & 0.0 & 0.8 & 0.9 \\ 
-  beta[1] & 1.0 & 3637 & -0.1 & 0.1 & -0.2 & -0.0 \\ 
-  beta[2] & 1.0 & 8170 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
-  beta[3] & 1.0 & 3987 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
-  beta[4] & 1.0 & 8531 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
-  beta[5] & 1.0 & 7619 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
-  beta[6] & 1.0 & 2999 & 0.2 & 0.1 & 0.1 & 0.3 \\ 
-  beta[7] & 1.0 & 3726 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
-  beta[8] & 1.0 & 7956 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
-  beta[9] & 1.0 & 5622 & 0.0 & 0.0 & -0.0 & 0.1 \\ 
-  beta[10] & 1.0 & 7530 & 0.1 & 0.0 & 0.1 & 0.2 \\ 
-  beta[11] & 1.0 & 4558 & -0.1 & 0.0 & -0.1 & 0.0 \\ 
-  beta[12] & 1.0 & 11150 & 0.0 & 0.0 & -0.0 & 0.0 \\ 
-  beta[13] & 1.0 & 8289 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
-  beta[14] & 1.0 & 5469 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
-  beta[15] & 1.0 & 4822 & 0.1 & 0.0 & -0.0 & 0.1 \\ 
-  beta[16] & 1.0 & 3863 & -0.1 & 0.0 & -0.1 & 0.0 \\ 
-  beta[17] & 1.0 & 2094 & 0.1 & 0.0 & -0.0 & 0.1 \\ 
-  beta[18] & 1.0 & 6949 & -0.0 & 0.0 & -0.1 & 0.0 \\ 
+beta0 & 1.000 & 5837 & -1.878 & 0.028 & -1.934 & -1.822 \\ 
+  sigma & 1.010 &  721 & 2.826 & 0.359 & 2.138 & 3.561 \\ 
+  theta[5] & 1.000 & 19566 & -0.188 & 0.787 & -1.722 & 1.349 \\ 
+  rho & 1.012 &  713 & 0.900 & 0.031 & 0.824 & 0.946 \\ 
+  phi[5] & 1.004 & 1652 & 5.974 & 1.573 & 3.264 & 9.527 \\ 
+  beta[1] & 1.001 & 3637 & -0.119 & 0.059 & -0.233 & -0.002 \\ 
+  beta[2] & 1.001 & 8170 & -0.005 & 0.027 & -0.069 & 0.049 \\ 
+  beta[3] & 1.000 & 3987 & 0.024 & 0.030 & -0.019 & 0.096 \\ 
+  beta[4] & 1.000 & 8531 & -0.004 & 0.022 & -0.057 & 0.040 \\ 
+  beta[5] & 1.000 & 7619 & -0.012 & 0.026 & -0.076 & 0.030 \\ 
+  beta[6] & 1.001 & 2999 & 0.164 & 0.055 & 0.051 & 0.268 \\ 
+  beta[7] & 1.000 & 3726 & 0.032 & 0.035 & -0.016 & 0.114 \\ 
+  beta[8] & 1.000 & 7956 & 0.010 & 0.025 & -0.032 & 0.074 \\ 
+  beta[9] & 1.000 & 5622 & 0.019 & 0.028 & -0.023 & 0.089 \\ 
+  beta[10] & 1.000 & 7530 & 0.144 & 0.033 & 0.079 & 0.208 \\ 
+  beta[11] & 1.000 & 4558 & -0.057 & 0.034 & -0.123 & 0.001 \\ 
+  beta[12] & 1.000 & 11150 & 0.004 & 0.018 & -0.031 & 0.046 \\ 
+  beta[13] & 1.000 & 8289 & -0.017 & 0.024 & -0.073 & 0.020 \\ 
+  beta[14] & 1.000 & 5469 & -0.028 & 0.031 & -0.100 & 0.015 \\ 
+  beta[15] & 1.000 & 4822 & 0.060 & 0.035 & -0.001 & 0.130 \\ 
+  beta[16] & 1.001 & 3863 & -0.060 & 0.038 & -0.134 & 0.003 \\ 
+  beta[17] & 1.000 & 2094 & 0.051 & 0.041 & -0.008 & 0.137 \\ 
+  beta[18] & 1.000 & 6949 & -0.010 & 0.021 & -0.061 & 0.026 \\ 
    \bottomrule
 \end{tabular}
-\caption{Stan output for Model 5} 
+\caption{Stan output for Model 5.} 
 \end{table}
 
-# Discussion
+The posterior means for $\sigma$ and $\rho$, as well as a "random" $\theta$ value are all similar to the posterior means of Model 4.
+
+The posterior predictive checks and mapped residuals for Model 5 are very similar to those for Model 4.
+Like Model 2, the slow sampling time is largely caused by autocorrelation between iterations.
+In this case, the autocorrelation appears to be introduced by the horseshoe prior rather than by the well-parameterized BYM2.
+
+The more interesting comparison here is with Model 3.
+The horseshoe priors in Model 5 fail to shrink any parameters *exactly* to zero.
+As previously discussed, running the model for more iterations would likely cause the coverage intervals on the $\beta_k$ estimates to tighten.
+
+# Discussion/Results?
 
@@
 
-- bymhrs is probably not "worth it", needs to be reparameterized/coded by someone better than me
-- incorporating overdispersion is the most important improvement in terms of model fit
+While the most complex model *does* converge and provide reasonable parameter estimates, its computational inefficiency may not be worth the addition of horseshoe priors to the BYM2 model, depending on the application.
+A more efficient approach may be to conduct variable selection using horseshoe priors on a simpler model (in our case, Model 1 or Model 3, for example) before incorporating the ICAR component.
+
+The results of all five models fit in this study provide insights the existing Transit Market Areas cannot.
+Many of the results confirm institutional or intuitive knowledge at Metro Transit: for example, that "impermanence", represented here by percent of residents renting, is a good predictor of transit ridership.
+
+**once latex is working, this is where the table of posterior estimates was**
+
+## Future Work
+
+The work in this study will be the foundation of a temporal study at Metro Transit.
+In the process of fitting the models discussed above, Metro Transit data and covariates were prepared for years 2015-2018.
+Future work will include applying these models to each year and/or incorporating explicitly temporal trends into the models.
+
+Many other covariates can be considered.
+The 18 selected for this study were selected largely based off of interest and perceived importance.
+Additional or more detailed demographic predictors could be incorporated: for example, percent of residents identified as immigrants (assuming recent immigrants may be transit reliant) or age or income breakdowns rather than simple medians.
+This study incorporated no land use characteristics or service-based predictors.
+Incorporating either could improve model fits or reveal new insights.
+
+In a different vein, there is interest at Metro Transit in developing a "transit-specific geography" to solve the bounding problem discussed in Section __ such that future models would not necessarily rely on geographic smoothing.
 

### 2020-02-17 07_figures.R

@@ -13,8 +13,16 @@ bgs <- block_groups("MN", counties, 2017)
 
 sfdat <- left_join(bgs, mod_dat, on = 'GEOID')
 
-ggplot(sfdat) + 
+post <- readRDS("~/Documents/honors/honors-project/final-fits/poisson_fit.RDS")
+post <- rstan::extract(post)
+f <- post$f   
+
+
+
+t <- ggplot(sfdat) + 
   geom_sf(aes(fill = log(daily_boards)), color = "grey", lwd = .1) +
   theme_minimal() +
   scale_fill_viridis(na.value = "transparent") +
   labs(fill = "Boardings", title = "Average Weekday Boardings, 2017", subtitle = "Log scale, enumerated by block group")
+
+ggsave(plot = t, "~/Documents/honors/honors-project/chapters/big_map.png", scale = 5)

### 2020-06-15 scripts/05_get-isochrones.R
first pass at indicators

@@ -13,11 +13,12 @@ invisible(lapply(packages, library, character.only = TRUE))
 
 rm(miss_pkgs, packages)
 
+
 options(tigris_class = 'sf')
 
 # first, download population-weighted centroids
 url <- "https://www2.census.gov/geo/docs/reference/cenpop2010/blkgrp/CenPop2010_Mean_BG27.txt"
-loc <- file.path("/Users/mcknigri/Documents/honors/honors-project/data/covariates", 'centrs.csv')
+loc <- file.path("/Users/raven/Documents/honors/old-honors/data/covariates", 'centrs.csv')
 download.file(url, loc)
 
 centrs <- read_csv('data/covariates/centrs.csv')
@@ -47,7 +48,7 @@ isos[, max_area := 2.01*10^6] # area of perfect r=800m circle
 isos[, walkability := area/max_area] 
 isos[, year  := as.character(3)]
 
-bg_isos <- isos[, .(area, max_area, walkability, id)]
+bg_isos <- isos[, .(area, max_area, walkability, id, year)]
 setnames(bg_isos, "id", 'GEOID')
 
 bg_isos[, walkability := as.numeric(area)]

### 2020-06-15 scripts/07_add-indicators.R

@@ -1,0 +1,49 @@
+## Goal: gather some indicator variables to try out
+
+## packages -----------------------------------------------
+packages <- c('data.table', 'tigris', 'ggplot2', 'dplyr', 'rgdal', 'sf')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+options(tigris_class = 'sf')
+
+## data ---------------------------------------------------
+
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+bgs <- block_groups("MN", counties, 2016)
+
+## gather -------------------------------------------------
+
+## college/university
+
+# read data from: https://hifld-geoplatform.opendata.arcgis.com/datasets/colleges-and-universities/data
+
+url <- "https://opendata.arcgis.com/datasets/0d7bedf9d582472e9ff7a6874589b545_0.zip?outSR=%7B%22latestWkid%22%3A3857%2C%22wkid%22%3A102100%7D"
+loc <- file.path(tempdir(), "cu.zip")
+download.file(url, loc)
+unzip(loc, exdir = file.path(tempdir(), "cu"), overwrite = TRUE)
+file.remove(loc)
+cu <- read_sf(file.path(tempdir(), "cu"), stringsAsFactors = FALSE)
+
+# transform
+cu <- st_transform(cu, st_crs(bgs))
+
+# just in counties
+cubg <- st_intersection(cu, bgs) # wow sf update is way faster
+ggplot() + geom_sf(data = bgs) + geom_sf(data = cubg)
+
+## light rail
+
+## airport
+
+## hospital
+
+## add to covariates --------------------------------------

### 2020-06-15 scripts/09-figures.R

@@ -1,65 +1,0 @@
-## goal: create lots of plots!
-
-## packages -----------------------------------------------
-
-# these lines make sure any user has all the necessary packages
-packages <- c('ggplot2', 'dplyr', 'sf', 'tigris', 'data.table')
-
-miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
-
-if(length(miss_pkgs) > 0){
-  install.packages(miss_pkgs)
-}
-
-invisible(lapply(packages, library, character.only = TRUE))
-
-rm(miss_pkgs, packages)
-
-## data ---------------------------------------------------
-
-counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
-options(tigris_class = "sf")
-bgs <- block_groups("MN", counties, 2016)
-
-# unscaled covariates for mapping etc
-cov <- readRDS("data/covariates/cleaned/all_covariates.RDS")
-tomap <- cov[, .(unemprate, perc_foreign, perc_transit_comm, perc_wfh, walkability, 
-                 avg_veh, perc_english_only, perc_rent, perc_hs, perc_bach, 
-                 estimate_median_age, estimate_median_hh_income, perc_only_white,
-                 w_perc_jobs_age_less30, w_perc_jobs_white, w_perc_jobs_men, 
-                 w_perc_jobs_less40, w_perc_jobs_no_college, genz, millenial,
-                 genx, boomer, emp_density, pop_density, GEOID)]
-
-# sf
-sfdat <- dplyr::left_join(bgs, tomap)
-
-
-## functions ----------------------------------------------
-source("functions/plotting.R")
-
-## PLOTS ##
-
-## histograms
-hist_titles <- c("Unemployment rate", "Percent foreign-born residents", "Percent of residents who commute via transit",
-                 "Percent of residents who work from home", "Walkability", "Average number of vehicles per household", 
-                 "Percent of residents who speak English only", "Percent of housing units occupied by renters", 
-                 "Percent of residents with high school diploma \n(highest educational attainment)", 
-                 "Percent of residents with bachelors degree \n(highest educational attainment)", 
-                 "Median age", "Median household income", "Percent of residents white alone", 
-                 "Percent of jobs for employees under age 30", "Percent of jobs for white employees", 
-                 "Percent of jobs for men", "Percent of jobs making less than $40,000", 
[…]

### 2020-06-15 scripts/10-figures.R

@@ -1,0 +1,86 @@
+## goal: create lots of plots!
+
+## packages -----------------------------------------------
+
+# these lines make sure any user has all the necessary packages
+packages <- c('ggplot2', 'dplyr', 'sf', 'tigris', 'data.table')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+## data ---------------------------------------------------
+
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+options(tigris_class = "sf")
+bgs <- block_groups("MN", counties, 2016)
+
+# unscaled covariates for mapping etc
+cov <- readRDS("data/covariates/cleaned/all_covariates.RDS")
+cov <- cov[year == "3"]
+tohist <- cov[, .(unemprate, perc_foreign, perc_transit_comm, perc_wfh, walkability, 
+                 avg_veh, perc_english_only, perc_rent, perc_hs, perc_bach, perc_no_veh,
+                 estimate_median_age, estimate_median_hh_income, perc_only_white,
+                 w_perc_jobs_age_less30, w_perc_jobs_white, w_perc_jobs_men, 
+                 w_perc_jobs_less40, w_perc_jobs_no_college, genz, millenial,
+                 genx, boomer, emp_density, pop_density, GEOID)]
+
+
+## functions ----------------------------------------------
+source("functions/plotting.R")
+
+## PLOTS ##
+
+## histograms
+hist_titles <- c("Unemployment rate", "Percent foreign-born residents", "Percent of residents who commute via transit",
+                 "Percent of residents who work from home", "Walkability", "Average number of vehicles per household", 
+                 "Percent of residents who speak English only", "Percent of housing units occupied by renters", 
+                 "Percent of residents with high school diploma \n(highest educational attainment)", 
+                 "Percent of residents with bachelors degree \n(highest educational attainment)", 
+                 "Percent of households with no vehicle", "Median age", "Median household income", 
+                 "Percent of residents white alone", "Percent of jobs for employees under age 30", 
+                 "Percent of jobs for white employees", "Percent of jobs for men", "Percent of jobs making less than $40,000", 
+                 "Percent of jobs for employees without \ncollege degrees", "Percent of population age 0-25", 
[…]

### 2020-06-15 scripts/07_add-indicators.R
add indicators

@@ -25,7 +25,6 @@ bgs <- block_groups("MN", counties, 2016)
 ## college/university
 
 # read data from: https://hifld-geoplatform.opendata.arcgis.com/datasets/colleges-and-universities/data
-
 url <- "https://opendata.arcgis.com/datasets/0d7bedf9d582472e9ff7a6874589b545_0.zip?outSR=%7B%22latestWkid%22%3A3857%2C%22wkid%22%3A102100%7D"
 loc <- file.path(tempdir(), "cu.zip")
 download.file(url, loc)
@@ -40,10 +39,51 @@ cu <- st_transform(cu, st_crs(bgs))
 cubg <- st_intersection(cu, bgs) # wow sf update is way faster
 ggplot() + geom_sf(data = bgs) + geom_sf(data = cubg)
 
+  
 ## light rail
 
+# data from https://gisdata.mn.gov/dataset/us-mn-state-metc-trans-stop-boardings-alightings
+lr <- fread('data/csv_trans_stop_boardings_alightings/TransitStopsBoardingsAndAlightings2018.csv') # year doesn't really matter
+lr <- lr[Route == "Blue Line" | Route == "North Star" | Route == "Green Line"]
+lr <- st_as_sf(lr, coords = c('longitude', 'latitude'), crs = st_crs(bgs))
+
+# just in counties
+lrbg <- st_intersection(lr, bgs) 
+ggplot() + geom_sf(data = bgs) + geom_sf(data = lrbg)
+
 ## airport
+abg <- 270539800001
 
 ## hospital
 
+# download data from http://gis-hennepin.opendata.arcgis.com/datasets/70e1331d41194f3fa8d48200a2a1af6a_1
+url <- "https://opendata.arcgis.com/datasets/70e1331d41194f3fa8d48200a2a1af6a_1.zip?outSR=%7B%22latestWkid%22%3A26915%2C%22wkid%22%3A26915%7D"
+loc <- file.path(tempdir(), "hos.zip")
+download.file(url, loc)
+unzip(loc, exdir = file.path(tempdir(), "hos"), overwrite = TRUE)
+file.remove(loc)
+hos <- read_sf(file.path(tempdir(), "hos"), stringsAsFactors = FALSE)
+
+# transform
+hos <- st_transform(hos, st_crs(bgs))
+
+# in counties
+hosbg <- st_intersection(hos, bgs)
+ggplot() + geom_sf(data = bgs) + geom_sf(data = hosbg)
+
[…]

### 2020-06-15 scripts/08_join-mod-dat.R

@@ -31,7 +31,7 @@ ridership <- ridership[, .(daily_boards = mean(daily_boards, na.rm = T),
                            daily_stops = mean(daily_stops, na.rm = T)), keyby = .(GEOID)]
 
 
-cov <- readRDS('data/covariates/cleaned/all_covariates_scaled.RDS')
+cov <- readRDS('data/covariates/cleaned/all_covariates_scaled_ind.RDS')
 setDT(cov)
 cov <- cov[year == "3"]
 

### 2020-06-15 scripts/10-figures.R

@@ -84,3 +84,5 @@ legend_titles <- c("rate", "percent", "percent", "percent", "area", "average", r
 for(i in 1:length(map_titles)){
   makeMap(sfdat, names(tomap)[i], map_titles[i], legend_titles[i])
 }
+
+## something a little different for indicators

### 2020-06-15 scripts/04_get-census.R
update to 2018 acs estimates

@@ -1,4 +1,4 @@
-# Goal: get census variables of interest, 2015-2017
+# Goal: get census variables of interest, 2015-2018
 # I'm sure there are more refined ways to do this 
 
 ## packages -----------------------------------------------
@@ -31,7 +31,7 @@ counties <- c('Anoka', 'Carver', 'Dakota', 'Hennepin', 'Ramsey', 'Scott', 'Washi
 bgs <- block_groups("MN", counties, 2016)
 
 # 2018 acs not out yet
-years <- list(2015, 2016, 2017)
+years <- list(2015, 2016, 2017, 2018)
 
 ## basics ---------------------------------------
 basics <- map_dfr(
@@ -297,6 +297,8 @@ saveRDS(unemp, "data/covariates/tract/unemp.RDS")
 
 ## employment data ----------------------------------------
 
+# this data is still only available thru 2017
+
 # i'll use the employment data i cleaned here:
 # https://github.com/ravenmcknight/LODES-analysis/blob/master/get-ts-data.R
 # add year restrictions here
@@ -325,6 +327,7 @@ l <- lapply(paste0('data/covariates/wac/', wac_files), fread)
 l[[1]]$year <- 3
 l[[2]]$year <- 2
 l[[3]]$year <- 1
+
 
 # this whole section should be cleaned up considerably
 

### 2020-06-15 scripts/05_get-isochrones.R

@@ -46,7 +46,7 @@ setDT(isos)
 isos[, area := st_area(geometry)]
 isos[, max_area := 2.01*10^6] # area of perfect r=800m circle
 isos[, walkability := area/max_area] 
-isos[, year  := as.character(3)]
+isos[, year  := as.character(4)]
 
 bg_isos <- isos[, .(area, max_area, walkability, id, year)]
 setnames(bg_isos, "id", 'GEOID')

### 2020-06-15 scripts/08_join-mod-dat.R

@@ -33,7 +33,17 @@ ridership <- ridership[, .(daily_boards = mean(daily_boards, na.rm = T),
 
 cov <- readRDS('data/covariates/cleaned/all_covariates_scaled_ind.RDS')
 setDT(cov)
-cov <- cov[year == "3"]
+
+# need year == '3' for lehd and year == '4' for acs
+late <- c("workers", "emp_density", "w_total_jobs_here", "w_perc_jobs_white", "w_perc_jobs_men",
+          "w_perc_jobs_no_college", "w_perc_jobs_less40", "w_perc_jobs_age_less30", "GEOID")
+late2 <- c("workers", "emp_density", "w_total_jobs_here", "w_perc_jobs_white", "w_perc_jobs_men",
+          "w_perc_jobs_no_college", "w_perc_jobs_less40", "w_perc_jobs_age_less30")
+
+emp <- cov[year == "3", ..late]
+notemp <- cov[year == "4", !..late2]
+
+cov <- emp[notemp, on = .(GEOID)]
 
 mod_dat <- cov[ridership, on = .(GEOID)]
 

### 2020-06-16 scripts/09_fitting.R
add script specific to mt fitting

@@ -72,7 +72,7 @@ poisson_hrs <- "stan/poisson_horseshoe.stan"
 
 poisson_theta <- "stan/poisson_theta.stan"
 
-#poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 20000, verbose = T, seed = 1997)
+poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 20000, verbose = T, seed = 1997)
 #saveRDS(poisson_theta_fit, "~fits/poisson_theta.RDS")
 
 ## and the bym2!! -------------------------------

### 2020-06-16 scripts/09_mt-fitting.R

@@ -1,0 +1,57 @@
+## Goal: fit models for mt!
+
+## packages -----------------------------------------------
+packages <- c('rstan', 'dplyr', 'data.table', 'shinystan', 'ggplot2',
+              'bayesplot', 'tigris', 'spdep', 'sf')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+options(tigris_use_cache = TRUE)
+options(tigris_class = 'sf')
+options(mc.cores = parallel::detectCores())
+rstan_options(autowrite = T)
+
+## data -----------------------------------------
+
+mod_dat <- readRDS('data/modeling-dat/mod_dat.RDS')
+setDT(mod_dat)
+
+vars <- c("emp_density", "w_perc_jobs_white", "w_perc_jobs_age_less30", 
+         "walkability", "avg_veh", "genz", "millenial", "genx", "boomer", 
+         "pop_density", "college", "hospital", "lightrail", "airport", 
+         "perc_wfh", "estimate_median_hh_income")
+
+mod_dat <- na.omit(mod_dat, cols = c("daily_boards", "daily_stops", vars))
+
+xdat <- mod_dat[, ..vars]
+setDT(xdat)
+
+###### FITTING ######
+
+## overdispersed poisson regression -------------
+
+y <- round(mod_dat$daily_boards)
+E <- mod_dat$daily_stops
+
+N <- nrow(mod_dat)
+
+K <- ncol(xdat)
+# ah factors have to be numeric
+x <- as.matrix(ncol = K, xdat[, `:=` (college = as.numeric(college), hospital = as.numeric(hospital),
+                                     airport = as.numeric(airport), lightrail = as.numeric(lightrail))])
+
+# data for basic poisson
+standat1 <- list(y = y, E = E, N = N, K = K, x = x)
+
+poisson_theta <- "stan/poisson_theta.stan"
+
+poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 100, verbose = T, seed = 1997)
+saveRDS(poisson_theta_fit, "fits/mt/poisson_theta.RDS")

### 2020-06-16 report/honors-translation.Rmd
add first go at report

---
title: "A Spatial Model of Bus Ridership"
author: "Raven McKnight"
output:
  html_document: 
    theme: paper
    toc: true
    toc_depth: 2
    toc_float: TRUE
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This document explores the results of a spatial model of Metro Transit's bus ridership. The model takes the spatial dimension of bus ridership into account and incorporates demographic predictors. This builds on existing methodology by leveraging more of what we know about ridership and the region to predict where ridership will occur.

The model and analysis presented here provide insight for the *propensity* to ride across space. The analysis is conducted at the Census block group level. Predictors are from the American Community Survey (ACS) 2017 5-year estimates and 2017 Longitudinal Employee Household Dynamic Survey (LODES). Predictors are selected based on prior understnading of ridership and riders as well as actual predictive significance. The goal of using demographic predictors is to identify groups of riders and locations where they are likely to ride.

The spatially explicit component of this model accounts for correlation across space. We naturally assume that neighboring block groups will have similar levels of ridership. This model formalizes this assumption into the analysis. The spatial component adds nuance to this model and **allows us to see patterns we cannot see in a non-spatial model.**

This document outlines key findings, implications, and uses for this model of bus ridership. More extensive motivations and methods are available at the end of the document.

# Findings

A key finding of this analysis is that spatial models allow us to see different, more nuanced patterns than we find in non-spatial models. The coefficients for some predictors change drastically when a spatial component is added. In other words, the relationship between predictors and ridership appear differently when we account for spatial correlation. Broadly, we see two types of changes:

**1. Decreased Coefficients: Predictors which are *less* important in a spatial model**  
Predictors which become less important with a spatial component often capture spatial patterns themselves. Population density is a prime example of such a predictor. Population density is included in practically all models of transit ridership because we know ridership can only exist where there is a sufficient density of people. In this way, measures of density (and several other predictors described below) are included in transit models largely because they identify areas which can support transit.

**2. Increased Coefficients: Predictors which are *more* important in a spatial model**  
The explanation for predictors which are more significant in spatial models is slightly more nuanced. These variables are inefficient at predicting ridership over a large area (ie, the 7 County Metro Area) but can be useufl in smaller areas with more things held constant. One example is median household income. We might assume that low-income areas will have higher ridership than high-income areas. However, a low income area fara way from transit will likely have lower ridership than a high-income area close to transit. The spatial model presented here allows us to see that higher median income *is* associated with lower ridership, but only when neighboring values are taken into consideration.

Between these two types of change between models, spatial models can show entirely different patterns than non-spatial counterparts. Often, the conclusions from spatial models are more consistent with our understanding of ridership (for example, median income). The following section will describe predictors used in this model including key insights and implications.

### Population Density

### Renters

### Median Household Income -> income brackets?

### Employment Density

### Median Age -> age cohort

#### Gen Z / Youth (Age 0-25)

#### Millennials (Age 26-40)

#### Gen X (Age 41-55)

#### Baby Boomers (Age 55+)

### White Employees

### Young Employees

### Walkability

### Households with No Vehicle -> number of vehicles per household

### Work From Home

### Employment Rate

### Median Gross Rent

### Public Benefits

### College/University (Indicator)

### Light Rail (Indicator)

### Airport (Indicator)

### Hospital? (Indicator)

# Implications and Future Work

# Motivation

- geographic smoothing
- bus riderhsip is spatial
- outliers
- geography bounding issues

There are several specific motivations for fitting demographic, spatial models of riderhsip. First, there is evidence to suggest that demographic predictors beyond simple population density may be informative in predicting bus ridership. Previous studies of Metro Transit's riders have shown that transit riders tend to be more diverse than the background population in terms of race, income, age, and so on (\cite{2019_transitsystem}). While it is challenging to collect demographics from individual bus riders, we can leverage demographic characteristics of the Twin Cities to predict areas of high ridership. Spatially aggregated data is the most feasible way to include such demographic information.

Including spatial information has other benefits. One primary goal of spatial modeling is to conduct geographic smoothing. In existing methods, such as the process used to determine Transit Market Areas, a model is fit and used to make predictions which are later smoothed using Geographic Information Systems (GIS). This method necessarily sacrifices some granularity in predictions. This model integrates spatial information into the structure of the model itself. This may be particularly helpful in the case of bus ridership data because boardings can only occur at existing bus stops. Geographic smoothing via spatial models can us help make predictions in block groups with no existing bus stops and avoid overfitting outliers.

Census geography itself is a limitation in modeling bus ridership. Geographies such as block groups tend to be bounded by major roadways. Unfortunately, bus routes tend to lie on those same roadways. Because of these bounds, bus stops may be split into block groups counterintuitively. For example, bus stops on either side of the same street travelling in opposite directions are often in different geographies. In these cases, boardings occur on the boundaries of block groups. There may be flaws in assigning demographics to ridership based on this method of spatial aggregation. The full implications of this limitation are not explored in this study, but some work has been done to define ``transit-specific geographies'' which could alleviate this problem (\cite{viggiano_bus_2017}). In this study, we use spatial smoothing via modeling to mitigate bounding issues.

In short, the motivations for this type of modelling are (1) to leverage more of what we know about ridership and our region and (2) to conduct geographic smoothing to handle known challenges with transit data.

<!-- Complex statistical models are likely more appropriate for internal use. There is still a place for simple modeles like the TMA model for communicating with the public.  -->

# Methods

This document provides a brief outline of the methodology used in this analysis. More detailed (and academic) descriptions of the general methodology are available in Morris et al (2019) and Riebler et al (2016). The code used to conduct this analysis and more descriptions of this specific model can be found on Github [here]().

The model used in this analysis is a *spatial Bayesian* model. Byaesian models incorporate prior information and produce distributions rather than point estimates as predictions. As described above, the spatial component is included to perform spatial smoothing. Specifically, the spatial aspect of this model accounts for spatial autocorrelation and adjusts predictions in individual block groups based on its neighboring values.

The specific model used is a Besag-York-Mollie (BYM) model. THe BYM has similar structure to a classic Bayesian Poisson regression wherein the response variable is assumed to follow a Poisson distribution determined by a set of predictors:

$$
Y_i \sim \text{Poisson}(\beta_1 x_1 + \beta_2x_2 + \ldots + \beta_n x_n + \varepsilon_i)
$$

In this notation, $Y_i$ is the response variable (ridership) in block group $i$. Each $x$ is a predictor, each $\beta$ is the corresponding coefficient, and $\varepsilon_i$ is an error term for each block group. Above,``non-spatial model'' refers to this type of Poisson regression. The BYM replaces $\varepsilon_i$ with two terms: non-spatial errors $\theta_i$ and spatial errors $\phi_i$ such that

$$
Y_i \sim \text{Poisson}(\beta_1 x_1 + \beta_2x_2 + \ldots + \beta_n x_n + \theta_i + \phi_i)
$$

The term $\phi_i$ is determined based on an adjacency matrix $W$ which encodes the spatial relationship between block groups. This is how spatial information comes into the model and adjusts predictions based on neighboring block groups. More on this matrix and the definition of $\phi_i$ is available in the resources cited above.

The specific model used in this analysis is a reparameterization of the classic BYM model above called the BYM2. The BYM2 retains the structure of the BYM but is significantly more computationally efficient to fit. The model is fit using Stan via the `rstan` R package.

### 2020-06-16 scripts/07_add-indicators.R
add unscaled covariates throughout for mapping

@@ -73,6 +73,7 @@ ggplot() + geom_sf(data = bgs) + geom_sf(data = hosbg)
 
 ## add to covariates --------------------------------------
 
+## SCALED ##
 cov <- readRDS('data/covariates/cleaned/all_covariates_scaled.RDS')
 setDT(cov)
 cov[, `:=` (airport = 0, college = 0, hospital = 0, lightrail = 0)]
@@ -87,3 +88,19 @@ cov[, `:=` (airport = as.factor(airport), college = as.factor(college),
             hospital = as.factor(hospital), lightrail = as.factor(lightrail))] 
 
 saveRDS(cov, 'data/covariates/cleaned/all_covariates_scaled_ind.RDS')
+
+## UNSCALED ##
+cov <- readRDS('data/covariates/cleaned/all_covariates.RDS')
+setDT(cov)
+cov[, `:=` (airport = 0, college = 0, hospital = 0, lightrail = 0)]
+
+# mark indicators
+cov[GEOID %in% cubg$GEOID, college := 1]
+cov[GEOID %in% hosbg$GEOID, hospital := 1]
+cov[GEOID %in% abg, airport := 1]
+cov[GEOID %in% lrbg$GEOID, lightrail := 1]
+
+cov[, `:=` (airport = as.factor(airport), college = as.factor(college),
+            hospital = as.factor(hospital), lightrail = as.factor(lightrail))] 
+
+saveRDS(cov, 'data/covariates/cleaned/all_covariates_ind.RDS')

### 2020-06-16 scripts/08_join-mod-dat.R

@@ -31,6 +31,7 @@ ridership <- ridership[, .(daily_boards = mean(daily_boards, na.rm = T),
                            daily_stops = mean(daily_stops, na.rm = T)), keyby = .(GEOID)]
 
 
+## SCALED ##
 cov <- readRDS('data/covariates/cleaned/all_covariates_scaled_ind.RDS')
 setDT(cov)
 
@@ -45,6 +46,25 @@ notemp <- cov[year == "4", !..late2]
 
 cov <- emp[notemp, on = .(GEOID)]
 
-mod_dat <- cov[ridership, on = .(GEOID)]
+mod_dat <- ridership[cov, on = .(GEOID)]
 
 saveRDS(mod_dat, 'data/modeling-dat/mod_dat.RDS')
+
+## UNSCALED ##
+cov <- readRDS('data/covariates/cleaned/all_covariates_ind.RDS')
+setDT(cov)
+
+# need year == '3' for lehd and year == '4' for acs
+late <- c("workers", "emp_density", "w_total_jobs_here", "w_perc_jobs_white", "w_perc_jobs_men",
+          "w_perc_jobs_no_college", "w_perc_jobs_less40", "w_perc_jobs_age_less30", "GEOID")
+late2 <- c("workers", "emp_density", "w_total_jobs_here", "w_perc_jobs_white", "w_perc_jobs_men",
+           "w_perc_jobs_no_college", "w_perc_jobs_less40", "w_perc_jobs_age_less30")
+
+emp <- cov[year == "3", ..late]
+notemp <- cov[year == "4", !..late2]
+
+cov <- emp[notemp, on = .(GEOID)]
+
+mod_dat <- ridership[cov, on = .(GEOID)]
+
+saveRDS(mod_dat, 'data/modeling-dat/mod_dat_unscaled.RDS')

### 2020-06-16 scripts/09_mt-fitting.R

@@ -55,3 +55,36 @@ poisson_theta <- "stan/poisson_theta.stan"
 
 poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 100, verbose = T, seed = 1997)
 saveRDS(poisson_theta_fit, "fits/mt/poisson_theta.RDS")
+
+
+## bym2 -----------------------------------------
+bym2 <- "stan/bym2.stan"
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+bgs <- block_groups("MN", counties, 2017)
+
+bg_dat <- left_join(mod_dat, bgs, on = 'GEOID')
+bg_dat <- st_as_sf(bg_dat)
+neighborhood <- poly2nb(bg_dat)
+source("functions/nb_data_funs.R")
+
+nbs <- nb2graph(neighborhood)
+N2 <- nbs$N
+node1 <- nbs$node1
+node2 <- nbs$node2
+N_edges <- nbs$N_edges
+y2 <- round(bg_dat$daily_boards)
+E2 <- bg_dat$daily_stops
+
+adj_matrix <- sparseMatrix(i = node1, j = node2, x = 1, symmetric = TRUE)
+Q <- Diagonal(nbs$N, rowSums(adj_matrix)) - adj_matrix
+Q_pert <- Q + Diagonal(nbs$N) * max(diag(Q)) * sqrt(.Machine$double.eps)
+Q_inv <- inla.qinv(Q_pert, constr = list(A = matrix(1, 1, nbs$N), e = 0))
+
+scaling_factor <- exp(mean(log(diag(Q_inv))))
+
+bym_dat <- list(y = y2, E = E2, x = x, K = K, 
+                N = N2, node1 = node1, node2 = node2, N_edges = N_edges, scaling_factor = scaling_factor)
+
+bym_fit <- stan(bym2, data = bym_dat, iter = 100, verbose = T, seed = 1997, 
+                control = list(max_treedepth = 12))
+#saveRDS(bym_fit, "fits/bym2.RDS")

### 2020-06-16 scripts/10-figures.R

@@ -22,9 +22,8 @@ options(tigris_class = "sf")
 bgs <- block_groups("MN", counties, 2016)
 
 # unscaled covariates for mapping etc
-cov <- readRDS("data/covariates/cleaned/all_covariates.RDS")
-cov <- cov[year == "3"]
-tohist <- cov[, .(unemprate, perc_foreign, perc_transit_comm, perc_wfh, walkability, 
+mod <- readRDS('data/modeling-dat/mod_dat.RDS')
+tohist <- mod[, .(unemprate, perc_foreign, perc_transit_comm, perc_wfh, walkability, 
                  avg_veh, perc_english_only, perc_rent, perc_hs, perc_bach, perc_no_veh,
                  estimate_median_age, estimate_median_hh_income, perc_only_white,
                  w_perc_jobs_age_less30, w_perc_jobs_white, w_perc_jobs_men, 
@@ -55,7 +54,7 @@ for(i in 1:length(hist_titles)){
 }
 
 ## maps
-tomap <- cov[perc_wfh < 0.3, .(unemprate, perc_foreign, perc_transit_comm, perc_wfh, walkability, 
+tomap <- mod[, .(unemprate, perc_foreign, perc_transit_comm, perc_wfh, walkability, 
                  avg_veh, perc_english_only, perc_rent, perc_hs, perc_bach, perc_no_veh,
                  estimate_median_age, estimate_median_hh_income, perc_only_white,
                  w_perc_jobs_age_less30, w_perc_jobs_white, w_perc_jobs_men, 
@@ -79,7 +78,7 @@ map_titles <- c("Unemployment rate", "Percent foreign-born residents", "Percent 
 
 
 legend_titles <- c("rate", "percent", "percent", "percent", "area", "average", rep("percent", 5),
-                   "years", "usd", rep("percent", 10), "density", "density")
+                   "years", "usd", rep("percent", 10), "log \ndensity", "log \ndensity")
 
 for(i in 1:length(map_titles)){
   makeMap(sfdat, names(tomap)[i], map_titles[i], legend_titles[i])

### 2020-06-16 functions/plotting.R
update scripts through creating all figs

@@ -19,12 +19,12 @@ rm(miss_pkgs, packages)
 
 ## MAKE HISTOGRAMS ##
 
-makeHist <- function(dat, var, titles) {
+makeHist <- function(dat, var, titles, subtitles) {
   ggobj <- ggplot(data = dat, aes_string(x = var)) + 
     geom_histogram(fill = "#33638dff") +
     theme_light() +
     labs(title = titles,
-         subtitle = "By Census Block Group, 2017",
+         subtitle = subtitles,
          x = "value") 
   ggsave(sprintf("img/hist/%s.png", var))
 }
@@ -33,22 +33,14 @@ makeHist <- function(dat, var, titles) {
 ## MAKE MAPS ##
 # theoretically should be basically the same
 
-makeMap <- function(dat, var, titles, legend_titles) {
+makeMap <- function(dat, var, titles, legend_titles, subtitles) {
   ggobj <- ggplot(data = dat) + 
     geom_sf(aes_string(fill = var), lwd = 0) +
     theme_light() +
     labs(title = paste0(titles),
-         subtitle = "By Census Block Group, 2017", 
+         subtitle = subtitles, 
          fill = legend_titles) +
-    scale_fill_viridis_c()
+    scale_fill_viridis_c(na.value = "transparent")
   ggsave(sprintf("img/map/%s.png", var))
 }
 
-
-testvars <- c('genz', 'millenial', 'genx', 'boomer')
-testtitles <- c("Genz", "Millenial", "Genx", "Boomer")
-ltitles <- c("perc", "perc", "perc", "perc")
-
-for(i in 1:length(testtitles)){
-  makeMap(sfdat, testvars[i], testtitles[i], ltitles[i])
-}

### 2020-06-16 scripts/10-figures.R

@@ -22,7 +22,7 @@ options(tigris_class = "sf")
 bgs <- block_groups("MN", counties, 2016)
 
 # unscaled covariates for mapping etc
-mod <- readRDS('data/modeling-dat/mod_dat.RDS')
+mod <- readRDS('data/modeling-dat/mod_dat_unscaled.RDS')
 tohist <- mod[, .(unemprate, perc_foreign, perc_transit_comm, perc_wfh, walkability, 
                  avg_veh, perc_english_only, perc_rent, perc_hs, perc_bach, perc_no_veh,
                  estimate_median_age, estimate_median_hh_income, perc_only_white,
@@ -48,9 +48,10 @@ hist_titles <- c("Unemployment rate", "Percent foreign-born residents", "Percent
                  "Percent of jobs for employees without \ncollege degrees", "Percent of population age 0-25", 
                  "Percent of population age 26-40", "Percent of population age 41-55", "Percent of population age 55+", 
                  "Employment density", "Population density")
+subtitle <- c(rep("By Census Block Group, 2018", 14), rep("By Census Block Group, 2017", 5), rep("By Census Block Group, 2018", 6))
 
 for(i in 1:length(hist_titles)){
-  makeHist(tohist, names(tohist)[i], hist_titles[i])
+  makeHist(tohist, names(tohist)[i], hist_titles[i], subtitle[i])
 }
 
 ## maps
@@ -81,7 +82,37 @@ legend_titles <- c("rate", "percent", "percent", "percent", "area", "average", r
                    "years", "usd", rep("percent", 10), "log \ndensity", "log \ndensity")
 
 for(i in 1:length(map_titles)){
-  makeMap(sfdat, names(tomap)[i], map_titles[i], legend_titles[i])
+  makeMap(sfdat, names(tomap)[i], map_titles[i], legend_titles[i], subtitle[i])
 }
 
-## something a little different for indicators
+## INDICATORS ##
+
+sfdat2 <- dplyr::left_join(bgs, mod)
+
+ggplot(sfdat2) +
+  geom_sf(aes(fill = college), lwd = 0.1) +
+  theme_light() +
+  scale_fill_manual(values = c("gray99", "#33638dff"), labels = c("no", "yes")) +
+  labs(title = "Block groups containing a college or university")
+ggsave("img/map/college.png")
+
+ggplot(sfdat2) +
[…]

### 2020-06-16 report/honors-translation.Rmd
all img to report

@@
 
+![](/Users/raven/Documents/honors/honors-project/img/map/log_pop_density.png){#id .class width=5.75in height=5.11in}
+
 ### Renters
 
-### Median Household Income -> income brackets?
+![](/Users/raven/Documents/honors/honors-project/img/map/perc_rent.png){#id .class width=5.75in height=5.11in}
+
+### Median Household Income
+
+![](/Users/raven/Documents/honors/honors-project/img/map/estimate_median_hh_income.png){#id .class width=5.75in height=5.11in}
 
@@
 
-### Median Age -> age cohort
+![](/Users/raven/Documents/honors/honors-project/img/map/log_emp_dens.png){#id .class width=5.75in height=5.11in}
+
+### Age Cohorts
 
@@
 
+![](/Users/raven/Documents/honors/honors-project/img/map/genz.png){#id .class width=5.75in height=5.11in}
+
 #### Millennials (Age 26-40)
+
+![](/Users/raven/Documents/honors/honors-project/img/map/millenial.png){#id .class width=5.75in height=5.11in}
 
@@
 
+![](/Users/raven/Documents/honors/honors-project/img/map/genx.png){#id .class width=5.75in height=5.11in}
+
 #### Baby Boomers (Age 55+)
+
+![](/Users/raven/Documents/honors/honors-project/img/map/boomer.png){#id .class width=5.75in height=5.11in}
 
@@
 
+![](/Users/raven/Documents/honors/honors-project/img/map/w_perc_jobs_white.png){#id .class width=5.75in height=5.11in}
+
 ### Young Employees
+
+![](/Users/raven/Documents/honors/honors-project/img/map/w_perc_jobs_age_less30.png){#id .class width=5.75in height=5.11in}
 
@@
 
-### Households with No Vehicle -> number of vehicles per household
+![](/Users/raven/Documents/honors/honors-project/img/map/walkability.png){#id .class width=5.75in height=5.11in}
+
+### Average Number of Vehicles per Household
+
+![](/Users/raven/Documents/honors/honors-project/img/map/avg_veh.png){#id .class width=5.75in height=5.11in}
 
@@
 
-### Employment Rate
-
-### Median Gross Rent
-
-### Public Benefits
+![](/Users/raven/Documents/honors/honors-project/img/map/perc_wfh.png){#id .class width=5.75in height=5.11in}
 
@@
 
+![](/Users/raven/Documents/honors/honors-project/img/map/college.png){#id .class width=5.75in height=5.11in}
+
 ### Light Rail (Indicator)
+
+![](/Users/raven/Documents/honors/honors-project/img/map/lightrail.png){#id .class width=5.75in height=5.11in}
 
@@
 
+![](/Users/raven/Documents/honors/honors-project/img/map/airport.png){#id .class width=5.75in height=5.11in}
+
 ### Hospital? (Indicator)
+
+![](/Users/raven/Documents/honors/honors-project/img/map/hospital.png){#id .class width=5.75in height=5.11in}
 

### 2020-06-16 scripts/09_mt-fitting.R
possible fix for rng/ysim issue

@@ -51,10 +51,10 @@ x <- as.matrix(ncol = K, xdat[, `:=` (college = as.numeric(college), hospital = 
 # data for basic poisson
 standat1 <- list(y = y, E = E, N = N, K = K, x = x)
 
-poisson_theta <- "stan/poisson_theta.stan"
+poisson_theta <- "stan/poisson_theta_ysim.stan"
 
-poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 100, verbose = T, seed = 1997)
-saveRDS(poisson_theta_fit, "fits/mt/poisson_theta.RDS")
+poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 100, verbose = T, seed = 1997, init_r = 1/10)
+#saveRDS(poisson_theta_fit, "fits/mt/poisson_theta.RDS")
 
 
 ## bym2 -----------------------------------------

### 2020-06-16 scripts/09_mt-fitting.R
ysim added to bym

@@ -58,7 +58,7 @@ poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 100, verbose = 
 
 
 ## bym2 -----------------------------------------
-bym2 <- "stan/bym2.stan"
+bym2 <- "stan/bym2_ysim.stan"
 counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
 bgs <- block_groups("MN", counties, 2017)
 
@@ -86,5 +86,5 @@ bym_dat <- list(y = y2, E = E2, x = x, K = K,
                 N = N2, node1 = node1, node2 = node2, N_edges = N_edges, scaling_factor = scaling_factor)
 
 bym_fit <- stan(bym2, data = bym_dat, iter = 100, verbose = T, seed = 1997, 
-                control = list(max_treedepth = 12))
+                control = list(max_treedepth = 12), init_r = 1/10)
 #saveRDS(bym_fit, "fits/bym2.RDS")

### 2020-06-17 scripts/09_mt-fitting.R
remove mt specific things

@@ -1,90 +1,0 @@
-## Goal: fit models for mt!
-
-## packages -----------------------------------------------
-packages <- c('rstan', 'dplyr', 'data.table', 'shinystan', 'ggplot2',
-              'bayesplot', 'tigris', 'spdep', 'sf')
-
-miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
-
-if(length(miss_pkgs) > 0){
-  install.packages(miss_pkgs)
-}
-
-invisible(lapply(packages, library, character.only = TRUE))
-
-rm(miss_pkgs, packages)
-
-options(tigris_use_cache = TRUE)
-options(tigris_class = 'sf')
-options(mc.cores = parallel::detectCores())
-rstan_options(autowrite = T)
-
-## data -----------------------------------------
-
-mod_dat <- readRDS('data/modeling-dat/mod_dat.RDS')
-setDT(mod_dat)
-
-vars <- c("emp_density", "w_perc_jobs_white", "w_perc_jobs_age_less30", 
-         "walkability", "avg_veh", "genz", "millenial", "genx", "boomer", 
-         "pop_density", "college", "hospital", "lightrail", "airport", 
-         "perc_wfh", "estimate_median_hh_income")
-
-mod_dat <- na.omit(mod_dat, cols = c("daily_boards", "daily_stops", vars))
-
-xdat <- mod_dat[, ..vars]
-setDT(xdat)
-
-###### FITTING ######
-
-## overdispersed poisson regression -------------
-
-y <- round(mod_dat$daily_boards)
-E <- mod_dat$daily_stops
-
-N <- nrow(mod_dat)
-
-K <- ncol(xdat)
-# ah factors have to be numeric
-x <- as.matrix(ncol = K, xdat[, `:=` (college = as.numeric(college), hospital = as.numeric(hospital),
-                                     airport = as.numeric(airport), lightrail = as.numeric(lightrail))])
-
-# data for basic poisson
-standat1 <- list(y = y, E = E, N = N, K = K, x = x)
-
-poisson_theta <- "stan/poisson_theta_ysim.stan"
-
-poisson_theta_fit <- stan(poisson_theta, data = standat1, iter = 100, verbose = T, seed = 1997, init_r = 1/10)
[…]

### 2020-06-17 scripts/04_get-census.R
a few disclaimers abt reproducibility

@@ -1,6 +1,8 @@
 # Goal: get census variables of interest, 2015-2018
 # I'm sure there are more refined ways to do this 
 
+# some of these were used in honors, others collected later for metro transit use/exploration
+
 ## packages -----------------------------------------------
 
 packages <- c('data.table', 'tigris', 'tidycensus', 'purrr', 'tidyr', 'ggplot2')

### 2020-06-17 scripts/07_add-indicators.R

@@ -1,5 +1,7 @@
 ## Goal: gather some indicator variables to try out
 
+# these were not used at all for the honors project
+
 ## packages -----------------------------------------------
 packages <- c('data.table', 'tigris', 'ggplot2', 'dplyr', 'rgdal', 'sf')
 

### 2020-06-17 scripts/08_join-mod-dat.R

@@ -32,6 +32,8 @@ ridership <- ridership[, .(daily_boards = mean(daily_boards, na.rm = T),
 
 
 ## SCALED ##
+
+## IMPORTANT: to reproduce honors project, use all covariates from 2017!! ##
 cov <- readRDS('data/covariates/cleaned/all_covariates_scaled_ind.RDS')
 setDT(cov)
 

### 2020-06-17 scripts/09_fitting.R

@@ -1,5 +1,7 @@
 ## Goal: fit models!
 
+# if you change 08 to use only 2017 covariates, this doc will reproduce honors results
+
 ## packages -----------------------------------------------
 packages <- c('rstan', 'dplyr', 'data.table', 'shinystan', 'ggplot2',
               'bayesplot', 'tigris', 'spdep', 'sf')

### 2021-01-26 2020-factbook-ideas.Rmd
reorganize 2019 & 2020 files

---
title: "2020 rider factbook ideas"
output:
  MetroTransitr::metro_html: default
  html_document:
    number_sections: no
    theme: paper
    toc: yes
    toc_float: yes
  pdf_document: default
---
<style type="text/css">
html {
  max-width: 10.5in;
  margin: 1.25in 0.75in;
  font-family: "Calibri", helvetica, sans-serif;
  font-size: 11pt;
}

body { width: 8.5in;
       max-width: 10.5in;
}
</style>

```{r setup, include=TRUE}
library(data.table)
knitr::opts_chunk$set(echo = TRUE)
```

```{r data, include = FALSE}
## 2019 data
dat19 <- readRDS("2019/data/modeling-dat/mod_dat_unscaled.RDS")
setDT(dat19)

dat20 <- readRDS("covid/data/modeling-dat/mod_dat_unscaled.RDS")
setDT(dat20)
```

```{r change}
setnames(dat20, "daily_boards", "daily_boards20")
boarding_change <- dat19[, .(GEOID, daily_boards)][dat20[, .(GEOID, daily_boards20)], on = "GEOID"]
boarding_change <- boarding_change[daily_boards != 0 & daily_boards20 != 0]
boarding_change[, change := daily_boards20 - daily_boards]
boarding_change[, perc_change := (change/daily_boards)*100]

avg_perc_change <- mean(boarding_change$perc_change, na.rm = TRUE)
```

```{r renters}
# predictors are the same in both 19 and 20

# get 75th percentile of renters?
#quantile(dat19$perc_rent, c(0.25, 0.5, 0.75, 0.90), na.rm = TRUE)
# 75th percentile rounds to 50% which is nice

renters75 <- dat19[perc_rent >= 0.50] # 447 bgs with ridership
renters50 <- dat19[perc_rent >= .21]
renters90 <- dat19[perc_rent >= .72]

avg_renters_perc_change <- mean(boarding_change[GEOID %in% renters75$GEOID]$perc_change, na.rm = TRUE)

avg_perc_change - avg_renters_perc_change
```

"Ridership in neighborhoods with a high percentage of renters decreased by ~9% less than the average block group"

## "Ridership in neighborhoods with at least 50% renters decreased by ~9% less than the average neighborhood."

high percentage = 75th percentile

```{r vehicles}
quantile(dat19$perc_no_veh, c(0.25, 0.5, 0.75, 0.90), na.rm = TRUE)

veh75 <- dat19[perc_no_veh >= 0.10] 
veh50 <- dat19[perc_no_veh >= .035]
veh90 <- dat19[perc_no_veh >= .21]

avg_veh_perc_change <- mean(boarding_change[GEOID %in% veh75$GEOID]$perc_change, na.rm = TRUE)

avg_perc_change - avg_veh_perc_change
```

"Ridership in neighborhoods with a high percentage of households with no vehicles decreased by ~9% less than the average block group"

## "Ridership in neighborhoods with at least 10% of households with no vehicle decreased by ~9% less than the average neighborhood"

```{r rail}
rail <- dat19[lightrail == 1]
avg_rail_perc_change <- mean(boarding_change[GEOID %in% rail$GEOID]$perc_change, na.rm = TRUE)

avg_perc_change 
avg_rail_perc_change
```
Only 2% less in block groups with rail stations

### 2021-01-26 reports/2019/2019-model-residuals.Rmd
reorganize 2019 & 2020 files

---
title: "2019 ridership model residuals"
author: "Raven McKnight"
date: "1/10/2021"
output:
  MetroTransitr::metro_html: default
---
<style type="text/css">
html {
  max-width: 10.5in;
  margin: 1.25in 0.75in;
  font-family: "Calibri", helvetica, sans-serif;
  font-size: 11pt;
}

body { width: 8.5in;
       max-width: 10.5in;
}
</style>

```{r setup, include=FALSE}
packages <- c("knitr", "sf", "ggplot2", "leaflet", "data.table", "tigris", "dplyr", "scales", 
              "ggnewscale", "htmltools")

miss_pkgs <- packages[!packages %in% installed.packages()[,1]]

if(length(miss_pkgs) > 0){
  install.packages(miss_pkgs)
}

invisible(lapply(packages, library, character.only = TRUE))

rm(miss_pkgs, packages)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r echo = FALSE}
 fit <- readRDS("~/Documents/honors/mt-translation/2019/fits/bym2_disagf.RDS")
bg_info <- readRDS("~/Documents/honors/mt-translation/2019/data/bg_info.RDS")
bg_info_unscaled <- readRDS("~/Documents/honors/mt-translation/2019/data/bg_info_unscaled.RDS")
bg_info_unscaled <- bg_info_unscaled[GEOID %in% bg_info$GEOID]

betas <- readRDS("~/Documents/honors/mt-translation/2019/fits/betas.RDS")
```

```{r echo = FALSE}
## get prediction for each bg using function from shiny app
predictDist <- function(selected_bg, betas, E, inputdat) {

  # model matrix
  mm <- model.matrix(~. , model.frame(inputdat))

  # predictions
  loglambda <- t(sapply(1:nrow(betas), function(x) rowSums(as.matrix(mm) * betas[x, ])))
  attr(loglambda, "dimnames") <- NULL
  loglambdare <- sapply(1:ncol(loglambda), function(x) loglambda[, x]  + unlist(selected_bg$convolved_re)[x])

  f <- t(sapply(1:length(loglambdare), function(x) E * exp(loglambdare[x])))

  y <- t(sapply(1:ncol(f), function(s) rpois(n = 1, lambda = f[s])))

  values <- list(lower = quantile(y, 0.025, na.rm = T),
                 upper = quantile(y, 0.975, na.rm = T),
                 mean = mean(y, na.rm = T),
                 median = median(y, na.rm = T))

  return(list(values = values, y = y))
}

## add predictions to bg info
for(i in 1:nrow(bg_info)){
  pred <- predictDist(selected_bg = bg_info[i,],
                      betas = betas,
                      E = bg_info_unscaled[i,]$daily_stops,
                      inputdat = bg_info[i, 4:14])$values
  bg_info[i, preds := list(pred)]
  bg_info[i, mean_pred := pred$mean]
}

# saveRDS(bg_info, "~/Documents/honors/mt-translation/2019/data/bg_info_pred.RDS")

#bg_info <- readRDS("~/Documents/honors/mt-translation/covid/data/bg_info_pred.RDS")

bg_resids <- bg_info_unscaled[, .(GEOID, daily_boards)][bg_info[, .(GEOID, preds, mean_pred)], on = "GEOID"]
bg_resids[, mean_resid := daily_boards - mean_pred]
bg_resids[, rel_resid := (mean_resid/daily_boards)*100]
```

```{r echo = FALSE, include = FALSE}
counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
bgs <- block_groups("MN", counties)

bg_resids <- dplyr::left_join(bgs, bg_resids, on = GEOID)
```

```{r echo = FALSE}
theme_map <- function(...) {
  theme_minimal() +
  theme(
    text = element_text(family = "Avenir Medium", color = "#22211d"),
    axis.line = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    # panel.grid.minor = element_line(color = "#ebebe5", size = 0.2),
    panel.grid.major = element_line(color = "#ebebe5", size = 0.2),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "#f5f5f2", color = NA), 
    panel.background = element_rect(fill = "#f5f5f2", color = NA), 
    legend.background = element_rect(fill = "#f5f5f2", color = NA),
    panel.border = element_blank(),
    ...
  )
}
```

# Static Map

```{r include = FALSE}
bg_resids <- bg_resids %>%
  mutate(bins = ifelse(rel_resid > 0, "1 - 25%", NA)) %>%
  mutate(bins = ifelse(rel_resid > 25, "26 - 50%", bins)) %>%
  mutate(bins = ifelse(rel_resid > 50, "51 - 75%", bins)) %>%
  mutate(bins = ifelse(rel_resid > 75, "75 - 100%", bins)) %>%
  mutate(bins = ifelse(rel_resid <= 0, "0 - 50%", bins)) %>%
  mutate(bins = ifelse(rel_resid < -50, "51 - 100%", bins)) %>%
  mutate(bins = ifelse(rel_resid < -100, "101 - 250%", bins)) %>%
  mutate(bins = ifelse(rel_resid < -250, "251 - 1000%", bins)) %>%
  mutate(bins = ifelse(rel_resid < -1001, "> 1001%", bins))

cols = c("> 1001%" = "#ff3e22", "251 - 1000%" = "#ff7352", "101 - 250%" = "#ff9e81", 
         "51 - 100%" = "#ffc6b2", "0 - 50%" = "#ffecaa", 
         "1 - 25%" = "#c6c6f7", "26 - 50%" = "#9696e0", "51 - 75%" = "#7e7ecf", 
         "75 - 100%" = "#3a3a98")

## mpls, stp outlines for mapping
url = 'ftp://ftp.gisdata.mn.gov/pub/gdrs/data/pub/us_mn_state_metc/bdry_census2010counties_ctus/shp_bdry_census2010counties_ctus.zip'
loc <- file.path(tempdir(), 'cities.zip')
download.file(url, loc)
unzip(loc, exdir = file.path(tempdir(), 'cities'), overwrite = TRUE)
file.remove(loc)
cities <- st_read(file.path(tempdir(), 'cities'), layer = 'Census2010CountiesAndCTUs', stringsAsFactors = FALSE)
cities <- st_transform(cities, "+proj=longlat +datum=NAD83 +no_defs")

msp <- cities %>% filter(CTU_NAME == "Minneapolis" | CTU_NAME == "St. Paul")
```

```{r}
ggplot() +
  geom_sf(data = bg_resids %>% filter(!is.na(rel_resid) & rel_resid > 0), 
          aes(fill = bins), color = "gray90", lwd = 0.2) +
  theme_map() + 
  scale_fill_manual(values = cols, name = "Positive Residuals\n (Overperforming)") +
  new_scale("fill") +
  geom_sf(data = bg_resids %>% filter(!is.na(rel_resid) & rel_resid < 0), 
          aes(fill = bins), color = "gray90", lwd = 0.2) +
    scale_fill_manual(values = cols, name = "Negative Residuals\n (Underperforming)", 
                      breaks = c("0 - 50%", "51 - 100%", "101 - 250%", 
                                 "251 - 1000%", "> 1001%")) +
  geom_sf(data = msp, fill = "transparent") +
  labs(title = "Relative Residuals of 2020 Ridership Model",
       subtitle = "By Census Block Group")

```

<!-- ```{r echo = FALSE, message = TRUE} -->
<!-- binpal <- colorFactor(cols, bg_resids$bins) -->

<!-- bg_resids_plot <- bg_resids %>% filter(!is.na(rel_resid)) -->

<!-- binlabs <- sprintf( -->
<!--   "<strong> Block Group ID: </strong> %s <br> -->
<!--   <strong> Relative Residual: </strong> %s percent <br><strong> Residual: </strong> %s daily boards <br> -->
<!--   <strong> Observed Daily Boardings: </strong> %s <br><strong> Predicted daily boardings: </strong> %s ", -->
<!--   bg_resids_plot$GEOID, round(bg_resids_plot$rel_resid, 2), round(bg_resids_plot$mean_resid, 2),  -->
<!--   round(bg_resids_plot$daily_boards, 2), round(bg_resids_plot$mean_pred, 2) -->
<!-- ) %>% lapply(htmltools::HTML) -->
<!-- ``` -->

<!-- ```{r message = TRUE, include = TRUE} -->
<!-- leaflet() %>% addProviderTiles(providers$CartoDB.Positron) %>%  -->
<!--   setView(lat = 44.963, lng = -93.22, zoom = 10) %>% -->
<!--   addPolygons(data = bg_resids_plot,  -->
<!--               fillColor = ~binpal(bg_resids_plot$bins), color = 'gray',  -->
<!--               weight = 1, fillOpacity = 0.7, label = binlabs) -->
<!-- ``` -->

### 2021-01-26 reports/2019/honors-translation-techdoc.Rmd
reorganize 2019 & 2020 files

---
title: "A Spatial Model of Bus Ridership: Technical Documentation - text does not match images here"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: Raven McKnight
output:
  MetroTransitr::metro_html: default
  html_document:
    number_sections: no
    theme: paper
    toc: yes
    toc_float: yes
---
<style type="text/css">
html {
  max-width: 10.5in;
  margin: 1.25in 0.75in;
  font-family: "Calibri", helvetica, sans-serif;
  font-size: 11pt;
}

body { width: 8.5in;
       max-width: 10.5in;
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(rstan)
library(ggplot2)
library(bayesplot)
library(data.table)

mod <- readRDS("~/Documents/honors/mt-translation/fits/mt/bym2-newdat.RDS")
mod_dat <- readRDS("~/Documents/honors/mt-translation/data/modeling-dat/mod_dat.RDS")
setDT(mod_dat)

vars <- c("emp_density", "walkability", "perc_no_veh", "age18to34",
          "college", "lightrail", "hospital", "airport",
          "estimate_median_hh_income", "perc_rent",
          "perc_only_white")

mod_dat <- na.omit(mod_dat, cols = c("daily_boards", "daily_stops", vars))
mod_dat <- mod_dat[daily_boards != 0]

xdat <- mod_dat[, ..vars]

y <- ceiling(mod_dat$daily_boards)
```

# Model Specification

[This case study](https://mc-stan.org/users/documentation/case-studies/icar_stan.html) and the accompanying journal article (Morris et al, 2019) offer a really good explanation and demonstration of this model. This is a very quick summary of the model and its parameters.

This model builds upon a typical Bayesian Poisson regression:

$$
\begin{aligned}
Y_i & \sim \text{Poisson}(E_i\lambda_i) \\
log(\lambda_i) & = \beta_0 + \sum_{k=1}^{18}x_{ik}\beta_k + \varepsilon_i\\
\beta_0, \beta_k, \varepsilon_i &\sim \text{Normal}(0, 1)
\end{aligned}
$$
Here, $Y_i$ is ridership (average weekday number of boardings) in block group $i$. $E_i$ is an exposure term (number of times a bus stops in block group $i$ on an average weekday) and $\lambda_i$ is the relative "risk" or likelihood of boardings. The errors $\varepsilon_i$ can account for overdispersion or extra-Poisson variance.

This model, the Besag-York-Mollie, decomposes $\varepsilon_i$ into nonspatial errors $\theta_i$ and spatial errors $\phi_i$. 
$$
log(\lambda_i) = \beta_0 + \sum_{k=1}^{K}x_{ik}\beta_k + \theta_i + \phi_i
$$

When $W$ is an adjacency matrix with entries $w_{ij}$ equal to 1 if block groups $i$ and $j$ share boundaries and 0 otherwise, each $\phi_i$ is defined

$$
\phi_i | \phi_j, j \neq i \sim \text{Normal}\left( \alpha \frac{\sum_{i=1}^{1495}w_{ij}}{d_{ii}}\phi_j, \sigma^2\right)
$$
Parameter $\alpha$ determines the amount of spatial correlation present in $Y$. The BYM simplifies the prior by setting $\alpha = 1$.

Alternatively, we can let matrix $Q$ equal $D - W$ where $W$ is the adjacency matrix and $D$ is a diagonal matrix with entries $d_{ii}$ equal to the number of neighbors block group $i$ has. Then, $\phi$ can be defined

$$
\phi \sim \text{Normal}(0, Q^{-1})
$$

The specific model fit here is actually a reparameterization of the classic BYM called the BYM2. The BYM2 replaces $\theta_i + \phi_i$ as follows:

$$
log(\lambda_i) = \beta_0 + \sum_{k=1}^{K}x_{ik}\beta_k + \left(\sqrt{\frac{\rho}{s}}\theta_i^{*} + \sqrt{1-\rho}\phi_i^{*}\right)\sigma
$$

\noindent where the new parameters $\rho, s, \text{and } \sigma$ are included to aid in sampling from the posterior. While the reparameterization appears significantly more complex than the BYM, the additional parameters do not change the structure of the BYM but rather allow Stan to explore the posterior more efficiently. In summary,

* $\rho \in [0, 1]$ determines how much variance or overdispersion is caused by spatial versus nonspatial error terms. $\rho = 1$ corresponds to all variance being spatial.
* The term $\left(\sqrt{\frac{\rho}{s}}\theta_i^{*} + \sqrt{1-\rho}\phi_i^{*}\right)\sigma$ is a set of convolved random effects. The convolution of $\theta_i$ and $\phi_i$ allows the two sets of random effects to interact more efficiently than in previous parameterizations. This is the primary improvement of the BYM2: the convolution significantly improves computational efficiency and decreases computation time. 
* $\sigma$ is the standard deviation of the entire convolved random effect
* $s$ is a scaling factor such that Var($\theta_i$) $\approx$ Var($\phi_i$) $\approx$ 1, an assumption which is necessary for $\sigma$ to truly be the standard deviation of the error terms.

\noindent In addition to improving sampling efficiency, the BYM2 takes a more "fully Bayesian" approach to hyperprior and hyperparameter selection. We follow the recommendations of Riebler et al (2016) and Morris et al (2019) in setting the following priors:

* $\theta_i \sim$ Normal($0, n)$ where $n$ is the number of connected subgraphs in the neighborhood graph. In many cases, such as modeling all counties in a state, $n=1$ and $\theta_i$ gets a simple Normal($0, 1$) prior. It is possible to fit the BYM2 with $n > 1$, although the computation time is significantly longer when $\sigma$ and $s$ must be calculated for each subgraph. 
* $\sigma \sim$ Normal($0, 1$)
* $\rho \sim$ Beta(1/2, 1/2). This prior is shaped like a horseshoe which suggests that variance will be mostly spatial or mostly nonspatial. This allows the sampler to avoid spending unnecessary time exploring unlikely combinations of $\theta_i$ and $\phi_i$.

# Code

The code used to fit this model is in the Github organization [here](https://github.com/Metropolitan-Council/spatial-ridership-mod). All of the files in the scripts/ folder were used to prepare the data and fit the model. In the stan/ folder, the file `bym2_ysim.stan` was used for this project. The code is below:

```{r}
cat(get_stancode(mod))
```

# Model Performance

For this project, the model was run for 2000 iterations on 4 chains with `max_treedepth = 12`. Here are the parameter estimates: 
```{r}
print(mod, pars = c("beta0", "beta", "sigma", "rho", "theta[5]", "phi[5]", "convolved_re[5]"), 
      probs = c(.025, .975))
```

These estimates correspond to the following covariates:

| Variable | Meaning                                      | Source        |
|----------|----------------------------------------------|---------------|
| $x_1$    | Employment density                           | 2017 LEHD     |
| $x_2$    | Percent of jobs for white employees          | 2017 LEHD     |
| $x_3$    | Percent of jobs for employees under age 30   | 2017 LEHD     |
| $x_4$    | Walkability                                  | OpenStreetMap |
| $x_5$    | Average number of vehicles per household     | 2018 ACS      |
| $x_6$    | Median age                                   | 2018 ACS      |
| $x_7$    | Presence of college campus                   | Homeland Infrastructure Foundation-Level Data (HIFLD)   |
| $x_8$    | Presence of rail station                     | MN Geospatial Commons              |
| $x_9$    | Presence of hospital campus                  | Hennepin County Open Data              |
| $x_{10}$ | Presence of MSP International Airport        | PolicyMap              |
| $x_{11}$ | Population density                           | 2018 ACS      |
| $x_{12}$ | Median household income                      | 2018 ACS      |
| $x_{13}$ | Percent of housing units occupied by renters | 2018 ACS      |

There are some issues with running the model for so few iterations. The effective sample size `n_eff` is very low for $\sigma$ and $\rho$ in particular. 46 transitions also reach maximum treedepth. For my honors thesis, I ran the model for 10000 iterations with `max_treedepth = 15` which dealt with both of those problems. The parameter estimates don't change significantly between 2000 and 10000 iterations but it does take about 5 times longer to run.

### ppcheck

Even with only 2000 iterations, the model fits well:

```{r}
bayesplot::color_scheme_set("purple")

y_sim <- as.matrix(mod, pars = c("y_sim"))

pp_check(y, y_sim[sample(50), ], ppc_dens_overlay) +
  theme_minimal() +
  xlim(0, 200) + ylim(0, 0.025) 
```

### Residuals
```{r}
params <- rstan::extract(mod, c('beta0', 'beta', 'convolved_re'))
E <- mod_dat$daily_stops

mm <- model.matrix(~. , model.frame(xdat), contrasts.arg = lapply(Filter(is.factor,
                  model.frame(xdat)), contrasts, contrasts=FALSE))

loglambda <- suppressWarnings(t(sapply(1:nrow(params$beta), function(x)  rowSums(as.matrix(mm)* params$beta[x,]))))
loglambda0 <- sapply(1:nrow(params$beta0), function(x) params$beta0[x] + loglambda[x, ])
loglambdare <- loglambda0 + t(params$convolved_re)
f <- sapply(1:length(E), function(x) E[x] * loglambdare[x, ])

# posterior predictive distribution
y <- t(sapply(1:nrow(f), function(s) rpois(n = 1, lambda = (f[s, ]))))

y <- vapply(f, function(s) rpois(n = 1, lambda = (f[s, s])), FUN.VALUE = 0)

  # get quantiles etc
values <- data.table(bg = rep(0, ncol(y)), mean = rep(0, ncol(y)), 
                     lower = rep(0, ncol(y)), upper = rep(0, ncol(y)), median = rep(0, ncol(y)))

# i know this could be done with apply but idk how right now
for(i in 1:ncol(y)){
  values[i, bg := i]
  values[i, mean := mean(y[, i], na.rm = T)]
  values[i, lower := quantile(y[, i], 0.025, na.rm = T)]
  values[i, upper := quantile(y[, i], 0.975, na.rm = T)]
  values[i, median := median(y[, i], na.rm = T)]

}

# everything is in the right shape but the values are WAY off lol
head(values)
```

# Sources

Morris, Mitzi, et al. 2019. “Bayesian hierarchical spatial models: Implementing the Besag York Molli ́e model in stan”. Spatial and Spatio-temporal Epidemiology 31 (). issn: 1877-5845, visited on 04/03/2020. doi:10.1016/j.sste.2019.100301. http://www.sciencedirect.com/science/ article/pii/S1877584518301175.

Riebler, Andrea, et al. 2016. “An intuitive Bayesian spatial model for disease mapping that accounts for scaling”. Statistical methods in medical research 25 (4): 1145–1165.

### 2021-01-26 reports/2019/honors-translation.Rmd
reorganize 2019 & 2020 files

---
title: "A Spatial Model of Bus Ridership"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  MetroTransitr::metro_html: default
  pdf_document: default
  html_document:
    number_sections: no
    theme: paper
    toc: yes
    toc_float: yes

---
<style type="text/css">
html {
  max-width: 10.5in;
  margin: 1.25in 0.75in;
  font-family: "Calibri", helvetica, sans-serif;
  font-size: 11pt;
}

body { width: 8.5in;
       max-width: 10.5in;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align="left", out.width = "100%", out.height = "100%")
```

# Introduction

This document explores the results of a spatial model of Metro Transit's bus ridership. The model takes the spatial dimension of bus ridership into account and incorporates demographic predictors. This builds on existing methodology by leveraging more of what we know about ridership and the region to predict where ridership will occur.

The model and analysis presented here provide insight to the likelihood of ridership across space. The analysis is conducted at the Census block group level. Ridership is defined as the average number of weekday boardings per block group in 2018. Predictors are from the 2018 American Community Survey (ACS) 5-year estimates and 2017 Longitudinal Employee Household Dynamic Survey (LEHD). The use of demographic predictors allows us to consider where ridership is likely to occur based on known characteristics of the region.

Bus ridership is inherently spatial. In other words, ridership (or boardings) can only occur at fixed bus stops. However, we intuitively expect neighboring block groups to have similar levels of ridership regardless of specific bus stop placement. For example, we might expect all block groups along the A Line to have similar levels of A Line ridership even though not all of those block groups contain an A Line stop. The model presented here addresses these spatial concerns by introducing spatial autocorrelation and geographic smoothing into the model. In other words, this model smooths estimates between block groups to create more realistic predictions of ridership across space. This spatial component adds nuance to the model and **allows us to see patterns we cannot see in non-spatial models.**

This document outlines key findings from this model of bus ridership. We discuss specific predictors and improvements over non-spatial models. More extensive motivations and methods are available at the end of the document.

# Findings

This model was tested using a wide variety of ACS variables. The model presented here uses thirteen of the most influential variables tested. Seven of these have a positive correlation with ridership. Areas with these seven characteristics have a higher "propensity for ridership" compared to other block groups. Similarly, variables negatively correlated with ridership can be thought of as decreasing an area's propensity for ridership. Below, we describe all thirteen variables and their associations with ridership.

## High Propensity for Ridership

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/perc_rent.png")
```
</div>

<div class = "col-md-4">
<br><br> **Percent renters** is the most influential positive predictor for ridership in this model. This variable tends to be correlated with dense, lower income, and less white areas. Areas with many renters are more tightly clustered around major transit corridors than other density measures, such as employment or population density, making it a powerful predictor of transit ridership.

</div>

</div>

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/log_emp_dens.png")
```
</div>

<div class = "col-md-4">
<br><br>**Employment density** is a quintessential predictor of ridership. Like percent renters, employment density is largely centered on the urban core with decreasing density in inner and outer suburbs. Note that the color scale in the map is logged to account for extreme outliers in downtown Minneapolis and Saint Paul.
</div>

</div>

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/w_perc_jobs_age_less30.png")
```
</div>

<div class = "col-md-4">
<br><br>**Percent of jobs for employees under age 30** is a specific employment characteristic from the LEHD data set. The spatial pattern of this variable is somewhat less clear but the strong association with ridership suggests that we can break employment density down into more nuanced predictors. Including such predictors builds on our prior understanding of employment density's importance in transit modeling. This predictor largely matches our intuition that younger people are more likely to take transit. 
</div>

</div>

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/walkability.png")
```
</div>

<div class = "col-md-4">
<br><br>**Walkability** is another classic predictor of ridership. Here, we define "walkability" as the area of a 10-minute walk isochrone from the population-weighted centroid of each block group. A walk isochrone indicates how far someone can walk along the existing street/sidewalk network in a set amount of time (here, 10 minutes). This predictor is also largely concentrated around the urban core and is slightly overshadowed by the "stronger" predictors above. Note that many rural block groups are missing a walkability value because their population-weighted centroid is far away from the existing street network. For the most part, block groups missing this data are outside of Metro Transit's service area. 
</div>

</div>

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/log_pop_density.png")
```
</div>

<div class = "col-md-4">
<br><br>**Population density** is a prerequisite for transit ridership. In this model, however, population density is somewhat overshadowed by percent renters and employment density. These three predictors are strongly correlated themselves and all represent a similar spatial pattern. Note that the color scale in this map is also logged. 
</div>

</div>

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/college.png")
```
</div>

<div class = "col-md-4">
<br><br> We include block groups containing **college or university campuses** as an indicator or dummy variable. These block groups have significantly higher ridership than we would expect due to the student population.  
</div>

</div>

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/lightrail.png")
```
</div>

<div class = "col-md-4">
<br><br> Block groups containing a **rail station** also have higher ridership than we would otherwise predict. This is likely due to increased transfers to and from rail lines. 
</div>

</div>

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/hospital.png")
```
</div>

<div class = "col-md-4">
<br><br> Finally, block groups with **hospitals** have higher-than-expected ridership. This is particularly notable in light of COVID-19 and Metro Mobility's essential health care worker rides. 
</div>

</div>

## Low Propensity for Ridership

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/estimate_median_hh_income.png")
```
</div>

<div class = "col-md-4">
<br><br>**Median household income** is the greatest negative predictor of ridership in this model. This relationship reflects our previous understanding of Metro Transit's ridership, which tends to be more low-income than the background population. Income is inversely related to key positive predictors such as employment density. 
</div>

</div>

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/w_perc_jobs_white.png")
```
</div>

<div class = "col-md-4">
<br><br>This variable is another example of a specific employment characteristic from the LEHD data set. Block groups with a majority of **jobs for white employees** tend to be outside of the urban core, in the suburbs. This variable is also associated with income levels and transit availability. 
</div>

</div>

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/estimate_median_age.png")
```
</div>

<div class = "col-md-4">
<br><br>Higher **median age** is associated with lower ridership. We expect this relationship based on previous studies of Metro Transit's ridership (Metro Transit, 2019). In general, the urban core is "younger" than the outer suburbs. 
</div>

</div>

<div class = "row">
We can break this predictor down into four generations to get a better sense of the spatial distribution of age in the region: Generation Z (Age 0-25), Millenials (Age 26-40), Generation X (Age 41-55), and Baby Boomers (Age 55+). Millenials are most tightly clustered in the urban core, while Generation X is the most concentrated in outer suburbs. As we might expect, Generation Z and Millenials are associated with higher ridership while Generation X and Baby Boomers are associated with lower ridership. These age cohorts are not included in the model, we map them here to clarify spatial patterns of age.

<div class = "col-md-6">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/genz.png")
```
</div>

\

<div class = "col-md-6">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/millenial.png")
```
</div>
</div>

<div class = "row">
<div class = "col-md-6">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/genx.png")
```
</div>

<div class = "col-md-6">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/boomer.png")
```
</div>

</div>

\

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/avg_veh.png")
```
</div>

<div class = "col-md-4">
<br><br>The proportion of households without vehicles is often used as a predictor for ridership. However, average **number of vehicles per household** is a slightly more informative predictor as it can differentiate between single versus multiple vehicle households. The majority of zero vehicle households are tightly clustered in Downtown Minneapolis or Saint Paul. Suburban households have an average closer to 2 or 3 vehicles. 
</div>

</div>

<div class = "row">

<div class = "col-md-8">
```{r}
knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/airport.png")
```
</div>

<div class = "col-md-4">
<br><br>Somewhat counterintutively, the **MSP International airport** is associated with significantly lower ridership. Recall this model only accounts for *bus* ridership. We would see a positive correlation if this model included light rail. 
</div>

</div>

## Improvement over Non-Spatial Models

A key finding of this analysis is that spatial models allow us to see different, more nuanced patterns than we find with non-spatial models. In other words, the relationship between predictors and ridership change when we account for spatial correlation. Broadly, we see two types of changes:

**1. Predictors which are *less* important in spatial models**  
Predictors which become less important in a spatial model often capture spatial patterns themselves. Population density is a prime example of such a predictor. Population density is included in practically all models of transit ridership because we know ridership can only exist where there is a sufficient density of people. When we use spatial models, we can focus less on "prerequisite" predictors like density because we rely less on them to define spatial patterns.

**2. Predictors which are *more* important in spatial models**  
The explanation for predictors which are more significant in spatial models is slightly more nuanced. These variables are inefficient at predicting ridership over a large area (for example, the 7 County Metro Area) but can be useful in smaller areas with more things held constant. One example is median household income. We might assume that low-income areas will have higher ridership than high-income areas. However, a low-income area far away from transit will likely have lower ridership than a high-income area close to transit. The spatial model presented here allows us to see that higher median income *is* associated with lower ridership, but only when neighboring values are taken into consideration.

Between these two types of change, spatial models can show entirely different patterns than their non-spatial counterparts. Often, the conclusions from spatial models are more consistent with our understanding of ridership (for example, median household income).

<!-- # Implications and Future Work -->

<!-- * include demographics of interest at various spatial scales -->
<!-- * use models like these to leverage demograhpics -->
<!-- * make predictions, understand dynamics -->
<!-- * future worK ?? -->

# Motivation

There are several specific motivations for fitting spatial, demographic models of ridership. First, there is evidence to suggest that demographic predictors beyond simple population density may be informative in predicting bus ridership. Previous studies of Metro Transit's ridership have shown that transit riders tend to be more diverse than the background population in terms of race, income, age, and so on (Metro Transit, 2019). While it is challenging to collect demographics from individual bus riders, we can leverage demographic characteristics of the Twin Cities to predict areas of high ridership.

Including spatial information has other benefits. One primary goal of spatial modeling is to conduct geographic smoothing. In existing methods, such as the process used to determine Transit Market Areas, a model is fit and used to make predictions which are later smoothed using Geographic Information Systems (GIS). This method necessarily sacrifices some granularity in predictions. The model presented here integrates spatial information into the structure of the model itself. This may be particularly helpful in the case of bus ridership data because boardings can only occur at existing bus stops. Geographic smoothing via spatial models can help make predictions in block groups with no existing bus stops and avoid overfitting outliers.

Census geography itself is a limitation in modeling bus ridership. Geographies such as block groups tend to be bounded by major roadways. Unfortunately, bus routes tend to lie on those same roadways. Because of these bounds, bus stops may be split into block groups counterintuitively. For example, bus stops on either side of the same street traveling in opposite directions are often aggregated into different block groups. The full implications of this limitation are not explored in this study, but some work has been done to define "transit-specific geographies" which could alleviate this problem (Viggiano, 2017). In this study, we use spatial smoothing via modeling to mitigate bounding issues.

In short, the motivations for this type of modeling are (1) to leverage more of what we know about ridership and our region and (2) to conduct geographic smoothing to handle known challenges with transit data.

<!-- Complex statistical models are likely more appropriate for internal use. There is still a place for simple modeles like the TMA model for communicating with the public.  -->

# Methods

This document provides a brief outline of the methodology used in this analysis. More detailed descriptions of the general methodology are available in Morris et al (2019) and Riebler et al (2016).

The model used in this analysis is a *spatial Bayesian* model. Bayesian models incorporate prior information and produce distributions rather than point estimates as predictions. As described above, the spatial component is included to perform spatial smoothing. Specifically, the spatial aspect of this model accounts for spatial autocorrelation and adjusts predictions in individual block groups based on values in neighboring block groups.

The specific model used is a reparameterized Besag-York-Mollie (BYM) model. The BYM has similar structure to a classic Bayesian Poisson regression wherein the response variable is assumed to follow a Poisson distribution determined by a set of predictors:

$$
Y_i \sim \text{Poisson}(\beta_1 x_1 + \beta_2x_2 + \ldots + \beta_n x_n + \varepsilon_i)
$$

In this notation, $Y_i$ is the response variable (ridership) in block group $i$. Each $x$ is a predictor (for example, population density), each $\beta$ is the corresponding coefficient, and $\varepsilon_i$ is an error term for each block group. Above,"non-spatial model" refers to this type of Poisson regression. The BYM replaces $\varepsilon_i$ with two terms: non-spatial errors $\theta_i$ and spatial errors $\phi_i$ such that

$$
Y_i \sim \text{Poisson}(\beta_1 x_1 + \beta_2x_2 + \ldots + \beta_n x_n + \theta_i + \phi_i)
$$

The term $\phi_i$ is determined based on an adjacency matrix $W$ which encodes spatial relationships between block groups. For block groups $i$ and $j$, entry $w_{ij} = 1$ if block groups $i$ and $j$ share a boundary and $0$ if they do not. This is how spatial information comes into the model and adjusts predictions based on neighbors. More on this matrix and the definition of $\phi_i$ is available in the resources cited above.

The specific model used in this analysis is a reparameterization of the classic BYM model above called the BYM2. The BYM2 retains the structure of the BYM but is significantly more computationally efficient to fit. The model is fit using Stan via the `rstan` R package.

# Citations

Metro Transit. 2019. “Transit System Performance Evaluation”. https://metrocouncil.org/ Transportation/Publications-And-Resources/Transit/2019-Transit-System-Performance- Evaluation.aspx.

Morris, Mitzi, et al. 2019. “Bayesian hierarchical spatial models: Implementing the Besag York Molli ́e model in stan”. Spatial and Spatio-temporal Epidemiology 31 (). issn: 1877-5845, visited on 04/03/2020. doi:10.1016/j.sste.2019.100301. http://www.sciencedirect.com/science/ article/pii/S1877584518301175.

Riebler, Andrea, et al. 2016. “An intuitive Bayesian spatial model for disease mapping that accounts for scaling”. Statistical methods in medical research 25 (4): 1145–1165.

Viggiano, Cecilia A. (Cecilia Ann). 2017. “Bus network sketch planning with origin-destination travel data”. Thesis, Massachusetts Institute of Technology. Visited on 04/03/2020. https: //dspace.mit.edu/handle/1721.1/111441.

### 2021-01-26 reports/2019/spatial-direct-demand-2019analysis.Rmd
reorganize 2019 & 2020 files

--- 2021-01-26 reports/2019/honors-translation.Rmd
@@
 ---
-title: "A Spatial Model of Bus Ridership"
-date: "`r format(Sys.time(), '%d %B %Y')`"
+title: "A Spatial Model Direct Demand Model: 2019 Analysis"
 output:
   MetroTransitr::metro_html: default
-  pdf_document: default
   html_document:
@@
     toc_float: yes
-
+  pdf_document: default
 ---
@@
 ```{r setup, include=FALSE}
+library(knitr)
+library(shiny)
+library(slickR)
 knitr::opts_chunk$set(echo = FALSE, fig.align="left", out.width = "100%", out.height = "100%")
@@
 
-This document explores the results of a spatial model of Metro Transit's bus ridership.
+This document explores the results of a spatial direct demand model of Metro Transit's bus ridership.
 The model takes the spatial dimension of bus ridership into account and incorporates demographic predictors.
@@
 
-The model and analysis presented here provide insight to the likelihood of ridership across space.
+The model and analysis presented here provide insight to the likelihood of ridership across space given the existing network.
 The analysis is conducted at the Census block group level.
@@
 The model presented here addresses these spatial concerns by introducing spatial autocorrelation and geographic smoothing into the model.
-In other words, this model smooths estimates between block groups to create more realistic predictions of ridership across space.
+In other words, this model smooths estimates between block groups to create more stable and realistic predictions of ridership across space.
 This spatial component adds nuance to the model and **allows us to see patterns we cannot see in non-spatial models.**
@@
 
-This model was tested using a wide variety of ACS variables.
-The model presented here uses thirteen of the most influential variables tested.
-Seven of these have a positive correlation with ridership.
-Areas with these seven characteristics have a higher "propensity for ridership" compared to other block groups.
-Similarly, variables negatively correlated with ridership can be thought of as decreasing an area's propensity for ridership.
-Below, we describe all thirteen variables and their associations with ridership.
+This model was tested with a wide variety of ACS variables.
+The model presented here uses eleven of the most influential variables tested.
+Five of these demographic predictors have a positive correlation with ridership.
+Areas with these characteristics have a higher "propensity for ridership" compared to other block groups.
+Similarly, the two demographic variables negatively correlated with ridership can be thought of as decreasing an area's propensity for ridership.
+Below, we describe all eleven variables, including four indicator variables, and their associations with ridership.
+A table of coefficient estimates is available after this discussion.
+
+##  Map of ridership
+
+<div class = "row">
+
+<div class = "col-md-8">
+```{r}
+knitr::include_graphics("/Users/raven/Documents/honors/mt-translation/precovid/img/map/boardings.png")
+```
+</div>
+</div>
 
@@
 ```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/perc_rent.png")
+knitr::include_graphics("/Users/raven/Documents/honors/mt-translation/precovid/img/map/perc_rent.png")
 ```
@@
 <div class = "col-md-4">
-<br><br> **Percent renters** is the most influential positive predictor for ridership in this model.
+<br><br> **Percent renters** is the most influential positive predictor of ridership in this model.
+It is more than two times as significant as the next most important predictor.
 This variable tends to be correlated with dense, lower income, and less white areas.
 Areas with many renters are more tightly clustered around major transit corridors than other density measures, such as employment or population density, making it a powerful predictor of transit ridership.
+This variable is likely associated with age, income, student status, and so on.
 
@@
 ```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/log_emp_dens.png")
+knitr::include_graphics("/Users/raven/Documents/honors/mt-translation/precovid/img/map/log_emp_dens.png")
 ```
@@
 <div class = "col-md-4">
-<br><br>**Employment density** is a quintessential predictor of ridership.
+<br><br>**Employment density** is a classic predictor of ridership.
 Like percent renters, employment density is largely centered on the urban core with decreasing density in inner and outer suburbs.
@@
 ```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/w_perc_jobs_age_less30.png")
-```
-</div>
-
-<div class = "col-md-4">
-<br><br>**Percent of jobs for employees under age 30** is a specific employment characteristic from the LEHD data set.
-The spatial pattern of this variable is somewhat less clear but the strong association with ridership suggests that we can break employment density down into more nuanced predictors.
-Including such predictors builds on our prior understanding of employment density's importance in transit modeling.
-This predictor largely matches our intuition that younger people are more likely to take transit.
-</div>
-
-</div>
-
-<div class = "row">
-
-<div class = "col-md-8">
-```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/walkability.png")
+knitr::include_graphics("/Users/raven/Documents/honors/mt-translation/precovid/img/map/walkability.png")
 ```
@@
 <br><br>**Walkability** is another classic predictor of ridership.
+The predictor is only slightly less significant than employment density.
 Here, we define "walkability" as the area of a 10-minute walk isochrone from the population-weighted centroid of each block group.
@@
 <div class = "col-md-8">
+
 ```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/log_pop_density.png")
+imgs <- list.files("/Users/raven/Documents/honors/mt-translation/precovid/img/map", pattern="age", full.names = TRUE)
+slickR(imgs, height = 500)
 ```
+
 </div>
@@
 <div class = "col-md-4">
-<br><br>**Population density** is a prerequisite for transit ridership.
-In this model, however, population density is somewhat overshadowed by percent renters and employment density.
-These three predictors are strongly correlated themselves and all represent a similar spatial pattern.
-Note that the color scale in this map is also logged.
+<br><br>**Percent of residents age 18-34** is equally as significant as walkability.
+This predictor reflects previous studies and surveys which show that this age group is over-represented in Metro Transit's riders.
+Note that the map to the left shows this predictor logged for visual clarity.
+Use the arrows to see raw values.
 </div>
@@
 <div class = "col-md-8">
+
 ```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/college.png")
+imgs2 <- list.files("/Users/raven/Documents/honors/mt-translation/precovid/img/map", pattern="veh", full.names = TRUE)
+slickR(imgs2, height = 500)
 ```
@@
 <div class = "col-md-4">
-<br><br> We include block groups containing **college or university campuses** as an indicator or dummy variable.
-These block groups have significantly higher ridership than we would expect due to the student population.
-</div>
-
-</div>
-
-<div class = "row">
-
-<div class = "col-md-8">
-```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/lightrail.png")
-```
-</div>
-
-<div class = "col-md-4">
-<br><br> Block groups containing a **rail station** also have higher ridership than we would otherwise predict. This is likely due to increased transfers to and from rail lines. 
-</div>
-
-</div>
-
-<div class = "row">
-
-<div class = "col-md-8">
-```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/hospital.png")
-```
-</div>
-
-<div class = "col-md-4">
-<br><br> Finally, block groups with **hospitals** have higher-than-expected ridership. This is particularly notable in light of COVID-19 and Metro Mobility's essential health care worker rides. 
+<br><br>**Percent of households without a vehicle** is slightly less significant than the age variable above.
+This is another classic predictor of ridership.
+As above, this map is logged for clarity.
+Click to see raw values.
 </div>
@@
 ```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/estimate_median_hh_income.png")
+knitr::include_graphics("/Users/raven/Documents/honors/mt-translation/precovid/img/map/estimate_median_hh_income.png")
 ```
@@
 <br><br>**Median household income** is the greatest negative predictor of ridership in this model.
-This relationship reflects our previous understanding of Metro Transit's ridership, which tends to be more low-income than the background population.
+It is almost twice as significant as the next most important predictor.
+The negative relationship here reflects our previous understanding of Metro Transit's ridership, which tends to be more low-income than the background population.
 Income is inversely related to key positive predictors such as employment density.
@@
 ```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/w_perc_jobs_white.png")
+knitr::include_graphics("/Users/raven/Documents/honors/mt-translation/precovid/img/map/perc_only_white.png")
 ```
@@
 <div class = "col-md-4">
-<br><br>This variable is another example of a specific employment characteristic from the LEHD data set.
-Block groups with a majority of **jobs for white employees** tend to be outside of the urban core, in the suburbs.
-This variable is also associated with income levels and transit availability.
+<br><br>**Percent of residents identified as white alone** by the ACS is the only other negative predictor in this model.
+This is another predictor derived from previoius studies of Metro Transit's ridership which show that transit riders tend to be more racially diverse than the background population.
+</div>
+
+</div>
+
+## Indicators
+
+This model also includes four indicator variables.
+These variables loosely represent special generators and account for "unusual" block groups.
+The following indicators are ordered by their significance.
+
+<div class = "row">
+
+<div class = "col-md-8">
+```{r}
+knitr::include_graphics("/Users/raven/Documents/honors/mt-translation/precovid/img/map/airport.png")
+```
+</div>
+
+<div class = "col-md-4">
+<br><br><br><br><br>The **MSP International** airport has the largest positive affect on ridership.
+This indicator is used in most models of ridership, including the model underpinning the Metropolitan Council's Transit Market Areas.
 </div>
@@
 ```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/estimate_median_age.png")
+knitr::include_graphics("/Users/raven/Documents/honors/mt-translation/precovid/img/map/rail.png")
 ```
@@
 <div class = "col-md-4">
-<br><br>Higher **median age** is associated with lower ridership.
-We expect this relationship based on previous studies of Metro Transit's ridership (Metro Transit, 2019).
-In general, the urban core is "younger" than the outer suburbs.
-</div>
-
-</div>
-
-<div class = "row">
-We can break this predictor down into four generations to get a better sense of the spatial distribution of age in the region: Generation Z (Age 0-25), Millenials (Age 26-40), Generation X (Age 41-55), and Baby Boomers (Age 55+).
-Millenials are most tightly clustered in the urban core, while Generation X is the most concentrated in outer suburbs.
-As we might expect, Generation Z and Millenials are associated with higher ridership while Generation X and Baby Boomers are associated with lower ridership.
-These age cohorts are not included in the model, we map them here to clarify spatial patterns of age.
-
-<div class = "col-md-6">
-```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/genz.png")
-```
-</div>
-
-\
-
-<div class = "col-md-6">
-```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/millenial.png")
-```
-</div>
-</div>
-
-<div class = "row">
-<div class = "col-md-6">
-```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/genx.png")
-```
-</div>
-
-<div class = "col-md-6">
-```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/boomer.png")
-```
-</div>
-
-</div>
-
-\
-
-<div class = "row">
-
-<div class = "col-md-8">
-```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/avg_veh.png")
-```
-</div>
-
-<div class = "col-md-4">
-<br><br>The proportion of households without vehicles is often used as a predictor for ridership.
-However, average **number of vehicles per household** is a slightly more informative predictor as it can differentiate between single versus multiple vehicle households.
-The majority of zero vehicle households are tightly clustered in Downtown Minneapolis or Saint Paul.
-Suburban households have an average closer to 2 or 3 vehicles.
+<br><br><br><br><br> Block groups containing a **rail station** also have higher ridership than we would otherwise predict. This is likely due to increased transfers to and from rail lines. 
 </div>
@@
 ```{r}
-knitr::include_graphics("/Users/raven/Documents/honors/honors-project/img/map/airport.png")
+knitr::include_graphics("/Users/raven/Documents/honors/mt-translation/precovid/img/map/college.png")
 ```
@@
 <div class = "col-md-4">
-<br><br>Somewhat counterintutively, the **MSP International airport** is associated with significantly lower ridership.
-Recall this model only accounts for *bus* ridership.
-We would see a positive correlation if this model included light rail.
+<br><br><br><br><br> We include block groups containing **college or university campuses** as an indicator or dummy variable.
+These block groups have significantly higher ridership than we would expect due to the student population.
+</div>
+
+</div>
+
+<div class = "row">
+
+<div class = "col-md-8">
+```{r}
+knitr::include_graphics("/Users/raven/Documents/honors/mt-translation/precovid/img/map/hospital.png")
+```
+</div>
+
+<div class = "col-md-4">
+<br><br><br><br><br> Finally, block groups with **hospitals** have higher-than-expected ridership.
+This is particularly notable in light of COVID-19 and Metro Mobility's essential health care worker rides.
 </div>
@@
 Population density is included in practically all models of transit ridership because we know ridership can only exist where there is a sufficient density of people.
+However, in the model presented here, when population density is included as a predictor, it receives a coefficient estimate of approximately zero.
 When we use spatial models, we can focus less on "prerequisite" predictors like density because we rely less on them to define spatial patterns.
@@
 However, a low-income area far away from transit will likely have lower ridership than a high-income area close to transit.
-The spatial model presented here allows us to see that higher median income *is* associated with lower ridership, but only when neighboring values are taken into consideration.
+The spatial model presented here allows us to see that higher median income *is* associated with lower ridership, but only when spatial characteristics are taken into consideration.
 
-Between these two types of change, spatial models can show entirely different patterns than their non-spatial counterparts.
+A third, more minor change of note, is that indicator variables become more signiciant in spatial models because geographic smoothing encourages underestimates in "unusual" block groups.
+
+Between these types of change, spatial models can show entirely different patterns than their non-spatial counterparts.
 Often, the conclusions from spatial models are more consistent with our understanding of ridership (for example, median household income).
 
-<!-- # Implications and Future Work -->
+# Coefficient Estimates
 
-<!-- * include demographics of interest at various spatial scales -->
-<!-- * use models like these to leverage demograhpics -->
-<!-- * make predictions, understand dynamics -->
-<!-- * future worK ?? -->
+The table below shows complete parameter estimates from a non-spatial model ($\hat\beta_{ns}$) **and** it's spatial counterpart ($\hat\beta_s$).
+
+| Variable | Meaning                                        | $\hat\beta_{ns}$    |$\hat\beta_s$    |
+|----------|------------------------------------------------|---------------------|-----------------|
+| $x_1$    | Employment density                             | 0.08                | 0.07            |
+| $x_2$    | Walkability                                    | 0.07                | 0.06            |
+| $x_3$    | Percent of households with no vehicle          | 0.09                | 0.04            |
+| $x_4$    | Percent of residents aged 18 to 34             | 0.09                | 0.06            |
+| $x_5$    | Presence of college or university campus       | 0.19                | 0.32            |
+| $x_6$    | Presence of rail station                       | 0.15                | 0.36            |
+| $x_7$    | Presence of hospital campus                    | 0.11                | 0.23            |
+| $x_8$    | Presence of MSP International Airport          | 0.04                | 0.52            |
+| $x_9$    | Median household income                        | -0.04               | -0.10           |
+| $x_{10}$ | Percent of housing units occupied by renters   | 0.17                | 0.15            |
+| $x_{11}$ | Percent of residents identified as white alone | -0.06               | -0.06           |
 
@@
 
-In short, the motivations for this type of modeling are (1) to leverage more of what we know about ridership and our region and (2) to conduct geographic smoothing to handle known challenges with transit data.
-
-<!-- Complex statistical models are likely more appropriate for internal use. There is still a place for simple modeles like the TMA model for communicating with the public.  -->
+**In short, the motivations for this type of modeling are (1) to leverage more of what we know about ridership and our region and (2) to conduct geographic smoothing to handle known challenges with transit data.**
 
@@
 $$
-Y_i \sim \text{Poisson}(\beta_1 x_1 + \beta_2x_2 + \ldots + \beta_n x_n + \varepsilon_i)
+\begin{align}
+Y_i &\sim \text{Poisson}(E_i \lambda_i) \\
+\text{log}(\lambda_i) &= \beta_1 x_1 + \beta_2x_2 + \ldots + \beta_n x_n + \varepsilon_i)
+\end{align}
 $$
@@
 In this notation, $Y_i$ is the response variable (ridership) in block group $i$.
-Each $x$ is a predictor (for example, population density), each $\beta$ is the corresponding coefficient, and $\varepsilon_i$ is an error term for each block group.
+Each $x$ is a predictor (for example, employment density), each $\beta$ is the corresponding coefficient, and $\varepsilon_i$ is an error term for each block group.
 Above,"non-spatial model" refers to this type of Poisson regression.
@@
 $$
-Y_i \sim \text{Poisson}(\beta_1 x_1 + \beta_2x_2 + \ldots + \beta_n x_n + \theta_i + \phi_i)
+\begin{align}
+Y_i &\sim \text{Poisson}(E_i \lambda_i) \\
+\text{log}(\lambda_i) &= \beta_1 x_1 + \beta_2x_2 + \ldots + \beta_n x_n + \theta_i + \phi_i)
+\end{align}
 $$

### 2021-01-26 reports/covid/covid-model-residuals.Rmd
reorganize 2019 & 2020 files

---
title: "COVID ridership model residuals"
author: "Raven McKnight"
output:
  MetroTransitr::metro_html: default
---
<style type="text/css">
html {
  max-width: 10.5in;
  margin: 1.25in 0.75in;
  font-family: "Calibri", helvetica, sans-serif;
  font-size: 11pt;
}

body { width: 8.5in;
       max-width: 10.5in;
}
</style>

```{r setup, include=FALSE}
packages <- c("knitr", "sf", "ggplot2", "leaflet", "data.table", "tigris", "dplyr", "scales", 
              "ggnewscale", "htmltools")

miss_pkgs <- packages[!packages %in% installed.packages()[,1]]

if(length(miss_pkgs) > 0){
  install.packages(miss_pkgs)
}

invisible(lapply(packages, library, character.only = TRUE))

rm(miss_pkgs, packages)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r echo = FALSE}
# fit <- readRDS("fits/covid_bymfit_nov25.RDS")
bg_info <- readRDS("~/Documents/honors/mt-translation/covid/data/bg_info.RDS")
bg_info_unscaled <- readRDS("~/Documents/honors/mt-translation/covid/data/bg_info_unscaled.RDS")
bg_info_unscaled <- bg_info_unscaled[GEOID %in% bg_info$GEOID]

# betas <- readRDS("fits/betas.RDS")
```

```{r echo = FALSE}
## get prediction for each bg using function from shiny app
predictDist <- function(selected_bg, betas, E, inputdat) {

  # model matrix
  mm <- model.matrix(~. , model.frame(inputdat))

  # predictions
  loglambda <- t(sapply(1:nrow(betas), function(x) rowSums(as.matrix(mm) * betas[x, ])))
  attr(loglambda, "dimnames") <- NULL
  loglambdare <- sapply(1:ncol(loglambda), function(x) loglambda[, x]  + unlist(selected_bg$convolved_re)[x])

  f <- t(sapply(1:length(loglambdare), function(x) E * exp(loglambdare[x])))

  y <- t(sapply(1:ncol(f), function(s) rpois(n = 1, lambda = f[s])))

  values <- list(lower = quantile(y, 0.025, na.rm = T),
                 upper = quantile(y, 0.975, na.rm = T),
                 mean = mean(y, na.rm = T),
                 median = median(y, na.rm = T))

  return(list(values = values, y = y))
}

## add predictions to bg info
# for(i in 1:nrow(bg_info)){
#   pred <- predictDist(selected_bg = bg_info[i,],
#                       betas = betas,
#                       E = bg_info_unscaled[i,]$daily_stops,
#                       inputdat = bg_info[i, 4:14])$values
#   bg_info[i, preds := list(pred)]
#   bg_info[i, mean_pred := pred$mean]
# }

# saveRDS(bg_info, "bg_info_pred.RDS")

bg_info <- readRDS("~/Documents/honors/mt-translation/covid/data/bg_info_pred.RDS")

bg_resids <- bg_info_unscaled[, .(GEOID, daily_boards)][bg_info[, .(GEOID, preds, mean_pred)], on = "GEOID"]
bg_resids[, mean_resid := daily_boards - mean_pred]
bg_resids[, rel_resid := (mean_resid/daily_boards)*100]
```

```{r echo = FALSE, include = FALSE}
counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
bgs <- block_groups("MN", counties)

bg_resids <- dplyr::left_join(bgs, bg_resids, on = GEOID)

bg_resids_plot <- bg_resids %>% filter(!is.na(rel_resid))
```

```{r echo = FALSE}
theme_map <- function(...) {
  theme_minimal() +
  theme(
    text = element_text(family = "Avenir Medium", color = "#22211d"),
    axis.line = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    # panel.grid.minor = element_line(color = "#ebebe5", size = 0.2),
    panel.grid.major = element_line(color = "#ebebe5", size = 0.2),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "#f5f5f2", color = NA), 
    panel.background = element_rect(fill = "#f5f5f2", color = NA), 
    legend.background = element_rect(fill = "#f5f5f2", color = NA),
    panel.border = element_blank(),
    ...
  )
}
```

# Introduction

These maps show the *residuals* from a spatial Bayesian direct demand model of 2020 bus ridership. In both maps, positive (blue) residuals represent areas where ridership is *higher* than the model predicts and negative (red) residuals represent areas where ridership is lower. These residuals exist after taking all of the predictors in Table 1 into account.

Table: Predictors used in ridership model

| Variable | Meaning                                        | coefficient     |
|----------|------------------------------------------------|-----------------|
| $x_1$    | Employment density                             | 0.06            |
| $x_2$    | Walkability                                    | 0.03            |
| $x_3$    | Percent of households with no vehicle          | 0.06            |
| $x_4$    | Percent of residents aged 18 to 34             | 0.04            |
| $x_5$    | Presence of college or university campus       | 0.08            |
| $x_6$    | Presence of rail station                       | 0.52            |
| $x_7$    | Presence of hospital campus                    | 0.00            |
| $x_8$    | Presence of MSP International Airport          | 0.00            |
| $x_9$    | Median household income                        | -0.20           |
| $x_{10}$ | Percent of housing units occupied by renters   | 0.15            |
| $x_{11}$ | Percent of residents identified as white alone | -0.07           |

# Static Map

```{r include = FALSE}
# bg_resids <- bg_resids %>%
#   mutate(bins = ifelse(rel_resid > 0, "1 - 25%", NA)) %>%
#   mutate(bins = ifelse(rel_resid > 25, "26 - 50%", bins)) %>%
#   mutate(bins = ifelse(rel_resid > 50, "51 - 75%", bins)) %>%
#   mutate(bins = ifelse(rel_resid > 75, "75 - 100%", bins)) %>%
#   mutate(bins = ifelse(rel_resid <= 0, "0 - 50%", bins)) %>%
#   mutate(bins = ifelse(rel_resid < -50, "51 - 100%", bins)) %>%
#   mutate(bins = ifelse(rel_resid < -100, "101 - 250%", bins)) %>%
#   mutate(bins = ifelse(rel_resid < -250, "251 - 1000%", bins)) %>%
#   mutate(bins = ifelse(rel_resid < -1001, "> 1001%", bins))
#   
# cols = c("> 1001%" = "#ff3e22", "251 - 1000%" = "#ff7352", "101 - 250%" = "#ff9e81", 
#          "51 - 100%" = "#ffc6b2", "0 - 50%" = "#ffecaa", 
#          "1 - 25%" = "#c6c6f7", "26 - 50%" = "#9696e0", "51 - 75%" = "#7e7ecf", 
#          "75 - 100%" = "#3a3a98")

## mpls, stp outlines for mapping
url = 'ftp://ftp.gisdata.mn.gov/pub/gdrs/data/pub/us_mn_state_metc/bdry_census2010counties_ctus/shp_bdry_census2010counties_ctus.zip'
loc <- file.path(tempdir(), 'cities.zip')
download.file(url, loc)
unzip(loc, exdir = file.path(tempdir(), 'cities'), overwrite = TRUE)
file.remove(loc)
cities <- st_read(file.path(tempdir(), 'cities'), layer = 'Census2010CountiesAndCTUs', stringsAsFactors = FALSE)
cities <- st_transform(cities, "+proj=longlat +datum=NAD83 +no_defs")

msp <- cities %>% filter(CTU_NAME == "Minneapolis" | CTU_NAME == "St. Paul")
```

```{r}
# 
# ggplot() +
#   geom_sf(data = bg_resids %>% filter(!is.na(rel_resid) & rel_resid > 0), 
#           aes(fill = bins), color = "gray90", lwd = 0.2) +
#   theme_map() + 
#   scale_fill_manual(values = cols, name = "Positive Residuals\n (Overperforming)") +
#   new_scale("fill") +
#   geom_sf(data = bg_resids %>% filter(!is.na(rel_resid) & rel_resid < 0), 
#           aes(fill = bins), color = "gray90", lwd = 0.2) +
#     scale_fill_manual(values = cols, name = "Negative Residuals\n (Underperforming)", 
#                       breaks = c("0 - 50%", "51 - 100%", "101 - 250%", 
#                                  "251 - 1000%", "> 1001%")) +
#   geom_sf(data = msp, fill = "transparent") +
#   labs(title = "Relative Residuals of 2020 Ridership Model",
#        subtitle = "By Census Block Group")
# 
```

```{r}
ggplot() +
  geom_sf(data = bg_resids_plot, aes(fill = mean_resid), 
          lwd = 0.2, color = "gray90") +
  scale_fill_fermenter(breaks = c(-1000, -500, -100, -50, 0,
                               50, 100, 500, 1000), 
                       palette = "RdBu", direction = 1, name = "Residuals") +
  geom_sf(data = msp, fill = "transparent") +
  theme_map() + 
  labs(title = "Residuals of 2020 Ridership Model",
       subtitle = "By Census Block Group")
```

```{r}
# ggplot() +
#   geom_sf(data = bg_resids_plot, aes(fill = rel_resid), lwd = 0) +
#   scale_fill_fermenter(breaks = c(-1000, -250, - 100,-50, 0, 
#                                   25, 50, 75), 
#                        palette = "RdBu", direction = 1, name = "Relative Residuals") +
#   theme_map() + 
#   labs(title = "Relative Residuals of 2020 Ridership Model",
#        subtitle = "By Census Block Group")
```

# Interactive Map

```{r echo = FALSE, message = TRUE}
binpal <- colorBin(bins = c(-24648, -1000, -500, -100, -50, 0, 50, 100, 500, 1000, 2405),
                   domain = bg_resids_plot$mean_resid, 
                   palette = "RdBu")

binlabs <- sprintf(
  "<strong> Block Group ID: </strong> %s <br>
  <strong> Observed Daily Boardings: </strong> %s <br><strong> Predicted daily boardings: </strong> %s
  <br><strong> Residual: </strong> %s daily boards <br> <strong> Relative Residual: </strong> %s percent 
   ",
  bg_resids_plot$GEOID, round(bg_resids_plot$daily_boards, 2), round(bg_resids_plot$mean_pred, 2), 
  round(bg_resids_plot$mean_resid, 2), round(bg_resids_plot$rel_resid, 2)
) %>% lapply(htmltools::HTML)
```

```{r message = TRUE, include = TRUE}
leaflet() %>% addProviderTiles(providers$CartoDB.Positron) %>% 
  setView(lat = 44.963, lng = -93.22, zoom = 10) %>%
  addPolygons(data = bg_resids_plot, 
              fillColor = ~binpal(bg_resids_plot$mean_resid), color = 'gray', 
              weight = 1, fillOpacity = 0.8, label = binlabs)
```

### 2021-01-26 reports/covid/spatial-direct-demand-covid-analysis.Rmd
reorganize 2019 & 2020 files

---
title: "A Spatial Direct Demand Model: Covid Analysis"
output:
  MetroTransitr::metro_html: default
  html_document:
    number_sections: no
    theme: paper
    toc: yes
    toc_float: yes
  pdf_document: default
---
<style type="text/css">
html {
  max-width: 10.5in;
  margin: 1.25in 0.75in;
  font-family: "Calibri", helvetica, sans-serif;
  font-size: 11pt;
}

body { width: 8.5in;
       max-width: 10.5in;
}
</style>

```{r setup, include=FALSE}
library(knitr)
library(sf)
library(ggplot2)
library(gridExtra)
library(data.table)
library(dplyr)
library(tidyr)
library(bayesplot)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align="left", out.width = "100%", out.height = "100%")
```

```{r include = FALSE}
fit <- readRDS("/Users/raven/Documents/honors/mt-translation/covid/fits/covid_bymfit_nov25.RDS")
bg_dat <- readRDS('/Users/raven/Documents/honors/mt-translation/covid/data/modeling-dat/mod_dat_unscaled.RDS')

bgs <- tigris::block_groups("MN", c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington"))

bg_dat <- dplyr::left_join(bgs, bg_dat, on = GEOID)

bg_dat19 <- readRDS('/Users/raven/Documents/honors/mt-translation/2019/data/modeling-dat/mod_dat_unscaled.RDS')
bg_dat19 <- dplyr::left_join(bgs, bg_dat19, on = GEOID)
```

```{r}
theme_map <- function(...) {
  theme_minimal() +
    theme(
      text = element_text(family = "Avenir Medium", color = "#22211d"),
      axis.line = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      # panel.grid.minor = element_line(color = "#ebebe5", size = 0.2),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = "white", color = NA), 
      panel.background = element_rect(fill = "white", color = NA), 
      legend.background = element_rect(fill = "white", color = NA),
      panel.border = element_blank(),
      ...
    )
}
```

<!-- ```{r} -->
<!-- print(fit, pars = c("beta0", "beta", "rho", "sigma", "theta[5]", "phi[5]")) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- color_scheme_set("purple") -->

<!-- y <- bg_dat$daily_boards -->
<!-- yrep <- as.matrix(fit, pars = "y_sim") -->

<!-- ppc_dens_overlay(y, yrep[sample(1:ncol(yrep), 50), ]) + -->
<!--   xlim(0, 100) + -->
<!--   labs(title = "posterior predictive check") -->
<!-- ``` -->

# Introduction

This document summarizes the results of a spatial direct demand model of Metro Transit’s bus ridership. The model takes the spatial dimension of bus ridership into account and incorporates demographic predictors. This builds on existing methodology by leveraging more of what we know about ridership and the region to predict where ridership will occur.

The model and analysis presented here provide insight to the likelihood of ridership across space given the existing network. The analysis is conducted at the Census block group level. Ridership is defined as the average number of weekday boardings per block group in 2020. Predictors are from the 2018 American Community Survey (ACS) 5-year estimates and 2017 Longitudinal Employee Household Dynamic Survey (LEHD). The use of demographic predictors allows us to consider where ridership is likely to occur based on known characteristics of the region.

Bus ridership is inherently spatial. In other words, ridership (or boardings) can only occur at fixed bus stops. However, we intuitively expect neighboring block groups to have similar levels of ridership regardless of specific bus stop placement. For example, we might expect all block groups along the A Line to have similar levels of A Line ridership even though not all of those block groups contain an A Line stop. The model presented here addresses these spatial concerns by introducing spatial autocorrelation and geographic smoothing into the model. In other words, this model smooths estimates between block groups to create more stable and realistic predictions of ridership across space. This spatial component adds nuance to the model and allows us to see patterns we cannot see in non-spatial models.

We first performed this analysis using bus ridership data from 2019. The 2019 analysis can be found [here](mtapshiny1p/MT/SIDev/spatial-direct-demand-2019analysis.html). This document outlines key findings from the model of 2020 ridership and compares findings to the pre-COVID 2019 analysis.

## Motivation for 2020 Model

We were motivated to refit this model using 2020 ridership data to understand who is still using transit under COVID-19 conditions. In many ways, we expect these riders to be Metro Transit's essential, core ridership. Performing this analysis gives insight into who those essential riders are.

Figure 1 shows average weekday boardings by Census block group in 2019 versus 2020 on the same scale.

```{r fig.cap = "Ridership in 2019 versus 2020"}
a <- ggplot(bg_dat19 %>% dplyr::filter(daily_boards != 0)) +
  geom_sf(aes(fill = log(daily_boards)), lwd = 0) +
  scale_fill_viridis_c(option = "plasma", 
                       limits = c(-5.541264, 9.990346)) +  # set limits based on 2019 ridership
  theme_map(legend.position = "bottom") +
  labs(title = "Average daily boardings, \nJan - December 2019", 
       guide = "log(boardings)")

match <- bg_dat %>%
  dplyr::filter(GEOID %in% (bg_dat19 %>% dplyr::filter(daily_boards != 0))$GEOID)

b <- ggplot(match) +
  geom_sf(aes(fill = log(daily_boards)), lwd = 0) +
  scale_fill_viridis_c(option = "plasma", 
                       limits = c(-5.541264, 9.990346), 
                       na.value = "grey70") +
  theme_map(legend.position = "None") +
  labs(title = "Average daily boardings, \nMarch - September 2020", 
       guide = "log(boardings)")

#extract legend
#https://github.com/hadley/ggplot2/wiki/Share-a-legend-between-two-ggplot2-graphs
g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}

legend <- g_legend(a)

gridExtra::grid.arrange(arrangeGrob(a + theme(legend.position="none"),
                                    b + theme(legend.position="none"),
                                    nrow=1),
                        legend, nrow=2,heights=c(10, 1))
```

# Findings

This model was tested with a wide variety of ACS variables. The model presented here uses eleven of the most influential variables tested. Five of these demographic predictors have a positive correlation with ridership. Areas with these characteristics have a higher “propensity for ridership” compared to other block groups. Similarly, the two demographic variables negatively correlated with ridership can be thought of as decreasing an area’s propensity for ridership.

Detailed descriptions and maps of the variables can be found in the [2019 report](). Table 1, below, summarizes the coefficient estimates for both the 2019 and 2020 models.

Table: Summary of coefficient estimates

| Variable | Meaning                                        | 2019            | covid           |
|----------|------------------------------------------------|-----------------|-----------------|
| $x_1$    | Employment density                             | 0.07            | 0.06            |
| $x_2$    | Walkability                                    | 0.06            | 0.03            |
| $x_3$    | Percent of households with no vehicle          | 0.04            | 0.06            |
| $x_4$    | Percent of residents aged 18 to 34             | 0.06            | 0.04            |
| $x_5$    | Presence of college or university campus       | 0.32            | 0.08            |
| $x_6$    | Presence of rail station                       | 0.36            | 0.52            |
| $x_7$    | Presence of hospital campus                    | 0.23            | 0.00            |
| $x_8$    | Presence of MSP International Airport          | 0.52            | 0.00            |
| $x_9$    | Median household income                        | -0.10           | -0.20           |
| $x_{10}$ | Percent of housing units occupied by renters   | 0.15            | 0.15            |
| $x_{11}$ | Percent of residents identified as white alone | -0.06           | -0.07           |

Table: Summary of change in coefficient estimates

| Change    | Variables                                                                                                          |
|-----------|--------------------------------------------------------------------------------------------------------------------|
| No change | Employment density, percent of housing units occupied by renters, percent of residents identified as white alone   |
| Increase  | Percent of residents with no vehicle, presence of rail station, median household income                            |
| Decrease  | Walkability, percent of residents age 18-34, presence of college campus, presence of hospital, presence of airport |

Many of these changes are fairly intuitive. For example, almost all of the "land use" predictors became less important in the 2020 model under COVID-19 conditions. For example, hospitals ($x_7$) and college campuses ($x_5$) are no longer generators of ridership. Additionally, ridership is no longer a strong positive predictor of ridership. While walkability does encourage ridership, the positives of walkability are easily outweighed by the risk of COVID-19.

We see some coefficients remain stable between 2019 and 2020. For example, employment density ($x_1$) and percent renters ($_10$). These predictors are equally important under normal and pandemic conditions. We might say that in block groups with high employment density, for example, we expect ridership to remain more stable than in block groups with average or low employment density.

Finally, we see some coefficients becoming larger in 2020. Notably, percent of households with no vehicle is a stronger predictor in 2020 than in 2019. We often expect this to be a strong predictor of ridership, even though the coefficient was relatively small in the 2019 model. In 2020, with much of the 9-to-5 commute removed, the effect of this predictor becomes more pronounced. In other words, the model -- and ridership -- have become less sensitive to the 9-to-5 commute and more sensitive to demographic differences.

<details>
  <summary> Compare distribution of 2019 vs 2020 coefficients </summary>

The figure below shows the distribution of possible coefficient estimates for each predictor in both 2019 and 2020. If the two distributions line up closely, the predictor has not changed significantly. The less the distributions overlap, the more confident we are that the coefficient has changed between 2019 and 2020. Note that all of these distributions have significant overlap. Therefore, a major limitation of this comparative analysis is relatively small and uncertain change year over year. The blue distribution represents 2019 and the red, 2020.

```{r}
covid_betas <- as.matrix(fit, pars = "beta")

fit19 <- readRDS("/Users/raven/Documents/honors/mt-translation/2019/fits/bym2_disagf.RDS")
betas_19 <- as.matrix(fit19, pars = "beta")

betas <- data.table(variable = c("emp_density", "walkability", "perc_no_veh", "age18to34",
          "college", "lightrail", "hospital", "airport", "estimate_median_hh_income", 
          "perc_rent", "perc_only_white"), 
          betas19 = apply(betas_19, 2, as.list), 
          betas20 = apply(covid_betas, 2, as.list))

betas <- betas %>%
  unnest(cols = c(betas19, betas20)) %>%
  mutate(betas19 = as.numeric(betas19),
         betas20 = as.numeric(betas20))
```

```{r fig.cap = "Comparison of coefficient distributions, 2019 vs 2020"}
colors <- c("2019" = "blue", "2020" = "red")

ggplot(data = betas) + 
  geom_histogram(aes(x = betas19, fill = "2019"), fill = "blue", alpha = 0.5) + 
  geom_histogram(aes(x = betas20, fill = "2020"), fill = "red", alpha = 0.5) +
  theme_light() +
  labs(x = NULL, y = NULL, fill = "Year") +
  scale_fill_manual(values = colors, name = "Year") +
  theme(legend.position = "top") +
  facet_wrap(~variable, scales = "free") 
```
</details>

### 2021-01-26 functions/nb_data_funs.R

@@ -1,0 +1,280 @@
+## credit to morris et al 2019
+
+library(INLA);
+library(spdep);
+
+
+# validateNb
+# check that nbObject is symmetric, has connected components
+#
+validateNb = function(x) {
+  if (is.symmetric.nb(x) && n.comp.nb(x)[[1]] < length(x)) return(TRUE);
+  return(FALSE);
+}
+
+
+# isDisconnected
+# check that nbObject has more than 1 component
+#
+isDisconnected = function(x) {
+  return(n.comp.nb(x)[[1]] > 1);
+}
+
+
+# indexByComponent
+#
+# input: vector of component ids
+# returns: vector of per-component consecutive node ids
+#
+indexByComponent = function(x) {
+  y = x;
+  comps = as.matrix(table(x));
+  num_comps = nrow(comps);
+  for (i in 1:nrow(comps)) {
+    idx = 1;
+    rel_idx = 1;
+    while (idx <= length(x)) {
+      if (x[idx] == i) {
+        y[idx] = rel_idx;
+        rel_idx = rel_idx + 1;
+      }
+      idx = idx + 1;
+    }
+  }
+  return(y);
+}
+
+
+# nb2graph
+#
+# input: nb_object
+# returns: dataframe containing num nodes, num edges,
+#          and a list of graph edges from node1 to node2.
+#
+nb2graph = function(x) {
+  N = length(x);
+  n_links = 0;
+  for (i in 1:N) {
+    if (x[[i]][1] != 0) {
+      n_links = n_links + length(x[[i]]);
+    }
+  }
+  N_edges = n_links / 2;
+  node1 = vector(mode="numeric", length=N_edges);
+  node2 = vector(mode="numeric", length=N_edges);
+  idx = 0;
+  for (i in 1:N) {
+    if (x[[i]][1] > 0) {
+      for (j in 1:length(x[[i]])) {
+        n2 = unlist(x[[i]][j]);
+        if (i < n2) {
+          idx = idx + 1;
+          node1[idx] = i;
+          node2[idx] = n2;
+        }
+      }
+    }
+  }
+  return (list("N"=N,"N_edges"=N_edges,"node1"=node1,"node2"=node2));
+}
+
+
+# nb2subgraph
+# for a given subcomponent, return graph as lists of node1, node2 pairs
+#
+# inputs:
+# x: nb object
+# c_id: subcomponent id
+# comp_ids: vector of subcomponent ids
+# offsets: vector of subcomponent node numberings
[…]

### 2021-01-26 functions/plotting.R

@@ -1,0 +1,46 @@
+## goal: speed up repetetive plotting
+
+## packages -----------------------------------------------
+
+# these lines make sure any user has all the necessary packages
+packages <- c('ggplot2', 'dplyr')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+## functions ----------------------------------------------
+
+## MAKE HISTOGRAMS ##
+
+makeHist <- function(dat, var, titles, subtitles) {
+  ggobj <- ggplot(data = dat, aes_string(x = var)) + 
+    geom_histogram(fill = "#33638dff") +
+    theme_light() +
+    labs(title = titles,
+         subtitle = subtitles,
+         x = "value") 
+  ggsave(sprintf("img/hist/%s.png", var))
+}
+
+
+## MAKE MAPS ##
+# theoretically should be basically the same
+
+makeMap <- function(dat, var, titles, legend_titles, subtitles) {
+  ggobj <- ggplot(data = dat) + 
+    geom_sf(aes_string(fill = var), lwd = 0) +
+    theme_light() +
+    labs(title = paste0(titles),
+         subtitle = subtitles, 
+         fill = legend_titles) +
+    scale_fill_viridis_c(na.value = "transparent")
+  ggsave(sprintf("img/map/%s.png", var))
+}
+

### 2021-01-26 scripts/01_data-pull.R

@@ -1,0 +1,116 @@
+# Goal: 
+#     FACT_APC to get observed trips
+#     FACT_AVL to get run trips
+#     schedule to fill in stops between time points
+# This script pulls data from Metro Transit data bases
+
+# Credit to Eric Lind
+
+## packages -----------------------------------------------
+
+# these lines make sure any user has all the necessary packages
+packages <- c('data.table', 'DBI', 'odbc', 'lubridate')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+
+## the connection -----------------------------------------
+
+ch_sdw <- dbConnect(odbc(), 'db_SDWarehouse')
+
+
+## picks --------------------------------------------------
+
+system.time(picks <- dbGetQuery(ch_sdw, "SELECT * from DIM_PICK_DATE"))
+# 0.002
+
+## actual trips (FACT_AVL) --------------------------------
+
+# query for pre-covid
+# avl_query <- paste0("
+#                     set transaction isolation level read uncommitted;
+#                     SELECT DISTINCT date_key, a.trip_tmsk, block_number, trip_number,
+#                     line_id, line_direction, service_id, time_bracket_start, DIM_TRIP_TM.ROUTE_PROVIDER
+#                     FROM FACT_AVL a
+#                     LEFT JOIN DIM_TRIP_TM
+#                     on DIM_TRIP_TM.TRIP_TMSK = a.TRIP_TMSK
+#                     LEFT JOIN DIM_STOP
+#                     on DIM_STOP.stopsk = a.stopsk
+#                     WHERE date_key BETWEEN 20150101 AND 20181231
+#                     ")
+
+avl_query <- paste0("
+                    set transaction isolation level read uncommitted;
+                    SELECT DISTINCT date_key, a.trip_tmsk, block_number, trip_number,
+                    line_id, line_direction, service_id, time_bracket_start, DIM_TRIP_TM.ROUTE_PROVIDER
+                    FROM FACT_AVL a
+                    LEFT JOIN DIM_TRIP_TM
+                    on DIM_TRIP_TM.TRIP_TMSK = a.TRIP_TMSK
[…]

### 2021-01-26 scripts/02_interpolate.R

@@ -1,0 +1,144 @@
+# Goal: Interpolate boarding & alighting for trips with no APC records
+# Reads data from 01 
+# This script also uses Metro Transit databases
+
+# Credit to Eric Lind
+
+## packages -----------------------------------------------
+
+packages <- c('data.table', 'DBI', 'odbc')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+## data ---------------------------------------------------
+
+# observed APCs
+apc <- readRDS('data/covid/raw/apc.RDS')
+setDT(apc)
+apctrips <- apc[, unique(trip_tmsk)]
+apctrips <- apc[ROUTE_PROVIDER == "Metro Transit"]
+
+# observed trips
+trips <- readRDS('data/covid/raw/trips.RDS')
+setDT(trips)
+trips <- trips[ROUTE_PROVIDER == "Metro Transit"]
+
+# picks to get unique trip IDs to gap fill
+picks <- readRDS('data/covid/raw/picks.RDS')
+setDT(picks)
+setkey(picks, PICK_START_DATE_KEY, PICK_END_DATE_KEY)
+apc <- picks[apc, on = .(PICK_START_DATE_KEY = date_key), roll = T]
+setnames(apc, 'PICK_START_DATE_KEY', 'date_key')
+
+# observed trips without APC
+noAPC_trips <- trips[!trip_tmsk %in% apctrips]
+noAPC_trips[, .N, line_id][order(-line_id)][1:20]
+
+# drop Northstar, training & testing routes, others with no APC
+noAPC_trips[as.integer(line_id) > 921, paste0(unique(line_id), collapse = "','")]
+noAPC_trips <- noAPC_trips[!line_id %in% c('888', '922', '771', '740', '791', '942', '906',
+                                           '741', '945','951','953','954','955','961','962','963','952',
+                                           '969','941','970','971','972','973')]
+
+# check range of dates before continuing!
+range(noAPC_trips$date_key)
+
+## pull schedule for missing trips ------------------------
+# get stop-level schedule for trips without APC
+ch_schedb <- dbConnect(odbc::odbc(), 'ScheduleDB')
+
+# direction look up table
[…]

### 2021-01-26 scripts/03_get-mod-dat.R

@@ -1,0 +1,127 @@
+## Goal: clean Metro Transit data from 01 and 02 into usable modeling data
+
+## packages -----------------------------------------------
+packages <- c('data.table', 'tigris', 'ggplot2', 'dplyr', 'rgdal', 'sf', 'lubridate')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+options(tigris_use_cache = TRUE)
+options(tigris_class = 'sf')
+
+## data ---------------------------------------------------
+# Read "raw" data, from "gap-fill.R" script
+apc <- readRDS('data/covid/raw/apc-interpolated.RDS')
+setDT(apc)
+
+apc[, site_id := as.character(site_id)]
+
+# assign stops to block groups ----------------------------
+url <- 'ftp://ftp.gisdata.mn.gov/pub/gdrs/data/pub/us_mn_state_metc/trans_transit_stops/shp_trans_transit_stops.zip'
+loc <- file.path(tempdir(), 'stops.zip')
+download.file(url, loc)
+unzip(loc, exdir = file.path(tempdir(), 'stops'), overwrite = TRUE)
+file.remove(loc)
+stops <- readOGR(file.path(tempdir(), 'stops'), layer = 'TransitStops', stringsAsFactors = FALSE)
+stops <- st_as_sf(stops)
+stops <- st_transform(stops, 4326)
+
+# block groups
+options(tigris_class = 'sf')
+counties <- c('Anoka', 'Carver', 'Dakota', 'Hennepin', 'Ramsey', 'Scott', 'Washington')
+bgs <- block_groups('MN', counties)
+bgs <- st_transform(bgs, 4326)
+
+apc[, site_id := as.character(site_id)]
+apc[, year := year(ymd(date_key))]
+
+# this step takes a while, save "stop_to_bg" and don't rerun
+apc_loc <- left_join(stops, apc) # dplyr joins are usually easier with sf objects
+saveRDS("data/covid/apc_loc.RDS")
+rm(apc) # just for space
+
+# intersect stops and block groups
+apc_bg <- st_join(apc_loc, bgs, st_intersects)
+setDT(apc_bg)
+
+# count bus stops
+bc <- apc_bg[rail == 0, .(count_bus_stops = length(unique(site_id))), by = 'GEOID']
[…]

### 2021-01-26 scripts/04_get-census.R

@@ -1,0 +1,378 @@
+# Goal: get census variables of interest, 2015-2018
+# I'm sure there are more refined ways to do this 
+
+## packages -----------------------------------------------
+
+packages <- c('data.table', 'tigris', 'tidycensus', 'purrr', 'tidyr', 'ggplot2')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+options(tigris_class = 'sf')
+
+
+## acs variables ------------------------------------------
+
+# look at acs 5yr estimates for 2015
+v18 <- load_variables(2018, "acs5", cache = TRUE)
+setDT(v15)
+# use this to look at variables
+View(v18)
+
+# get counties
+counties <- c('Anoka', 'Carver', 'Dakota', 'Hennepin', 'Ramsey', 'Scott', 'Washington')
+bgs <- block_groups("MN", counties, 2016)
+
+years <- list(2018)
+
+## basics ---------------------------------------
+basics <- map_dfr(
+  years,
+  ~ get_acs(
+    geography = "block group",
+    variables = c(tot_pop = "B01003_001",
+                  median_age = "B01002_001",
+                  median_hh_income = "B19013_001",
+                  white_alone = "B02001_002"),
+    state = "MN",
+    county = counties,
+    year = .x,
+    survey = "acs5"
+  ),
+  .id = "year"
+)
+
+basics <- basics %>% pivot_wider(names_from = 'variable', values_from = c('estimate', 'moe'))
+setDT(basics)
+basics[, perc_only_white := estimate_white_alone/estimate_tot_pop]
+basics <- basics[, .(year, GEOID, estimate_median_age, estimate_tot_pop, estimate_median_hh_income, perc_only_white)]
+
+saveRDS(basics, 'data/covariates/basic_acs.RDS')
+
+## age cohorts ----------------------------------
+
+agecohorts <- map_dfr(
+  years, 
+  ~ get_acs(
+    geography = "block group",
+    table = "B01001",
+    summary_var = "B01001_001",  # should have used this throughout!
+    state = "MN",
+    county = counties,
+    year = .x,
+    survey = "acs5"
+  ),
[…]

### 2021-01-26 scripts/05_get-isochrones.R

@@ -1,0 +1,59 @@
+## goal: get area of walk isochrones for each bg
+
+## packages -------------------------------------
+packages <- c('readr', 'sf', 'tigris', 'data.table', 'otptools', 'dplyr', 'ggplot2')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+
+options(tigris_class = 'sf')
+
+# first, download population-weighted centroids
+url <- "https://www2.census.gov/geo/docs/reference/cenpop2010/blkgrp/CenPop2010_Mean_BG27.txt"
+loc <- file.path("/Users/raven/Documents/honors/old-honors/data/covariates", 'centrs.csv')
+download.file(url, loc)
+
+centrs <- read_csv('data/covariates/centrs.csv')
+setDT(centrs)
+
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+bgs <- block_groups("MN", counties, 2016)
+
+centrs <- centrs[COUNTYFP %in% bgs$COUNTYFP]
+centrs[, bgid := paste0(STATEFP, COUNTYFP, TRACTCE, BLKGRPCE)]
+
+centrs_loc <- centrs[, c("bgid", "LATITUDE", "LONGITUDE")]
+setnames(centrs_loc, c("bgid", "LATITUDE", "LONGITUDE"), c("id", "lat", "lon"))
+
+# then, use opentripplanner to get 10min walk isos
+isos <- queryIsochrone(location = centrs_loc,
+                       otp_params = otp_params(cutoffSec = c(600), mode = "WALK", 
+                                               maxWalkDistance = 800, walkSpeed = 5000/3600),
+                       host = "localhost", port = 8080)
+
+saveRDS(isos, 'data/covariates/raw_isochrones.RDS')
+
+setDT(isos)
+
+isos[, area := st_area(geometry)]
+isos[, max_area := 2.01*10^6] # area of perfect r=800m circle
+isos[, walkability := area/max_area] 
+isos[, year  := as.character(4)]
+
+bg_isos <- isos[, .(area, max_area, walkability, id, year)]
+setnames(bg_isos, "id", 'GEOID')
+
[…]

### 2021-01-26 scripts/06_clean-covariates.R

@@ -1,0 +1,100 @@
+## Goal: get covariate data
+
+## packages -----------------------------------------------
+packages <- c('data.table', 'tigris', 'ggplot2', 'dplyr', 'rgdal', 'sf')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+options(tigris_class = 'sf')
+
+## read all covariates into df together -------------------
+# this is very imperfect
+
+file_list <- paste0("data/covariates/", list.files(path = "data/covariates/", pattern = ".RDS"))
+
+for (file in file_list){
+  
+  # if the merged dataset doesn't exist, create it
+  if (!exists("dataset")){
+    dataset <- readRDS(file)
+  }
+  
+  # if the merged dataset does exist, merge to it
+  if (exists("dataset")){
+    temp_dataset <- readRDS(file)
+    dataset <- temp_dataset[dataset, on = .(year, GEOID)]
+    rm(temp_dataset)
+  }
+  
+}
+
+setDT(dataset)
+
+# remove duplicates
+dataset[, `:=` (i.summary_est = NULL, i.summary_est.1 = NULL, i.genz = NULL, 
+                i.millenial = NULL, i.genx = NULL, i.boomer = NULL, i.under40 = NULL,
+                i.age18to34 = NULL)]
+
+dataset[, tract_GEOID := stringr::str_sub(GEOID, end = -2)]
+
+# only available at tract level
+tract_list <- paste0("data/covariates/tract/", list.files(path = "data/covariates/tract/", pattern = ".RDS"))
+
+for (file in tract_list){
+  
+  # if the merged dataset doesn't exist, create it
+  if (!exists("tdataset")){
+    tdataset <- readRDS(file)
+  }
+  
+  # if the merged dataset does exist, merge to it
+  if (exists("tdataset")){
+    temp_tdataset <- readRDS(file)
+    tdataset <- temp_tdataset[tdataset, on = .(year, GEOID)]
+    rm(temp_tdataset)
+  }
+  
+}
+
+setDT(tdataset)
[…]

### 2021-01-26 scripts/07_add-indicators.R

@@ -1,0 +1,104 @@
+## Goal: gather some indicator variables to try out
+
+## packages -----------------------------------------------
+packages <- c('data.table', 'tigris', 'ggplot2', 'dplyr', 'rgdal', 'sf')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+options(tigris_class = 'sf')
+
+## data ---------------------------------------------------
+
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+bgs <- block_groups("MN", counties, 2016)
+
+## gather -------------------------------------------------
+
+## college/university
+
+# read data from: https://hifld-geoplatform.opendata.arcgis.com/datasets/colleges-and-universities/data
+url <- "https://opendata.arcgis.com/datasets/0d7bedf9d582472e9ff7a6874589b545_0.zip?outSR=%7B%22latestWkid%22%3A3857%2C%22wkid%22%3A102100%7D"
+loc <- file.path(tempdir(), "cu.zip")
+download.file(url, loc)
+unzip(loc, exdir = file.path(tempdir(), "cu"), overwrite = TRUE)
+file.remove(loc)
+cu <- read_sf(file.path(tempdir(), "cu"), stringsAsFactors = FALSE)
+
+# transform
+cu <- st_transform(cu, st_crs(bgs))
+
+# just in counties
+cubg <- st_intersection(cu, bgs) # wow sf update is way faster
+ggplot() + geom_sf(data = bgs) + geom_sf(data = cubg)
+
+  
+## light rail
+
+# data from https://gisdata.mn.gov/dataset/us-mn-state-metc-trans-stop-boardings-alightings
+lr <- fread('data/csv_trans_stop_boardings_alightings/TransitStopsBoardingsAndAlightings2018.csv') # year doesn't really matter
+lr <- lr[Route == "Blue Line" | Route == "North Star" | Route == "Green Line"]
+lr <- st_as_sf(lr, coords = c('longitude', 'latitude'), crs = st_crs(bgs))
+
+# just in counties
[…]

### 2021-01-26 scripts/08_join-mod-dat.R

@@ -1,0 +1,64 @@
+# Goal: export the data needed for each model I'll fit
+
+## packages -------------------------------------
+
+packages <- c('data.table', 'ggplot2', 'lubridate')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+## data -----------------------------------------
+
+ridership <- readRDS('data/covid/aggregations/apc_daily_bg.RDS')
+setDT(ridership)
+ridership[, year := year(ymd(date_key))]
+ridership[, wday := wday(ymd(date_key), label = TRUE)]
+# get 2018 -- used 2017 for honors
+ridership <- ridership[year == 2020]
+# and set to match w covariates
+ridership[, year := as.character(4)]
+# remove weekends
+ridership <- ridership[wday != "Sat" & wday != "Sun"]
+# avg weekday
+ridership <- ridership[, .(daily_boards = mean(daily_boards, na.rm = T),
+                           daily_stops = mean(daily_stops, na.rm = T)), keyby = .(GEOID)]
+
+
+## SCALED ##
+cov <- readRDS('data/covariates/cleaned/all_covariates_scaled_ind.RDS')
+setDT(cov)
+
+# need year == '3' for lehd and year == '4' for acs
+lehd<- c("workers", "emp_density", "w_total_jobs_here", "w_perc_jobs_white", "w_perc_jobs_men",
+          "w_perc_jobs_no_college", "w_perc_jobs_less40", "w_perc_jobs_age_less30", "GEOID")
+lehd2 <- c("workers", "emp_density", "w_total_jobs_here", "w_perc_jobs_white", "w_perc_jobs_men",
+          "w_perc_jobs_no_college", "w_perc_jobs_less40", "w_perc_jobs_age_less30")
+
+emp <- cov[year == "3", ..lehd]
+notemp <- cov[year == "4", !..lehd2]
+
+cov <- emp[notemp, on = .(GEOID)]
+
+mod_dat <- ridership[cov, on = .(GEOID)]
+
+saveRDS(mod_dat, 'data/covid/modeling-dat/mod_dat.RDS')
+
+## UNSCALED ##
[…]

### 2021-01-26 scripts/09_fitting2019.R

@@ -1,0 +1,88 @@
+## Goal: fit models!
+## packages -----------------------------------------------
+packages <- c('rstan', 'dplyr', 'data.table', 'shinystan', 'ggplot2',
+              'bayesplot', 'tigris', 'spdep', 'sf')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+options(tigris_use_cache = TRUE)
+options(tigris_class = 'sf')
+options(mc.cores = parallel::detectCores())
+rstan_options(autowrite = T)
+
+## data -----------------------------------------
+
+mod_dat <- readRDS('data/2019/modeling-dat/mod_dat.RDS')
+setDT(mod_dat)
+
+vars <- c("emp_density", "w_perc_jobs_white", "w_perc_jobs_age_less30",
+          "walkability", "avg_veh", "under40",
+          "college", "lightrail", "hospital", "airport",
+          "pop_density", "estimate_median_hh_income", "perc_rent")
+
+## bym2 -----------------------------------------
+bym2 <- "stan/bym2_ysim.stan"
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+bgs <- block_groups("MN", counties, 2018)
+
+vars <- c("emp_density", "walkability", "perc_no_veh", "age18to34",
+           "college", "lightrail", "hospital", "airport",
+           "estimate_median_hh_income", "perc_rent",
+           "perc_only_white")
+
+mod_dat <- na.omit(mod_dat, cols = c("daily_boards", "daily_stops", vars))
+mod_dat <- mod_dat[daily_boards != 0]
+
+xdat <- mod_dat[, ..vars]
+setDT(xdat)
+
+xdat[, `:=` (college = as.numeric(as.character(college)), hospital = as.numeric(as.character(hospital)),
+              lightrail = as.numeric(as.character(lightrail)), airport = as.numeric(as.character(airport)))]
+
+bg_dat <- left_join(mod_dat, bgs, on = 'GEOID')
[…]

### 2021-01-26 scripts/10_explore-fitted.R

@@ -1,0 +1,167 @@
+## data and packages ------------------
+
+library(data.table)
+library(tidyverse)
+library(sf)
+library(spdep)
+library(tigris)
+
+options(tigris_class = "sf")
+# get morans I
+
+
+mod_dat <- readRDS('data/covid/modeling-dat/mod_dat.RDS')
+setDT(mod_dat)
+
+vars <- c("emp_density", "walkability", "perc_no_veh", "age18to34",
+          "college", "lightrail", "hospital", "airport",
+          "estimate_median_hh_income", "perc_rent",
+          "perc_only_white")
+mod_dat <- na.omit(mod_dat, cols = c("daily_boards", "daily_stops", vars))
+mod_dat <- mod_dat[daily_boards != 0]
+xdat <- mod_dat[, ..vars]
+
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+bgs <- block_groups("MN", counties)
+sfdat <- left_join(bgs, mod_dat)
+
+
+fit <- readRDS("fits/covid/covid_bymfit_nov25.RDS")
+fit2 <- readRDS("fits/covid/pois_theta.RDS")
+
+
+## residuals --------------------------
+
+# get predictive distributions
+yrep <- as.matrix(fit, pars = "y_sim")
+yrep2 <- as.matrix(fit2, pars = "y_sim")
+
+spatial_values <- data.table(predy = colMeans(yrep), 
+                             lower = apply(yrep, 2, quantile, 0.025), 
+                             upper = apply(yrep, 2, quantile, 0.975),
+                             truey = mod_dat$daily_boards, 
+                             GEOID = mod_dat$GEOID)
+spatial_values[, resid := predy - truey]
+
+nonspatial_values <- data.table(predy = colMeans(yrep2),
+                                lower = apply(yrep2, 2, quantile, 0.025),
+                                upper = apply(yrep, 2, quantile, 0.975),
+                                truey = mod_dat$daily_boards,
+                                GEOID = mod_dat$GEOID)
+nonspatial_values[, resid := predy - truey]
+
+## true y in 95% CI
+
[…]

### 2021-01-26 scripts/10_fittingcovid.R

@@ -1,0 +1,120 @@
+## Goal: fit models for mt!
+
+## packages -----------------------------------------------
+packages <- c('rstan', 'dplyr', 'data.table', 'shinystan', 'ggplot2',
+              'bayesplot', 'tigris', 'spdep', 'sf')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+options(tigris_use_cache = TRUE)
+options(tigris_class = 'sf')
+options(mc.cores = parallel::detectCores())
+rstan_options(autowrite = T)
+
+## data -----------------------------------------
+
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+bgs <- block_groups("MN", counties)
+
+mod_dat <- readRDS('data/covid/modeling-dat/mod_dat.RDS')
+setDT(mod_dat)
+
+vars <- c("emp_density", "walkability", "perc_no_veh", "age18to34",
+          "college", "lightrail", "hospital", "airport",
+          "estimate_median_hh_income", "perc_rent",
+          "perc_only_white")
+
+mod_dat <- na.omit(mod_dat, cols = c("daily_boards", "daily_stops", vars))
+mod_dat <- mod_dat[daily_boards != 0]
+
+bg_dat <- left_join(mod_dat, bgs, on = 'GEOID')
+bg_dat <- st_as_sf(bg_dat) 
+neighborhood <- poly2nb(bg_dat)
+
+## REMOVE UNCONNECTED SUBGRAPHS
+
+# first, remove bgs with 0 neighbors
+setDT(bg_dat)
+for(i in 1:nrow(bg_dat)){
+  bg_dat[i, neighbors := list(neighborhood[[i]])]
+  bg_dat[i, num_neighbors := length(unlist(neighbors))]
+  
+  if(neighborhood[[i]] == 0){
+    bg_dat[i, num_neighbors := 0]
+  }
+}
+
+bg_dat <- bg_dat[num_neighbors != 0]
+
+# I'm sure there's a better way to do this but I just grabbed geoids from QGIS
+subgraph_geoids <- c(271630712063, 271630712062, 271630712071, 271630712073, 
[…]

### 2021-01-26 scripts/11-figures.R

@@ -1,0 +1,151 @@
+## goal: create lots of plots!
+
+## packages -----------------------------------------------
+
+# these lines make sure any user has all the necessary packages
+packages <- c('ggplot2', 'dplyr', 'sf', 'tigris', 'data.table')
+
+miss_pkgs <- packages[!packages %in% installed.packages()[,1]]
+
+if(length(miss_pkgs) > 0){
+  install.packages(miss_pkgs)
+}
+
+invisible(lapply(packages, library, character.only = TRUE))
+
+rm(miss_pkgs, packages)
+
+## data ---------------------------------------------------
+
+counties <- c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")
+options(tigris_class = "sf")
+bgs <- block_groups("MN", counties, 2016)
+
+# unscaled covariates for mapping etc
+mod <- readRDS('data/modeling-dat/mod_dat_unscaled.RDS')
+# multiply by 100 for labeling percents
+# just maps I might actually use
+mod[perc_no_veh == 0, perc_no_veh := 0.01]
+mod[age18to34 == 0, age18to34 := 0.01]
+mod[, `:=` (perc_rent = 100*perc_rent, w_perc_jobs_age_less30 = w_perc_jobs_age_less30*100,
+            w_perc_job_white = w_perc_jobs_white*100, under40 = under40*100, rail = lightrail, 
+            age18to34 = age18to34*100, perc_no_veh = perc_no_veh*100, perc_only_white = perc_only_white*100,
+            perc_english_only = perc_english_only*100, genz = genz*100, millenial = millenial*100,
+            genx = genx*100, boomer = boomer*100, log_veh = log(perc_no_veh), log_age = log(age18to34))]
+
+mod[, `:=` (rail = as.factor(rail), college = as.factor(college), hospital = as.factor(hospital),
+            airport = as.factor(airport))]
+
+tohist <- mod[, .(unemprate, perc_foreign, perc_transit_comm, perc_wfh, walkability, 
+                 avg_veh, perc_english_only, perc_rent, perc_hs, perc_bach, perc_no_veh,
+                 estimate_median_age, estimate_median_hh_income, perc_only_white,
+                 w_perc_jobs_age_less30, w_perc_jobs_white, w_perc_jobs_men, 
+                 w_perc_jobs_less40, w_perc_jobs_no_college, genz, millenial,
[…]
